### 2018-02-01 01-intro.tex
- Primeiro commit - Apenas compila - Semântica antiga (Tese e artigo TECS) - Texto novo (SBLP)

\section{Introduction}
\label{sec.intro}

\CEU~\cite{ceu.sensys13} is a programming language for embedded soft real-time
systems that aims to offer a concurrent and expressive alternative to C and C++
with the characteristics that follow:

\begin{description}
\item [Reactive:] code only executes in reactions to events.
\item [Structured:] programs use structured control mechanisms, such as
    \code{await} (to suspend a line of execution), and \code{par} (to combine
    multiple lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}

Structured reactive programming eliminates the
\emph{callback hell}~\cite{TODO}, letting programmers write code in
direct/sequential style \cite{TODO}.
%
Previous work evaluates the expressiveness of \CEU in comparison to
event-driven code in C and attests a reduction in source code size (around
25\%) with a small increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
\CEU has been used in the context of wireless sensor
networks~\cite{ceu.sensys13,ceu.terra},
multimedia systems~\cite{ceu.media.webmedia16}, and
games \footnote{\fs{TODO}}.
It has also been used as an alternative language in an undergradutate-level
course on embedded systems for the past 5 years.

\fs{Embedded systems requirements, determinism, termination}

\begin{itemize}
\item \CEU
    \begin{itemize}
    \item variaveis compartilhadas
    \item concorrencia segura
    \item terminacao e determinismo
    \end{itemize}
\item Modelo Sincrono
    \begin{itemize}
    \item conhecido como ammenable to verification
    \item valido para \CEU?
    \end{itemize}
\end{itemize}

%only the static semantics for embedded systems

### 2018-02-01 02-ceu.tex
- Primeiro commit - Apenas compila - Semântica antiga (Tese e artigo TECS) - Texto novo (SBLP)

\section{\CEU}
\label{sec.ceu}

\CEU is a synchronous reactive language inspired by
Esterel~\cite{esterel.ieee91} in which programs advance in a sequence of
discrete reactions to external events.
%
\CEU is designed for control-intensive applications, supporting concurrent
lines of execution, known as \emph{trails}, and broadcast communication
through events.
%
Computations within a reaction (such as expressions, assignments, and
C~calls) execute in no time in accordance to the synchronous
hypothesis~\cite{rp.hypothesis}.
%
The \code{await} is the only \CEU statement that actually ``consumes'' time.
%%
An~\code{await} statement blocks the current trail allowing the program to
advance its other trails; when all trails are blocked, the reaction
terminates and control returns to the environment.

In \CEU, every execution path within a loop must contain at least one
\code{await} statement~\cite{ceu.sensys13,esterel.primer}.
%
This restriction, which is checked statically by the \CEU compiler, ensures
that every reaction runs in bounded time, eventually terminating with all
program trails blocked in \code{await} statements.
%
\CEU has a further restriction which it shares with Esterel and synchronous
languages in general~\cite{esterel.preemption}: computations that actually
take a non-negligible time to run (e.g., cryptography or image processing
algorithms) violate the zero-delay hypothesis, and thus cannot be directly
implemented.

Listing~\ref{lst.syntax} below shows a compact reference of~\CEU.

\bgroup
\def\T<#1>{\langle\mathit{#1}\rangle}
\def\C#1#2{\hfill\rmfamily\itshape\makebox[#1\columnwidth][l]{//~#2}}
\begin{lstlisting}[
  basicstyle=\ttfamily\footnotesize,
  caption={The concrete syntax of \CEU.},
  label={lst.syntax},
]
|\C{1.}{Declarations:}|
input  $\T<type>$ $\T<ids>$;|\C{.6}{declare external input events}|
output $\T<type>$ $\T<ids>$;|\C{.6}{declare external output events}|
event  $\T<type>$ $\T<ids>$;|\C{.6}{declare internal events}|
var    $\T<type>$ $\T<id>$ = $\T<exp>$;|\C{.6}{declare and initialize variable}|

|\C{1.}{Event handling:}|
$\T<id>$ = await $\T<id>$;|\C{.6}{await event and assign the received value}|
$\T<id>$ = await $\T<time>$;|\C{.6}{await time and assign the delayed delta}|
emit $\T<id>$($\T<exp>$);|\C{.6}{emit event passing a value}|

|\C{1.}{Control-flow:}|
$\T<stmt>$ ; $\T<stmt>$|\C{.45}{sequence}|
if $\T<exp>$ then $\T<stmts>$ else $\T<stmts>$ end|\C{.45}{conditional}|
loop do $\T<stmts>$ end|\C{.45}{repetition}|
finalize $\T<stmts>$ with $\T<stmts>$ end|\C{.45}{finalization}|

par/or  do $\T<stmts>$ with $\T<stmts>$ end|\C{.45}{aborts when any side terminates}|
par/and do $\T<stmts>$ with $\T<stmts>$ end|\C{.45}{terminates when all sides terminate}|
par     do $\T<stmts>$ with $\T<stmts>$ end|\C{.45}{never terminates}|

|\C{1.}{Assignment \& integration with C:}|
$\T<id>$ = $\T<exp>$;|\C{.45}{assign value to variable}|
_$\T<id>$($\T<exps>$)|\C{.45}{call C function (id starts with `\_'\,)}|
\end{lstlisting}
\egroup

To make matters concrete, consider the program of Listing~\ref{lst.ceu}.
%
This program continuously turns a LED~on for~2 seconds and off for~1 second,
and terminates after~1 minute of activity with the LED off.
%
The implementation uses a \code{par/or} to run two activities in parallel:
an endless loop that blinks the LED on and off (lines~2--7), and a single
statement that waits for~1 minute before terminating (line~9).
%
The \code{par/or} block stands for a \emph{parallel-or} composition; when
executed it creates~$n$ parallel trails (in this case, $n=2$) and rejoins
them when any of these~$n$ trails terminates, automatically aborting the
other trails.

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={\CEU program that blinks a LED during 1 minute.},
  label={lst.ceu},
]
par/or do
    loop do
        _led(1);
        await 2s;
        _led(0);
        await 1s;
    end
with
    await 1min;
end
_led(0);
\end{lstlisting}

In \CEU, any identifier prefixed with an underscore (e.g., \code{_led}) is
passed unchanged to the underlying~C compiler.
%
Therefore, access to~C is straightforward and syntactically traceable.
%
To ensure that programs operate synchronously, the compiler environment
should only provide access to~C operations that are assumed run in zero
time, such as non-blocking~I/O and access to \code{struct}'s.

\subsection{External and Internal Events}
\label{sec.ceu.evts}

\CEU defines time as a discrete sequence of reactions to unique external
input events.
%
These external input events are received from the environment; each of them
delimits a new logical unit of time and triggers a corresponding reaction.
%
The life-cycle of a program in \CEU can be summarized as
follows~\cite{ceu.sensys13}.
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail
  (parallel constructs may create new trails).
  %
\item Active trails execute until they await or terminate, one after
  another.  This step is called a \emph{reaction chain}, and always runs in
  bounded time.
  %
\item When all its trails are blocked, the program goes idle and the
  environment takes control.
  %
\item On the occurrence of a new external input event, the environment
  awakes \emph{all} trails awaiting that event, and the program goes back to
  step~(ii).
\end{enumerate}

A program must react to an event completely before handling the next one.
%
By the synchronous hypothesis, the time the program spends in step~(ii)
above is conceptually zero (in practice, negligible).
%
Thus, from the point of view of the environment, the program is always
idle---on step~(iii).
%
In practice, if a new external input event occurs while a reaction is being
computed, the event is saved on a queue, which effectively schedules it to
be processed by a subsequent reaction.

\subsubsection*{External events and discrete time}

\begin{figure}[b]
\centering
\includegraphics[width=\columnwidth]{tick}
\caption{The discrete notion of time in \CEU.}
\label{fig.ticks}
\end{figure}

The processing of external input events induces a discrete notion of time in
\CEU.
%
Figure~\ref{fig.ticks} illustrates this notion.
%
The continuous timeline shows an absolute reference clock with ``physical
timestamps'' for the event occurrences (e.g., event~\code{C} occurs at
$17ms521us$).
%
The discrete timeline shows how the same occurring events fit in the logical
notion of time of \CEU.
%
The boot reaction \code{boot-0} happens before any input on program startup.
%
Event~\code{A} ``physically'' occurs during \code{boot-0} but, because time
is discrete, its corresponding reaction can only execute afterwards, at
logical instant~\code{A-1}.
%
Similarly, event~\code{B} occurs during~\code{A-1} and its reaction is
postponed to execute at~\code{B-2}.
%
Event~\code{C} also occurs during~\code{A-1} but its reaction must also wait
for~\code{B-2} to execute and so it is postponed to execute at~\code{C-3}.
%
Event~\code{D} occurs during an idle period and can start immediately at
\code{D-4}.
%
Finally, two instances of event~\code{E} occur during~\code{D-4}; they are
handled in the subsequent reactions~\code{E-5} and~\code{E-6}.

Unique input events imply mutually exclusive reactions, which execute
atomically and never overlap.
%
Automatic mutual exclusion is a prerequisite for deterministic reactions as
we discuss in Section~\ref{sec.sem}.
%
%8<- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
% \gl{A não ser que seja desenvolvido (e.g., explicado e comparado com
%   Esterel) eu acho que o parágrafo anterior é dispensável.}
% \fs{Tirei o "simplifies resoning about concurrency".
%     O resto do parágrafo é absoluto e tenta dar a intuição das condições para
%     ter determinismo.}
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - ->8

In practice, the synchronous hypothesis for \CEU holds if reaction rate is
faster than the rate of incoming input events.
%
Otherwise, the program continuously accumulates a delay between the real
occurrence time and actual reaction time of events.
%
In the soft real-time systems targeted by \CEU (e.g., sensor networks,
multimedia systems, interactive games, etc.) such delay and postponed
reactions might be tolerated by users as long as they are infrequent and the
application does not take too long to catch up with real time.

\subsubsection*{Internal events as subroutines}

In \CEU, the queue-based processing of events described previously applies
only to external input events, i.e., those events submitted to the program
by the environment.
%
Internal events, which are those events generated internally by the program
via \code{emit} statements, are processed in a stack-based manner.
%
These internal events provide a fine-grained execution control and, because
of their stack-based processing, can be used to implement a limited form of
subroutine, as illustrated in Listing~\ref{lst.sub} below.

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={A \CEU program with a ``subroutine''.},
  label={lst.sub},
]
event int* inc;     // subroutine |`|inc|'|
par/or do
    loop do         // definitions are loops
        var int* p = await inc;
        *p = *p + 1;
    end
with
    var int v = 1;
    $\cdots$
    emit inc(&v);   // call |`|inc|'|
    _assert(v==2);  // after return
end
\end{lstlisting}
\vskip-\baselineskip

In Listing~\ref{lst.sub}, the ``subroutine'' \code{inc} is defined as a loop
(lines~3--6) that continuously awaits its identifying event (line~4), and
increments the value passed to it by reference (lines~5).
%
A trail in parallel (lines~8--11) invokes the subroutine through the
\code{emit} statement at line~10.
%
Given the stack-based execution mode of internal events, after the emit
statement at line~10 is executed, the calling trail pauses, the subroutine
awakes (line~4), runs its body (yielding \code{v=2}), loops, and awaits the
next ``call'' (line~4, again).
%
Only after this sequence does the calling trail resumes and moves on to
execute the assertion on line~11.

\CEU also supports nested \code{emit} invocations for internal events.
%
For instance, the body of the subroutine \code{inc} in Listing~\ref{lst.sub}
could \code{emit} another internal event after awaking (line~4), creating a
new level in the stack.
%
We can think of the stack as a record of the nested, fine-grained internal
reactions that happen inside the same bigger reaction to some external
event.

This form of subroutine has a significant limitation though: it cannot
express recursion (an \code{emit} to itself is always ignored as a running
trail cannot be waiting on itself).
%
That said, it is this very limitation that brings important safety
properties to subroutines.
%
First, such subroutines are guaranteed to react in bounded time.
%
Second, memory for locals is also bounded, not requiring data stacks.
%
Third, \CEU subroutines can be safely used by the other primitives of the
language, such as parallel compositions and the \code{await} statement,
without breaking the programming model.
%
In particular, after calling a subroutine these primitives wait while
keeping context information, such as locals and the program counter, which
makes the calls behave similarly to those of
coroutines~\cite{lua.coroutines}.
%
Finally, in previous work, we built other advanced control mechanisms on top
of internal events, such as resumable exceptions and reactive
variables~\cite{ceu.rem13}.
%
%In Section~\ref{sec.adv.excpt} we show how to use them to implement
%exceptions.
%
%8<- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\gl{Revisei do início da Seção~2 até aqui.}
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - ->8

\subsection{Shared-Memory Concurrency}
    - referenciar warnings

Embedded applications make extensive use of global memory and shared resources,
such as through memory-mapped registers and system calls to device drivers.
Hence, an important goal of \CEU is to ensure a reliable behavior for programs
with concurrent lines of execution sharing memory and interacting with the
environment.

\begin{figure}[h]
\begin{minipage}[h]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax Accesses to \code{x} are never concurrent.}
\end{minipage}
%
\begin{minipage}[h]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\centering\small{\bx Accesses to \code{y} are concurrent but deterministic.}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure}

In \CEU, when multiple trails are active during the same reaction, they are
scheduled in lexical order, i.e., in the order they appear in the program
source code.
%
For instance, consider the two examples in Figure~\ref{lst.shared}, both
defining a shared variable (ln. 2), and assigning to it in parallel trails (ln.
5, 8).

In the example \ax, the two assignments to \code{x} can only execute in
reactions to different events \code{A} and \code{B}, which cannot occur
simultaneously by definition (Section~\ref{sec.ceu.evts}).
Hence, for the sequence of events \code{A->B}, \code{x} becomes \code{4}
(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes \code{3}
(\code{(1*2)+1}).

In the example \bx, the two assignments to \code{y} are simultaneous because
they execute in reaction to the same event \code{A}.
Since \CEU employs lexical order for intra-reaction statements, the execution
is still deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).
%
However, that an (apparently innocuous) change in the order of trails modifies
the behavior of the program.
%
To mitigate this threat, \CEU performs concurrency checks at compile time to
detect conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment
cannot read or write to that variable~\cite{ceu.sensys13}.
%
Nonetheless, the static checks are optional and do not affect the semantics of
the language.

\subsection{Abortion and Finalization}

The \code{par/or} of \CEU is an \emph{orthogonal abortion
mechanism}~\cite{esterel.preemption} because the two sides in the composition
need not be tweaked with synchronization primitives or state variables in order
to affect each other.
%
In addition, abortion is \emph{immediate} in the sense that it executes
atomically in the current micro reaction.
%
Immediate orthogonal abortion is a distinctive feature of synchronous languages
and cannot be expressed effectively in traditional (asynchronous)
multi-threaded languages~\cite{esterel.preemption,sync_async.threadsstop}.

However, aborting lines of execution that deal with external resources may lead
to inconsistencies.
%
For this reason, \CEU provides a \code{finalize} construct to unconditionally
execute a series of statements even if the enclosing block is aborted.

\CEU also enforces the use of \code{finalize} for system calls that deal with
pointers representing resources, as illustrated in the two examples of
Figure~\ref{lst.fin.ceu}:
%
\begin{itemize}
\item If \CEU \textbf{passes} a pointer to a system call (ln. \ax:5), the
pointer represents a \textbf{local} resource (ln. \ax:2) that requires finalization
(ln. \ax:7).
\item If \CEU \textbf{receives} a pointer from a system call return (ln. \bx:4),
the pointer represents an \textbf{external} resource (ln. \bx:2) that requires
finalization (ln. \bx:6).
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires
finalization clauses to accompany them.
%
In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request}
(ln. 5), which is an asynchronous call that transmits the buffer in the
background.
If the block aborts (ln. 11) before receiving an acknowledge from the
environment (ln. 9), the local \code{msg} goes out of scope and the external
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln. 7).
%
In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
4) returns an external file resource as a pointer.
If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln. 6).
%
In both cases, the code does not compile without the \code{finalize}
construct.%
\footnote{
The compiler only forces the programmer to write the finalization clause, but
cannot check if it actually handles the resource properly.
}

The finalization mechanism of \CEU is fundamental to preserve the orthogonality
of the \code{par/or} construct since the clean up code is encapsulated in the
aborted trail itself.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
par/or do
   var _buffer_t msg;
   <...> // prepare msg
   finalize
      _send_request(&msg);
   with
      _send_cancel(&msg);
   end
   await SEND_ACK;
with
   <...>
end
.
\end{lstlisting}
\centering\small{\ax Local resource finalization}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
par/or do
   var _FILE* f;
   finalize
      f = _fopen(...);
   with
      _fclose(f);
   end
   _fwrite(..., f);
   await A;
   _fwrite(..., f);
with
   <...>
end
\end{lstlisting}
\centering\small{\bx External resource finalization}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{
\CEU enforces the use of finalization to prevent \emph{dangling pointers} for
local resources and \emph{memory leaks} for external resources.
\label{lst.fin.ceu}
}
\end{figure}

### 2018-02-01 03-semantics.tex
- Primeiro commit - Apenas compila - Semântica antiga (Tese e artigo TECS) - Texto novo (SBLP)

\newcommand{\ST}{\1\xrightarrow[~n~]{}\1}
\newcommand{\LL}{\langle}
\newcommand{\RR}{\rangle}
\newcommand{\DS}{\displaystyle}

\newcommand{\1}{\;}
\newcommand{\2}{\;\;}
\newcommand{\3}{\;\;\;}
\newcommand{\5}{\;\;\;\;\;}

\section{Formal Semantics}
\label{sec.sem}

%\begin{document}

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel highlighting the peculiarities of \CEU, 
in particular the stack-based execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and system calls (which behave like in conventional 
imperative languages).

\subsection{Abstract Syntax}
\label{sec.sem.syntax}

\begin{figure}
\begin{lstlisting}[mathescape=true]
                                   // primary expressions
  p ::= mem(id)                   (any memory access to `id')
      $|$ await(id)                  (await event `id')
      $|$ emit(id)                   (emit event `id')
      $|$ break                      (loop escape)
                                   // compound expressions
      $|$ if mem(id) then p else p   (conditional)
      $|$ p ; p                      (sequence)
      $|$ loop p                     (repetition)
      $|$ p and p                    (par/and)
      $|$ p or p                     (par/or)
      $|$ fin p                      (finalization)
                                   // derived by semantic rules
      $|$ awaiting(id,n)             (awaiting `id' since sequence number `n')
      $|$ emitting(n)                (emitting on stack level `n')
      $|$ p @ loop p                 (unwinded loop)
      $|$ nop                        (terminated expression)
\end{lstlisting}
\caption{
    Reduced syntax of \CEU.
\label{lst.formal.syntax}
}
\end{figure}

Figure~\ref{lst.formal.syntax} shows the syntax for a subset of \CEU that is 
sufficient to describe all semantic peculiarities of the language.
%
Except for $fin$ and the expressions used internally by the semantics (i.e., 
$awating$, $emitting$, $p~@~loop; p$, and $nop$), all other expressions are 
equivalent to their counterparts in the concrete language.

The $mem(id)$ primitive represents all accesses, assignments, system calls, and 
output events that affect a memory location identified by $id$.
%
According to the synchronous hypothesis of \CEU, $mem$ expressions are 
considered to be atomic and instantaneous.
%
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression 
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
%unused identifier).
%
Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
an identifier is either a variable or an event.

\subsection{Operational Semantics}

The core of our semantics describes how a program reacts to a single external 
input event, i.e., starting from the input event, how the program behaves and 
becomes idle again to proceed to the subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way 
that at most one transition is possible at any time, resulting in deterministic 
reactions.
%
Each reaction is identified by a ever-increasing $n$ that remains constant 
during the entire reaction.
The transition rules map a program $p$ and a stack of events $S$ in a single 
step to a modified program and stack:
%
\begin{align*}
\LL S, p \RR &\ST
\LL S', p' \RR
    & \textbf{(rule-inner)}
\end{align*}
%
where
%
\begin{align*}
S, S' &\in id^*
    &&(stack~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
p, p' &\in P
    && (program~as~described~in~Figure~\ref{lst.formal.syntax}) \\
n     &\in \mathds{N}
    && (unique~identifier~for~the~entire~reaction)
\end{align*}
%
At the beginning of a reaction, the stack is initialized with the occurring 
external event $ext$ ($S=[ext]$), but $emit$ expressions can push new events on 
top of it (we discuss how they are popped further).
The sequence number $n$, which is incremented each reaction, prevents that 
awaiting expressions awake in the same reaction they are reached (the 
\emph{delayed awaits} as explained in Section~\ref{sec.ceu.ints}).

The transition rules for the primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL S,\1mem(id) \RR &\ST
\LL S,\1nop \RR
    & \textbf{(mem)}        \\
%%%
\LL S,\1await(id) \RR &\ST
\LL S,\1awaiting(id,n+1) \RR
    & \textbf{(await)}      \\
%%%
\LL id:S,\1awaiting(id,m) \RR &\ST
\LL id:S,\1nop \RR, \2if~m\leq n
    & \textbf{(awake)}   \\
%%%
\LL S,\1emit(id) \RR &\ST
\LL id:S,\1emitting(|S|) \RR
    & \textbf{(emit)}       \\
%%%
\LL S,\1emitting(k) \RR &\ST
\LL S,\1nop \RR, \2 if~k=|S|
    & \textbf{(pop)}        %\\
\end{align*}
}
%
A $mem$ operation executes immediately and becomes a $nop$ to indicate 
termination (rule \textbf{mem}).
An $await$ is transformed into an $awaiting$ (rule \textbf{await}) as an 
artifice to remember the external sequence number $n+1$ it can awake:
an $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
event $id$ matches the top of the stack and it was reached in a previous 
reaction (i.e., sequence number $m \leq n$).
%
%Remember that in \CEU, the \code{await} statement returns the value associated 
%with the corresponding event: the yielded $mem$ represents the operation to 
%query that value.
%
An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
stands for the stack length), and pushing the referred event on the stack (rule 
\textbf{emit}).
With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
rule \textbf{pop} expects its parameter $k$ to match the current stack level.
This trick provides the desired stack-based semantics for internal events.

Proceeding to compound expressions, the rules for conditionals and sequences 
are straightforward:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS val(id,n) \neq 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, p \RR }
    & \textbf{(if-true)}       \\
%%%
& \frac
    { \DS val(id,n) = 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, q \RR }
    & \textbf{(if-false)}       \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~;~q) \RR \ST \LL S', (p'~;~q) \RR }
    & \textbf{(seq-adv)}      \\
%%%
& \LL S, (nop~;~q) \RR \ST  \LL S, q \RR
    & \textbf{(seq-nop)}      \\
%%%
& \LL S, (break~;~q) \RR \ST \LL S, break \RR
    & \textbf{(seq-brk)}
\end{eqnarray*}
}
%
Given that our semantics focuses on control, rules \textbf{if-true} and 
\textbf{if-false} are the only to query $mem$ expressions.
%
The ``magical'' function $val$ receives a memory identifier and the current 
reaction sequence number, returning the current memory value.
%
Although the value here is arbitrary, it is unique in a reaction, because a 
given expression can execute only once within it (remember that $loops$ must 
contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

The rules for loops are analogous to sequences, but use \code{`@'} as 
separators to properly bind breaks to their enclosing loops:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (loop~p) \RR \ST \LL S, (p~@~loop~p) \RR
    & \textbf{(loop-expd)}       \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
% -----------------------------------------------------------
    {\DS \LL S, (p~@~loop~q) \RR \ST \LL S', (p'~@~loop~q) \RR }
    & \textbf{(loop-adv)}    \\
%%%
& \LL S, (nop~@~loop~p) \RR \ST \LL S, loop~p \RR
    & \textbf{(loop-nop)}    \\
%%%
& \LL S, (break~@~loop~p) \RR \ST \LL S, nop \RR
    & \textbf{(loop-brk)}
\end{eqnarray*}
}
%
When a program encounters a $loop$, it first expands its body in sequence with 
itself (rule \textbf{loop-expd}).
Rules \textbf{loop-adv} and \textbf{loop-nop} are similar to rules 
\textbf{seq-adv} and \textbf{seq-nop}, advancing the loop until they reach a 
$nop$.
However, what follows the loop is the loop itself (rule \textbf{loop-nop}).
Note that if we used \code{`;'} as a separator in loops, rules 
\textbf{loop-brk} and \textbf{seq-brk} would conflict.
%
Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything into 
a $nop$.
%Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything 
%into a $clear(p)$.
%We cannot simply transform the loop into a $nop$ because its body may be a 
%parallel composition containing finalization blocks.

Proceeding to parallel compositions, the semantic rules for $and$ and $or$ 
always force transitions on their left branches $p$ to occur before their right 
branches $q$:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p'~and~q) \RR }
    & \textbf{(and-adv1)}      \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p~and~q') \RR }
    & \textbf{(and-adv2)}      \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~q) \RR \ST \LL S', (p'~or~q) \RR }
    & \textbf{(or-adv1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S (p~or~q) \RR \ST \LL S', (p~or~q') \RR }
    & \textbf{(or-adv2)}   %\\
\end{eqnarray*}
}
%
The deterministic behavior of the semantics relies on the \emph{isBlocked} 
predicate, which is defined in Figure~\ref{fig.isBlocked} and used in rules 
\textbf{and-adv2} and \textbf{or-adv2}.
These rules require the left branch $p$ to be blocked in order to allow the 
right transition from $q$ to $q'$.
%
Basically, the \emph{isBlocked} predicate determines that an expression becomes 
blocked when all of its trails in parallel hang in $awaiting$ and $emitting$ 
expressions that cannot advance.

\begin{figure}
{\small
\begin{align*}
  isBlocked(n,a:S, awaiting(b,m)) &= (a \neq b \1\vee\1 m > n)   \\
  isBlocked(n,S, emitting(s))    &= (|S| \neq s)                     \\
  isBlocked(n,S, (p~;~q))        &= isBlocked(n,S,p)             \\
  isBlocked(n,S, (p~@~loop~q))   &= isBlocked(n,S,p)             \\
  isBlocked(n,S, (p~and~q))      &= isBlocked(n,S,p) \wedge
                                    isBlocked(n,S,q)             \\
  isBlocked(n,S, (p~or~q))       &= isBlocked(n,S,p) \wedge
                                    isBlocked(n,S,q)             \\
  isBlocked(n,S, \_)             &= false \2  (nop,await,      \\
                                  &    \5\5\5\2 emit,break,if,loop)   %\\
\end{align*}
}%
%\rule{14cm}{0.37pt}
\caption{
The recursive predicate $isBlocked$ is true only if all branches in parallel 
are hanged in $awaiting$ or $emitting$ expressions that cannot transit.
\label{fig.isBlocked}
}
\end{figure}

%%%

For a parallel $and$, if one of the sides terminates, the composition is simply 
substituted by the other side (rules \textbf{and-nop1} and \textbf{and-nop2}, 
as follows).
%
For a parallel $or$, if one of the sides terminates, the whole composition 
terminates, also applying the $clear$ function to properly finalize the aborted 
side (rules \textbf{or-nop1} and \textbf{or-nop2}):
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (nop~and~q) \RR \ST \LL S, q \RR
    & \textbf{(and-nop1)}   \\
%%%
& \LL S, (p~and~nop) \RR \ST \LL S, p \RR
    & \textbf{(and-nop2)}   \\
%%%
& \LL S, (nop~or~q) \RR \ST \LL S, clear(q) \RR
    & \textbf{(or-nop1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~nop) \RR \ST \LL S, clear(p) \RR }
    & \textbf{(or-nop2)}   %\\
\end{eqnarray*}
}
%
The $clear$ function, defined in Figure~\ref{fig.formal.clear}, concatenates 
all active $fin$ bodies of the side being aborted, so that they execute before 
the composition rejoins.
Note that there are no transition rules for $fin$ expressions.
This is because once reached, a $fin$ expression halts and will only execute 
when it is aborted by a trail in parallel and is expanded by the $clear$ 
function.
In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
the concrete language to a $fin$ in the formal semantics.
%
Note also that there is a syntactic restriction that a $fin$ body can only 
contain $mem$ expressions in sequence, i.e., they are guaranteed to execute 
entirely within a reaction.

\begin{figure}[b]
{\small
\begin{align*}
  clear( fin~p )       &= p                   \\
  clear( p~;~q )       &= clear(p)            \\
  clear( p~@~loop~q) ) &= clear(p)            \\
  clear( p~and~q )     &= clear(p)~;~clear(q) \\
  clear( p~or~q )      &= clear(p)~;~clear(q) \\
  clear( \_ )          &= nop
\end{align*}
}%
%\rule{14cm}{0.37pt}
\caption{
The function $clear$ extracts $fin$ expressions in parallel and put their 
bodies in sequence.
\label{fig.formal.clear}
}
\end{figure}

Finally, a $break$ in one of the sides in parallel escapes the closest 
enclosing $loop$, properly aborting the other side by applying the $clear$ 
function:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (break~and~q) \RR \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(and-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(and-brk2)}   \\
%%%
& \LL S, (break~or~q) \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(or-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(or-brk2)}   %\\
\end{eqnarray*}
}
%
A reaction eventually blocks in $awaiting$ and $emitting$ expressions in 
parallel trails.
%
If all trails hangs only in $awaiting$ expressions, it means that the program 
cannot advance in the current reaction.
%
However, $emitting$ expressions are pending in lower stack indexes and should 
eventually resume in the ongoing reaction (see rule \textbf{pop}).
%
Therefore, we define another rule that behaves as \textbf{rule-inner} 
(presented above) if the program can advance, and, otherwise, pops the stack to 
resume the lower level:
%
$$
\frac
    { \DS \LL S,p \RR \ST                   \LL S',p' \RR }
%   -----------------------------------------------------------
    {     \LL S,p \RR \xRightarrow[~~n~~]{} \LL S',p' \RR }
%
\5\5\5
%
\frac
    { \DS isBlocked(n,\1s:S,\1p) }
%   -----------------------------------------------------------
    { \LL s:S,p \RR \xRightarrow[~~n~~]{} \LL S,p \RR }
\5\5\5\textbf{(rule-outer)}
$$
%
To describe a \emph{reaction} in \CEU, i.e., how a program behaves in reaction 
to a single external event, we use the reflexive transitive closure of 
\textbf{rule-outer}:
%
$$
    \LL S,p \RR \xRightarrow[~~n~~]{*} \LL S',p' \RR
$$
%
Finally, to describe the complete execution of a program, we trigger multiple 
``invocations'' of reactions in sequence:
%
\begin{align*}
\LL [e1], p \RR
    & \xRightarrow[~~1~~]{*}
\LL [  ], p' \RR
\\
\LL [e2], p' \RR
    & \xRightarrow[~~2~~]{*}
\LL [  ], p'' \RR
\\
\LL [e3], p'' \RR
    & \xRightarrow[~~3~~]{*}
\LL [  ], p''' \RR
\\
& ...
\end{align*}
%
Each invocation starts with the occurring external event at the top of the 
stack and finishes with a modified program and an empty stack.
After each invocation, we increment the sequence number.

\subsection{Concrete Language Mapping}

Most statements from \CEU (``concrete \CEU'') map directly to those presented 
in the reduced syntax in Figure~\ref{lst.formal.syntax} (``abstract \CEU'').
%, namely, the \code{if}, \code{`;'}, \code{loop}, \code{par/and}, and 
%\code{par/or}.
For instance, the \code{if} in the concrete language behaves exactly like the 
$if$ in the formal semantics.
However, there are some significant mismatches between the concrete and 
abstract \CEU, and we (informally) present appropriate mappings in this 
section.
%
Again, we are not considering side-effects, which are all mapped to the $mem$ 
semantic construct.

\begin{comment}
$
    \code{if}      \mapsto if   ,\2
    \code{';'}     \mapsto ~';' ,\2
    \code{loop}    \mapsto loop ,\2
    \code{par/and} \mapsto and  ,\2
    \code{par/or}  \mapsto or
$.
\end{comment}

\subsubsection{await and emit}

The concrete \code{await} and \code{emit} primitives support communication of 
values between them.
In the two-step translation in Figure~\ref{lst.map.emit.await}, we start with 
the concrete program in \CEU (a), which communicates the value $1$ between the 
\code{emit} and \code{await} in parallel.
In the intermediate translation (b), we include the shared variable \code{e\_} 
to hold the value being communicated between the two trails in order to 
simplify the \code{emit}.
Finally, we convert the program into the equivalent in the abstract syntax (c), 
translating side-effect statements into $mem$ expressions.
External events have a similar translation, i.e., each external event requires 
a corresponding variable that is explicitly set by the environment before each 
reaction.

\begin{figure}[t]
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  emit e => 1;
with
  v = await e;
  _printf("%d\n",v);
end

.
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  e_ = 1;
  emit e;
with
  await e;
  v = e_;
  _printf("%d\n",v);
end
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
<...> ; mem ; emit(e)
        or
await(e) ; mem ; mem

.
\end{lstlisting}
\centering\small{(c)}
\end{minipage}
\caption{
Two-step translation from concrete to abstract \code{emit} and \code{await} 
expressions.
The concrete code in (a) communicates the value \code{1} from the \code{emit} 
to the \code{await}.
The abstract code in (c) uses a shared variable to hold the value.
\label{lst.map.emit.await}
}
\end{figure}

\subsubsection{First-class Timers}

To encompass first-class timers, we introduce a special external event 
\code{DT} that is intercalated with each other event occurrence in an 
application (e.g.  \emph{e1, e2}):

\begin{align*}
\LL [DT], p \RR
    & \xRightarrow[~~1~~]{*}
\LL [    ], p' \RR
\\
\LL [e1], p' \RR
    & \xRightarrow[~~2~~]{*}
\LL [  ], p'' \RR
\\
\LL [DT], p'' \RR
    & \xRightarrow[~~3~~]{*}
\LL [    ], p''' \RR
\\
\LL [e2], p''' \RR
    & \xRightarrow[~~4~~]{*}
\LL [  ], p'''' \RR
\\
& ...
\end{align*}

The event \code{DT} has an associated variable \code{DT\_} carrying the 
wall-clock time elapsed between two occurrences in sequence, as depicted by the 
two-step translation in Figure~\ref{lst.map.timers.2}.
In the concrete program (a), the variable \code{dt} holds the residual delta 
time (as described in Section~\ref{sec.ceu.timers}) after awaking from the 
timer.
In the first step of the translation (b), we expand the \code{await 10ms} to a 
\code{loop} that decrements the elapsed number of microseconds for each 
occurrence of \code{DT}.
When the variable \code{tot} reaches zero, we escape the \code{loop} setting 
the variable \code{dt} to contain the appropriate delta.
In the last step (c), we convert the program to the abstract syntax.

\begin{figure}
\begin{minipage}[t]{0.30\linewidth}
\begin{lstlisting}
dt = await 10ms;

.
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.37\linewidth}
\begin{lstlisting}
var int tot = 10000; // 10ms
loop do
    await DT;
    tot = tot - DT_;
    if tot <= 0 then
        dt = -tot;
        break;
    end
end
.
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%
\begin{minipage}[t]{0.30\linewidth}
\begin{lstlisting}
mem;
loop(
    await(DT);
    mem;
    if mem then
        mem;
        break
    else
        nop
)
\end{lstlisting}
\centering\small{(c)}
\end{minipage}
\caption{
Two-step translation from concrete to abstract timer.
\label{lst.map.timers.2}
}
\end{figure}

\subsubsection{Finalization Blocks}
\label{sec.formal.fins}

The biggest mismatch between concrete and abstract \CEU is regarding the 
\code{finalize} blocks, which require more complex modifications in the program
for a proper mapping using $fin$ expressions.
In the three-step translation in Figure~\ref{lst.map.fin}, we start with a 
concrete program (a) that uses a \code{finalize} to safely \code{\_release} the 
reference to \code{ptr} kept after the call to \code{\_hold}.
In the translation, we first need to catch the outermost \code{do-end} 
termination to run the finalization code.
For this, we translate the block into a \code{par/or} (b) with the original 
body in parallel with a $fin$ expression to run the finalization code.
Note that the $fin$ has no transition rules in the semantics, keeping the 
\code{par/or} alive.
This way, the $fin$ body only executes when the \code{par/or} terminates either 
normally (after the \code{await B}), or aborted from an outer composition.
%
However, the $fin$ still (incorrectly) executes even if the call to 
\code{\_hold} is not reached in the body due to an abort before awaking from 
the \code{await A}.
%
To deal with this issue, for each $fin$ we need a corresponding flag to keep 
track of code that needs to be finalized (c).
%
The flag is initially set to false, avoiding the finalization code to execute.
Only after the call to \code{\_hold} that we set the flag to true and enable 
the $fin$ body to execute.
%
The complete translation substitutes the side-effect operations with $mem$ 
expressions (d).

\begin{figure}
\begin{minipage}[t]{0.26\linewidth}
\begin{lstlisting}
do
  var int* ptr = <...>;
  await A;
  finalize
    _hold(ptr);
  with
    _release(ptr);
  end
  await B;
end

.
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.26\linewidth}
\begin{lstlisting}
par/or do
  var int* ptr = <...>;
  await A;
  _hold(ptr);
  await B;
with
  { fin
      _release(ptr); }
end

.
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%
\begin{minipage}[t]{0.26\linewidth}
\begin{lstlisting}
f_ = 0;
par/or do
  var int* ptr = <...>;
  await A;
  _hold(ptr);
  f_ = 1;
  await B;
with
  { fin
      if f_ then
        _release(ptr);
      end }
end
.
\end{lstlisting}
\centering\small{(c)}
\end{minipage}
%
\begin{minipage}[t]{0.18\linewidth}
\begin{lstlisting}
mem;
(
   mem;
   await(A);
   mem;
   mem;
   await(B);
or
   fin
     if mem then
       mem
     else
       nop
)
\end{lstlisting}
\centering\small{(d)}
\end{minipage}
%
\caption{
Three-step translation from concrete to abstract finalization.
\label{lst.map.fin}
}
\end{figure}

### 2018-02-01 04-related.tex
- Primeiro commit - Apenas compila - Semântica antiga (Tese e artigo TECS) - Texto novo (SBLP)

\section{Related Work}

\CEU was strongly influenced by Esterel but they differ in the most fundamental 
aspect of the notion of time~\cite{ceu.sensys13}.
%
Esterel defines time as a discrete sequence of logical unit instants or 
``ticks''.
At each tick, the program reacts to an arbitrary number of simultaneous input 
events from the environment.
%
The presence of multiple inputs requires careful static analysis to detect and
reject programs with \emph{causality cycles} and \emph{schizophrenia problems}%
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
In contrast, \CEU defines time as a discrete sequence of reactions to
unique input events.
%
In the formal semantics, ... \fs{como isso aparece na semantica} ...
\CEU also rejects some syntactically correct programs to avoid infinite
execution, but with simple restrictions in the abstract syntax tree.

Another distinction is that, in Esterel, the behavior of internal and external
events is equivalent, while in \CEU internal events introduce stack-based micro
reactions within external reactions, providing more fine-grained control for
intra-reaction execution.
%
In the formal semantics, ... \fs{como isso aparece na semantica} ...
%
Some variants of the Statecharts synchronous visual language also distinguish
internal from external events~\cite{statecharts.variants}.
In Statemate~\cite{statecharts.statemate}, \emph{``reactions to external and
internal events (...) can be sensed only after completion of the step''},
implying queue-based execution.
In Stateflow~\cite{statecharts.stateflow}, \emph{``the receiving state (of the
event) acts here as a function''}, which is similar to \CEU's
stack-based execution.
We are not aware of formalizations for these ideas for a deeper comparison with
\CEU.

Like other synchronous languages
(%
    \emph{Reactive C}~\cite{rp.rc},                 % 1991
    \emph{Protothreads}~\cite{wsn.protothreads},    % 2006
    \emph{SOL}~\cite{wsn.sol},                      % 2007
    \emph{SC}~\cite{rp.synchc},                     % 2009
    and
    \emph{PRET-C}~\cite{rp.pretc}%                  % 2010
),
\CEU relies on deterministic scheduling to preserve intra-reaction determinism.
%
In addition, it also performs concurrency checks to detect trails that, when
reordered, change the observable behavior of the program, i.e., trails that
actually rely on deterministic scheduling~\cite{ceu.sensys13}.
%
Esterel is only deterministic with respect to external behavior: \emph{``the
same sequence of inputs always produces the same sequence of 
outputs''}~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: \emph{``if there is no control dependency, as in
\code{(call~f1()~||~call~f2())},
the order is unspecified and it would be an error to rely on 
it''}~\cite{esterel.primer}.
%
For this reason, Esterel, does not support shared-memory concurrency:
\emph{``if a variable is written by some thread, then it can neither be read
nor be written by concurrent threads''}~\cite{esterel.primer}.

Esterel describes a finalization mechanism in a standardization
proposal~\cite{esterel.v7} that is similar to \CEU's.
However, we are not aware of an open implementation or a formal semantics for a
deeper comparison.

\fs{outras linguagens sincronas}
\fs{outras linguagens deterministicas}
\fs{outras linguagens com terminacao}

### 2018-02-01 05-conclusion.tex
- Primeiro commit - Apenas compila - Semântica antiga (Tese e artigo TECS) - Texto novo (SBLP)

\section{Conclusion}
\label{sec.conclusion}

\fs{TODO}

### 2018-02-01 index.tex
- Primeiro commit - Apenas compila - Semântica antiga (Tese e artigo TECS) - Texto novo (SBLP)

\documentclass[sigplan, anonymous, review]{acmart}

\input{header}
% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\acmDOI{10.475/123_4}

% ISBN
\acmISBN{123-4567-24-567/08/06}

%Conference
\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El
  Paso, Texas USA}
\acmYear{1997}
\copyrightyear{2016}

\acmPrice{15.00}

%\acmBadgeL[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
%\acmBadgeR[http://ctuning.org/ae/ppopp2016.html]{ae-logo}

\begin{document}

\title[A Deterministic and Terminating Semantics for \CEU]
{A Deterministic and Terminating Semantics
  for the Synchronous Programming Language~\CEU}

\author[R.\,C.\,M.~Santos]{Rodrigo C.\,M.~Santos}
\email{rsantos@inf.puc-rio.br}
\affiliation{PUC-Rio, Rio de Janeiro, Brazil}
%
\author[G.\,F.~Lima]{Guilherme F.~Lima}
\email{glima@inf.puc-rio.br}
\affiliation{PUC-Rio, Rio de Janeiro, Brazil}
%
\author[F.~Sant'Anna]{Francisco Sant'Anna}
\email{francisco@ime.uerj.br}
\affiliation{UERJ, Rio de Janeiro, Brazil}
%
\author[R.~Ierusalimschy]{Roberto Ierusalimschy}
\email{roberto@inf.puc-rio.br}
\affiliation{PUC-Rio, Rio de Janeiro, Brazil}
%
\author[E.\,H.~Haeusler]{Edward Hermann Haeusler}
\email{hermann@inf.puc-rio.br}
\affiliation{PUC-Rio, Rio de Janeiro, Brazil}

\begin{abstract}
\CEU is a synchronous programming language for embedded soft real-time systems.
%
It supports advanced features, such as safe shared-memory concurrency and
abortion of execution lines, while enforcing internal determinism and
time-bounded reactions.
%
In this paper, we present a small-step structural operational semantics for
\CEU and prove that under it program reactions are deterministic and always
terminate.
%
Thus, given the same timeline of input events, the same \CEU program in the
same initial memory state will always react in bounded time and arrive at the
same final memory state.
%
We also discuss some equivalence results and the relation between the proposed
semantics and its actual implementation.
\end{abstract}

\terms{Languages, design, theory}
\keywords{Operational semantics, \CEU, Synchronous languages}

\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10003752.10010124.10010131.10010134</concept_id>
  <concept_desc>Theory of computation~Operational semantics</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10011007.10011006.10011008.10011009.10011014</concept_id>
  <concept_desc>Software and its
                engineering~Concurrent programming languages</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010562.10010564</concept_id>
  <concept_desc>Computer systems
                organization~Embedded software</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[500]{Theory of computation~Operational semantics}
\ccsdesc[500]{Software and its engineering~Concurrent programming languages}
\ccsdesc[300]{Computer systems organization~Embedded software}

\maketitle

\input{01-intro}
\input{02-ceu}
\input{03-semantics}
\input{04-related}
\input{05-conclusion}

\bibliographystyle{ACM-Reference-Format}
\bibliography{my,other}

\appendix
%\include{proofs}

\end{document}

### 2018-02-02 index.tex
Abstract reescrito. Tirei a frase com "some equivalence results". Guilherme, que equivalências são essas?

@@
 %
-It supports advanced features, such as safe shared-memory concurrency and
-abortion of execution lines, while enforcing internal determinism and
-time-bounded reactions.
+It focus on control-flow safety features, such as safe shared-memory
+concurrency and safe abortion of lines of execution, while enforcing internal
+determinism and time-bounded reactions to the environment.
 %
-In this paper, we present a small-step structural operational semantics for
-\CEU and prove that under it program reactions are deterministic and always
-terminate.
+In this work, we present a small-step structural operational semantics for
+\CEU and a proof that reactions are deterministic and always terminate:
 %
-Thus, given the same timeline of input events, the same \CEU program in the
-same initial memory state will always react in bounded time and arrive at the
-same final memory state.
+For a given timeline of input events, multiple executions of the same program
+always react in bounded time and arrive at the same final memory state.
 %
-We also discuss some equivalence results and the relation between the proposed
-semantics and its actual implementation.
+%We also discuss some equivalence results and the relation between the proposed
+%semantics and its actual implementation.
 \end{abstract}

### 2018-02-02 01-intro.tex
Revisão da Introdução.

@@
 
-\CEU~\cite{ceu.sensys13} is a programming language for embedded soft real-time
-systems that aims to offer a concurrent and expressive alternative to C and C++
-with the characteristics that follow:
-
+\CEU~\cite{ceu.sensys13,ceu.tecs17} is a Esterel-based~\cite{esterel.ieee91}
+programming language for embedded soft real-time systems that aims to offer a
+concurrent, safe, and expressive alternative to C with the characteristics that
+follow:
+%
 \begin{description}
@@
 \end{description}
+%
+Structured reactive programming let developers write code in direct/sequential
+style, recovering from the inversion of control imposed by event-driven
+execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
+%
+Synchronous languages offer a simple run-to-completion model of execution that
+enable deterministic execution and make formal reasoning tractable.
+For this reason, it has been successfully adopted in safety-critical real-time
+embedded systems.~\cite{rp.twelve}
 
-Structured reactive programming eliminates the
-\emph{callback hell}~\cite{TODO}, letting programmers write code in
-direct/sequential style \cite{TODO}.
+Previous work in the context of embedded sensor networks evaluates the
+expressiveness of \CEU in comparison to event-driven code in C and attests a
+reduction in source code size (around 25\%) with a small increase in memory
+usage (around 5--10\% for \emph{text} and \emph{data})~\cite{ceu.sensys13}.
 %
-Previous work evaluates the expressiveness of \CEU in comparison to
-event-driven code in C and attests a reduction in source code size (around
-25\%) with a small increase in memory usage (around 5--10\% for \emph{text} and 
-\emph{data})~\cite{ceu.sensys13}.
+\CEU has also been used in the context of multimedia
+systems~\cite{ceumedia.webmedia16} and games~\cite{ceu.mod15}, and as an
+alternative language in an undergraduate-level course on embedded systems for
+the past 6 years.
+
+\CEU inherits the synchronous and imperative mindset of Esterel but adopts a
+simpler semantics with fine-grained execution control.~\cite{ceu.tecs17}
 %
-\CEU has been used in the context of wireless sensor
-networks~\cite{ceu.sensys13,ceu.terra},
-multimedia systems~\cite{ceu.media.webmedia16}, and
-games \footnote{\fs{TODO}}.
-It has also been used as an alternative language in an undergradutate-level
-course on embedded systems for the past 5 years.
-
-\fs{Embedded systems requirements, determinism, termination}
-
+The list that follows summarizes the semantic peculiarities of \CEU:
+%
 \begin{itemize}
-\item \CEU
-    \begin{itemize}
-    \item variaveis compartilhadas
-    \item concorrencia segura
-    \item terminacao e determinismo
-    \end{itemize}
-\item Modelo Sincrono
-    \begin{itemize}
-    \item conhecido como ammenable to verification
-    \item valido para \CEU?
-    \end{itemize}
+    \item Stack-based execution for internal events, which provides a limited
+          form of coroutines.
+    \item Fine-grained, intra-reaction deterministic execution, which allows
+          programs to safely share memory.
+    \item Finalization mechanism for abortion of lines of execution, which
+          safely release external resources.
+    \item First-class synchronized timers.
 \end{itemize}
 
-%only the static semantics for embedded systems
+In this work, we present a formal semantics for a subset of \CEU that focus on
+its peculiarities in comparison to other synchronous languages.
+\begin{itemize}
+    \item qual a abordagem / operational semantics / dois passos
+    \item quais os resultados / provas
+    \item quais os desafios e limitações
+    \item \gl{TODO}
+\end{itemize}
+
+\fs{Descrever seções.}

### 2018-02-02 index.tex
Revisão da Introdução.

@@
 %
-For a given timeline of input events, multiple executions of the same program
-always react in bounded time and arrive at the same final memory state.
+For a given arbitrary timeline of input events, multiple executions of the same
+program always react in bounded time and arrive at the same final memory state.
 %

### 2018-02-02 02-ceu.tex
Início da Seção 2 revisado: - tirei "output" - mudei o exemplo para usar "input" e "finalize" - falei sobre o mecanismo ortogonal de abortamento - faltou falar sobre timers, por alguma razão não está em lugar nenhum do artigo (falta de espaço?)

@@
 
-\CEU is a synchronous reactive language inspired by
-Esterel~\cite{esterel.ieee91} in which programs advance in a sequence of
-discrete reactions to external events.
+\CEU is a synchronous reactive language in which programs advance in a sequence
+of discrete reactions to external events.
 %
-\CEU is designed for control-intensive applications, supporting concurrent
-lines of execution, known as \emph{trails}, and broadcast communication
+It is designed for control-intensive applications, supporting concurrent lines
+of execution, known as \emph{trails}, and instantaneous broadcast communication
 through events.
@@
 Computations within a reaction (such as expressions, assignments, and
-C~calls) execute in no time in accordance to the synchronous
+system calls) are also instantaneous in accordance to the synchronous
 hypothesis~\cite{rp.hypothesis}.
 %
-The \code{await} is the only \CEU statement that actually ``consumes'' time.
-%%
-An~\code{await} statement blocks the current trail allowing the program to
-advance its other trails; when all trails are blocked, the reaction
+\CEU provides an \code{await} statement which is the only that actually
+``consumes'' time.
+%
+An \code{await} statement blocks the current running trail allowing the program
+to advance its other trails; when all trails are blocked, the reaction
 terminates and control returns to the environment.
 
-In \CEU, every execution path within a loop must contain at least one
+In \CEU, every execution path within loops must contain at least one
 \code{await} statement~\cite{ceu.sensys13,esterel.primer}.
 %
-This restriction, which is checked statically by the \CEU compiler, ensures
-that every reaction runs in bounded time, eventually terminating with all
-program trails blocked in \code{await} statements.
+This restriction, which is statically checked by the compiler, ensures that
+every reaction runs in bounded time, eventually terminating with all trails
+blocked in \code{await} statements.
 %
-\CEU has a further restriction which it shares with Esterel and synchronous
-languages in general~\cite{esterel.preemption}: computations that actually
+\CEU has an additional restriction, which it shares with Esterel and
+synchronous languages in general~\cite{esterel.preemption}: computations that
 take a non-negligible time to run (e.g., cryptography or image processing
@@
 
-Listing~\ref{lst.syntax} below shows a compact reference of~\CEU.
+Listing~\ref{lst.syntax} shows a compact reference of~\CEU:
 
@@
 ]
-|\C{1.}{Declarations:}|
-input  $\T<type>$ $\T<ids>$;|\C{.6}{declare external input events}|
-output $\T<type>$ $\T<ids>$;|\C{.6}{declare external output events}|
-event  $\T<type>$ $\T<ids>$;|\C{.6}{declare internal events}|
-var    $\T<type>$ $\T<id>$ = $\T<exp>$;|\C{.6}{declare and initialize variable}|
+|\C{1.}{Declarations}|
+input $\T<type>$ $\T<id>$; | \C{.6}{declares external input event} |
+event $\T<type>$ $\T<id>$; | \C{.6}{declares internal event}       |
+var   $\T<type>$ $\T<id>$; | \C{.6}{declares variable}             |
 
-|\C{1.}{Event handling:}|
-$\T<id>$ = await $\T<id>$;|\C{.6}{await event and assign the received value}|
-$\T<id>$ = await $\T<time>$;|\C{.6}{await time and assign the delayed delta}|
-emit $\T<id>$($\T<exp>$);|\C{.6}{emit event passing a value}|
+|\C{1.}{Event handling}|
+$\T<id>$ = await $\T<id>$;   | \C{.6}{awaits event and assigns received value} |
+$\T<id>$ = await $\T<time>$; | \C{.6}{awaits time and assigns delayed delta}   |
+emit $\T<id>$($\T<exp>$);    | \C{.6}{emits event passing value}               |
 
-|\C{1.}{Control-flow:}|
-$\T<stmt>$ ; $\T<stmt>$|\C{.45}{sequence}|
-if $\T<exp>$ then $\T<stmts>$ else $\T<stmts>$ end|\C{.45}{conditional}|
-loop do $\T<stmts>$ end|\C{.45}{repetition}|
-finalize $\T<stmts>$ with $\T<stmts>$ end|\C{.45}{finalization}|
+|\C{1.}{Control flow}|
+$\T<stmt>$ ; $\T<stmt>$                       | \C{.45}{sequence}     |
+if $\T<exp>$ then $\T<stmts>$ else $\T<stmts>$ end | \C{.45}{conditional}  |
+loop do $\T<stmts>$ end                   | \C{.45}{repetition}   |
+finalize [$\T<stmts>$] with $\T<stmts>$ end    | \C{.45}{finalization} |
 
-par/or  do $\T<stmts>$ with $\T<stmts>$ end|\C{.45}{aborts when any side terminates}|
-par/and do $\T<stmts>$ with $\T<stmts>$ end|\C{.45}{terminates when all sides terminate}|
-par     do $\T<stmts>$ with $\T<stmts>$ end|\C{.45}{never terminates}|
+|\C{1.}{Logical parallelism}|
+par/or  do $\T<stmts>$ with $\T<stmts>$ end | \C{.45}{aborts when any side ends}      |
+par/and do $\T<stmts>$ with $\T<stmts>$ end | \C{.45}{terminates when all sides ends} |
+par     do $\T<stmts>$ with $\T<stmts>$ end | \C{.45}{never terminates}               |
 
-|\C{1.}{Assignment \& integration with C:}|
-$\T<id>$ = $\T<exp>$;|\C{.45}{assign value to variable}|
-_$\T<id>$($\T<exps>$)|\C{.45}{call C function (id starts with `\_'\,)}|
+|\C{1.}{Assignment \& Integration with C}|
+$\T<id>$ = $\T<exp>$; | \C{.6}{assigns value to variable}                |
+_$\T<id>$($\T<exps>$) | \C{.6}{calls C function (id starts with `\_'\,)} |
 \end{lstlisting}
@@
 
-To make matters concrete, consider the program of Listing~\ref{lst.ceu}.
+Listing~\ref{lst.ceu} shows a complete example in \CEU that blinks a LED with a
+frequency of 1 second, terminating with a button press always with the LED off.
 %
-This program continuously turns a LED~on for~2 seconds and off for~1 second,
-and terminates after~1 minute of activity with the LED off.
+The implementation first declares the \code{BUTTON} as an input event (ln.~1).
+Then, it uses a \code{par/or} composition to run two activities in parallel:
+a single statement that waits for a button press before terminating (ln.~3),
+and an endless loop that blinks the LED on and off (ln.~8--13).
+The \code{finalize} clause (ln.~5--7) ensures that, no matter how its enclosing
+trail terminates, the LED will be unconditionally turned off (ln.~6).
+
+The \code{par/or} composition, which stands for a \emph{parallel-or}, provides
+an orthogonal abortion mechanism in which its composed trails do not know when
+and how they are aborted (i.e., abortion is external to them).
 %
-The implementation uses a \code{par/or} to run two activities in parallel:
-an endless loop that blinks the LED on and off (lines~2--7), and a single
-statement that waits for~1 minute before terminating (line~9).
+This is possible to do safely in synchronous languages due to the accurate
+control of concurrent activities, i.e., in between every reaction, the whole
+system is idle and consistent~\cite{esterel.preemption}.
 %
-The \code{par/or} block stands for a \emph{parallel-or} composition; when
-executed it creates~$n$ parallel trails (in this case, $n=2$) and rejoins
-them when any of these~$n$ trails terminates, automatically aborting the
-other trails.
+The finalization mechanism extends orthogonal abortion to also work with
+activities that use stateful resources from the environment (such as files and
+network handlers), as we discuss in Section~\ref{sec.ceu.fin}.
+%
+\fs{First-class timers.}
 
@@
   float=h,
-  caption={\CEU program that blinks a LED during 1 minute.},
+  caption={A program in \CEU that blinks a LED every second, terminating on a
+           button press in a consistent state.},
   label={lst.ceu},
 ]
+input void BUTTON;
 par/or do
+    await BUTTON;
+with
+    finalize with
+        _led(0);
+    end
     loop do
         _led(1);
-        await 2s;
+        await 1s;
         _led(0);
@@
     end
-with
-    await 1min;
 end
-_led(0);
 \end{lstlisting}
@@
 %
-To ensure that programs operate synchronously, the compiler environment
-should only provide access to~C operations that are assumed run in zero
-time, such as non-blocking~I/O and access to \code{struct}'s.
+To ensure that programs operate under the synchronous hypothesis, the compiler
+environment should only provide access to~C operations that can be assumed to
+be instantaneous, such as non-blocking~I/O and simple accesses to data
+structures.
 
@@
 In Listing~\ref{lst.sub}, the ``subroutine'' \code{inc} is defined as a loop
-(lines~3--6) that continuously awaits its identifying event (line~4), and
-increments the value passed to it by reference (lines~5).
+(ln.~3--6) that continuously awaits its identifying event (ln.~4), and
+increments the value passed to it by reference (ln.~5).
 %
-A trail in parallel (lines~8--11) invokes the subroutine through the
-\code{emit} statement at line~10.
+A trail in parallel (ln.~8--11) invokes the subroutine through an \code{emit}
+statement (ln.~10).
 %
-Given the stack-based execution mode of internal events, after the emit
-statement at line~10 is executed, the calling trail pauses, the subroutine
-awakes (line~4), runs its body (yielding \code{v=2}), loops, and awaits the
-next ``call'' (line~4, again).
+Given the stack-based execution mode of internal events, as the emit executes,
+the calling trail pauses, the subroutine awakes (ln.~4), runs its body
+(yielding \code{v=2}), loops, and awaits the next ``call'' (ln.~4, again).
 %
 Only after this sequence does the calling trail resumes and moves on to
-execute the assertion on line~11.
+execute the assertion (ln.~11).
 
@@
 For instance, the body of the subroutine \code{inc} in Listing~\ref{lst.sub}
-could \code{emit} another internal event after awaking (line~4), creating a
+could \code{emit} another internal event after awaking (ln.~4), creating a
 new level in the stack.
@@
 \begin{lstlisting}[numbers=left,xleftmargin=3em]
-input void A, B;
+input void A;
+input void B;
 var int x = 1;
@@
 \subsection{Abortion and Finalization}
+\label{sec.ceu.fin}
 

### 2018-02-02 02-ceu.tex
Revisão 2.1

@@
 %
-These external input events are received from the environment; each of them
-delimits a new logical unit of time and triggers a corresponding reaction.
+External input events are received from the environment, and each delimits a
+new logical unit of time that triggers an associated reaction.
 %
 The life-cycle of a program in \CEU can be summarized as
-follows~\cite{ceu.sensys13}.
+follows~\cite{ceu.sensys13}:
 %
-\begin{enumerate}
-\item The program initiates the ``boot reaction'' in a single trail
-  (parallel constructs may create new trails).
-  %
+\begin{enumerate}[i]
+\item The program initiates a ``boot reaction'' in a single trail (parallel
+      constructs may create new trails).
 \item Active trails execute until they await or terminate, one after
@@
       bounded time.
-  %
-\item When all its trails are blocked, the program goes idle and the
-  environment takes control.
-  %
+\item When all trails are blocked, the program goes idle and the environment
+      takes control.
 \item On the occurrence of a new external input event, the environment
       awakes \emph{all} trails awaiting that event, and the program goes back to
-  step~(ii).
+      step~(i).
 \end{enumerate}
@@
 %
-By the synchronous hypothesis, the time the program spends in step~(ii)
-above is conceptually zero (in practice, negligible).
+By the synchronous hypothesis, the time the program spends in step~(ii) is
+conceptually zero (in practice, negligible).
 %
-Thus, from the point of view of the environment, the program is always
-idle---on step~(iii).
+Hence, from the point of view of the environment, the program is always
+idle on step~(iii).
 %
-In practice, if a new external input event occurs while a reaction is being
-computed, the event is saved on a queue, which effectively schedules it to
-be processed by a subsequent reaction.
+In practice, if a new external input event occurs while a reaction executes,
+the event is saved on a queue, which effectively schedules it to be processed
+in a subsequent reaction.
 
@@
 
-The processing of external input events induces a discrete notion of time in
-\CEU.
-%
-Figure~\ref{fig.ticks} illustrates this notion.
+The sequential processing of external input events induces a discrete notion of
+time in \CEU, as illustrated in Figure~\ref{fig.ticks}.
 %
@@
 %
-The boot reaction \code{boot-0} happens before any input on program startup.
+The boot reaction \code{boot-0} happens before any input, at program startup.
 %
 Event~\code{A} ``physically'' occurs during \code{boot-0} but, because time
-is discrete, its corresponding reaction can only execute afterwards, at
-logical instant~\code{A-1}.
+is discrete, its corresponding reaction only executes afterwards, at logical
+instant~\code{A-1}.
 %
@@
 
-In practice, the synchronous hypothesis for \CEU holds if reaction rate is
+In practice, the synchronous hypothesis for \CEU holds if reactions execute
 faster than the rate of incoming input events.
 %
-Otherwise, the program continuously accumulates a delay between the real
-occurrence time and actual reaction time of events.
+Otherwise, the program would continuously accumulate delays between physical
+occurrences and actual reactions for the input events.
 %
@@
 application does not take too long to catch up with real time.
+%
+Note that the synchronous semantics is the norm in typical event-driven
+systems, such as event dispatching in UI toolkits, game loops in game engines,
+and clock ticks in embedded systems.
 

### 2018-02-02 02-ceu.tex
Revisão 2.1.1: - adicionado event iterator "every", vai ser importante para forçar terminação

@@
 loop do $\T<stmts>$ end                   | \C{.45}{repetition}   |
+every $\T<id>$ in $\T<id>$ do $\T<stmts>$ end      | \C{.45}{event iteration}   |
 finalize [$\T<stmts>$] with $\T<stmts>$ end    | \C{.45}{finalization} |
@@
 
-In \CEU, the queue-based processing of events described previously applies
-only to external input events, i.e., those events submitted to the program
-by the environment.
+In \CEU, queue-based processing of events applies only to external input
+events, i.e., events submitted to the program by the environment.
 %
-Internal events, which are those events generated internally by the program
-via \code{emit} statements, are processed in a stack-based manner.
+Internal events, which are events generated internally by the program via
+\code{emit} statements, are processed in a stack-based manner.
 %
-These internal events provide a fine-grained execution control and, because
-of their stack-based processing, can be used to implement a limited form of
-subroutine, as illustrated in Listing~\ref{lst.sub} below.
+Internal events provide a fine-grained execution control, and, because of their
+stack-based processing, can be used to implement a limited form of subroutines,
+as illustrated in Listing~\ref{lst.sub}:
 
@@
 ]
-event int* inc;     // subroutine |`|inc|'|
+event int* inc;         // declares subroutine "inc"
 par/or do
-    loop do         // definitions are loops
-        var int* p = await inc;
+    var int* p;
+    every p in inc do   // implements subroutine with an event iterator
         *p = *p + 1;
@@
     var int v = 1;
-    $\cdots$
-    emit inc(&v);   // call |`|inc|'|
-    _assert(v==2);  // after return
+    emit inc(&v);       // calls "inc"
+    emit inc(&v);       // calls "inc"
+    _assert(v==3);      // asserts result after returns
 end
@@
 
-In Listing~\ref{lst.sub}, the ``subroutine'' \code{inc} is defined as a loop
-(ln.~3--6) that continuously awaits its identifying event (ln.~4), and
-increments the value passed to it by reference (ln.~5).
+In the example, the ``subroutine'' \code{inc} is defined as an event iterator
+(ln.~4--6) that continuously awaits its identifying event (ln.~4), and
+increments the value passed by reference (ln.~5).
 %
-A trail in parallel (ln.~8--11) invokes the subroutine through an \code{emit}
-statement (ln.~10).
+A trail in parallel (ln.~8--11) invokes the subroutine through two consecutive
+\code{emit} statements (ln.~9--10).
 %
-Given the stack-based execution mode of internal events, as the emit executes,
-the calling trail pauses, the subroutine awakes (ln.~4), runs its body
-(yielding \code{v=2}), loops, and awaits the next ``call'' (ln.~4, again).
+Given the stack-based execution for internal events, as the first emit
+executes, the calling trail pauses (ln.~9), the subroutine awakes (ln.~4), runs
+its body (yielding \code{v=2}), iterates, and awaits the next ``call'' (ln.~4,
+again).
 %
-Only after this sequence does the calling trail resumes and moves on to
-execute the assertion (ln.~11).
+Only after this sequence does the calling trail resumes (ln.~9), makes a new
+invocation (ln.~10), and passes the assertion test (ln.~11).
+
+%- multiple executions
+%- footnote to distinguish from TECS
 
@@
 We can think of the stack as a record of the nested, fine-grained internal
-reactions that happen inside the same bigger reaction to some external
+reactions that happen inside the same outer reaction to a single external
 event.
 
-This form of subroutine has a significant limitation though: it cannot
-express recursion (an \code{emit} to itself is always ignored as a running
-trail cannot be waiting on itself).
+This form of subroutines has a significant limitation though: it cannot
+express recursion, since an \code{emit} to itself is always ignored as a
+running trail cannot be waiting on itself.
 %
-That said, it is this very limitation that brings important safety
+That being said, it is this very limitation that brings important safety
 properties to subroutines.
 %
-First, such subroutines are guaranteed to react in bounded time.
+First, they guaranteed to react in bounded time.
 %
 Second, memory for locals is also bounded, not requiring data stacks.
+
+At first sight, an event iterator for event \code{A} seems like a syntactic
+sugar for a \code{loop} followed by and \code{await A}.
+However, event iterators have an important restriction to enforce reaction
+termination, as we discuss in Section~\ref{sec.sem}:
+    they cannot contain \code{break} statements.
+
+\begin{comment}
 %
@@
 %
-%In Section~\ref{sec.adv.excpt} we show how to use them to implement
-%exceptions.
-%
-%8<- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-\gl{Revisei do início da Seção~2 até aqui.}
-%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - ->8
+\end{comment}
 

### 2018-02-02 02-ceu.tex
Revisão Seção 2.2

@@
 \subsection{Shared-Memory Concurrency}
-    - referenciar warnings
 
@@
 \begin{minipage}[h]{0.45\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
+\begin{lstlisting}[numbers=right]
 input void A;
@@
 %
-\begin{minipage}[h]{0.45\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
+\begin{minipage}[h]{0.53\linewidth}
+%\begin{lstlisting}
+\begin{lstlisting}[xleftmargin=2em]
 input void A;
+//
 var int y = 1;
@@
 end
+
 \end{lstlisting}
@@
 For instance, consider the two examples in Figure~\ref{lst.shared}, both
-defining a shared variable (ln. 2), and assigning to it in parallel trails (ln.
-5, 8).
+defining shared variables (ln. 3), and assigning to them in parallel trails
+(ln. 6, 9).
 
@@
 %
-However, that an (apparently innocuous) change in the order of trails modifies
-the behavior of the program.
+However, note that an apparently innocuous change in the order of trails
+modifies the behavior of the program.
 %

### 2018-02-02 02-ceu.tex
(>) Revisão 2.3

@@
 The \code{par/or} composition, which stands for a \emph{parallel-or}, provides
-an orthogonal abortion mechanism in which its composed trails do not know when
-and how they are aborted (i.e., abortion is external to them).
+an orthogonal abortion mechanism~\cite{esterel.preemption} in which its
+composed trails do not know when and how they are aborted (i.e., abortion is
+external to them).
 %
@@
 
-The \code{par/or} of \CEU is an \emph{orthogonal abortion
-mechanism}~\cite{esterel.preemption} because the two sides in the composition
-need not be tweaked with synchronization primitives or state variables in order
-to affect each other.
-%
-In addition, abortion is \emph{immediate} in the sense that it executes
-atomically in the current micro reaction.
-%
-Immediate orthogonal abortion is a distinctive feature of synchronous languages
-and cannot be expressed effectively in traditional (asynchronous)
-multi-threaded languages~\cite{esterel.preemption,sync_async.threadsstop}.
-
-However, aborting lines of execution that deal with external resources may lead
-to inconsistencies.
-%
-For this reason, \CEU provides a \code{finalize} construct to unconditionally
-execute a series of statements even if the enclosing block is aborted.
-
-\CEU also enforces the use of \code{finalize} for system calls that deal with
-pointers representing resources, as illustrated in the two examples of
-Figure~\ref{lst.fin.ceu}:
-%
-\begin{itemize}
-\item If \CEU \textbf{passes} a pointer to a system call (ln. \ax:5), the
-pointer represents a \textbf{local} resource (ln. \ax:2) that requires finalization
-(ln. \ax:7).
-\item If \CEU \textbf{receives} a pointer from a system call return (ln. \bx:4),
-the pointer represents an \textbf{external} resource (ln. \bx:2) that requires
-finalization (ln. \bx:6).
-\end{itemize}
-%
-\CEU tracks the interaction of system calls with pointers and requires
-finalization clauses to accompany them.
-%
-In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
-(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request}
-(ln. 5), which is an asynchronous call that transmits the buffer in the
-background.
-If the block aborts (ln. 11) before receiving an acknowledge from the
-environment (ln. 9), the local \code{msg} goes out of scope and the external
-transmission now holds a \emph{dangling pointer}.
-The finalization ensures that the transmission also aborts (ln. 7).
-%
-In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
-4) returns an external file resource as a pointer.
-If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file
-remains open as a \emph{memory leak}.
-The finalization ensures that the file closes properly (ln. 6).
-%
-In both cases, the code does not compile without the \code{finalize}
-construct.%
-\footnote{
-The compiler only forces the programmer to write the finalization clause, but
-cannot check if it actually handles the resource properly.
-}
-
-The finalization mechanism of \CEU is fundamental to preserve the orthogonality
-of the \code{par/or} construct since the clean up code is encapsulated in the
-aborted trail itself.
-
 \begin{figure}
-\begin{minipage}[t]{0.48\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
+\begin{minipage}[t]{0.45\linewidth}
+\begin{lstlisting}[numbers=right]
 par/or do
-   var _buffer_t msg;
+   var _msg_t msg;
    <...> // prepare msg
    finalize
-      _send_request(&msg);
+      _send(&msg);
    with
-      _send_cancel(&msg);
+      _cancel(&msg);
    end
@@
 end
-.
+//
 \end{lstlisting}
@@
 %
-\begin{minipage}[t]{0.48\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
+\begin{minipage}[t]{0.53\linewidth}
+\begin{lstlisting}[xleftmargin=2em]
 par/or do
@@
 \end{figure}
+
+The \code{par/or} of \CEU is an orthogonal abortion mechanism because the two
+sides in the composition need not be tweaked with synchronization primitives or
+state variables in order to affect each other.
+%
+In addition, abortion is \emph{immediate} in the sense that it executes
+atomically in the current micro reaction.
+%
+Immediate orthogonal abortion is a distinctive feature of synchronous languages
+and cannot be expressed effectively in traditional (asynchronous)
+multi-threaded languages~\cite{esterel.preemption,sync_async.threadsstop}.
+
+However, aborting lines of execution that deal with external resources may lead
+to inconsistencies.
+%
+For this reason, \CEU provides a \code{finalize} construct to unconditionally
+execute a series of statements even if the enclosing block is aborted.
+
+\CEU also enforces the use of \code{finalize} for system calls that deal with
+pointers representing resources, as illustrated in the two examples of
+Figure~\ref{lst.fin.ceu}:
+%
+\begin{itemize}
+    \item If \CEU \textbf{passes} a pointer to a system call (ln. \ax:5), the
+          pointer represents a \textbf{local} resource (ln. \ax:2) that
+          requires finalization (ln. \ax:7).
+    \item If \CEU \textbf{receives} a pointer from a system call return
+          (ln.  \bx:4), the pointer represents an \textbf{external} resource
+          (ln. \bx:2) that requires finalization (ln. \bx:6).
+\end{itemize}
+%
+\CEU tracks the interaction of system calls with pointers and requires
+finalization clauses to accompany them.
+%
+In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
+(ln. 2) is an internal resource passed as a pointer to \code{_send} (ln. 5),
+which is an asynchronous call that transmits the buffer in the background.
+If the block aborts (ln. 11) before receiving an acknowledge from the
+environment (ln. 9), the local \code{msg} goes out of scope and the external
+transmission now holds a \emph{dangling pointer}.
+The finalization ensures that the transmission also aborts (ln. 7).
+%
+In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{_fopen} (ln.
+4) returns an external file resource as a pointer.
+If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file
+remains open as a \emph{memory leak}.
+The finalization ensures that the file closes properly (ln. 6).
+%
+In both cases, the code does not compile without the \code{finalize}
+construct.%
+\footnote{
+The compiler only forces the programmer to write the finalization clause, but
+cannot check if it actually handles the resource properly.
+}
+
+The finalization mechanism of \CEU is fundamental to preserve the orthogonality
+of the \code{par/or} construct since the clean up code is encapsulated in the
+aborted trail itself.

### 2018-02-07 03-semantics.tex
Semantics: major review

@@
-\newcommand{\ST}{\1\xrightarrow[~n~]{}\1}
+\newcommand{\NST}{\1\xrightarrow[nst]\1}
+\newcommand{\OUT}{\1\xrightarrow[out]\1}
 \newcommand{\LL}{\langle}
@@
 
-\begin{figure}
-\begin{lstlisting}[mathescape=true]
+\begin{lstlisting}[
+  %numbers=left,
+  basicstyle=\ttfamily\footnotesize,
+  float=h,
+  caption={Reduced syntax of \CEU.},
+  label={lst.formal.syntax},
+  mathescape=true
+]
                                    // primary expressions
   p ::= mem(id)                    (any memory access to `id')
-      $|$ await(id)                  (await event `id')
-      $|$ emit(id)                   (emit event `id')
+      $|$ awaitExt(id)               (await external event `id')
+      $|$ awaitInt(id)               (await internal event `id')
+      $|$ emitInt(id)                (emit internal event `id')
       $|$ break                      (loop escape)
@@
                                    // derived by semantic rules
-      $|$ awaiting(id,n)             (awaiting `id' since sequence number `n')
-      $|$ emitting(n)                (emitting on stack level `n')
-      $|$ p @ loop p                 (unwinded loop)
-      $|$ nop                        (terminated expression)
+      $|$ p @loop p                  (unwinded loop)
+      $|$ p @and q                   (unwinded par/and)
+      $|$ p @or q                    (unwinded par/or)
+      $|$ @canrun(n)                 (can run on stack level `n')
+      $|$ @nop                       (terminated expression)
 \end{lstlisting}
-\caption{
-    Reduced syntax of \CEU.
-\label{lst.formal.syntax}
-}
-\end{figure}
 
-Figure~\ref{lst.formal.syntax} shows the syntax for a subset of \CEU that is 
+Listing~\ref{lst.formal.syntax} shows the syntax for a subset of \CEU that is 
 sufficient to describe all semantic peculiarities of the language.
 %
-Except for $fin$ and the expressions used internally by the semantics (i.e., 
-$awating$, $emitting$, $p~@~loop; p$, and $nop$), all other expressions are 
-equivalent to their counterparts in the concrete language.
+Except for $fin$ and the expressions used internally by the semantics, which
+are prefixed with a character $@$, all other expressions are equivalent to
+their counterparts in the concrete language.
 
@@
 %
-Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
-an identifier is either a variable or an event.
+Note that $mem$, $awaitExt$, and $awaitInt$ / $emitInt$ expressions do not
+share identifiers, i.e., an identifier is either a variable, an external event,
+or an internal event.
 
@@
 %
-Each reaction is identified by a ever-increasing $n$ that remains constant 
-during the entire reaction.
-The transition rules map a program $p$ and a stack of events $S$ in a single 
-step to a modified program and stack:
+The transition rules map a triple with a program $p$, a stack level $n$, and an
+emitted event $e$ to a modified triple as follows:
 %
 \begin{align*}
-\LL S, p \RR &\ST
-\LL S', p' \RR
-    & \textbf{(rule-inner)}
+\LL p,  n,  e  \RR &\1\xrightarrow\1
+\LL p', n', e' \RR
+    & \textbf{(rule-general)}
 \end{align*}
@@
 \begin{align*}
-S, S' &\in id^*
-    &&(stack~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
 p, p' &\in P
-    && (program~as~described~in~Figure~\ref{lst.formal.syntax}) \\
-n     &\in \mathds{N}
-    && (unique~identifier~for~the~entire~reaction)
+    && (program~as~described~in~Listing~\ref{lst.formal.syntax})
+\\
+n, n' &\in \mathds{N}
+    && (current~stack~level)
+\\
+e, e' &\in \mathds{E \cup \{\epsilon\}}
+    && (emitted~event,~possibly~none)
 \end{align*}
 %
-At the beginning of a reaction, the stack is initialized with the occurring 
-external event $ext$ ($S=[ext]$), but $emit$ expressions can push new events on 
-top of it (we discuss how they are popped further).
-The sequence number $n$, which is incremented each reaction, prevents that 
-awaiting expressions awake in the same reaction they are reached (the 
-\emph{delayed awaits} as explained in Section~\ref{sec.ceu.ints}).
+At the beginning of a reaction to an input event, the triple is initialized at
+stack level $0$ ($n=0$) and with the emitted event $id$ ($e=id$).
+%, but \code{emitInt} expressions can increase the stack level.
+At the end of a reaction, after a number of transitions, the triple will block
+with a (possibly) modified program $p'$, at stack level $0$, with no event
+emitted $\epsilon$:
+%
+\begin{align*}
+\LL p,0,ext \RR
+    & \1\xrightarrow{*}\1
+\LL p',0,\epsilon \RR
+\end{align*}
+%
+We also define two transition rule subtypes $\OUT$ (for \emph{outermost}) and
+$\NST$ (for \emph{nested}).
 
-The transition rules for the primary expressions are as follows:
+The $\OUT$ rules \textbf{push} and \textbf{pop} are non-recursive definitions
+that manipulate the stack level as follows:
+%
+{ \setlength{\jot}{20pt}
+\begin{eqnarray*}
+& \frac
+    { \DS e \neq \epsilon }
+%   -----------------------------------------------------------
+    { \DS \LL p,n,e \RR \OUT \LL bcast(p),n+1,\epsilon \RR }
+    & \textbf{(push)}   \\
+%%%
+& \frac
+    { \DS n>0, \2 ((p=@nop) \wedge isblocked(n,p)) }
+%   -----------------------------------------------------------
+    { \DS \LL p,n,\epsilon \RR \OUT \LL p,n-1,\epsilon \RR }
+    & \textbf{(pop)}    \\
+%%%
+%& \LL p,0,\epsilon \RR \1\xrightarrow\1 \bot
+    %& \textbf{(end)}    %\\
+\end{eqnarray*}
+}
+%
+Rule \textbf{push} matches whenever there is an emitted event in the triple,
+and instantly broadcasts the event to the program, which means
+    (a) awaking active \code{await} expressions altogether (see $bcast(p)$ in
+        Figure~\ref{fig.bcast}),
+    (b) creating a nested reaction by increasing the stack level,
+    (c) and, at the same time, consuming the event ($e$ becomes $\epsilon$).
+%
+Note that rule \textbf{push} is the only one in the semantics that matches an
+emitted event and also immediately consumes it.
+%
+The rule \textbf{pop} only decreases the stack level, not affecting the
+program, and does not apply if the program is not blocked (see $isblocked(n,p)$
+in Figure~\ref{fig.isblocked}).
+This condition ensures that an \code{emit} only resumes after its internal
+reaction completes, as discussed in Section~\ref{sec.ceu.evts}.
+
+Note that at the beginning of a reaction, an event $ext$ is emitted, which
+will trigger rule \textbf{push}, which will immediately raise the stack level
+to $1$.
+Since rule \textbf{pop} is the only to decrease the stack level and only
+applies to a blocked or terminated program, the semantics guarantees that a
+reaction only terminates with a blocked program at stack level $0$.
+
+The $\NST$ rules are recursive definitions but do not affect the stack level
+and never have an emitted event as a precondition:
+%
+\begin{align*}
+\LL p, n,\epsilon \RR &\NST
+\LL p',n,e        \RR
+    & \textbf{(rule-inner)}
+\end{align*}
+%
+The distinction between $\OUT$ and $\NST$ prevents rules \textbf{push} and
+\textbf{pop} to match and inadvertently modify the current stack level before
+completing a nested reaction.
+%
+A complete reaction behaves as follows:
+%
+\begin{align*}
+a) &\5\5
+    \LL p,0,ext \RR
+        \1\xrightarrow[out]{push}\1
+    \LL q,1,\epsilon \RR
+\\
+b) &\5\5 \1[ \1\xrightarrow[in]{*}\1
+    \LL r,i,e \RR
+        \1\xrightarrow[out]\1
+    \LL s,j,\epsilon \RR \1]*
+\\
+c) &\5\5 \1\xrightarrow[in]{*}\1
+    \LL t,k,\epsilon \RR
+        \1\xrightarrow[out]{pop}\1
+    \LL u,0,\epsilon \RR
+\end{align*}
+%
+First, a $\xrightarrow[out]{push}$ starts a nested reaction at level $1$ (case
+$a$).
+Then, a series of alternations between $\NST$ transitions followed by a single
+$\OUT$ transition represent nested reactions and stack operations (case $b$).
+Finally, at some point, a $\xrightarrow[out]{pop}$ lowers the stack level to
+$0$ and terminates the reaction.
+
+The $\NST$ transition rules for the primary expressions are as follows:
 %
@@
 \begin{align*}
-\LL S,\1mem(id) \RR &\ST
-\LL S,\1nop \RR
+\LL mem(id), n, \epsilon \RR &\NST
+\LL @nop, n, \epsilon \RR
     & \textbf{(mem)}        \\
 %%%
-\LL S,\1await(id) \RR &\ST
-\LL S,\1awaiting(id,n+1) \RR
-    & \textbf{(await)}      \\
+\LL emit(id), n, \epsilon \RR &\NST
+\LL @canrun(n), n, id \RR
+    & \textbf{(emitInt)}    \\
 %%%
-\LL id:S,\1awaiting(id,m) \RR &\ST
-\LL id:S,\1nop \RR, \2if~m\leq n
-    & \textbf{(awake)}   \\
-%%%
-\LL S,\1emit(id) \RR &\ST
-\LL id:S,\1emitting(|S|) \RR
-    & \textbf{(emit)}       \\
-%%%
-\LL S,\1emitting(k) \RR &\ST
-\LL S,\1nop \RR, \2 if~k=|S|
-    & \textbf{(pop)}        %\\
+\LL @canrun(n), n, \epsilon \RR &\NST
+\LL @nop, n, \epsilon \RR
+    & \textbf{(canrun)}     \\
 \end{align*}
@@
 %
-A $mem$ operation executes immediately and becomes a $nop$ to indicate 
-termination (rule \textbf{mem}).
-An $await$ is transformed into an $awaiting$ (rule \textbf{await}) as an 
-artifice to remember the external sequence number $n+1$ it can awake:
-an $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
-event $id$ matches the top of the stack and it was reached in a previous 
-reaction (i.e., sequence number $m \leq n$).
-%
-%Remember that in \CEU, the \code{await} statement returns the value associated 
-%with the corresponding event: the yielded $mem$ represents the operation to 
-%query that value.
-%
-An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
-stands for the stack length), and pushing the referred event on the stack (rule 
-\textbf{emit}).
-With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
-rule \textbf{pop} expects its parameter $k$ to match the current stack level.
-This trick provides the desired stack-based semantics for internal events.
+A $mem$ operation becomes a $nop$ which indicates termination (rule
+\textbf{mem}).
+An $emitInt(id)$ generates an event $id$ and transits to a $@canrun(n)$ which
+can only resume at level $n$ (rule \textbf{emitInt}).
+Since nested rules cannot transit with $e \neq \epsilon$, this rule will cause
+rule \textbf{push} to execute at the outer level, creating a new level $n+1$ on
+the stack.
+Also, with the new stack level, the resulting $@canrun(n)$ itself cannot
+transit, providing the desired stack-based semantics for internal events.
 
@@
 & \frac
-    { \DS val(id,n) \neq 0 }
+    { \DS val(id) \neq 0 }
 %   -----------------------------------------------------------
-    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
-          \LL S, p \RR }
+    { \DS \LL (if~mem(id)~then~p~else~q),n,\epsilon \RR \NST
+          \LL p, n, \epsilon \RR }
     & \textbf{(if-true)}       \\
@@
 %   -----------------------------------------------------------
-    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
-          \LL S, q \RR }
+    { \DS \LL (if~mem(id)~then~p~else~q),n,\epsilon \RR \NST
+          \LL q,n,\epsilon \RR }
     & \textbf{(if-false)}       \\
@@
 & \frac
-    {\DS \LL S,p \RR \ST \LL S',p' \RR }
+    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
 %   -----------------------------------------------------------
-    {\DS \LL S, (p~;~q) \RR \ST \LL S', (p'~;~q) \RR }
+    { \DS \LL (p~;~q), n, \epsilon \RR \NST \LL (p'~;~q), n, e \RR }
     & \textbf{(seq-adv)}      \\
 %%%
-& \LL S, (nop~;~q) \RR \ST  \LL S, q \RR
+& \LL (@nop~;~q),n,\epsilon \RR \NST  \LL q,n,\epsilon \RR
     & \textbf{(seq-nop)}      \\
 %%%
-& \LL S, (break~;~q) \RR \ST \LL S, break \RR
+& \LL (break~;~q),n,\epsilon \RR \NST \LL break,n,\epsilon \RR
     & \textbf{(seq-brk)}
@@
 %
-The ``magical'' function $val$ receives a memory identifier and the current 
-reaction sequence number, returning the current memory value.
+Function $val$ receives a memory identifier and returns the current memory
+value.
 %
-Although the value here is arbitrary, it is unique in a reaction, because a 
-given expression can execute only once within it (remember that $loops$ must 
-contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
-reaction they are reached).
+%Although the value here is arbitrary, it is unique in a reaction, because a 
+%given expression can execute only once within it (remember that $loops$ must 
+%contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
+%reaction they are reached).
 %For all other rules, we omit these values (e.g., \textbf{seq-nop}).
+
+As determined for nested rules, compound expressions also can only have
+$\epsilon$ as a precondition and they never modify $n$.
+However, they can still emit an event to nest another reaction.
+For instance, in rule \textbf{seq-adv}, if the sub-expression $p$ emits event
+$e$, the whole composition also emits $e$.
+%However, rules \textbf{push} and \textbf{pop} can only match at the outermost
+%level.
 
@@
 \begin{eqnarray*}
-& \LL S, (loop~p) \RR \ST \LL S, (p~@~loop~p) \RR
+& \LL (loop~p),n,\epsilon \RR \NST \LL (p~@loop~p), n, \epsilon \RR
     & \textbf{(loop-expd)}       \\
@@
 & \frac
-    {\DS \LL S,p \RR \ST \LL S',p' \RR }
+    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
 % -----------------------------------------------------------
-    {\DS \LL S, (p~@~loop~q) \RR \ST \LL S', (p'~@~loop~q) \RR }
+    { \DS \LL (p~@loop~q),n,\epsilon \RR \NST \LL (p'~@loop~q), n, e \RR }
     & \textbf{(loop-adv)}    \\
 %%%
-& \LL S, (nop~@~loop~p) \RR \ST \LL S, loop~p \RR
+& \LL (@nop~@loop~p), n, \epsilon \RR \NST \LL (loop~p), n, \epsilon \RR
     & \textbf{(loop-nop)}    \\
 %%%
-& \LL S, (break~@~loop~p) \RR \ST \LL S, nop \RR
+& \LL (break~@loop~p), n, \epsilon \RR \NST \LL @nop, n, \epsilon \RR
     & \textbf{(loop-brk)}
@@
 }
+
 %
@@
 \textbf{seq-adv} and \textbf{seq-nop}, advancing the loop until they reach a 
-$nop$.
+$@nop$.
 However, what follows the loop is the loop itself (rule \textbf{loop-nop}).
@@
 Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything into 
-a $nop$.
+a $@nop$.
 %Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything 
@@
 \begin{eqnarray*}
+& \LL (p~and~q),n,\epsilon \RR \NST \LL (p~@and~(@canrun(n)~;~q)),n,\epsilon \RR
+    & \textbf{(and-expd)}       \\
+%%%
 & \frac
-    {\DS \LL S,p \RR \ST \LL S',p' \RR }
+    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
 %   -----------------------------------------------------------
-    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p'~and~q) \RR }
+    { \DS \LL (p~@and~q),n,\epsilon \NST \LL (p'~@and~q),n,e \RR }
     & \textbf{(and-adv1)}      \\
@@
 & \frac
-    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
+    { \DS isblocked(n,p) \1,\2 \LL q,n,\epsilon \RR \NST \LL q',n,e \RR }
 %   -----------------------------------------------------------
-    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p~and~q') \RR }
+    { \DS \LL (p~@and~q),n,\epsilon \RR \NST \LL (p~@and~q'), n, e \RR }
     & \textbf{(and-adv2)}      \\
 %%%
+& \LL (p~or~q), n, \epsilon \RR \NST \LL (p~@or~(@canrun(n)~;~q)), n, \epsilon \RR
+    & \textbf{(or-expd)}       \\
+%%%
 & \frac
-    {\DS \LL S,p \RR \ST \LL S',p' \RR }
+    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
 %   -----------------------------------------------------------
-    {\DS \LL S, (p~or~q) \RR \ST \LL S', (p'~or~q) \RR }
+    { \DS \LL (p~@or~q),n,\epsilon \RR \NST \LL (p'~@or~q), n, e \RR }
     & \textbf{(or-adv1)}   \\
@@
 & \frac
-    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
+    { \DS isblocked(n,p) \1,\2 \LL q,n,\epsilon \RR \NST \LL q',n,e \RR }
 %   -----------------------------------------------------------
-    {\DS \LL S (p~or~q) \RR \ST \LL S', (p~or~q') \RR }
+    { \DS \LL (p~@or~q),n,\epsilon \RR \NST \LL (p~@or~q'), n, e \RR }
     & \textbf{(or-adv2)}   %\\
@@
 %
-The deterministic behavior of the semantics relies on the \emph{isBlocked} 
-predicate, which is defined in Figure~\ref{fig.isBlocked} and used in rules 
-\textbf{and-adv2} and \textbf{or-adv2}.
+Rules \textbf{and-expd} and \textbf{or-expd} insert a $@canrun(n)$ at the
+beginning of the right branch.
+This ensures that an $emitInt$ on the left branch, which would transit to a
+$@canrun(n)$, still resumes before the right branch starts.
+%
+The deterministic behavior of the semantics relies on the \emph{isblocked} 
+predicate (Figure~\ref{fig.isblocked}) which appears in rules \textbf{and-adv2}
+and \textbf{or-adv2}.
 These rules require the left branch $p$ to be blocked in order to allow the 
-right transition from $q$ to $q'$.
-%
-Basically, the \emph{isBlocked} predicate determines that an expression becomes 
-blocked when all of its trails in parallel hang in $awaiting$ and $emitting$ 
-expressions that cannot advance.
+right branch transition from $q$ to $q'$.
 
-\begin{figure}
-{\small
-\begin{align*}
-  isBlocked(n,a:S, awaiting(b,m)) &= (a \neq b \1\vee\1 m > n)   \\
-  isBlocked(n,S, emitting(s))    &= (|S| \neq s)                     \\
-  isBlocked(n,S, (p~;~q))        &= isBlocked(n,S,p)             \\
-  isBlocked(n,S, (p~@~loop~q))   &= isBlocked(n,S,p)             \\
-  isBlocked(n,S, (p~and~q))      &= isBlocked(n,S,p) \wedge
-                                    isBlocked(n,S,q)             \\
-  isBlocked(n,S, (p~or~q))       &= isBlocked(n,S,p) \wedge
-                                    isBlocked(n,S,q)             \\
-  isBlocked(n,S, \_)             &= false \2  (nop,await,      \\
-                                  &    \5\5\5\2 emit,break,if,loop)   %\\
-\end{align*}
-}%
-%\rule{14cm}{0.37pt}
-\caption{
-The recursive predicate $isBlocked$ is true only if all branches in parallel 
-are hanged in $awaiting$ or $emitting$ expressions that cannot transit.
-\label{fig.isBlocked}
-}
-\end{figure}
-
-%%%
-
-For a parallel $and$, if one of the sides terminates, the composition is simply 
-substituted by the other side (rules \textbf{and-nop1} and \textbf{and-nop2}, 
-as follows).
+For a parallel $@and$, if one of the sides terminates, the composition is
+simply substituted by the other side (rules \textbf{and-nop1} and
+\textbf{and-nop2}.
 %
@@
 \begin{eqnarray*}
-& \LL S, (nop~and~q) \RR \ST \LL S, q \RR
+& \LL (@nop~@and~q), n, \epsilon \RR \NST \LL q,n,\epsilon \RR
     & \textbf{(and-nop1)}   \\
 %%%
-& \LL S, (p~and~nop) \RR \ST \LL S, p \RR
+& \LL (p~@and~@nop), n, \epsilon \RR \NST \LL p,n,\epsilon \RR
     & \textbf{(and-nop2)}   \\
 %%%
-& \LL S, (nop~or~q) \RR \ST \LL S, clear(q) \RR
+& \LL (@nop~@or~q), n, \epsilon \RR \NST \LL clear(q),n,\epsilon \RR
     & \textbf{(or-nop1)}   \\
@@
 & \frac
-    {\DS isBlocked(n,S,p) }
+    { \DS isblocked(n,p) }
 %   -----------------------------------------------------------
-    {\DS \LL S, (p~or~nop) \RR \ST \LL S, clear(p) \RR }
+    { \DS \LL (p~@or~@nop), n, \epsilon \RR \NST \LL clear(p),n,\epsilon \RR }
     & \textbf{(or-nop2)}   %\\
@@
 }
+
 %
-The $clear$ function, defined in Figure~\ref{fig.formal.clear}, concatenates 
-all active $fin$ bodies of the side being aborted, so that they execute before 
-the composition rejoins.
+The $clear$ function (Figure~\ref{fig.formal.clear}) concatenates all active
+$fin$ bodies of the side being aborted, so that they execute before the
+composition rejoins.
 Note that there are no transition rules for $fin$ expressions.
@@
 function.
-In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
-the concrete language to a $fin$ in the formal semantics.
+%In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
+%the concrete language to a $fin$ in the formal semantics.
 %
-Note also that there is a syntactic restriction that a $fin$ body can only 
-contain $mem$ expressions in sequence, i.e., they are guaranteed to execute 
-entirely within a reaction.
-
-\begin{figure}[b]
-{\small
-\begin{align*}
-  clear( fin~p )       &= p                   \\
-  clear( p~;~q )       &= clear(p)            \\
-  clear( p~@~loop~q) ) &= clear(p)            \\
-  clear( p~and~q )     &= clear(p)~;~clear(q) \\
-  clear( p~or~q )      &= clear(p)~;~clear(q) \\
-  clear( \_ )          &= nop
-\end{align*}
-}%
-%\rule{14cm}{0.37pt}
-\caption{
-The function $clear$ extracts $fin$ expressions in parallel and put their 
-bodies in sequence.
-\label{fig.formal.clear}
-}
-\end{figure}
+Note also that there is a syntactic restriction that $fin$ bodies cannot
+contain awaiting expressions ($awaitExt$, $awaitInt$, and $fin$), i.e., they
+are guaranteed to execute entirely within a reaction.
 
@@
 \begin{eqnarray*}
-& \LL S, (break~and~q) \RR \ST \LL S, (clear(q)~;~break) \RR
+& \LL (break~@and~q), n, \epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
     & \textbf{(and-brk1)}   \\
@@
 & \frac
-    {\DS isBlocked(n,S,p) }
+    { \DS isblocked(n,p) }
 %   -----------------------------------------------------------
-    {\DS \LL S, (p~and~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
+    { \DS \LL (p~@and~break), n, \epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
     & \textbf{(and-brk2)}   \\
 %%%
-& \LL S, (break~or~q) \ST \LL S, (clear(q)~;~break) \RR
+& \LL (break~@or~q),n,\epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
     & \textbf{(or-brk1)}   \\
@@
 & \frac
-    {\DS isBlocked(n,S,p) }
+    { \DS isblocked(n,p) }
 %   -----------------------------------------------------------
-    {\DS \LL S, (p~or~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
+    { \DS \LL (p~@or~break),n,\epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
     & \textbf{(or-brk2)}   %\\
@@
 %
-A reaction eventually blocks in $awaiting$ and $emitting$ expressions in 
-parallel trails.
+A reaction eventually blocks in $awaitExt$, $awaitInt$, $fin$, and $@canrun$
+expressions in parallel trails.
 %
-If all trails hangs only in $awaiting$ expressions, it means that the program 
+If no trails are blocked in $@canrun$ expressions, it means that the program 
 cannot advance in the current reaction.
 %
-However, $emitting$ expressions are pending in lower stack indexes and should 
-eventually resume in the ongoing reaction (see rule \textbf{pop}).
-%
-Therefore, we define another rule that behaves as \textbf{rule-inner} 
-(presented above) if the program can advance, and, otherwise, pops the stack to 
-resume the lower level:
-%
-$$
-\frac
-    { \DS \LL S,p \RR \ST                   \LL S',p' \RR }
-%   -----------------------------------------------------------
-    {     \LL S,p \RR \xRightarrow[~~n~~]{} \LL S',p' \RR }
-%
-\5\5\5
-%
-\frac
-    { \DS isBlocked(n,\1s:S,\1p) }
-%   -----------------------------------------------------------
-    { \LL s:S,p \RR \xRightarrow[~~n~~]{} \LL S,p \RR }
-\5\5\5\textbf{(rule-outer)}
-$$
-%
-To describe a \emph{reaction} in \CEU, i.e., how a program behaves in reaction 
-to a single external event, we use the reflexive transitive closure of 
-\textbf{rule-outer}:
-%
-$$
-    \LL S,p \RR \xRightarrow[~~n~~]{*} \LL S',p' \RR
-$$
-%
-Finally, to describe the complete execution of a program, we trigger multiple 
-``invocations'' of reactions in sequence:
-%
+However, $@canrun$ expressions can still resume in lower stack indexes and will
+eventually resume in the current reaction (see rule \textbf{pop}).
+
+\begin{figure}
+{\small
 \begin{align*}
-\LL [e1], p \RR
-    & \xRightarrow[~~1~~]{*}
-\LL [  ], p' \RR
-\\
-\LL [e2], p' \RR
-    & \xRightarrow[~~2~~]{*}
-\LL [  ], p'' \RR
-\\
-\LL [e3], p'' \RR
-    & \xRightarrow[~~3~~]{*}
-\LL [  ], p''' \RR
-\\
-& ...
+  bcast(e, awaitExt(e)) &= @nop                         \\
+  bcast(e, awaitInt(e)) &= @nop                         \\
+  bcast(e, @canrun(n))  &= @canrun(n)                   \\
+  bcast(e, fin~p)       &= fin~p                        \\
+  bcast(e, p~;~q)       &= bcast(e,p)~;~q               \\
+  bcast(e, p~@loop~q)   &= bcast(e,p)~@loop~q           \\
+  bcast(e, p~@and~q)    &= bcast(e,p)~@and~bcast(e,q)   \\
+  bcast(e, p~@or~q)     &= bcast(e,p)~@or~bcast(e,q)    \\
+  bcast(e, \_)          &= \bot \2 (mem,emitInt,break,if,  \\
+                                 & \5\5 loop,and,or,@nop) %\\
 \end{align*}
-%
-Each invocation starts with the occurring external event at the top of the 
-stack and finishes with a modified program and an empty stack.
-After each invocation, we increment the sequence number.
+}
+\caption{
+The function $bcast$ awakes awaiting trails matching the event by converting
+\code{awaitExt} and \code{awaitInt} to \code{@nop} expressions.
+\label{fig.bcast}
+}
+\end{figure}
+
+\begin{figure}
+{\small
+\begin{align*}
+  isblocked(n, \1 awaitExt(id)) &= true                                   \\
+  isblocked(n, \1 awaitInt(id)) &= true                                   \\
+  isblocked(n, \1 @canrun(m))   &= (n > m)                                \\
+  isblocked(n, \1 fin~p)        &= true                                   \\
+  isblocked(n, \1 p~;~q)        &= isblocked(n,p)                         \\
+  isblocked(n, \1 p~@loop~q)    &= isblocked(n,p)                         \\
+  isblocked(n, \1 p~@and~q)     &= isblocked(n,p) \wedge isblocked(n,q)   \\
+  isblocked(n, \1 p~@or~q)      &= isblocked(n,p) \wedge isblocked(n,q)   \\
+  isblocked(n, \1 \_)           &= false \2 (mem,emitInt,break,if,        \\
+                                & \5\5\5\1 loop,and,or,@nop)   %\\
+\end{align*}
+}
+\caption{
+The predicate $isblocked$ is true only if all branches in parallel are blocked
+waiting for events, finalization clauses, or certain stack levels.
+\label{fig.isblocked}
+}
+\end{figure}
+
+\begin{figure}[b]
+{\small
+\begin{align*}
+  clear( awaitExt(e) ) &= @nop                  \\
+  clear( awaitInt(e) ) &= @nop                  \\
+  clear( @canrun(n) )  &= @nop                  \\
+  clear( fin~p )       &= p                     \\
+  clear( p~;~q )       &= clear(p)              \\
+  clear( p~@loop~q )   &= clear(p)              \\
+  clear( p~@and~q )    &= clear(p)~;~clear(q)   \\
+  clear( p~@or~q )     &= clear(p)~;~clear(q)   \\
+  clear( \_ )          &= \bot \2 (mem,emitInt,break,if, \\
+                                  & \5\5 loop,and,or,@nop) %\\ 
+\end{align*}
+}
+\caption{
+The function $clear$ extracts $fin$ expressions in parallel and put their 
+bodies in sequence.
+\label{fig.formal.clear}
+}
+\end{figure}
 

### 2018-02-07 02-ceu.tex
Remove timers from the paper.

@@
 
-Listing~\ref{lst.ceu} shows a complete example in \CEU that blinks a LED with a
-frequency of 1 second, terminating with a button press always with the LED off.
+Listing~\ref{lst.ceu} shows a complete example in \CEU that blinks a LED
+whenever a radio packet is received, terminating with a button press always
+with the LED off.
 %
-The implementation first declares the \code{BUTTON} as an input event (ln.~1).
+The implementation first declares the \code{BUTTON} and \code{RADIO_RECV} as
+input events (ln.~1--2).
 Then, it uses a \code{par/or} composition to run two activities in parallel:
-a single statement that waits for a button press before terminating (ln.~3),
-and an endless loop that blinks the LED on and off (ln.~8--13).
-The \code{finalize} clause (ln.~5--7) ensures that, no matter how its enclosing
-trail terminates, the LED will be unconditionally turned off (ln.~6).
+a single statement that waits for a button press before terminating (ln.~4),
+and an endless loop that blinks the LED on and off (ln.~9--14).
+The \code{finalize} clause (ln.~6--8) ensures that, no matter how its enclosing
+trail terminates, the LED will be unconditionally turned off (ln.~7).
 
@@
 %
-\fs{First-class timers.}
 
@@
 input void BUTTON;
+input void RADIO_RECV;
 par/or do
@@
         _led(1);
-        await 1s;
+        await RADIO_RECV;
         _led(0);
-        await 1s;
+        await RADIO_RECV;
     end

### 2018-02-07 03-semantics.tex
Remove timers from the paper.

@@
 
-\subsubsection{First-class Timers}
-
-To encompass first-class timers, we introduce a special external event 
-\code{DT} that is intercalated with each other event occurrence in an 
-application (e.g.  \emph{e1, e2}):
-
-\begin{align*}
-\LL [DT], p \RR
-    & \xRightarrow[~~1~~]{*}
-\LL [    ], p' \RR
-\\
-\LL [e1], p' \RR
-    & \xRightarrow[~~2~~]{*}
-\LL [  ], p'' \RR
-\\
-\LL [DT], p'' \RR
-    & \xRightarrow[~~3~~]{*}
-\LL [    ], p''' \RR
-\\
-\LL [e2], p''' \RR
-    & \xRightarrow[~~4~~]{*}
-\LL [  ], p'''' \RR
-\\
-& ...
-\end{align*}
-
-The event \code{DT} has an associated variable \code{DT\_} carrying the 
-wall-clock time elapsed between two occurrences in sequence, as depicted by the 
-two-step translation in Figure~\ref{lst.map.timers.2}.
-In the concrete program (a), the variable \code{dt} holds the residual delta 
-time (as described in Section~\ref{sec.ceu.timers}) after awaking from the 
-timer.
-In the first step of the translation (b), we expand the \code{await 10ms} to a 
-\code{loop} that decrements the elapsed number of microseconds for each 
-occurrence of \code{DT}.
-When the variable \code{tot} reaches zero, we escape the \code{loop} setting 
-the variable \code{dt} to contain the appropriate delta.
-In the last step (c), we convert the program to the abstract syntax.
-
-\begin{figure}
-\begin{minipage}[t]{0.30\linewidth}
-\begin{lstlisting}
-dt = await 10ms;
-
-.
-\end{lstlisting}
-\centering\small{(a)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.37\linewidth}
-\begin{lstlisting}
-var int tot = 10000; // 10ms
-loop do
-    await DT;
-    tot = tot - DT_;
-    if tot <= 0 then
-        dt = -tot;
-        break;
-    end
-end
-.
-\end{lstlisting}
-\centering\small{(b)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.30\linewidth}
-\begin{lstlisting}
-mem;
-loop(
-    await(DT);
-    mem;
-    if mem then
-        mem;
-        break
-    else
-        nop
-)
-\end{lstlisting}
-\centering\small{(c)}
-\end{minipage}
-\caption{
-Two-step translation from concrete to abstract timer.
-\label{lst.map.timers.2}
-}
-\end{figure}
-
 \subsubsection{Finalization Blocks}

### 2018-02-08 03-semantics.tex
Small adjusts

@@
 %
-First, a $\xrightarrow[out]{push}$ starts a nested reaction at level $1$ (case
-$a$).
-Then, a series of alternations between $\NST$ transitions followed by a single
-$\OUT$ transition represent nested reactions and stack operations (case $b$).
-Finally, at some point, a $\xrightarrow[out]{pop}$ lowers the stack level to
-$0$ and terminates the reaction.
+First (case $a$), a $\xrightarrow[out]{push}$ starts a nested reaction at level
+$1$.
+Then (case $b$), a series of alternations between $\NST$ transitions and a
+single $\OUT$ transition represent, respectively, nested reactions and stack
+operations.
+Finally (case $c$), at some point, a $\xrightarrow[out]{pop}$ decrements the
+stack level to $0$ and terminates the reaction.
 
-The $\NST$ transition rules for the primary expressions are as follows:
+The $\NST$ transition rules for primary expressions are as follows:
 %
@@
 %
-A $mem$ operation becomes a $nop$ which indicates termination (rule
+A $mem$ operation becomes a $@nop$ which indicates termination (rule
 \textbf{mem}).

### 2018-02-08 02-ceu.tex
Many changes:  - \code definition (problems with the font)  - "every" as a primitive in the concrete and formal language  - removed "Concrete Language Mapping" section (no space), included the most relevant parts in the semantics discussion

@@
 In \CEU, every execution path within loops must contain at least one
-\code{await} statement~\cite{ceu.sensys13,esterel.primer}.
+\code{await} statement to an external input
+event~\cite{ceu.sensys13,esterel.primer}.
+% TODO: or every
 %
@@
 %
-The implementation first declares the \code{BUTTON} and \code{RADIO_RECV} as
+The implementation first declares the \code{BUTTON} and \code{RADIO\_RECV} as
 input events (ln.~1--2).
@@
 
-In \CEU, any identifier prefixed with an underscore (e.g., \code{_led}) is
+In \CEU, any identifier prefixed with an underscore (e.g., \code{\_led}) is
 passed unchanged to the underlying~C compiler.
@@
 
-At first sight, an event iterator for event \code{A} seems like a syntactic
-sugar for a \code{loop} followed by and \code{await A}.
-However, event iterators have an important restriction to enforce reaction
-termination, as we discuss in Section~\ref{sec.sem}:
-    they cannot contain \code{break} statements.
+At first sight, event iteration such as ``\code{every e do <...> end}'' seems
+to be equivalent to ``\code{loop do await e; <...> end}''.
+However, the \code{loop} variation would not compile because it does not
+contain a path to an external input \code{await} (\code{e} is an internal
+event).
+Event iterators have syntactic restrictions and cannot contain \code{await} or
+\code{break} statements.
+The absence of \code{break} guarantees that an iterator never terminates from
+itself, essentially behaving as safe blocking point in the program.
 
@@
 In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
-(ln. 2) is an internal resource passed as a pointer to \code{_send} (ln. 5),
+(ln. 2) is an internal resource passed as a pointer to \code{\_send} (ln. 5),
 which is an asynchronous call that transmits the buffer in the background.
@@
 %
-In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{_fopen} (ln.
+In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
 4) returns an external file resource as a pointer.

### 2018-02-08 03-semantics.tex
Many changes:  - \code definition (problems with the font)  - "every" as a primitive in the concrete and formal language  - removed "Concrete Language Mapping" section (no space), included the most relevant parts in the semantics discussion

@@
       $|$ loop p                     (repetition)
+      $|$ every id p                 (event iteration)
       $|$ p and p                    (par/and)
@@
 %
-Except for $fin$ and the expressions used internally by the semantics, which
-are prefixed with a character $@$, all other expressions are equivalent to
-their counterparts in the concrete language.
-
 The $mem(id)$ primitive represents all accesses, assignments, system calls, and
@@
 or an internal event.
+
+In the concrete language, the \code{await} and \code{emit} primitives support
+communication of values between them, e.g., an \code{emit a(10)} awakes
+\code{v=await a} setting \code{v=10}.
+To reproduce this functionality in the formal semantics, we can use a shared
+variable to hold the value of an \code{emit} and access it after the
+corresponding \code{await}.
+
+A ``\code{finalize <A> with <B> end ; <C>}'' in the concrete language is
+equivalent to ``\code{<A> ; ((fin <B>) or <C>)}'' in the abstract syntax.
+In the concrete language, \code{<A>} and \code{<C>} execute in sequence, and
+the finalization code \code{<B>} is implicitly suspended waiting for \code{<C>}
+termination.
+In the formal language, \code{fin <B>} suspends forever when reached (it is an
+awaiting expression that never awakes).
+Hence, we need an explicit \code{or} to execute \code{<C>} in parallel, whose
+termination aborts \code{fin <B>}, which finally causes \code{<B>} to execute
+(by the semantic rules to be discussed).
 
@@
 can only resume at level $n$ (rule \textbf{emitInt}).
-Since nested rules cannot transit with $e \neq \epsilon$, this rule will cause
-rule \textbf{push} to execute at the outer level, creating a new level $n+1$ on
+Since nested rules can only transit with $e=\epsilon$, this rule causes rule
+\textbf{push} to execute at the outer level, creating a new level $n+1$ on
 the stack.
@@
 }
-
 %
@@
 Note also that there is a syntactic restriction that $fin$ bodies cannot
-contain awaiting expressions ($awaitExt$, $awaitInt$, and $fin$), i.e., they
-are guaranteed to execute entirely within a reaction.
+contain awaiting expressions ($awaitExt$, $awaitInt$, $every$, and $fin$),
+i.e., they are guaranteed to execute entirely within a reaction.
 
@@
 %
-A reaction eventually blocks in $awaitExt$, $awaitInt$, $fin$, and $@canrun$
-expressions in parallel trails.
+A reaction eventually blocks in $awaitExt$, $awaitInt$, $every$, $fin$, and
+$@canrun$ expressions in parallel trails.
 %
@@
   bcast(e, awaitInt(e)) &= @nop                         \\
+  bcast(e, every~e~p)   &= p; every e p                 \\
   bcast(e, @canrun(n))  &= @canrun(n)                   \\
@@
 The function $bcast$ awakes awaiting trails matching the event by converting
-\code{awaitExt} and \code{awaitInt} to \code{@nop} expressions.
+$awaitExt$ and $awaitInt$ to $@nop$ expressions, and by unwinding an $every$
+expression.
 \label{fig.bcast}
@@
   isblocked(n, \1 awaitInt(id)) &= true                                   \\
+  isblocked(n, \1 every~e~p)    &= true                                   \\
   isblocked(n, \1 @canrun(m))   &= (n > m)                                \\
@@
   clear( awaitInt(e) ) &= @nop                  \\
+  clear( every~e~p )   &= @nop                  \\
   clear( @canrun(n) )  &= @nop                  \\
@@
 \end{figure}
-
-\subsection{Concrete Language Mapping}
-
-Most statements from \CEU (``concrete \CEU'') map directly to those presented 
-in the reduced syntax in Figure~\ref{lst.formal.syntax} (``abstract \CEU'').
-%, namely, the \code{if}, \code{`;'}, \code{loop}, \code{par/and}, and 
-%\code{par/or}.
-For instance, the \code{if} in the concrete language behaves exactly like the 
-$if$ in the formal semantics.
-However, there are some significant mismatches between the concrete and 
-abstract \CEU, and we (informally) present appropriate mappings in this 
-section.
-%
-Again, we are not considering side-effects, which are all mapped to the $mem$ 
-semantic construct.
-
-\begin{comment}
-$
-    \code{if}      \mapsto if   ,\2
-    \code{';'}     \mapsto ~';' ,\2
-    \code{loop}    \mapsto loop ,\2
-    \code{par/and} \mapsto and  ,\2
-    \code{par/or}  \mapsto or
-$.
-\end{comment}
-
-\subsubsection{await and emit}
-
-The concrete \code{await} and \code{emit} primitives support communication of 
-values between them.
-In the two-step translation in Figure~\ref{lst.map.emit.await}, we start with 
-the concrete program in \CEU (a), which communicates the value $1$ between the 
-\code{emit} and \code{await} in parallel.
-In the intermediate translation (b), we include the shared variable \code{e\_} 
-to hold the value being communicated between the two trails in order to 
-simplify the \code{emit}.
-Finally, we convert the program into the equivalent in the abstract syntax (c), 
-translating side-effect statements into $mem$ expressions.
-External events have a similar translation, i.e., each external event requires 
-a corresponding variable that is explicitly set by the environment before each 
-reaction.
-
-\begin{figure}[t]
-\begin{minipage}[t]{0.32\linewidth}
-\begin{lstlisting}
-par/or do
-  <...>
-  emit e => 1;
-with
-  v = await e;
-  _printf("%d\n",v);
-end
-
-.
-\end{lstlisting}
-\centering\small{(a)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.32\linewidth}
-\begin{lstlisting}
-par/or do
-  <...>
-  e_ = 1;
-  emit e;
-with
-  await e;
-  v = e_;
-  _printf("%d\n",v);
-end
-\end{lstlisting}
-\centering\small{(b)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.32\linewidth}
-\begin{lstlisting}
-<...> ; mem ; emit(e)
-        or
-await(e) ; mem ; mem
-
-.
-\end{lstlisting}
-\centering\small{(c)}
-\end{minipage}
-\caption{
-Two-step translation from concrete to abstract \code{emit} and \code{await} 
-expressions.
-The concrete code in (a) communicates the value \code{1} from the \code{emit} 
-to the \code{await}.
-The abstract code in (c) uses a shared variable to hold the value.
-\label{lst.map.emit.await}
-}
-\end{figure}
-
-\subsubsection{Finalization Blocks}
-\label{sec.formal.fins}
-
-The biggest mismatch between concrete and abstract \CEU is regarding the 
-\code{finalize} blocks, which require more complex modifications in the program
-for a proper mapping using $fin$ expressions.
-In the three-step translation in Figure~\ref{lst.map.fin}, we start with a 
-concrete program (a) that uses a \code{finalize} to safely \code{\_release} the 
-reference to \code{ptr} kept after the call to \code{\_hold}.
-In the translation, we first need to catch the outermost \code{do-end} 
-termination to run the finalization code.
-For this, we translate the block into a \code{par/or} (b) with the original 
-body in parallel with a $fin$ expression to run the finalization code.
-Note that the $fin$ has no transition rules in the semantics, keeping the 
-\code{par/or} alive.
-This way, the $fin$ body only executes when the \code{par/or} terminates either 
-normally (after the \code{await B}), or aborted from an outer composition.
-%
-However, the $fin$ still (incorrectly) executes even if the call to 
-\code{\_hold} is not reached in the body due to an abort before awaking from 
-the \code{await A}.
-%
-To deal with this issue, for each $fin$ we need a corresponding flag to keep 
-track of code that needs to be finalized (c).
-%
-The flag is initially set to false, avoiding the finalization code to execute.
-Only after the call to \code{\_hold} that we set the flag to true and enable 
-the $fin$ body to execute.
-%
-The complete translation substitutes the side-effect operations with $mem$ 
-expressions (d).
-
-\begin{figure}
-\begin{minipage}[t]{0.26\linewidth}
-\begin{lstlisting}
-do
-  var int* ptr = <...>;
-  await A;
-  finalize
-    _hold(ptr);
-  with
-    _release(ptr);
-  end
-  await B;
-end
-
-.
-\end{lstlisting}
-\centering\small{(a)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.26\linewidth}
-\begin{lstlisting}
-par/or do
-  var int* ptr = <...>;
-  await A;
-  _hold(ptr);
-  await B;
-with
-  { fin
-      _release(ptr); }
-end
-
-.
-\end{lstlisting}
-\centering\small{(b)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.26\linewidth}
-\begin{lstlisting}
-f_ = 0;
-par/or do
-  var int* ptr = <...>;
-  await A;
-  _hold(ptr);
-  f_ = 1;
-  await B;
-with
-  { fin
-      if f_ then
-        _release(ptr);
-      end }
-end
-.
-\end{lstlisting}
-\centering\small{(c)}
-\end{minipage}
-%
-\begin{minipage}[t]{0.18\linewidth}
-\begin{lstlisting}
-mem;
-(
-   mem;
-   await(A);
-   mem;
-   mem;
-   await(B);
-or
-   fin
-     if mem then
-       mem
-     else
-       nop
-)
-\end{lstlisting}
-\centering\small{(d)}
-\end{minipage}
-%
-\caption{
-Three-step translation from concrete to abstract finalization.
-\label{lst.map.fin}
-}
-\end{figure}

### 2018-02-08 01-intro.tex
Pequenas revisões.

@@
 For this reason, it has been successfully adopted in safety-critical real-time
-embedded systems.~\cite{rp.twelve}
+embedded systems~\cite{rp.twelve}.
 
@@
 \CEU has also been used in the context of multimedia
-systems~\cite{ceumedia.webmedia16} and games~\cite{ceu.mod15}, and as an
-alternative language in an undergraduate-level course on embedded systems for
-the past 6 years.
+systems~\cite{ceumedia.webmedia16} and games~\cite{ceu.mod15}.
+%, and as an
+%alternative language in an undergraduate-level course on embedded systems for
+%the past 6 years.
 
@@
 \begin{itemize}
+    \item Fine-grained, intra-reaction deterministic execution, which makes
+          \CEU fully deterministic.
     \item Stack-based execution for internal events, which provides a limited
-          form of coroutines.
-    \item Fine-grained, intra-reaction deterministic execution, which allows
-          programs to safely share memory.
-    \item Finalization mechanism for abortion of lines of execution, which
-          safely release external resources.
-    \item First-class synchronized timers.
+          but memory-bounded form of subroutines.
+    \item Finalization mechanism for lines of execution, which makes abortion
+          safe with regards to external resources.
 \end{itemize}

### 2018-02-08 02-ceu.tex
Pequenas revisões.

@@
 event~\cite{ceu.sensys13,esterel.primer}.
-% TODO: or every
 %
@@
 event).
+%
 Event iterators have syntactic restrictions and cannot contain \code{await} or
@@
 itself, essentially behaving as safe blocking point in the program.
-
-\begin{comment}
 %
-Third, \CEU subroutines can be safely used by the other primitives of the
-language, such as parallel compositions and the \code{await} statement,
-without breaking the programming model.
-%
-In particular, after calling a subroutine these primitives wait while
-keeping context information, such as locals and the program counter, which
-makes the calls behave similarly to those of
-coroutines~\cite{lua.coroutines}.
-%
-Finally, in previous work, we built other advanced control mechanisms on top
-of internal events, such as resumable exceptions and reactive
-variables~\cite{ceu.rem13}.
-%
-\end{comment}
+For this reason, the restriction that execution paths within loops must
+contains at least one external \code{await} is extended to contain an
+\code{every} statement alternatively.
 

### 2018-02-08 03-semantics.tex
Pequenas revisões.

@@
 %
-At the beginning of a reaction to an input event, the triple is initialized at
-stack level $0$ ($n=0$) and with the emitted event $id$ ($e=id$).
+At the beginning of a reaction to an input event $id$, the triple is
+initialized at stack level $0$ ($n=0$) and with the emitted event $id$
+($e=id$).
 %, but \code{emitInt} expressions can increase the stack level.
-At the end of a reaction, after a number of transitions, the triple will block
-with a (possibly) modified program $p'$, at stack level $0$, with no event
-emitted $\epsilon$:
+At the end of a reaction, after an arbitrary but finite number of transitions,
+the triple will block with a (possibly) modified program $p'$, at stack level
+$0$, with no event emitted $\epsilon$:
 %
@@
 %
-We also define two transition rule subtypes $\OUT$ (for \emph{outermost}) and
-$\NST$ (for \emph{nested}).
+We also distinguish between two transition rule subtypes:
+    $\OUT$, for \emph{outermost} transitions; and
+    $\NST$, for \emph{nested} transitions.
 
 The $\OUT$ rules \textbf{push} and \textbf{pop} are non-recursive definitions
-that manipulate the stack level as follows:
+that only apply to the program as a whole and manipulate the stack level as
+follows:
 %

### 2018-02-08 03-semantics.tex
Provas informais no fim da Secao 3.

@@
 & \frac
-    { \DS n>0, \2 ((p=@nop) \wedge isblocked(n,p)) }
+    { \DS n>0, \2 ((p=@nop) \vee isblocked(n,p)) }
 %   -----------------------------------------------------------
@@
 \end{figure}
+
+\subsection{Determinism, Termination, and Memory Bounds}
+
+- informal discussion
+
+\subsubsection*{Determinism}
+
+The proof for determinism relies on the fact all semantic rules are mutually
+exclusive, i.e., their preconditions are unique in the set of rules.
+This can be verified by simple rule inspections.
+
+Rule \textbf{push} is the only one with $e \neq \epsilon$ as a precondition,
+and is trivially mutually exclusive with all other rules.
+
+Rule \textbf{pop} either has $p=@nop$ or $isblocked(n,p)$ as preconditions.
+%
+Note that rule \textbf{pop} only syntactically applies to top-level
+transitions.
+For instance, it can never match $\NST$ rules for subprograms as in rule
+\textbf{seq-adv}.
+%
+Hence, for the first case, rule \textbf{pop} only applies, and is the only one
+to apply, to $nop$ as the whole program (i.e., a $nop$ not surrounded by other
+expressions, such as in rule \textbf{seq-nop}).
+%
+For the second case, we need to show that given $\LL p,n,\epsilon \RR$, no
+$\NST$ transitions apply with $isblocked(n,p)$ and vice versa.
+Except for $@canrun$, there are no $\NST$ transitions for the other blocking
+expressions ($awaitExt$, $awaitInt$, $every$, and $fin$).
+However, considering the precondition $\LL p,n,\epsilon \RR$,
+$isblocked(n,@canrun(n))$ is false.
+Hence, given the preconditions for rule \textbf{pop}, no $\NST$ transitions can
+occur.
+Conversely, if a $\NST$ transition is possible, then $isblocked(n,p)$ must be
+false.
+Again, except for $@canrun$, all other transitions do not involve blocking
+expressions, hence, for these transitions, $isblocked(n,p)$ must be false.
+For rule \textbf{canrun}, a transition can only occur if the current stack
+level matches $@canrun(n)$.
+In this case, $isblocked(n,@canrun(n))$ is false.
+
+Finally, we need to show that $\NST$ transitions are mutually exclusive among
+themselves.
+%
+Note that most rules have unique syntactic prefixes, e.g., $(@nop~@and~q)$
+(rule \textbf{and-nop1}) is trivially mutually exclusive with $(@nop~@loop~p)$
+(rule \textbf{or-nop1}).
+%
+The only exceptions are rules \textbf{and-adv1} vs. \textbf{and-adv2}, and
+\textbf{or-adv1} vs. \textbf{or-adv2}.
+In both cases, we need to show that if the left branch can advance, then it
+cannot be blocked and vice-versa, i.e., that 
+$\LL p,n,\epsilon \RR \NST \LL p',n,e \RR$ and $isblocked(n,p)$ are mutually
+exclusive.
+This is the same reasoning for rule \textbf{pop} above.
+
+\subsubsection*{Termination}
+
+- there is always a possible transition until n=0
+
+every cannot restart itself
+    - break disallowed
+    - emit ignored
+
+\subsubsection*{Memory Bounds}
+
+- program is finite
+- lexical scope
+    - no heap allocation
+- no code reentrancy
+    - reexecution only due to loops
+    - loop reuse nested vars
+-

### 2018-02-08 01-intro.tex
Revisão até antes da semântica.

@@
 %
-Structured reactive programming let developers write code in direct/sequential
-style, recovering from the inversion of control imposed by event-driven
+Structured reactive programming lets developers write code in direct style,
+recovering from the inversion of control imposed by event-driven
 execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
 %
-Synchronous languages offer a simple run-to-completion model of execution that
-enable deterministic execution and make formal reasoning tractable.
+Synchronous languages offer a simple run-to-completion execution model that
+enables deterministic execution and make formal reasoning tractable.
 For this reason, it has been successfully adopted in safety-critical real-time
@@
 \CEU inherits the synchronous and imperative mindset of Esterel but adopts a
-simpler semantics with fine-grained execution control.~\cite{ceu.tecs17}
+simpler semantics with fine-grained execution control~\cite{ceu.tecs17}.
 %

### 2018-02-08 02-ceu.tex
Revisão até antes da semântica.

@@
 Computations within a reaction (such as expressions, assignments, and
-system calls) are also instantaneous in accordance to the synchronous
+system calls) are also instantaneous considering the synchronous
 hypothesis~\cite{rp.hypothesis}.
 %
-\CEU provides an \code{await} statement which is the only that actually
-``consumes'' time.
+%\CEU provides an \code{await} statement which is the only that actually
+%``consumes'' time.
 %
-An \code{await} statement blocks the current running trail allowing the program
-to advance its other trails; when all trails are blocked, the reaction
-terminates and control returns to the environment.
+\CEU provides an \code{await} statement that blocks the current running trail
+allowing the program to advance its other trails; when all trails are blocked,
+the reaction terminates and control returns to the environment.
 
@@
 |\C{1.}{Declarations}|
-input $\T<type>$ $\T<id>$; | \C{.6}{declares external input event} |
-event $\T<type>$ $\T<id>$; | \C{.6}{declares internal event}       |
-var   $\T<type>$ $\T<id>$; | \C{.6}{declares variable}             |
+input $\T<type>$ $\T<id>$; | \C{.6}{declares an external input event} |
+event $\T<type>$ $\T<id>$; | \C{.6}{declares an internal event}       |
+var   $\T<type>$ $\T<id>$; | \C{.6}{declares a variable}              |
 
 |\C{1.}{Event handling}|
-$\T<id>$ = await $\T<id>$;   | \C{.6}{awaits event and assigns received value} |
-$\T<id>$ = await $\T<time>$; | \C{.6}{awaits time and assigns delayed delta}   |
-emit $\T<id>$($\T<exp>$);    | \C{.6}{emits event passing value}               |
+$\T<id>$ = await $\T<id>$;   | \C{.6}{awaits an event and assigns the received value} |
+emit $\T<id>$($\T<exp>$);    | \C{.6}{emits an event passing a value} |
 
@@
 |\C{1.}{Assignment \& Integration with C}|
-$\T<id>$ = $\T<exp>$; | \C{.6}{assigns value to variable}                |
-_$\T<id>$($\T<exps>$) | \C{.6}{calls C function (id starts with `\_'\,)} |
+$\T<id>$ = $\T<exp>$; | \C{.6}{assigns a value to a variable}              |
+_$\T<id>$($\T<exps>$) | \C{.6}{calls a C function (id starts with `\_'\,)} |
 \end{lstlisting}
@@
 
-Listing~\ref{lst.ceu} shows a complete example in \CEU that blinks a LED
+Listing~\ref{lst.ceu} shows a complete example in \CEU that toggles a LED
 whenever a radio packet is received, terminating with a button press always
@@
 a single statement that waits for a button press before terminating (ln.~4),
-and an endless loop that blinks the LED on and off (ln.~9--14).
+and an endless loop that toggles the LED on and off on radio receives
+(ln.~9--14).
 The \code{finalize} clause (ln.~6--8) ensures that, no matter how its enclosing
@@
 %
-This is possible to do safely in synchronous languages due to the accurate
-control of concurrent activities, i.e., in between every reaction, the whole
-system is idle and consistent~\cite{esterel.preemption}.
+%This is possible to do safely in synchronous languages due to the accurate
+%control of concurrent activities, i.e., in between every reaction, the whole
+%system is idle and consistent~\cite{esterel.preemption}.
 %
@@
   float=h,
-  caption={A program in \CEU that blinks a LED every second, terminating on a
-           button press in a consistent state.},
+  caption={A program in \CEU that toggles a LED on every radio receive,
+           terminating on a button press in a consistent state.},
   label={lst.ceu},
@@
       awakes \emph{all} trails awaiting that event, and the program goes back to
-      step~(i).
+      step~(ii).
 \end{enumerate}
@@
 %
-In the soft real-time systems targeted by \CEU (e.g., sensor networks,
-multimedia systems, interactive games, etc.) such delay and postponed
-reactions might be tolerated by users as long as they are infrequent and the
-application does not take too long to catch up with real time.
+Considering the context soft real-time systems,
+%targeted by \CEU (e.g., sensor networks, multimedia systems, interactive games, etc.)
+such delays and postponed reactions might be tolerated as long as they are
+infrequent and the application does not take too long to catch up with real
+time.
 %
@@
     var int* p;
-    every p in inc do   // implements subroutine with an event iterator
+    every p in inc do   // implements "inc" through an event iterator
         *p = *p + 1;
@@
     emit inc(&v);       // calls "inc"
-    _assert(v==3);      // asserts result after returns
+    _assert(v==3);      // asserts result after the two returns
 end
@@
 
-\CEU also supports nested \code{emit} invocations for internal events.
-%
-For instance, the body of the subroutine \code{inc} in Listing~\ref{lst.sub}
-could \code{emit} another internal event after awaking (ln.~4), creating a
-new level in the stack.
+\CEU also supports nested \code{emit} invocations, e.g., the body of the
+subroutine \code{inc} (ln.~5) could \code{emit} an event targeting another
+subroutine, creating a new level in the stack.
 %
@@
 %
-First, they guaranteed to react in bounded time.
+First, they are guaranteed to react in bounded time.
 %
@@
 
-At first sight, event iteration such as ``\code{every e do <...> end}'' seems
+At first sight, event iteration such as in ``\code{every e do <...> end}'' seems
 to be equivalent to ``\code{loop do await e; <...> end}''.
@@
 %
-Event iterators have syntactic restrictions and cannot contain \code{await} or
-\code{break} statements.
+However, event iterators enforce syntactic restrictions and cannot contain
+\code{await} or \code{break} statements.
 The absence of \code{break} guarantees that an iterator never terminates from
-itself, essentially behaving as safe blocking point in the program.
+itself, essentially behaving as a safe blocking point in the program.
 %
 For this reason, the restriction that execution paths within loops must
-contains at least one external \code{await} is extended to contain an
-\code{every} statement alternatively.
+contain at least one external \code{await} is extended to alternatively contain
+an \code{every} statement.
 
@@
 input void A;
-//
+// (empty line)
 var int y = 1;
@@
 %
-Nonetheless, the static checks are optional and do not affect the semantics of
-the language.
+Nonetheless, the static checks are optional and are not a prerequisite for the
+deterministic semantics of the language.
 
@@
 For this reason, \CEU provides a \code{finalize} construct to unconditionally
-execute a series of statements even if the enclosing block is aborted.
+execute a series of statements even if the enclosing block is externally
+aborted.
 

### 2018-02-08 index.tex
Revisão até antes da semântica.

--- 2018-02-02 index.tex
@@
 
-\title[A Deterministic and Terminating Semantics for \CEU]
+\title[A Memory-Bounded, Deterministic, and Terminating Semantics for \CEU]
 {A Deterministic and Terminating Semantics
@@
 It focus on control-flow safety features, such as safe shared-memory
-concurrency and safe abortion of lines of execution, while enforcing internal
-determinism and time-bounded reactions to the environment.
+concurrency and safe abortion of lines of execution, while enforcing
+memory-bounded, deterministic, and terminating reactions to the environment.
 %
 In this work, we present a small-step structural operational semantics for
-\CEU and a proof that reactions are deterministic and always terminate:
+\CEU and a proof that reactions have the properties enumerated above:
 %
-For a given timeline of input events, multiple executions of the same program
-always react in bounded time and arrive at the same final memory state.
+that for a given arbitrary timeline of input events, multiple executions of the
+same program always react in bounded time and arrive at the same final finite
+memory state.
 %

### 2018-02-08 03-semantics.tex
Revisão da semântica.

@@
 %
-The $mem(id)$ primitive represents all accesses, assignments, system calls, and
-output events that affect a memory location identified by $id$.
+The $mem(id)$ primitive represents all accesses, assignments, and system calls,
+that affect a memory location identified by $id$.
 %
@@
 %
-Note that $mem$, $awaitExt$, and $awaitInt$ / $emitInt$ expressions do not
+Note that $mem$, $awaitExt$, and $awaitInt$/$emitInt$ expressions do not
 share identifiers, i.e., an identifier is either a variable, an external event,
@@
 
-In the concrete language, the \code{await} and \code{emit} primitives support
-communication of values between them, e.g., an \code{emit a(10)} awakes
-\code{v=await a} setting \code{v=10}.
+Regarding mismatches between the concrete and formal languages, the concrete
+\code{await} and \code{emit} primitives support communication of values between
+them, e.g., an ``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
+\code{v=10}.
 To reproduce this functionality in the formal semantics, we can use a shared
@@
 corresponding \code{await}.
-
-A ``\code{finalize <A> with <B> end ; <C>}'' in the concrete language is
+%
+Also, a ``\code{finalize <A> with <B> end ; <C>}'' in the concrete language is
 equivalent to ``\code{<A> ; ((fin <B>) or <C>)}'' in the abstract syntax.
@@
 termination.
-In the formal language, \code{fin <B>} suspends forever when reached (it is an
-awaiting expression that never awakes).
+In the formal language, ``\code{fin <B>}'' suspends forever when reached (it is
+an awaiting expression that never awakes).
 Hence, we need an explicit \code{or} to execute \code{<C>} in parallel, whose
-termination aborts \code{fin <B>}, which finally causes \code{<B>} to execute
-(by the semantic rules to be discussed).
+termination aborts ``\code{fin <B>}'', which finally causes \code{<B>} to
+execute (by the semantic rules to be discussed).
 
@@
 The core of our semantics describes how a program reacts to a single external 
-input event, i.e., starting from the input event, how the program behaves and 
-becomes idle again to proceed to the subsequent reaction.
+input event, i.e., starting from an input event, how the program behaves and 
+becomes idle again to proceed to a subsequent reaction.
 %
@@
 We also distinguish between two transition rule subtypes:
-    $\OUT$, for \emph{outermost} transitions; and
-    $\NST$, for \emph{nested} transitions.
+    \emph{outermost} transitions $\OUT$; and
+    \emph{nested} transitions $\NST$.
 
@@
         Figure~\ref{fig.bcast}),
-    (b) creating a nested reaction by increasing the stack level,
-    (c) and, at the same time, consuming the event ($e$ becomes $\epsilon$).
+    (b) creating a nested reaction by increasing the stack level, and, at the same time,
+    (c) consuming the event ($e$ becomes $\epsilon$).
 %
@@
 The rule \textbf{pop} only decreases the stack level, not affecting the
-program, and does not apply if the program is not blocked (see $isblocked(n,p)$
-in Figure~\ref{fig.isblocked}).
+program, and only applies if the program is blocked (see $isblocked(n,p)$ in
+Figure~\ref{fig.isblocked}).
 This condition ensures that an \code{emit} only resumes after its internal
-reaction completes, as discussed in Section~\ref{sec.ceu.evts}.
+reaction completes and blocks, as discussed in Section~\ref{sec.ceu.evts}.
 
@@
 %
-A $mem$ operation becomes a $@nop$ which indicates termination (rule
+A $mem$ operation becomes a $@nop$ which indicates the memory access (rule
 \textbf{mem}).
@@
 can only resume at level $n$ (rule \textbf{emitInt}).
-Since nested rules can only transit with $e=\epsilon$, this rule causes rule
-\textbf{push} to execute at the outer level, creating a new level $n+1$ on
+Since all $\NST$ rules can only transit with $e=\epsilon$, an $emitInt$ causes
+rule \textbf{push} to execute at the outer level, creating a new level $n+1$ on
 the stack.
@@
 %
-Function $val$ receives a memory identifier and returns the current memory
-value.
+Function $val$ receives a memory identifier and returns its current value.
 %
@@
 
-As determined for nested rules, compound expressions also can only have
-$\epsilon$ as a precondition and they never modify $n$.
-However, they can still emit an event to nest another reaction.
-For instance, in rule \textbf{seq-adv}, if the sub-expression $p$ emits event
-$e$, the whole composition also emits $e$.
+%As determined for nested rules, compound expressions also can only have
+%$\epsilon$ as a precondition and they never modify $n$.
+%However, they can still emit an event to nest another reaction.
+%For instance, in rule \textbf{seq-adv}, if the sub-expression $p$ emits event
+%$e$, the whole composition also emits $e$.
 %However, rules \textbf{push} and \textbf{pop} can only match at the outermost
@@
 beginning of the right branch.
-This ensures that an $emitInt$ on the left branch, which would transit to a
+This ensures that an $emitInt$ on the left branch, which transits to a
 $@canrun(n)$, still resumes before the right branch starts.
@@
 The deterministic behavior of the semantics relies on the \emph{isblocked} 
-predicate (Figure~\ref{fig.isblocked}) which appears in rules \textbf{and-adv2}
-and \textbf{or-adv2}.
+predicate (see Figure~\ref{fig.isblocked}) which appears in rules
+\textbf{and-adv2} and \textbf{or-adv2}.
 These rules require the left branch $p$ to be blocked in order to allow the 
@@
 %%%
-& \LL (p~@and~@nop), n, \epsilon \RR \NST \LL p,n,\epsilon \RR
+& \frac
+    { \DS isblocked(n,p) }
+%   -----------------------------------------------------------
+    { \DS \LL (p~@and~@nop), n, \epsilon \RR \NST \LL p,n,\epsilon \RR }
     & \textbf{(and-nop2)}   \\
@@
 %
-The $clear$ function (Figure~\ref{fig.formal.clear}) concatenates all active
-$fin$ bodies of the side being aborted, so that they execute before the
+The $clear$ function (see Figure~\ref{fig.formal.clear}) concatenates all
+active $fin$ bodies of the side being aborted, so that they execute before the
 composition rejoins.
@@
   bcast(e, awaitInt(e)) &= @nop                         \\
-  bcast(e, every~e~p)   &= p; every e p                 \\
+  bcast(e, every~e~p)   &= p;~every~e~p                 \\
   bcast(e, @canrun(n))  &= @canrun(n)                   \\
@@
 The function $bcast$ awakes awaiting trails matching the event by converting
-$awaitExt$ and $awaitInt$ to $@nop$ expressions, and by unwinding an $every$
-expression.
+$awaitExt$ and $awaitInt$ to $@nop$ expressions, and by unwinding $every$
+expressions.
 \label{fig.bcast}
@@
 exclusive, i.e., their preconditions are unique in the set of rules.
-This can be verified by simple rule inspections.
+This can be verified by direct inspection of rules.
 
@@
 %
-Note that rule \textbf{pop} only syntactically applies to top-level
+Note that rule \textbf{pop} only applies syntactically to top-level
 transitions.
@@
 $\LL p,n,\epsilon \RR \NST \LL p',n,e \RR$ and $isblocked(n,p)$ are mutually
-exclusive.
-This is the same reasoning for rule \textbf{pop} above.
+exclusive, which is exactly the same reasoning for rule \textbf{pop} above.
 

### 2018-02-15 03-semantics.tex
Formatação da sintaxe abstrata na Seção 3
Adicionei macros em header.tex para formatar os comandos da sintaxe
abstrata.  Reescrevi a gramática para usar essas macros.

@@
-\newcommand{\NST}{\1\xrightarrow[nst]\1}
-\newcommand{\OUT}{\1\xrightarrow[out]\1}
+\newcommand{\NST}{\1\xrightarrow[\mathit{nst}]\1}
+\newcommand{\OUT}{\1\xrightarrow[\mathit{out}]\1}
 \newcommand{\LL}{\langle}
@@
 
-%\begin{document}
-
-In this section, we introduce a reduced syntax of \CEU and propose an 
+In this section, we introduce a reduced syntax for \CEU and propose an 
 operational semantics to formally describe the language.
 We describe a small synchronous kernel highlighting the peculiarities of \CEU, 
-in particular the stack-based execution for internal events.
+in particular, the stack-based execution for internal events.
 %
 For the sake of simplicity, we focus on the control aspects of the language, 
-leaving out side effects and system calls (which behave like in conventional 
+leaving out side-effects and system calls (which behave like in conventional 
 imperative languages).
@@
 
-\begin{lstlisting}[
-  %numbers=left,
-  basicstyle=\ttfamily\footnotesize,
-  float=h,
-  caption={Reduced syntax of \CEU.},
-  label={lst.formal.syntax},
-  mathescape=true
-]
-                                   // primary expressions
-  p ::= mem(id)                    (any memory access to `id')
-      $|$ awaitExt(id)               (await external event `id')
-      $|$ awaitInt(id)               (await internal event `id')
-      $|$ emitInt(id)                (emit internal event `id')
-      $|$ break                      (loop escape)
-                                   // compound expressions
-      $|$ if mem(id) then p else p   (conditional)
-      $|$ p ; p                      (sequence)
-      $|$ loop p                     (repetition)
-      $|$ every id p                 (event iteration)
-      $|$ p and p                    (par/and)
-      $|$ p or p                     (par/or)
-      $|$ fin p                      (finalization)
-                                   // derived by semantic rules
-      $|$ p @loop p                  (unwinded loop)
-      $|$ p @and q                   (unwinded par/and)
-      $|$ p @or q                    (unwinded par/or)
-      $|$ @canrun(n)                 (can run on stack level `n')
-      $|$ @nop                       (terminated expression)
-\end{lstlisting}
+% \begin{lstlisting}[
+%   %numbers=left,
+%   basicstyle=\ttfamily\footnotesize,
+%   float=h,
+%   caption={Reduced syntax of \CEU.},
+%   label={lst.formal.syntax},
+%   mathescape=true
+% ]
+%                                    // primary expressions
+%   p ::= mem(id)                    (any memory access to `id')
+%       $|$ awaitExt(id)               (await external event `id')
+%       $|$ awaitInt(id)               (await internal event `id')
+%       $|$ emitInt(id)                (emit internal event `id')
+%       $|$ break                      (loop escape)
+%                                    // compound expressions
+%       $|$ if mem(id) then p else p   (conditional)
+%       $|$ p ; p                      (sequence)
+%       $|$ loop p                     (repetition)
+%       $|$ every id p                 (event iteration)
+%       $|$ p and p                    (par/and)
+%       $|$ p or p                     (par/or)
+%       $|$ fin p                      (finalization)
+%                                    // derived by semantic rules
+%       $|$ p @loop p                  (unwinded loop)
+%       $|$ p @and q                   (unwinded par/and)
+%       $|$ p @or q                    (unwinded par/or)
+%       $|$ @canrun(n)                 (can run on stack level `n')
+%       $|$ @nop                       (terminated expression)
+% \end{lstlisting}
 
-Listing~\ref{lst.formal.syntax} shows the syntax for a subset of \CEU that is 
+The grammar below defines the syntax of a subset of \CEU that is
 sufficient to describe all semantic peculiarities of the language.
+\bgroup
+\def\lbl#1{\qquad\text{\emph{#1}}}%
+\newdimen\X
+\X=-1\jot
+\begin{alignat*}{2}
+  p\Coloneqq
+      &\enspace\ceu{\Mem(\Id)}
+      &&\lbl{any memory access to~``$\ceu\Id$''}\\[\X]
+      %%
+  \mid&\enspace\ceu{\AwaitExt(\Id)}
+      &&\lbl{await external event~``$\ceu\Id$''}\\[\X]
+      %%
+  \mid&\enspace\ceu{\AwaitInt(\Id)}
+      &&\lbl{await internal event~``$\ceu\Id$''}\\[\X]
+      %%
+  \mid&\enspace\ceu{\EmitInt(\Id)}
+      &&\lbl{emit internal event~``$\ceu\Id$''}\\[\X]
+      %%
+  \mid&\enspace\ceu{\Break}
+      &&\lbl{loop escape}\\[\X]
+      %%
+  \mid&\enspace\ceu{\Ifelse{b}{p_1}{p_2}}
+      &&\lbl{conditional}\\[\X]
+      %%
+  \mid&\enspace\ceu{p1\,;\,p2}
+      &&\lbl{sequence}\\[\X]
+      %%
+  \mid&\enspace\ceu{\Loop p_1}
+      &&\lbl{repetition}\\[\X]
+      %%
+  \mid&\enspace\ceu{\Every{\Id}\ p_1}
+      &&\lbl{event iteration}\\[\X]
+      %%
+  \mid&\enspace\ceu{p_1\And p_2}
+      &&\lbl{par/and}\\[\X]
+      %%
+  \mid&\enspace\ceu{p_2\Or p_2}
+      &&\lbl{par/or}\\[\X]
+      %%
+  \mid&\enspace\ceu{\Fin p}
+      &&\lbl{finalization}\\[\X]
+      %%
+  \mid&\enspace\ceu{p_1\AtLoop p_2}
+      &&\lbl{unwinded loop}\\[\X]
+      %%
+  \mid&\enspace\ceu{p_1\AtAnd\ p_2}
+      &&\lbl{unwinded par/and}\\[\X]
+      %%
+  \mid&\enspace\ceu{p_1\AtOr\ p_2}
+      &&\lbl{unwinded par/or}\\[\X]
+      %%
+  \mid&\enspace\ceu{\CanRun(n)}
+      &&\lbl{can run on stack level~$n$}\\[\X]
+      %%
+  \mid&\enspace\ceu{\Nop}
+      &&\lbl{terminated program}
+\end{alignat*}
+\egroup
+
+The~$\ceu{\Mem(id)}$ primitive represents all accesses, assignments, and system calls
+that affect a memory location identified by~$id$.
 %
-The $mem(id)$ primitive represents all accesses, assignments, and system calls,
-that affect a memory location identified by $id$.
-%
-According to the synchronous hypothesis of \CEU, $mem$ expressions are 
+According to the synchronous hypothesis of \CEU, $\ceu{\Mem}$ expressions are 
 considered to be atomic and instantaneous.
@@
 %
-Note that $mem$, $awaitExt$, and $awaitInt$/$emitInt$ expressions do not
+Note that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, and~$\ceu{\AwaitInt}$ and~$\ceu{\EmitInt}$ expressions do not
 share identifiers, i.e., an identifier is either a variable, an external event,
@@
 
-Regarding mismatches between the concrete and formal languages, the concrete
+Regarding mismatches between the concrete and abstract languages, the concrete
 \code{await} and \code{emit} primitives support communication of values between
@@
 \\
-n, n' &\in \mathds{N}
+n, n' &\in \mathbf{N}
     && (current~stack~level)
 \\
-e, e' &\in \mathds{E \cup \{\epsilon\}}
+e, e' &\in \mathbf{E} \cup \{\epsilon\}
     && (emitted~event,~possibly~none)

### 2018-02-15 03-semantics.tex
Reformatação do inicio da Seção 3.2
Comprimi a definição dos conjuntos sintáticos (P, N e E).

@@
 %unused identifier).
-%
-Note that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, and~$\ceu{\AwaitInt}$ and~$\ceu{\EmitInt}$ expressions do not
-share identifiers, i.e., an identifier is either a variable, an external event,
+
+We assume that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, and~$\ceu{\AwaitInt}$ and~$\ceu{\EmitInt}$ expressions do not
+share identifiers: any identifier is either a variable, an external event,
 or an internal event.
 
-Regarding mismatches between the concrete and abstract languages, the concrete
+Most expressions in the abstract language are mapped to
+their counterparts in the concrete language.  The exceptions are the
+finalization block~$\ceu{\Fin{p}}$ and the \texttt{@}-expressions, which are
+internal expressions that result from the
+expansion of awaits, emits, and loops by the transition rules to be discussed.
+
+Regarding other mismatches between the concrete and abstract languages, the concrete
 \code{await} and \code{emit} primitives support communication of values between
 them, e.g., an ``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
-\code{v=10}.
+variable~\code{v} to~10.
 To reproduce this functionality in the formal semantics, we can use a shared
-variable to hold the value of an \code{emit} and access it after the
-corresponding \code{await}.
+variable to hold the value of an $\ceu{\EmitInt}$ and access it after the
+corresponding $\ceu{\AwaitInt}$.
 %
-Also, a ``\code{finalize <A> with <B> end ; <C>}'' in the concrete language is
-equivalent to ``\code{<A> ; ((fin <B>) or <C>)}'' in the abstract syntax.
-In the concrete language, \code{<A>} and \code{<C>} execute in sequence, and
-the finalization code \code{<B>} is implicitly suspended waiting for \code{<C>}
+Also, a ``\code{finalize $A$ with $B$ end; $C$}'' in the concrete language is
+equivalent to ``\ceu{A;\;((\Fin{B})\ \Or\ C)}'' in the abstract language.
+In the concrete language, $A$ and~$C$ execute in sequence, and
+the finalization code~$B$ is implicitly suspended waiting for~$C$
 termination.
-In the formal language, ``\code{fin <B>}'' suspends forever when reached (it is
+In the abstract language, ``$\ceu{\Fin B}$'' suspends forever when reached (it is
 an awaiting expression that never awakes).
-Hence, we need an explicit \code{or} to execute \code{<C>} in parallel, whose
-termination aborts ``\code{fin <B>}'', which finally causes \code{<B>} to
+Hence, we need an explicit \code{or} to execute~$C$ in parallel, whose
+termination aborts ``$\ceu{\Fin B}$'', which finally causes~$B$ to
 execute (by the semantic rules to be discussed).
@@
 %
-The transition rules map a triple with a program $p$, a stack level $n$, and an
-emitted event $e$ to a modified triple as follows:
+The transition rules map a triple with a program~$p$, a stack level~$n$, and an
+emitted event~$e$ to a modified triple as follows:
+\[
+  \<p, n, e>\to\<p',n',e'>
+\]
+where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are
+nonnegative integers representing the current stack level,
+and~$e,e'\in\E\cup\{\nil\}$ are the events emitted before and after
+the transition, both possibly the empty event~($\nil$).
+
+% \begin{align*}
+% p, p' &\in\P
+%     && (program~as~described~in~Listing~\ref{lst.formal.syntax})
+% \\
+% n, n' &\in\N
+%     && (current~stack~level)
+% \\
+% e, e' &\in\E \cup \{\epsilon\}
+%     && (emitted~event,~possibly~none)
+% \end{align*}
 %
-\begin{align*}
-\LL p,  n,  e  \RR &\1\xrightarrow\1
-\LL p', n', e' \RR
-    & \textbf{(rule-general)}
-\end{align*}
-%
-where
-%
-\begin{align*}
-p, p' &\in P
-    && (program~as~described~in~Listing~\ref{lst.formal.syntax})
-\\
-n, n' &\in \mathbf{N}
-    && (current~stack~level)
-\\
-e, e' &\in \mathbf{E} \cup \{\epsilon\}
-    && (emitted~event,~possibly~none)
-\end{align*}
-%
-At the beginning of a reaction to an input event $id$, the triple is
-initialized at stack level $0$ ($n=0$) and with the emitted event $id$
+At the beginning of a reaction to an input event~$id$, the triple is
+initialized at stack level $0$ ($n=0$) and with the emitted event
 ($e=id$).
@@
 At the end of a reaction, after an arbitrary but finite number of transitions,
-the triple will block with a (possibly) modified program $p'$, at stack level
-$0$, with no event emitted $\epsilon$:
+the triple will block with a (possibly) modified program~$p'$, at stack
+level~$0$, and with no event emitted~$(\nil)$:
 %
-\begin{align*}
-\LL p,0,ext \RR
-    & \1\xrightarrow{*}\1
-\LL p',0,\epsilon \RR
-\end{align*}
-%
-We also distinguish between two transition rule subtypes:
-    \emph{outermost} transitions $\OUT$; and
+\[
+  \<p,0,e>\xrightarrow{*}\<p',0,\nil>
+\]
+
+We distinguish between two types of transition rules:
+    \emph{outermost} transitions $\OUT$ and
     \emph{nested} transitions $\NST$.
@@
 The $\OUT$ rules \textbf{push} and \textbf{pop} are non-recursive definitions
-that only apply to the program as a whole and manipulate the stack level as
-follows:
+that only apply to the program as a whole and manipulate the stack level:
 %

### 2018-02-16 03-semantics.tex
Reformatação ainda do início da Seção 3.2
Reescrevi as regras para utilizem o pacote busproofs.  Meu objetivo foi
comprimir as definições e torná-las uniformes.

Adicionei duas subsubseções novas para separar a parte que define transições
out da parte que define transições nst.

Redefini as macros das transições, agora \out e \nst em header.tex.  As
macros originais, \OUT e \NST, usam \xrightarrow, o que acaba introduzindo
um gap entre linhas adjacentes em modo texto.

@@
 
+%-
 % \begin{lstlisting}[
@@
 % \end{lstlisting}
+%-
 
@@
 \[
-  \<p, n, e>\to\<p',n',e'>
+  \<p,n,e>\trans\<p',n',e'>\,,
 \]
@@
 nonnegative integers representing the current stack level,
-and~$e,e'\in\E\cup\{\nil\}$ are the events emitted before and after
-the transition, both possibly the empty event~($\nil$).
+and~$e,e'\in\E\cup\{\nil\}$ are the events emitted before and after the
+transition (both possibly the empty event~$\nil$).
 
+We will refer to the triples on the left-hand and right-hand sides of
+symbol~$\to$ as \emph{descriptions} (denoted~$\delta$).  The triple on the
+left-hand side of symbol~$\to$ is called the \emph{input description}, and
+the triple on its right-hand side is called the \emph{output description}.
+
+%-
 % \begin{align*}
@@
 % \end{align*}
-%
-At the beginning of a reaction to an input event~$id$, the triple is
-initialized at stack level $0$ ($n=0$) and with the emitted event
+%-
+
+At the beginning of a reaction to an input event~$id$, the input description is
+initialized with stack level~0 ($n=0$) and with the emitted event
 ($e=id$).
@@
 At the end of a reaction, after an arbitrary but finite number of transitions,
-the triple will block with a (possibly) modified program~$p'$, at stack
-level~$0$, and with no event emitted~$(\nil)$:
-%
+the last output description will block with a (possibly) modified program~$p'$, at stack
+level~0, and with no event emitted~($\nil$):
 \[
-  \<p,0,e>\xrightarrow{*}\<p',0,\nil>
+  \<p,0,e>\mathbin{\trans[*]}\<p',0,\nil>\,.
 \]
@@
 We distinguish between two types of transition rules:
-    \emph{outermost} transitions $\OUT$ and
-    \emph{nested} transitions $\NST$.
+    \emph{outermost} transitions $\out$ and
+    \emph{nested} transitions $\nst$\,.
 
-The $\OUT$ rules \textbf{push} and \textbf{pop} are non-recursive definitions
+\subsubsection*{Outermost transitions}
+
+The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions
 that only apply to the program as a whole and manipulate the stack level:
+\begin{align*}
+  &\AxiomC{$e\ne\nil$}
+  \UnaryInfC{$\<p,n,e>\out\<\bcast(p,e),n+1,\nil>$}
+  \DisplayProof
+  \Rtag{push}\\[2\jot]
+  %%
+  &\AxiomC{$n>0$}
+  \AxiomC{$\ceu{p=\Nop}\vee\isblocked(p,n)$}
+  \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
+  \DisplayProof
+  \Rtag{pop}
+\end{align*}
+
+%-
+% { \setlength{\jot}{20pt}
+% \begin{eqnarray*}
+% & \frac
+%     { \DS e \neq \epsilon }
+% %   -----------------------------------------------------------
+%     { \DS \LL p,n,e \RR \OUT \LL bcast(p),n+1,\epsilon \RR }
+%     & \textbf{(push)}   \\
+% %%%
+% & \frac
+%     { \DS n>0, \2 ((p=@nop) \vee isblocked(n,p)) }
+% %   -----------------------------------------------------------
+%     { \DS \LL p,n,\epsilon \RR \OUT \LL p,n-1,\epsilon \RR }
+%     & \textbf{(pop)}    \\
+% %%%
+% %& \LL p,0,\epsilon \RR \1\xrightarrow\1 \bot
+%     %& \textbf{(end)}    %\\
+% \end{eqnarray*}
+% }
+%-
+
+Rule \R{push} matches whenever there is an emitted event in the input
+description,
+and instantly broadcasts the event to the program, which means
+    (a)~awaking active $\ceu{\AwaitExt}$ or $\ceu{\AwaitInt}$ expressions altogether (see function~$\bcast$ in
+        Figure~\ref{fig.bcast}),
+    (b)~creating a nested reaction by increasing the stack level, and, at the same time, and
+    (c)~consuming the event ($e$ becomes~$\nil$).
 %
-{ \setlength{\jot}{20pt}
-\begin{eqnarray*}
-& \frac
-    { \DS e \neq \epsilon }
-%   -----------------------------------------------------------
-    { \DS \LL p,n,e \RR \OUT \LL bcast(p),n+1,\epsilon \RR }
-    & \textbf{(push)}   \\
-%%%
-& \frac
-    { \DS n>0, \2 ((p=@nop) \vee isblocked(n,p)) }
-%   -----------------------------------------------------------
-    { \DS \LL p,n,\epsilon \RR \OUT \LL p,n-1,\epsilon \RR }
-    & \textbf{(pop)}    \\
-%%%
-%& \LL p,0,\epsilon \RR \1\xrightarrow\1 \bot
-    %& \textbf{(end)}    %\\
-\end{eqnarray*}
-}
-%
-Rule \textbf{push} matches whenever there is an emitted event in the triple,
-and instantly broadcasts the event to the program, which means
-    (a) awaking active \code{await} expressions altogether (see $bcast(p)$ in
-        Figure~\ref{fig.bcast}),
-    (b) creating a nested reaction by increasing the stack level, and, at the same time,
-    (c) consuming the event ($e$ becomes $\epsilon$).
-%
-Note that rule \textbf{push} is the only one in the semantics that matches an
+Rule \R{push} is the only rule in the semantics that matches an
 emitted event and also immediately consumes it.
-%
-The rule \textbf{pop} only decreases the stack level, not affecting the
-program, and only applies if the program is blocked (see $isblocked(n,p)$ in
+
+Rule \R{pop} only decreases the stack level, not affecting the
+program, and only applies if the program is blocked (see function~$\isblocked$ in
 Figure~\ref{fig.isblocked}).
-This condition ensures that an \code{emit} only resumes after its internal
+This condition ensures that an $\ceu{\EmitInt}$ only resumes after its internal
 reaction completes and blocks, as discussed in Section~\ref{sec.ceu.evts}.
 
-Note that at the beginning of a reaction, an event $ext$ is emitted, which
+At the beginning of the reaction, an external event is emitted, which
 will trigger rule \textbf{push}, which will immediately raise the stack level
-to $1$.
-Since rule \textbf{pop} is the only to decrease the stack level and only
-applies to a blocked or terminated program, the semantics guarantees that a
-reaction only terminates with a blocked program at stack level $0$.
+to~1.
+At the end of the reaction, the program will block or terminate and
+successive applications of
+rule~\R{pop} will eventually lead to a description containing this
+same program at stack level~0.
+(Rule \R{pop} is the only rule that decreases the stack level.)
 
-The $\NST$ rules are recursive definitions but do not affect the stack level
-and never have an emitted event as a precondition:
+\subsubsection*{Nested transitions}
+
+The~$\nst$ rules are recursive definitions with the following general
+format:
+\[
+\<p,n,\nil>\nst\<p',n,e>.
+\]
 %
+%-
+% \begin{align*}
+% \LL p, n,\epsilon \RR &\NST
+% \LL p',n,e        \RR
+%     & \textbf{(rule-inner)}
+% \end{align*}
+%-
+%
+Nested transitions do not affect the stack level and never have an emitted
+event as a precondition.  The distinction between~$\out$ and~$\nst$ prevents
+rules \R{push} and \R{pop} from matching and, consequently, from
+inadvertently modifying the current stack level before the nested reaction
+is complete.
+
+A complete reaction consists of the following series of transitions:
 \begin{align*}
-\LL p, n,\epsilon \RR &\NST
-\LL p',n,e        \RR
-    & \textbf{(rule-inner)}
+  \<p,0,e_\ext>\out[push]\<p_1,1,\nil>
+  \Big[\null\nst[*]\null\out\null\Big]\!\!\ast
+  \null\nst[*]\null\out[pop]\<p',0,\nil>\,.
 \end{align*}
 %
-The distinction between $\OUT$ and $\NST$ prevents rules \textbf{push} and
-\textbf{pop} to match and inadvertently modify the current stack level before
-completing a nested reaction.
+%-
+% \begin{align*}
+% a) &\5\5
+%     \LL p,0,ext \RR
+%         \1\xrightarrow[out]{push}\1
+%     \LL q,1,\epsilon \RR
+% \\
+% b) &\5\5 \1[ \1\xrightarrow[in]{*}\1
+%     \LL r,i,e \RR
+%         \1\xrightarrow[out]\1
+%     \LL s,j,\epsilon \RR \1]*
+% \\
+% c) &\5\5 \1\xrightarrow[in]{*}\1
+%     \LL t,k,\epsilon \RR
+%         \1\xrightarrow[out]{pop}\1
+%     \LL u,0,\epsilon \RR
+% \end{align*}
+%-
 %
-A complete reaction behaves as follows:
-%
-\begin{align*}
-a) &\5\5
-    \LL p,0,ext \RR
-        \1\xrightarrow[out]{push}\1
-    \LL q,1,\epsilon \RR
-\\
-b) &\5\5 \1[ \1\xrightarrow[in]{*}\1
-    \LL r,i,e \RR
-        \1\xrightarrow[out]\1
-    \LL s,j,\epsilon \RR \1]*
-\\
-c) &\5\5 \1\xrightarrow[in]{*}\1
-    \LL t,k,\epsilon \RR
-        \1\xrightarrow[out]{pop}\1
-    \LL u,0,\epsilon \RR
-\end{align*}
-%
-First (case $a$), a $\xrightarrow[out]{push}$ starts a nested reaction at level
-$1$.
-Then (case $b$), a series of alternations between $\NST$ transitions and a
-single $\OUT$ transition represent, respectively, nested reactions and stack
-operations.
-Finally (case $c$), at some point, a $\xrightarrow[out]{pop}$ decrements the
-stack level to $0$ and terminates the reaction.
+First, a~$\out[push]$ starts a nested reaction at level~1.
+Then, a series of alternations between zero or more~$\nst$ transitions (nested reactions) and a
+single~$\out$ transition (stack operation) takes place.
+Finally, a last~$\out[pop]$ transition decrements the
+stack level to~0 and terminates the reaction.
 
-The $\NST$ transition rules for primary expressions are as follows:
+The~$\nst$ transition rules for atomic expressions are defined as follows:
 %

### 2018-02-16 03-semantics.tex
Reformatação do final da Seção 3.2
Reformatei as regras e definições de funções restantes.  Mantive os
originais em comentários.

@@
 The~$\nst$ transition rules for atomic expressions are defined as follows:
-%
-{ \setlength{\jot}{20pt}
 \begin{align*}
-\LL mem(id), n, \epsilon \RR &\NST
-\LL @nop, n, \epsilon \RR
-    & \textbf{(mem)}        \\
-%%%
-\LL emit(id), n, \epsilon \RR &\NST
-\LL @canrun(n), n, id \RR
-    & \textbf{(emitInt)}    \\
-%%%
-\LL @canrun(n), n, \epsilon \RR &\NST
-\LL @nop, n, \epsilon \RR
-    & \textbf{(canrun)}     \\
+  \<\ceu{\Mem(\Id)},n,\nil>
+  &\nst\<\ceu{\Nop},n,\nil>\Rtag{mem}\\
+  %%
+  \<\ceu{\EmitInt(\Id)},n,\nil>
+  &\nst\<\ceu{\CanRun(n)},n,\ceu{\Id}>\Rtag{emit-int}\\
+  %%
+  \<\ceu{\CanRun(n)},n,\nil>
+  &\nst\<\ceu{\Nop},n,\nil>\Rtag{can-run}
 \end{align*}
-}
-%
-A $mem$ operation becomes a $@nop$ which indicates the memory access (rule
-\textbf{mem}).
-An $emitInt(id)$ generates an event $id$ and transits to a $@canrun(n)$ which
-can only resume at level $n$ (rule \textbf{emitInt}).
-Since all $\NST$ rules can only transit with $e=\epsilon$, an $emitInt$ causes
-rule \textbf{push} to execute at the outer level, creating a new level $n+1$ on
+
+%-
+% { \setlength{\jot}{20pt}
+% \begin{align*}
+% \LL mem(id), n, \epsilon \RR &\NST
+% \LL @nop, n, \epsilon \RR
+%     & \textbf{(mem)}        \\
+% %%%
+% \LL emit(id), n, \epsilon \RR &\NST
+% \LL @canrun(n), n, id \RR
+%     & \textbf{(emitInt)}    \\
+% %%%
+% \LL @canrun(n), n, \epsilon \RR &\NST
+% \LL @nop, n, \epsilon \RR
+%     & \textbf{(canrun)}     \\
+% \end{align*}
+% }
+%-
+
+A $\ceu{\Mem}$ operation becomes a $\ceu{\Nop}$ which indicates the memory access (rule
+\R{mem}).
+An $\ceu{\EmitInt(id)}$ generates an event $\ceu{\Id}$ and transits to a $\ceu{\CanRun(n)}$ which
+can only resume at level~$n$ (rule \R{emit-int}).
+Since all~$\nst$ rules can only transit with $e=\nil$, an $\ceu{\EmitInt}$ causes
+rule \R{push} to execute at the outer level, creating a new level~$n+1$ on
 the stack.
-Also, with the new stack level, the resulting $@canrun(n)$ itself cannot
-transit, providing the desired stack-based semantics for internal events.
+Also, with the new stack level, the resulting $\ceu{\CanRun}(n)$ itself cannot
+transit (rule~\R{can-run}), providing the desired stack-based semantics for internal events.
 
@@
 %
-{ \setlength{\jot}{20pt}
-\begin{eqnarray*}
-& \frac
-    { \DS val(id) \neq 0 }
-%   -----------------------------------------------------------
-    { \DS \LL (if~mem(id)~then~p~else~q),n,\epsilon \RR \NST
-          \LL p, n, \epsilon \RR }
-    & \textbf{(if-true)}       \\
-%%%
-& \frac
-    { \DS val(id,n) = 0 }
-%   -----------------------------------------------------------
-    { \DS \LL (if~mem(id)~then~p~else~q),n,\epsilon \RR \NST
-          \LL q,n,\epsilon \RR }
-    & \textbf{(if-false)}       \\
-%%%
-& \frac
-    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
-%   -----------------------------------------------------------
-    { \DS \LL (p~;~q), n, \epsilon \RR \NST \LL (p'~;~q), n, e \RR }
-    & \textbf{(seq-adv)}      \\
-%%%
-& \LL (@nop~;~q),n,\epsilon \RR \NST  \LL q,n,\epsilon \RR
-    & \textbf{(seq-nop)}      \\
-%%%
-& \LL (break~;~q),n,\epsilon \RR \NST \LL break,n,\epsilon \RR
-    & \textbf{(seq-brk)}
-\end{eqnarray*}
-}
+\begin{gather*}
+  \AxiomC{$\eval(\ceu{\Mem(\Id)})$}
+  \UnaryInfC{$\<\ceu{\Ifelse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<p,n,\nil>$}
+  \DisplayProof
+  \Rtag{if-true}\\[2\jot]
+  %%
+  \AxiomC{$\lnot\eval(\ceu{\Mem(\Id)})$}
+  \UnaryInfC{$\<\ceu{\Ifelse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<q,n,\nil>$}
+  \DisplayProof
+  \Rtag{if-false}\\[2\jot]
+  %%
+  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
+  \UnaryInfC{$\<\ceu{p\,;\,q},n,\nil>\nst\<\ceu{p';\,q},n,e>$}
+  \DisplayProof
+  \Rtag{seq-adv}
+\end{gather*}
+\vskip-\belowdisplayskip
+\vskip-\abovedisplayskip
+\vskip2\jot
+\begin{align*}
+  \<\ceu{\Nop;\,q},n,\nil>&\nst\<q,n,\nil>\Rtag{seq-nop}\\[2\jot]
+  %%
+  \<\ceu{\Break;\,q},n,\nil>&\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
+\end{align*}
+
+%-
+% { \setlength{\jot}{20pt}
+% \begin{eqnarray*}
+% & \frac
+%     { \DS val(id) \neq 0 }
+% %   -----------------------------------------------------------
+%     { \DS \LL (if~mem(id)~then~p~else~q),n,\epsilon \RR \NST
+%           \LL p, n, \epsilon \RR }
+%     & \textbf{(if-true)}       \\
+% %%%
+% & \frac
+%     { \DS val(id,n) = 0 }
+% %   -----------------------------------------------------------
+%     { \DS \LL (if~mem(id)~then~p~else~q),n,\epsilon \RR \NST
+%           \LL q,n,\epsilon \RR }
+%     & \textbf{(if-false)}       \\
+% %%%
+% & \frac
+%     { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
+% %   -----------------------------------------------------------
+%     { \DS \LL (p~;~q), n, \epsilon \RR \NST \LL (p'~;~q), n, e \RR }
+%     & \textbf{(seq-adv)}      \\
+% %%%
+% & \LL (@nop~;~q),n,\epsilon \RR \NST  \LL q,n,\epsilon \RR
+%     & \textbf{(seq-nop)}      \\
+% %%%
+% & \LL (break~;~q),n,\epsilon \RR \NST \LL break,n,\epsilon \RR
+%     & \textbf{(seq-brk)}
+% \end{eqnarray*}
+% }
+%-
+
+Given that our semantics focuses on control, rules \R{if-true} and
+\R{if-false} are the only to query~$\ceu{\Mem}$ expressions.
 %
-Given that our semantics focuses on control, rules \textbf{if-true} and 
-\textbf{if-false} are the only to query $mem$ expressions.
-%
-Function $val$ receives a memory identifier and returns its current value.
+Function~$\eval$ evaluates a given~$\ceu{\Mem}$ expression.
 %
@@
 
-The rules for loops are analogous to sequences, but use \code{`@'} as 
+The rules for loops are analogous to sequences, but use ``\code{@}'' as
 separators to properly bind breaks to their enclosing loops:
+\begin{align*}
+  \<\ceu{\Loop{p}},n,\nil>
+  &\nst\<\ceu{p\AtLoop{p}},n,\nil>\Rtag{loop-expd}\\[2\jot]
+  %%
+  &\hskip-6.35em
+  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
+  \UnaryInfC{$\<\ceu{p\AtLoop{p}},n,\nil>\nst\<\ceu{p'\AtLoop{p}},n,e>$}
+  \DisplayProof
+  \Rtag{loop-adv}\\[2\jot]
+  %%
+  \<\ceu{\Nop\AtLoop{p}},n,\nil>
+  &\nst\<\ceu{\Loop{p}},n,\nil>\Rtag{loop-nop}\\[2\jot]
+  %%
+  \<\ceu{\Break\AtLoop{p}},n,\nil>
+  &\nst\<\ceu{\Nop},n,\nil>\Rtag{loop-brk}
+\end{align*}
+
+%-
+% %
+% { \setlength{\jot}{20pt}
+% \begin{eqnarray*}
+% & \LL (loop~p),n,\epsilon \RR \NST \LL (p~@loop~p), n, \epsilon \RR
+%     & \textbf{(loop-expd)}       \\
+% %%%
+% & \frac
+%     { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
+% % -----------------------------------------------------------
+%     { \DS \LL (p~@loop~q),n,\epsilon \RR \NST \LL (p'~@loop~q), n, e \RR }
+%     & \textbf{(loop-adv)}    \\
+% %%%
+% & \LL (@nop~@loop~p), n, \epsilon \RR \NST \LL (loop~p), n, \epsilon \RR
+%     & \textbf{(loop-nop)}    \\
+% %%%
+% & \LL (break~@loop~p), n, \epsilon \RR \NST \LL @nop, n, \epsilon \RR
+%     & \textbf{(loop-brk)}
+% \end{eqnarray*}
+% }
+%-
+
+When a program encounters a $\ceu{\Loop}$, it first expands its body in sequence with 
+itself (rule \R{loop-expd}).
+Rules \R{loop-adv} and \R{loop-nop} are similar to rules 
+\R{seq-adv} and \R{seq-nop}, advancing the loop until a~$\ceu{\Nop}$ is reached.
+However, what follows the loop is the loop itself (rule \R{loop-nop}).
+Note that if we used ``\code{;}'' as a separator in loops, rules
+\R{loop-brk} and \R{seq-brk} would conflict.
 %
-{ \setlength{\jot}{20pt}
-\begin{eqnarray*}
-& \LL (loop~p),n,\epsilon \RR \NST \LL (p~@loop~p), n, \epsilon \RR
-    & \textbf{(loop-expd)}       \\
-%%%
-& \frac
-    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
-% -----------------------------------------------------------
-    { \DS \LL (p~@loop~q),n,\epsilon \RR \NST \LL (p'~@loop~q), n, e \RR }
-    & \textbf{(loop-adv)}    \\
-%%%
-& \LL (@nop~@loop~p), n, \epsilon \RR \NST \LL (loop~p), n, \epsilon \RR
-    & \textbf{(loop-nop)}    \\
-%%%
-& \LL (break~@loop~p), n, \epsilon \RR \NST \LL @nop, n, \epsilon \RR
-    & \textbf{(loop-brk)}
-\end{eqnarray*}
-}
-
-%
-When a program encounters a $loop$, it first expands its body in sequence with 
-itself (rule \textbf{loop-expd}).
-Rules \textbf{loop-adv} and \textbf{loop-nop} are similar to rules 
-\textbf{seq-adv} and \textbf{seq-nop}, advancing the loop until they reach a 
-$@nop$.
-However, what follows the loop is the loop itself (rule \textbf{loop-nop}).
-Note that if we used \code{`;'} as a separator in loops, rules 
-\textbf{loop-brk} and \textbf{seq-brk} would conflict.
-%
-Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything into 
-a $@nop$.
+Rule \R{loop-brk} escapes the enclosing loop, transforming everything into 
+a~$\ceu{\Nop}$.
 %Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything 
@@
 
-Proceeding to parallel compositions, the semantic rules for $and$ and $or$ 
-always force transitions on their left branches $p$ to occur before their right 
-branches $q$:
-%
-{ \setlength{\jot}{20pt}
-\begin{eqnarray*}
-& \LL (p~and~q),n,\epsilon \RR \NST \LL (p~@and~(@canrun(n)~;~q)),n,\epsilon \RR
-    & \textbf{(and-expd)}       \\
-%%%
-& \frac
-    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
-%   -----------------------------------------------------------
-    { \DS \LL (p~@and~q),n,\epsilon \NST \LL (p'~@and~q),n,e \RR }
-    & \textbf{(and-adv1)}      \\
-%%%
-& \frac
-    { \DS isblocked(n,p) \1,\2 \LL q,n,\epsilon \RR \NST \LL q',n,e \RR }
-%   -----------------------------------------------------------
-    { \DS \LL (p~@and~q),n,\epsilon \RR \NST \LL (p~@and~q'), n, e \RR }
-    & \textbf{(and-adv2)}      \\
-%%%
-& \LL (p~or~q), n, \epsilon \RR \NST \LL (p~@or~(@canrun(n)~;~q)), n, \epsilon \RR
-    & \textbf{(or-expd)}       \\
-%%%
-& \frac
-    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
-%   -----------------------------------------------------------
-    { \DS \LL (p~@or~q),n,\epsilon \RR \NST \LL (p'~@or~q), n, e \RR }
-    & \textbf{(or-adv1)}   \\
-%%%
-& \frac
-    { \DS isblocked(n,p) \1,\2 \LL q,n,\epsilon \RR \NST \LL q',n,e \RR }
-%   -----------------------------------------------------------
-    { \DS \LL (p~@or~q),n,\epsilon \RR \NST \LL (p~@or~q'), n, e \RR }
-    & \textbf{(or-adv2)}   %\\
-\end{eqnarray*}
-}
-%
-Rules \textbf{and-expd} and \textbf{or-expd} insert a $@canrun(n)$ at the
+The semantic rules for~$\ceu{\And}$ and~$\ceu{\Or}$ compositions
+force transitions on their left branches~$p$ to occur before their right
+branches~$q$:
+\begin{gather*}
+  \hskip-.5em
+  \<\ceu{p\And{q}},n,\nil>
+  \nst\<\ceu{p\And(\CanRun(n);q)},n,\nil>
+  \Rtag{and-expd}\\[2\jot]
+  %%
+  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
+  \UnaryInfC{$\<\ceu{p\AtAnd{q}},n,\nil>\nst\<\ceu{p'\AtAnd{q}},n,e>$}
+  \DisplayProof
+  \Rtag{and-adv1}\\[2\jot]
+  %%
+  \AxiomC{$\isblocked(p,n)$}
+  \AxiomC{$\<q,n,\nil>\nst\<q',n,e>$}
+  \BinaryInfC{$\<\ceu{p\AtAnd{q}},n,\nil>\nst\<\ceu{p\AtAnd{q'}},n,e>$}
+  \DisplayProof
+  \Rtag{and-adv2}\\[2\jot]
+  %%
+  \<\ceu{p\Or{q}},n,\nil>
+  \nst\<\ceu{p\Or(\CanRun(n);q)},n,\nil>
+  \Rtag{or-expd}\\[2\jot]
+  %%
+  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
+  \UnaryInfC{$\<\ceu{p\AtOr{q}},n,\nil>\nst\<\ceu{p'\AtOr{q}},n,e>$}
+  \DisplayProof
+  \Rtag{or-adv1}\\[2\jot]
+  %%
+  \AxiomC{$\isblocked(p,n)$}
+  \AxiomC{$\<q,n,\nil>\nst\<q',n,e>$}
+  \BinaryInfC{$\<\ceu{p\AtOr{q}},n,\nil>\nst\<\ceu{p\AtOr{q'}},n,e>$}
+  \DisplayProof
+  \Rtag{or-adv2}
+\end{gather*}
+
+%-
+% { \setlength{\jot}{20pt}
+% \begin{eqnarray*}
+% & \LL (p~and~q),n,\epsilon \RR \NST \LL (p~@and~(@canrun(n)~;~q)),n,\epsilon \RR
+%     & \textbf{(and-expd)}       \\
+% %%%
+% & \frac
+%     { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
+% %   -----------------------------------------------------------
+%     { \DS \LL (p~@and~q),n,\epsilon \NST \LL (p'~@and~q),n,e \RR }
+%     & \textbf{(and-adv1)}      \\
+% %%%
+% & \frac
+%     { \DS isblocked(n,p) \1,\2 \LL q,n,\epsilon \RR \NST \LL q',n,e \RR }
+% %   -----------------------------------------------------------
+%     { \DS \LL (p~@and~q),n,\epsilon \RR \NST \LL (p~@and~q'), n, e \RR }
+%     & \textbf{(and-adv2)}      \\
+% %%%
+% & \LL (p~or~q), n, \epsilon \RR \NST \LL (p~@or~(@canrun(n)~;~q)), n, \epsilon \RR
+%     & \textbf{(or-expd)}       \\
+% %%%
+% & \frac
+%     { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
+% %   -----------------------------------------------------------
+%     { \DS \LL (p~@or~q),n,\epsilon \RR \NST \LL (p'~@or~q), n, e \RR }
+%     & \textbf{(or-adv1)}   \\
+% %%%
+% & \frac
+%     { \DS isblocked(n,p) \1,\2 \LL q,n,\epsilon \RR \NST \LL q',n,e \RR }
+% %   -----------------------------------------------------------
+%     { \DS \LL (p~@or~q),n,\epsilon \RR \NST \LL (p~@or~q'), n, e \RR }
+%     & \textbf{(or-adv2)}   %\\
+% \end{eqnarray*}
+% }
+%-
+
+Rules \R{and-expd} and \R{or-expd} insert a~$\ceu{\CanRun(n)}$ at the
 beginning of the right branch.
-This ensures that an $emitInt$ on the left branch, which transits to a
-$@canrun(n)$, still resumes before the right branch starts.
+This ensures that~any $\ceu{\EmitInt}$ on the left branch, which transits to a~$\ceu{\CanRun(n)}$, still resumes before the right branch starts.
 %
@@
 predicate (see Figure~\ref{fig.isblocked}) which appears in rules
-\textbf{and-adv2} and \textbf{or-adv2}.
-These rules require the left branch $p$ to be blocked in order to allow the 
-right branch transition from $q$ to $q'$.
+\R{and-adv2} and \R{or-adv2}.
+These rules require the left branch~$p$ to be blocked for the
+right branch to transition from~$q$ to~$q'$.
 
-For a parallel $@and$, if one of the sides terminates, the composition is
-simply substituted by the other side (rules \textbf{and-nop1} and
-\textbf{and-nop2}.
+In a parallel~$\ceu{\AtAnd}$, if one of the sides terminates, the composition is
+simply substituted by the other side (rules \R{and-nop1} and
+\R{and-nop2} below).
 %
-For a parallel $or$, if one of the sides terminates, the whole composition 
-terminates, also applying the $clear$ function to properly finalize the aborted 
-side (rules \textbf{or-nop1} and \textbf{or-nop2}):
-%
-{ \setlength{\jot}{20pt}
-\begin{eqnarray*}
-& \LL (@nop~@and~q), n, \epsilon \RR \NST \LL q,n,\epsilon \RR
-    & \textbf{(and-nop1)}   \\
-%%%
-& \frac
-    { \DS isblocked(n,p) }
-%   -----------------------------------------------------------
-    { \DS \LL (p~@and~@nop), n, \epsilon \RR \NST \LL p,n,\epsilon \RR }
-    & \textbf{(and-nop2)}   \\
-%%%
-& \LL (@nop~@or~q), n, \epsilon \RR \NST \LL clear(q),n,\epsilon \RR
-    & \textbf{(or-nop1)}   \\
-%%%
-& \frac
-    { \DS isblocked(n,p) }
-%   -----------------------------------------------------------
-    { \DS \LL (p~@or~@nop), n, \epsilon \RR \NST \LL clear(p),n,\epsilon \RR }
-    & \textbf{(or-nop2)}   %\\
-\end{eqnarray*}
-}
-%
-The $clear$ function (see Figure~\ref{fig.formal.clear}) concatenates all
-active $fin$ bodies of the side being aborted, so that they execute before the
+In a parallel~$\ceu{\AtOr}$, however, if one of the sides terminates, the whole composition
+terminates and function~$\clear$ is used to properly finalize the aborted
+side (rules \R{or-nop1} and \R{or-nop2}).
+\begin{gather*}
+  \<\ceu{{\Nop}\AtAnd{q}},n,\nil>\nst\<q,n,\nil>\Rtag{and-nop1}\\[2\jot]
+  %%
+  \AxiomC{$\isblocked(p,n)$}
+  \UnaryInfC{$\<\ceu{p\AtAnd{\Nop}},n,\nil>\nst\<p,n,\nil>$}
+  \DisplayProof
+  \Rtag{and-nop2}\\[2\jot]
+  %%
+  \<\ceu{{\Nop}\AtOr{q}},n,\nil>\nst\<\clear(q),n,\nil>\Rtag{or-nop1}\\[2\jot]
+  %%
+  \AxiomC{$\isblocked(p,n)$}
+  \UnaryInfC{$\<\ceu{p\AtOr{\Nop}},n,\nil>\nst\<\clear(p),n,\nil>$}
+  \DisplayProof
+  \Rtag{or-nop2}
+\end{gather*}
+
+%-
+% { \setlength{\jot}{20pt}
+% \begin{eqnarray*}
+% & \LL (@nop~@and~q), n, \epsilon \RR \NST \LL q,n,\epsilon \RR
+%     & \textbf{(and-nop1)}   \\
+% %%%
+% & \frac
+%     { \DS isblocked(n,p) }
+% %   -----------------------------------------------------------
+%     { \DS \LL (p~@and~@nop), n, \epsilon \RR \NST \LL p,n,\epsilon \RR }
+%     & \textbf{(and-nop2)}   \\
+% %%%
+% & \LL (@nop~@or~q), n, \epsilon \RR \NST \LL clear(q),n,\epsilon \RR
+%     & \textbf{(or-nop1)}   \\
+% %%%
+% & \frac
+%     { \DS isblocked(n,p) }
+% %   -----------------------------------------------------------
+%     { \DS \LL (p~@or~@nop), n, \epsilon \RR \NST \LL clear(p),n,\epsilon \RR }
+%     & \textbf{(or-nop2)}   %\\
+% \end{eqnarray*}
+% }
+%-
+
+The~$\clear$ function (see Figure~\ref{fig.clear}) concatenates all
+active~$\ceu{\Fin}$ bodies of the side being aborted, so that they execute before the
 composition rejoins.
-Note that there are no transition rules for $fin$ expressions.
-This is because once reached, a $fin$ expression halts and will only execute 
-when it is aborted by a trail in parallel and is expanded by the $clear$ 
+Note that there are no transition rules for~$\ceu{\Fin}$ expressions.
+This is because once reached, a $\ceu{\Fin}$ expression halts and will only execute 
+when it is aborted by a parallel trail and is expanded by the~$\clear$ 
 function.
@@
 %
-Note also that there is a syntactic restriction that $fin$ bodies cannot
-contain awaiting expressions ($awaitExt$, $awaitInt$, $every$, and $fin$),
-i.e., they are guaranteed to execute entirely within a reaction.
+Note also that there is a syntactic restriction that postulates that~$\ceu{\Fin}$ bodies cannot
+contain awaiting expressions ($\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$,
+$\ceu{\Every}$, or $\ceu{\Fin}$),
+i.e., the result of a~$\clear$ call is guaranteed to execute entirely within a reaction.
 
-Finally, a $break$ in one of the sides in parallel escapes the closest 
-enclosing $loop$, properly aborting the other side by applying the $clear$ 
+Finally, a~$\ceu{\Break}$ in one of the sides in parallel escapes the closest
+enclosing~$\ceu{\Loop}$, properly aborting the other side by applying the~$\clear$
 function:
+\begin{gather*}
+  \hskip-.5em
+  \<\ceu{{\Break}\AtAnd{q}},n,\nil>\nst\<\ceu{\clear(q);\Break},n,\nil>
+  \Rtag{and-brk1}\\[2\jot]
+  %%
+  \hskip-.5em
+  \AxiomC{$\isblocked(p,n)$}
+  \UnaryInfC{$\<\ceu{p\AtAnd{\Break}},n,\nil>
+    \nst\<\ceu{\clear(p);\Break},n,\nil>$}
+  \DisplayProof
+  \Rtag{and-brk2}\\[2\jot]
+  %%
+  \<\ceu{{\Break}\AtOr{q}},n,\nil>\nst\<\ceu{\clear(q);\Break},n,\nil>
+  \Rtag{or-brk1}\\[2\jot]
+  %%
+  \AxiomC{$\isblocked(p,n)$}
+  \UnaryInfC{$\<\ceu{p\AtOr{\Break}},n,\nil>
+    \nst\<\ceu{\clear(p);\Break},n,\nil>$}
+  \DisplayProof
+  \Rtag{or-brk2}
+\end{gather*}
+
+%-
+% { \setlength{\jot}{20pt}
+% \begin{eqnarray*}
+% & \LL (break~@and~q), n, \epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
+%     & \textbf{(and-brk1)}   \\
+% %%%
+% & \frac
+%     { \DS isblocked(n,p) }
+% %   -----------------------------------------------------------
+%     { \DS \LL (p~@and~break), n, \epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
+%     & \textbf{(and-brk2)}   \\
+% %%%
+% & \LL (break~@or~q),n,\epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
+%     & \textbf{(or-brk1)}   \\
+% %%%
+% & \frac
+%     { \DS isblocked(n,p) }
+% %   -----------------------------------------------------------
+%     { \DS \LL (p~@or~break),n,\epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
+%     & \textbf{(or-brk2)}   %\\
+% \end{eqnarray*}
+% }
+%-
+
+A reaction eventually blocks in~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$, $\ceu{\Fin}$,
+and~$\ceu{\CanRun}$ expressions in parallel trails.
 %
-{ \setlength{\jot}{20pt}
-\begin{eqnarray*}
-& \LL (break~@and~q), n, \epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
-    & \textbf{(and-brk1)}   \\
-%%%
-& \frac
-    { \DS isblocked(n,p) }
-%   -----------------------------------------------------------
-    { \DS \LL (p~@and~break), n, \epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
-    & \textbf{(and-brk2)}   \\
-%%%
-& \LL (break~@or~q),n,\epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
-    & \textbf{(or-brk1)}   \\
-%%%
-& \frac
-    { \DS isblocked(n,p) }
-%   -----------------------------------------------------------
-    { \DS \LL (p~@or~break),n,\epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
-    & \textbf{(or-brk2)}   %\\
-\end{eqnarray*}
-}
-%
-A reaction eventually blocks in $awaitExt$, $awaitInt$, $every$, $fin$, and
-$@canrun$ expressions in parallel trails.
-%
-If no trails are blocked in $@canrun$ expressions, it means that the program 
+If no trails are blocked in~$\ceu{\CanRun}$ expressions, it means that the program
 cannot advance in the current reaction.
 %
-However, $@canrun$ expressions can still resume in lower stack indexes and will
-eventually resume in the current reaction (see rule \textbf{pop}).
+However, $\ceu{\CanRun}$ expressions can still resume in lower stack indexes and will
+eventually resume in the current reaction (see rule \R{pop}).
 
-\begin{figure}
-{\small
+\begin{figure}[h]
+\small
+\begin{gather*}
+  \boxed{
     \begin{align*}
-  bcast(e, awaitExt(e)) &= @nop                         \\
-  bcast(e, awaitInt(e)) &= @nop                         \\
-  bcast(e, every~e~p)   &= p;~every~e~p                 \\
-  bcast(e, @canrun(n))  &= @canrun(n)                   \\
-  bcast(e, fin~p)       &= fin~p                        \\
-  bcast(e, p~;~q)       &= bcast(e,p)~;~q               \\
-  bcast(e, p~@loop~q)   &= bcast(e,p)~@loop~q           \\
-  bcast(e, p~@and~q)    &= bcast(e,p)~@and~bcast(e,q)   \\
-  bcast(e, p~@or~q)     &= bcast(e,p)~@or~bcast(e,q)    \\
-  bcast(e, \_)          &= \bot \2 (mem,emitInt,break,if,  \\
-                                 & \5\5 loop,and,or,@nop) %\\
-\end{align*}
+      %%
+      %%-
+      \intertext{\llap{(i)~}Function~$\bcast$:}
+      %%-
+      %%
+      \bcast(\ceu{\AwaitExt(e)},e)
+      &=\ceu{\Nop}\\[-1\jot]
+      %%
+      \bcast(\ceu{\AwaitInt(e)},e)
+      &=\ceu{\Nop}\\[-1\jot]
+      %%
+      \bcast(\ceu{\Every{e}\ {p}},e)
+      &=\ceu{p;\,\Every{e}\ {p}}\\[-1\jot]
+      %%
+      \bcast(\ceu{\CanRun(n)},e)
+      &=\ceu{\CanRun(n)}\\[-1\jot]
+      %%
+      \bcast(\ceu{\Fin{p}},e)
+      &=\ceu{\Fin{p}}\\[-1\jot]
+      %%
+      \bcast(\ceu{p;\,q})
+      &=\ceu{\bcast(p,e);\,q}\\[-1\jot]
+      %%
+      \bcast(\ceu{p\AtLoop{q}},e)
+      &=\ceu{\bcast(p,e)\AtLoop{q}}\\[-1\jot]
+      %%
+      \bcast(\ceu{p\AtAnd{q}},e)
+      &=\ceu{{\bcast(p,e)}\AtAnd{\bcast(q,e)}}\\[-1\jot]
+      %%
+      \bcast(\ceu{p\AtOr{q}},e)
+      &=\ceu{{\bcast(p,e)}\AtOr{\bcast(q,e)}}\\[-1\jot]
+      %%
+      bcast(\_,e)
+      &=\bot\enspace
+        (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
+      &\quad\ceu{\Ifelse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
+      \\[1\jot]
+      %%
+      %%-
+      \intertext{\llap{(ii)~}Predicate~$\isblocked$:}
+      %%-
+      %%
+      \isblocked(\ceu{\AwaitExt(\Id)},n)
+      &=\mathit{true}\\[-1\jot]
+      %%
+      \isblocked(\ceu{\AwaitInt(\Id)},n)
+      &=\mathit{true}\\[-1\jot]
+      %%
+      \isblocked(\ceu{\Every{e}\ {p}},n)
+      &=\mathit{true}\\[-1\jot]
+      %%
+      \isblocked(\ceu{\CanRun(m)},n)
+      &=(n>m)\\[-1\jot]
+      %%
+      \isblocked(\ceu{\Fin{p}},n)
+      &=\mathit{true}\\[-1\jot]
+      %%
+      \isblocked(\ceu{p;\,q},n)
+      &=\isblocked(p,n)\\[-1\jot]
+      %%
+      \isblocked(\ceu{p\AtLoop{q}},n)
+      &=\isblocked(p,n)\\[-1\jot]
+      %%
+      \isblocked(\ceu{p\AtAnd{q}},n)
+      &=\isblocked(p,n)\land\isblocked(q,n)\\[-1\jot]
+      %%
+      \isblocked(\ceu{p\AtOr{q}},n)
+      &=\isblocked(p,n)\land\isblocked(q,n)\\[-1\jot]
+      %%
+      \isblocked(\_,n)
+      &=\mathit{false}\enspace
+        (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
+      &\quad\ceu{\Ifelse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
+        \\[1\jot]
+      %%
+      %%-
+      \intertext{\llap{(iii)~}Function~$\clear$:}
+      %%-
+      %%
+      \clear(\ceu{\AwaitExt(e)})
+      &=\ceu{\Nop}\\[-1\jot]
+      %%
+      \clear(\ceu{\AwaitInt(e)})
+      &=\ceu{\Nop}\\[-1\jot]
+      %%
+      \clear(\ceu{\Every{e}\ p})
+      %%
+      &=\ceu{\Nop}\\[-1\jot]
+      %%
+      \clear(\ceu{\CanRun(n)})
+      &=\ceu{\Nop}\\[-1\jot]
+      %%
+      \clear(\ceu{\Fin{p}})
+      &=p\\[-1\jot]
+      %%
+      \clear(\ceu{p;\,q})
+      &=\clear(p)\\[-1\jot]
+      %%
+      \clear(\ceu{p\AtLoop{q}})
+      &=\clear(p)\\[-1\jot]
+      %%
+      \clear(\ceu{p\AtAnd{q}})
+      &=\ceu{\clear(p);\,\clear(q)}\\[-1\jot]
+      %%
+      \clear(\ceu{p\AtOr{q}})
+      &=\ceu{\clear(p);\,\clear(q)}\\[-1\jot]
+      %%
+      \clear(\_)
+      &=\bot\enspace
+        (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
+      &\quad\ceu{\Ifelse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
+    \end{align*}}
+\end{gather*}
+\vskip-2\belowdisplayskip
+\caption{%
+  (i)~Function~$\bcast$ awakes awaiting trails matching the event by
+  converting~$\ceu{\protect\AwaitExt}$ and~$\ceu{\protect\AwaitInt}$
+  to~$\ceu{\protect\Nop}$ expressions, and by unwinding $\ceu{\protect\Every}$
+  expressions.
+  %%
+  \enspace(ii)~Predicate~$\isblocked$ is true only if all branches in parallel
+  are blocked waiting for events, for finalization clauses, or for certain
+  stack levels.
+  %%
+  \enspace(iii)~Function~$\clear$ extracts~$\ceu{\protect\Fin}$ expressions in
+  parallel and put their bodies in sequence.
 }
-\caption{
-The function $bcast$ awakes awaiting trails matching the event by converting
-$awaitExt$ and $awaitInt$ to $@nop$ expressions, and by unwinding $every$
-expressions.
 \label{fig.bcast}
-}
+\label{fig.isblocked}
+\label{fig.clear}
 \end{figure}
 
-\begin{figure}
-{\small
-\begin{align*}
-  isblocked(n, \1 awaitExt(id)) &= true                                   \\
-  isblocked(n, \1 awaitInt(id)) &= true                                   \\
-  isblocked(n, \1 every~e~p)    &= true                                   \\
-  isblocked(n, \1 @canrun(m))   &= (n > m)                                \\
-  isblocked(n, \1 fin~p)        &= true                                   \\
-  isblocked(n, \1 p~;~q)        &= isblocked(n,p)                         \\
-  isblocked(n, \1 p~@loop~q)    &= isblocked(n,p)                         \\
-  isblocked(n, \1 p~@and~q)     &= isblocked(n,p) \wedge isblocked(n,q)   \\
-  isblocked(n, \1 p~@or~q)      &= isblocked(n,p) \wedge isblocked(n,q)   \\
-  isblocked(n, \1 \_)           &= false \2 (mem,emitInt,break,if,        \\
-                                & \5\5\5\1 loop,and,or,@nop)   %\\
-\end{align*}
-}
-\caption{
-The predicate $isblocked$ is true only if all branches in parallel are blocked
-waiting for events, finalization clauses, or certain stack levels.
-\label{fig.isblocked}
-}
-\end{figure}
+%-
+% {\small
+% \begin{align*}
+%   bcast(e, awaitExt(e)) &= @nop                         \\
+%   bcast(e, awaitInt(e)) &= @nop                         \\
+%   bcast(e, every~e~p)   &= p;~every~e~p                 \\
+%   bcast(e, @canrun(n))  &= @canrun(n)                   \\
+%   bcast(e, fin~p)       &= fin~p                        \\
+%   bcast(e, p~;~q)       &= bcast(e,p)~;~q               \\
+%   bcast(e, p~@loop~q)   &= bcast(e,p)~@loop~q           \\
+%   bcast(e, p~@and~q)    &= bcast(e,p)~@and~bcast(e,q)   \\
+%   bcast(e, p~@or~q)     &= bcast(e,p)~@or~bcast(e,q)    \\
+%   bcast(e, \_)          &= \bot \2 (mem,emitInt,break,if,  \\
+%                                  & \5\5 loop,and,or,@nop) %\\
+% \end{align*}
+% }
+%-
 
-\begin{figure}[b]
-{\small
-\begin{align*}
-  clear( awaitExt(e) ) &= @nop                  \\
-  clear( awaitInt(e) ) &= @nop                  \\
-  clear( every~e~p )   &= @nop                  \\
-  clear( @canrun(n) )  &= @nop                  \\
-  clear( fin~p )       &= p                     \\
-  clear( p~;~q )       &= clear(p)              \\
-  clear( p~@loop~q )   &= clear(p)              \\
-  clear( p~@and~q )    &= clear(p)~;~clear(q)   \\
-  clear( p~@or~q )     &= clear(p)~;~clear(q)   \\
-  clear( \_ )          &= \bot \2 (mem,emitInt,break,if, \\
-                                  & \5\5 loop,and,or,@nop) %\\ 
-\end{align*}
-}
-\caption{
-The function $clear$ extracts $fin$ expressions in parallel and put their 
-bodies in sequence.
-\label{fig.formal.clear}
-}
-\end{figure}
+%
+%-
+% {\small
+% \begin{align*}
+%   isblocked(n, \1 awaitExt(id)) &= true                                   \\
+%   isblocked(n, \1 awaitInt(id)) &= true                                   \\
+%   isblocked(n, \1 every~e~p)    &= true                                   \\
+%   isblocked(n, \1 @canrun(m))   &= (n > m)                                \\
+%   isblocked(n, \1 fin~p)        &= true                                   \\
+%   isblocked(n, \1 p~;~q)        &= isblocked(n,p)                         \\
+%   isblocked(n, \1 p~@loop~q)    &= isblocked(n,p)                         \\
+%   isblocked(n, \1 p~@and~q)     &= isblocked(n,p) \wedge isblocked(n,q)   \\
+%   isblocked(n, \1 p~@or~q)      &= isblocked(n,p) \wedge isblocked(n,q)   \\
+%   isblocked(n, \1 \_)           &= false \2 (mem,emitInt,break,if,        \\
+%                                 & \5\5\5\1 loop,and,or,@nop)   %\\
+% \end{align*}
+% }
+%-
+
+%-
+% {\small
+% \begin{align*}
+%   clear( awaitExt(e) ) &= @nop                  \\
+%   clear( awaitInt(e) ) &= @nop                  \\
+%   clear( every~e~p )   &= @nop                  \\
+%   clear( @canrun(n) )  &= @nop                  \\
+%   clear( fin~p )       &= p                     \\
+%   clear( p~;~q )       &= clear(p)              \\
+%   clear( p~@loop~q )   &= clear(p)              \\
+%   clear( p~@and~q )    &= clear(p)~;~clear(q)   \\
+%   clear( p~@or~q )     &= clear(p)~;~clear(q)   \\
+%   clear( \_ )          &= \bot \2 (mem,emitInt,break,if, \\
+%                                   & \5\5 loop,and,or,@nop) %\\ 
+% \end{align*}
+% }
+%-
 

### 2018-02-16 03-semantics.tex
Substituição de "id" por "e" na Figura 4
Remoção do negrito na definição dos conjuntos N, P, E
Acho que letra maiúscula é suficiente para indicar que é um conjunto.
Não usei o \mathbb por que fica estranho para P e E.  E é comum usar N para
naturais em semântica, já que você está definindo a sintaxe e não os números
propriamente ditos.

@@
 At the beginning of the reaction, an external event is emitted, which
-will trigger rule \textbf{push}, which will immediately raise the stack level
+will trigger rule \R{push}, which will immediately raise the stack level
 to~1.
@@
       %%
-      \isblocked(\ceu{\AwaitExt(\Id)},n)
+      \isblocked(\ceu{\AwaitExt(e)},n)
       &=\mathit{true}\\[-1\jot]
       %%
-      \isblocked(\ceu{\AwaitInt(\Id)},n)
+      \isblocked(\ceu{\AwaitInt(e)},n)
       &=\mathit{true}\\[-1\jot]

### 2018-02-17 06-appendix.tex
Adiciona lemas inicias sobre determinismo no apêndice

\appendix
\section{Proofs}
\label{sec.proofs}

\theoremstyle{remark}
\newtheorem{case}{Case}
\makeatletter
\newtheoremstyle{subcase}%
  {\thm@preskip\topsep \divide\thm@preskip\tw@}% space above
  {\thm@postskip\thm@preskip}% space below
  {\addtolength{\@totalleftmargin}{\parindent}%
    \addtolength{\linewidth}{-\parindent}%
    \parshape 1 \parindent \linewidth
    \normalfont}% body font
  {\z@}% indent amount
  {\itshape}% head font
  {.}% punctuation after head
  {.5em}% spacing after head
  {}% head spec
\makeatletter
\theoremstyle{subcase}
\newtheorem{subcase}{Case}
\numberwithin{subcase}{case}
\AtBeginEnvironment{proof}{\setcounter{case}{0}}
\AtBeginEnvironment{case}{\setcounter{subcase}{0}}

% \makeatletter
% \newcommand\case{\@startsection{paragraph}{4}{\z@}%
%    {-.5\baselineskip \@plus -2\p@ \@minus -.2\p@}%
%    {-3.5\p@}%
%    {\@parfont}}
%  \def\@parfont{\itshape}
% \makeatother

\begin{lemma}
  If~$\delta\out\delta_1$ and~$\delta\out\delta_2$ then~$\delta_1=\delta_2$.
\end{lemma}
\begin{proof}
  The lemma is vacuously true if~$\delta$ cannot be advanced by~$\out$
  transitions.  Suppose that is not the case and let~$\delta=\<p,n,e>$,
  $\delta_1=\<p_1,n_1,e_1>$ and~$\delta_2=\<p_2,n_2,e_2>$.  Then, there are
  two possibilities.
  \begin{case}
    $e\ne\nil$.  Both transitions are applications of rule~\R{push}.
    Hence~$p_1=p_2=\bcast(p,e)$, $n_1=n_2=n+1$, and~$e_1=e_2=\nil$.
  \end{case}
  \begin{case}
    $e=\nil$.  Both transitions are applications of rule~\R{pop}.
    Hence~$p_1=p_2=p$, $n_1=n_2=n-1$, and~$e_1=e_2=\nil$.\qedhere
  \end{case}
\end{proof}

\begin{lemma}
  If~$\delta\nst\delta_1$ and~$\delta\nst\delta_2$ then~$\delta_1=\delta_2$.
\end{lemma}
\begin{proof}
  By induction on the structure of~$\nst$ derivations.  The lemma is
  vacuously true if~$\delta$ cannot be advanced by~$\nst$ transitions.
  Suppose that is not the case and let~$\delta=\<p,n,e>$,
  $\delta_1=\<p_1,n_1,e_1>$ and~$\delta_2=\<p_2,n_2,e_2>$.  Then, by the
  hypothesis of the lemma, there are derivations~$\pi_1$ and~$\pi_2$ such
  that
  \begin{align*}
    \pi_1&\Vdash\<p,n,e>\nst\<p_1,n_1,e_1>\\
    \pi_2&\Vdash\<p,n,e>\nst\<p_2,n_2,e_2>
  \end{align*}
  i.e., the conclusion of~$\pi_1$ is~$\<p,n,e>\nst\<p_1,n_1,e_1>$ and the
  conclusion of~$\pi_2$ is~$\<p,n,e>\nst\<p_2,n_2,e_2>$.

  By the definition of transition~$\nst$, we have that~$e=\nil$ and
  $n_1=n_2=n$.  It remains to be shown that~$p_1=p_2$ and~$e_1=e_2$.

  Depending on the structure of program~$p$, the following~11 cases are
  possible.  (Note that~$p$ cannot be an~$\ceu{\AwaitExt}$,
  $\ceu{\AwaitInt}$, $\ceu{\Break}$, $\ceu{\Every}$, $\ceu{\Fin}$,
  or~$\ceu{\Nop}$ expression as there are no~$\nst$ rules to transition such
  programs.)

  \begin{case}
    $p=\ceu{\Mem(\Id)}$.
    %%
    Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{mem}, i.e.,
    their conclusions are obtained by an application of this rule.
    Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
  \end{case}

  \begin{case}
    $p=\ceu{\EmitInt(e')}$.
    %%
    Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{emit-int}.
    Hence~$p_1=p_2=\ceu{\CanRun(n)}$ and~$e_1=e_2=e'$.
  \end{case}

  \begin{case}
    $p=\ceu{\CanRun(n)}$.
    %%
    Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{can-run}.
    Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
  \end{case}

  \begin{case}
    $p=\ceu{\Ifelse{\Mem(\Id)}{p'}{p''}}$.  There are two subcases.
    \begin{subcase}
      $\eval(\ceu{\Mem(\Id)})$ is true.
      %%
      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{if-true}.
      Hence~$p_1=p_2=p'$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $\eval(\ceu{\Mem(\Id)})$ is false.
      %%
      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{if-false}.
      Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{p';\,p''}$.  There are three subcases.
    \begin{subcase}
      $p'=\ceu{\Nop}$.
      %%
      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{seq-nop}.
    \end{subcase}
    \begin{subcase}
      $p'=\ceu{\Break}$.
      %%
      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{seq-brk}.
    \end{subcase}
    \begin{subcase}
      $p'\ne\ceu{\Nop},\ceu{\Break}$.
      %%
      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{seq-adv}.
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{\Loop{p'}}$.
    %%
    Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{loop-expd}.
  \end{case}

  \begin{case}
    $p=\ceu{p'\AtLoop{p''}}$.  There are three subcases.
    \begin{subcase}
      $p'=\ceu{\Nop}$.
      %%
      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{loop-nop}.
    \end{subcase}
    \begin{subcase}
      $p'=\ceu{\Break}$.
      %%
      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{loop-break}.
    \end{subcase}
    \begin{subcase}
      $p'\ne\ceu{\Nop},\ceu{\Break}$.
      %%
      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{loop-adv}.
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{p'\And{p''}}$.
    %%
    Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{and-expd}.
  \end{case}

  \begin{case}
    $p=\ceu{p'\AtAnd{p''}}$.
  \end{case}

  \begin{case}
    $p=\ceu{p'\Or{p''}}$.
    %%
    Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{and-expd}.
  \end{case}

  \begin{case}
    $p=\ceu{p'\AtOr{p''}}$.
  \end{case}
\end{proof}

% LocalWords:  subcase subcases

### 2018-02-18 06-appendix.tex
Completa a prova do lema A.2

@@
 
-\theoremstyle{remark}
-\newtheorem{case}{Case}
 \makeatletter
@@
   {}% head spec
+\newtheoremstyle{subsubcase}%
+  {\thm@preskip\topsep \divide\thm@preskip\tw@}% space above
+  {\thm@postskip\thm@preskip}% space below
+  {\addtolength{\@totalleftmargin}{\parindent}%
+    \addtolength{\linewidth}{-\parindent}%
+    \parshape 1 2\parindent \linewidth
+    \normalfont}% body font
+  {\z@}% indent amount
+  {\itshape}% head font
+  {.}% punctuation after head
+  {.5em}% spacing after head
+  {}% head spec
 \makeatletter
+
+\theoremstyle{remark}
+\newtheorem{case}{Case}
+
 \theoremstyle{subcase}
@@
 \numberwithin{subcase}{case}
+
+\theoremstyle{subsubcase}
+\newtheorem{subsubcase}{Case}
+\numberwithin{subsubcase}{subcase}
+
 \AtBeginEnvironment{proof}{\setcounter{case}{0}}
 \AtBeginEnvironment{case}{\setcounter{subcase}{0}}
+\AtBeginEnvironment{subcase}{\setcounter{subsubcase}{0}}
 
@@
 
-\begin{lemma}
+\begin{lemma}\label{lem.det-out}
   If~$\delta\out\delta_1$ and~$\delta\out\delta_2$ then~$\delta_1=\delta_2$.
@@
 
-\begin{lemma}
+\begin{lemma}\label{lem.det-nst}
   If~$\delta\nst\delta_1$ and~$\delta\nst\delta_2$ then~$\delta_1=\delta_2$.
@@
     %%
-    Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{mem}, i.e.,
-    their conclusions are obtained by an application of this rule.
+    Then derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{mem},
+    i.e., their conclusions are obtained by an application of this rule.
     Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
@@
     %%
-    Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{emit-int}.
+    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{emit-int}.
     Hence~$p_1=p_2=\ceu{\CanRun(n)}$ and~$e_1=e_2=e'$.
@@
     %%
-    Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{can-run}.
+    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{can-run}.
     Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
@@
       %%
-      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{if-true}.
+      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{if-true}.
       Hence~$p_1=p_2=p'$ and~$e_1=e_2=\nil$.
@@
       %%
-      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{if-false}.
+      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{if-false}.
       Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
@@
       %%
-      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{seq-nop}.
+      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-nop}.
+      Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
     \end{subcase}
@@
       %%
-      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{seq-brk}.
+      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-brk}.
+      Hence~$p_1=p_2=\ceu{\Break}$ and~$e_1=e_2=\nil$.
     \end{subcase}
@@
       %%
-      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{seq-adv}.
+      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-adv}.
+      Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
+      \begin{align*}
+        \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
+        \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
+      \end{align*}
+      for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By induction
+      hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
+      Hence~$p_1=\ceu{p_1';p''}=\ceu{p_2';p''}=p_2$ and~$e_1=e_1'=e_2'=e_2$.
     \end{subcase}
@@
     %%
-    Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{loop-expd}.
+    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-expd}.
+    Hence~$p_1=p_2=\ceu{p'\AtLoop{p'}}$ and~$e_1=e_2=\nil$.
   \end{case}
@@
       %%
-      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{loop-nop}.
+      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-nop}.
+      Hence~$p_1=p_2=\ceu{\Loop{p''}}$ and~$e_1=e_2=\nil$.
     \end{subcase}
@@
       %%
-      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{loop-break}.
+      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-break}.
+      Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
     \end{subcase}
@@
       %%
-      Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{loop-adv}.
+      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-adv}.
+      Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
+      \begin{align*}
+        \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
+        \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
+      \end{align*}
+      for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By induction
+      hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
+      Hence~$p_1=\ceu{p_1'\AtLoop{p''}}=\ceu{p_2'\AtLoop{p''}}=p_2$
+      and~$e_1=e_1'=e_2'=e_2$.
     \end{subcase}
@@
     %%
-    Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{and-expd}.
+    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-expd}.
+    Hence~$p_1=p_2=\ceu{{p'}\And{(\CanRun(n);\,p'')}}$ and~$e_1=e_2=\nil$.
   \end{case}
@@
   \begin{case}
-    $p=\ceu{p'\AtAnd{p''}}$.
+    $p=\ceu{p'\AtAnd{p''}}$.  There are two subcases.
+    \begin{subcase}
+      $\isblocked(p',n)$~is false.  There are three subcases.
+      \begin{subsubcase}
+        $p'=\ceu{\Nop}$.
+        %%
+        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop1}.
+        Hence~$p_1=p_2=p''$ and~$e_1=e_2\nil$.
+      \end{subsubcase}
+      \begin{subsubcase}\label{lem.det-nst.and-brk1}
+        $p'=\ceu{\Break}$.
+        %%
+        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-brk1}.
+        Hence~$p_1=p_2=\ceu{\clear(p'');\Break}$ and~$e_1=e_2\nil$.
+      \end{subsubcase}
+      \begin{subsubcase}\label{lem.det-nst.and-adv1}
+        $p'\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv1}.
+        Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
+        \begin{align*}
+          \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
+          \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
+        \end{align*}
+        for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By induction
+        hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
+        Hence~$p_1=\ceu{{p_1'}\And{p''}}=\ceu{{p_2'}\And{p''}}=p_2$
+        and~$e_1=e_1'=e_2'=e_2$.
+      \end{subsubcase}
+    \end{subcase}
+    \begin{subcase}
+      $\isblocked(p',n)$~is true.  There are three subcases.
+      \begin{subsubcase}
+        $p''=\ceu{\Nop}$.
+        %%
+        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop2}.
+        Hence~$p_1=p_2=p'$ and~$e_1=e_2\nil$.
+      \end{subsubcase}
+      \begin{subsubcase}\label{lem.det-nst.and-brk2}
+        $p''=\ceu{\Break}$.
+        %%
+        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-brk2}.
+        Hence~$p_1=p_2=\ceu{\clear(p');\Break}$ and~$e_1=e_2=\nil$.
+      \end{subsubcase}
+      \begin{subsubcase}\label{lem.det-nst.and-adv2}
+        $p''\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv2}.  Thus there
+        are derivations~$\pi_1''$ and~$\pi_2''$ such that
+        \begin{align*}
+          \pi_1''&\Vdash\<p'',n,\nil>\nst\<p_1'',n,e_1''>\\
+          \pi_2''&\Vdash\<p'',n,\nil>\nst\<p_2'',n,e_2''>
+        \end{align*}
+        for some~$p_1'',p_2''\in\P$ and~$e_1'',e_2''\in\E$.  By induction
+        hypothesis, $p_1''=p_2''$ and~$e_1''=e_2''$.
+        Hence~$p_1=\ceu{{p'}\And{p_1''}}=\ceu{{p'}\And{p_2''}}=p_2$
+        and~$e_1=e_1''=e_2''=e_2$.
+      \end{subsubcase}
+    \end{subcase}
   \end{case}
@@
     %%
-    Derivations~$\pi_1$ and~$\pi_2$ are instances of rule~\R{and-expd}.
+    Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-expd}.
+    Hence~$p_1=p_2=\ceu{{p'}\Or{(\CanRun(n);\,p'')}}$ and~$e_1=e_2=\nil$.
   \end{case}
@@
   \begin{case}
-    $p=\ceu{p'\AtOr{p''}}$.
+    $p=\ceu{p'\AtOr{p''}}$.  There are two subcases.
+    \begin{subcase}
+      $\isblocked(p',n)$~is false.  There are three subcases.
+      \begin{subsubcase}
+        $p'=\ceu{\Nop}$.
+        %%
+        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-nop1}.
+        Hence~$p_1=p_2=\clear(p'')$ and~$e_1=e_2\nil$.
+      \end{subsubcase}
+      \begin{subsubcase}
+        $p'=\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{lem.det-nst.and-brk1}.
+      \end{subsubcase}
+      \begin{subsubcase}
+        $p'\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{lem.det-nst.and-adv1}.
+      \end{subsubcase}
+    \end{subcase}
+    \begin{subcase}
+      $\isblocked(p',n)$~is true.  There are three subcases.
+      \begin{subsubcase}
+        $p''=\ceu{\Nop}$.
+        %%
+        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-nop1}.
+        Hence~$p_1=p_2=\clear(p')$ and~$e_1=e_2\nil$.
+      \end{subsubcase}
+      \begin{subsubcase}
+        $p''=\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{lem.det-nst.and-brk2}.
+      \end{subsubcase}
+      \begin{subsubcase}
+        $p''\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{lem.det-nst.and-adv2}.
+        %%
+        \qedhere
+      \end{subsubcase}
+    \end{subcase}
   \end{case}

### 2018-02-18 03-semantics.tex
Corrige regra loop-adv na Seção 3.2
Adiciona as macros \outpush e \outpop à header.tex
Essas macros corrigem a tipografia de \out[push] e \out[pop].

@@
 \begin{align*}
-  \<p,0,e_\ext>\out[push]\<p_1,1,\nil>
+  \<p,0,e_\ext>\outpush\<p_1,1,\nil>
   \Big[\null\nst[*]\null\out\null\Big]\!\!\ast
-  \null\nst[*]\null\out[pop]\<p',0,\nil>\,.
+  \null\nst[*]\null\outpop\<p',0,\nil>\,.
 \end{align*}
@@
 %
-First, a~$\out[push]$ starts a nested reaction at level~1.
+First, a~$\outpush$ starts a nested reaction at level~1.
 Then, a series of alternations between zero or more~$\nst$ transitions (nested reactions) and a
 single~$\out$ transition (stack operation) takes place.
-Finally, a last~$\out[pop]$ transition decrements the
+Finally, a last~$\outpop$ transition decrements the
 stack level to~0 and terminates the reaction.
@@
   &\hskip-6.35em
-  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
-  \UnaryInfC{$\<\ceu{p\AtLoop{p}},n,\nil>\nst\<\ceu{p'\AtLoop{p}},n,e>$}
+  \AxiomC{$\<q,n,\nil>\nst\<q',n,e>$}
+  \UnaryInfC{$\<\ceu{q\AtLoop{p}},n,\nil>\nst\<\ceu{q'\AtLoop{p}},n,e>$}
   \DisplayProof

### 2018-02-18 06-appendix.tex
Adiciona dois teoremas sobre determinismo ao apêndice
Determinismo de n aplicações de \out e \nst

@@
   \end{case}
+\end{proof}
+
+\begin{theorem}\label{thm.det-out-pop-n}
+  If~$\delta\out[n]\delta_1$ and~$\delta\out[n]\delta_2$
+  then~$\delta_1=\delta_2$.
+\end{theorem}
+\begin{proof}
+  By induction on~$n$.
+  %%
+  The theorem is trivially true for~$n=0$ and follows directly from
+  Lemma~\ref{lem.det-out} for~$n=1$.  (Note that, for~$n>1$, the theorem is
+  vacuously true for~$\outpush$ transitions.  By the format of
+  rules~\R{push} and~\R{pop}, transitions~$\outpush$ cannot be applied more
+  than once in a row and cannot occur after a~$\outpop$ transition.)
+
+  Suppose
+  \[
+    \delta\out[1]\delta_1'\out[n-1]\delta_1
+    \quad\text{and}\quad
+    \delta\out[1]\delta_2'\out[n-1]\delta_2,
+  \]
+  for some~$n>1$ and~$\delta_1'$, $\delta_2'\in\Delta$.
+  %%
+  By Lemma~\ref{lem.det-out}, $\delta_1'=\delta_2'$.  By induction
+  hypothesis, $\delta_1=\delta_2$.\qedhere
 \end{proof}
@@
 
+\begin{theorem}\label{thm.det-nst-n}
+  If~$\delta\nst[n]\delta_1$ and~$\delta\nst[n]\delta_2$
+  then~$\delta_1=\delta_2$.
+\end{theorem}
+\begin{proof}
+  Similar to the proof of Theorem~\ref{thm.det-out-pop-n}.
+\end{proof}
+
 % LocalWords:  subcase subcases

### 2018-02-18 06-appendix.tex
Adiciona lema sobre irredutibilidade NST ao apêndice
I.e., "case", "subcase", e "subsubcase".

@@
 \label{sec.proofs}
-
-\makeatletter
-\newtheoremstyle{subcase}%
-  {\thm@preskip\topsep \divide\thm@preskip\tw@}% space above
-  {\thm@postskip\thm@preskip}% space below
-  {\addtolength{\@totalleftmargin}{\parindent}%
-    \addtolength{\linewidth}{-\parindent}%
-    \parshape 1 \parindent \linewidth
-    \normalfont}% body font
-  {\z@}% indent amount
-  {\itshape}% head font
-  {.}% punctuation after head
-  {.5em}% spacing after head
-  {}% head spec
-\newtheoremstyle{subsubcase}%
-  {\thm@preskip\topsep \divide\thm@preskip\tw@}% space above
-  {\thm@postskip\thm@preskip}% space below
-  {\addtolength{\@totalleftmargin}{\parindent}%
-    \addtolength{\linewidth}{-\parindent}%
-    \parshape 1 2\parindent \linewidth
-    \normalfont}% body font
-  {\z@}% indent amount
-  {\itshape}% head font
-  {.}% punctuation after head
-  {.5em}% spacing after head
-  {}% head spec
-\makeatletter
-
-\theoremstyle{remark}
-\newtheorem{case}{Case}
-
-\theoremstyle{subcase}
-\newtheorem{subcase}{Case}
-\numberwithin{subcase}{case}
-
-\theoremstyle{subsubcase}
-\newtheorem{subsubcase}{Case}
-\numberwithin{subsubcase}{subcase}
-
-\AtBeginEnvironment{proof}{\setcounter{case}{0}}
-\AtBeginEnvironment{case}{\setcounter{subcase}{0}}
-\AtBeginEnvironment{subcase}{\setcounter{subsubcase}{0}}
 
@@
   The theorem is trivially true for~$n=0$ and follows directly from
-  Lemma~\ref{lem.det-out} for~$n=1$.  (Note that, for~$n>1$, the theorem is
-  vacuously true for~$\outpush$ transitions.  By the format of
+  Lemma~\ref{lem.det-out} for~$n=1$.  (Note that for~$n>1$ the theorem is
+  vacuously true for~$\outpush$ transitions---by the format of
   rules~\R{push} and~\R{pop}, transitions~$\outpush$ cannot be applied more
@@
 
+\begin{definition}\label{def.Hnst}
+  A description~$\delta=\<p,n,e>$ is \emph{nested-irre\-ducible}
+  iff~$p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p,n)$~is true.
+  Nested-irreducible descriptions serve as normal forms for~$\nst$
+  transitions: their program~$p$ and stack-level~$n$ embody the result of an
+  exhaustive number of~$\nst$ transitions.  We will write~$\delta_\Hnst$ to
+  indicate that description~$\delta$ is nested-irreducible.
+\end{definition}
+
+The next lemma justifies the use of qualifier ``irreducible'' in
+Definition~\ref{def.Hnst}.
+
+\begin{lemma}\label{lem.irr-nst-n}
+  If~$\delta\nst[n]\delta_\Hnst'$ then, for all~$i\ne{n}$, there is
+  no~$\delta_\Hnst''$ such that~$\delta\nst[i]\delta''_\Hnst$.
+\end{lemma}
+\begin{proof}
+  By contradiction.  Let~$\delta\nst[n]\delta'_\Hnst$, for some~$n\ge0$.
+  There are two cases.
+  \begin{case}\label{lem.irr-nst-n-case1}
+    Suppose there are~$i>n$ and~$\delta''_\Hnst$ such
+    that~$\delta\nst[i]\delta''$.
+    %%
+    Then, by definition of~$\nst[i]$,
+    \begin{equation}\label{lem.irr-nst-n-eq1}
+      \delta\nst[n]\delta'\nst[n+1]\delta_1'\nst[n+2]\cdots\nst[i]\delta''.
+    \end{equation}
+    Since~$\delta'=\<p',n,e'>$ is nested-irreducible,
+    $p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p',n)$.  In either case, by
+    the definition of~$\nst$, there is no~$\delta_1'$ such
+    that~$\delta'\nst[1]\delta_1'$, which
+    contradicts~\eqref{lem.irr-nst-n-eq1}.  Therefore, no such~$i$ can
+    exist.
+  \end{case}
+  \begin{case}
+    Suppose there are~$0\le{i}<n$ and~$\delta''_\Hnst$ such
+    that~$\delta\nst[i]\delta''$.  Then since~$n>i$, by
+    Case~\ref{lem.irr-nst-n-case1}, $\delta'$~could not exist, which is
+    absurd.  Therefore, the assumption that there is such~$i$ is
+    false.\qedhere
+  \end{case}
+\end{proof}
+
 % LocalWords:  subcase subcases

### 2018-02-18 06-appendix.tex
Adiciona teorema sobre terminação NST ao apêndice

@@
   \begin{case}
-    $e\ne\nil$.  Both transitions are applications of rule~\R{push}.
+    $e\ne\nil$.  Both transitions are applications of~\R{push}.
     Hence~$p_1=p_2=\bcast(p,e)$, $n_1=n_2=n+1$, and~$e_1=e_2=\nil$.
@@
   \begin{case}
-    $e=\nil$.  Both transitions are applications of rule~\R{pop}.
+    $e=\nil$.  Both transitions are applications of~\R{pop}.
     Hence~$p_1=p_2=p$, $n_1=n_2=n-1$, and~$e_1=e_2=\nil$.\qedhere
@@
   A description~$\delta=\<p,n,e>$ is \emph{nested-irre\-ducible}
-  iff~$p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p,n)$~is true.
-  Nested-irreducible descriptions serve as normal forms for~$\nst$
-  transitions: their program~$p$ and stack-level~$n$ embody the result of an
-  exhaustive number of~$\nst$ transitions.  We will write~$\delta_\Hnst$ to
-  indicate that description~$\delta$ is nested-irreducible.
+  iff~$e\ne\nil$ or~$p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p,n)$~is
+  true.  Nested-irreducible descriptions serve as normal forms for~$\nst$
+  transitions: they embody the result of an exhaustive number of~$\nst$
+  transitions.  We will write~$\delta_\Hnst$ to indicate that
+  description~$\delta$ is nested-irreducible.
 \end{definition}
@@
 \begin{proof}
-  By contradiction.  Let~$\delta\nst[n]\delta'_\Hnst$, for some~$n\ge0$.
+  By contradiction on the hypothesis that there is such~$i$.
+  %%
+  Let~$\delta\nst[n]\delta'_\Hnst$, for some~$n\ge0$.
   There are two cases.
@@
 
+The next lemma establishes some basic properties of sequences of~$\nst$
+transitions.
+
+\begin{lemma}\label{lem.props-nst-n}
+  If~$\<p_1,n,e>\nst[n]\<p_1',n,e'>$ then, for any~$p_2$:
+  \begin{enumerate}[(a)]
+  \item\label{lem.props-nst-n.a}
+    $\<\ceu{p_1;\,p_2},n,e>\nst[n]\<p_1';p_2,n,e'>$;
+    %%
+  \item\label{lem.props-nst-n.b}
+    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[n]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$;
+    %%
+  \item\label{lem.props-net-n.c}
+    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[n]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$;
+    %%
+  \item\label{lem.props-net-n.d}
+    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[n]\<\ceu{{p_1}'\AtOr{p_2}},n,e'>$.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+  By induction on~$n$.
+  %%
+  \begin{enumerate}[(a)]
+  \item The lemma is trivially true for~$n=0$, as~$p_1=p_1'$, and follows
+    directly from~\R{seq-adv} for~$n=1$.  Suppose
+    \begin{equation}
+      \label{lem.props-nst-n.a.eq1}
+      \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>,
+    \end{equation}
+    for some~$n>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible, i.e.,
+    $e=\nil$ and~$p\ne{\ceu{\Nop},\ceu{\Break}}$ and~$\isblocked(p_1,n)$ is
+    false.  By~\eqref{lem.props-nst-n.a.eq1} and by~\R{seq-adv},
+    \begin{equation}
+      \label{lem.props-nst-n.a.eq2}
+      \<\ceu{p_1;\,p_2},n,e>\nst[1]\<\ceu{p_1'';\,p_2},n,e''>.
+    \end{equation}
+    From~\eqref{lem.props-nst-n.a.eq1}, by induction hypothesis,
+    \begin{equation}
+      \label{lem.props-nst-n.a.eq3}
+      \<\ceu{p_1'';\,p_2},n,e''>\nst[n-1]\<\ceu{p_1';\,p_2},n,e'>.
+    \end{equation}
+    From~\eqref{lem.props-nst-n.a.eq2} and~\eqref{lem.props-nst-n.a.eq3},
+    \[
+      \<\ceu{p_1;\,p_2},n,e>\nst[n]\<\ceu{p_1';\,p_2},n,e'>.
+    \]
+  \end{enumerate}
+\end{proof}
+
+\begin{theorem}\label{thm.term-nst-*}
+  For any~$\delta$ there is a~$\delta'_\Hnst$ such
+  that~$\delta\nst[*]\delta'_\Hnst$.
+\end{theorem}
+\begin{proof}
+  By induction on the structure of programs.
+  %%
+  Let~$\delta=\<p,n,\nil>$.  The theorem is trivially true if~$\delta$ is
+  nested-irreducible itself, as by definition~$\delta\nst[0]\delta_\Hnst$.
+  Suppose that is not the case.  Then, depending on the structure of~$p$,
+  there are~11 possibilities.  In each one of them, we show that
+  such~$\delta'_\Hnst$ indeed exists.
+  \begin{case}
+    $p=\ceu{\Mem(\Id)}$.
+    %%
+    Then, by~\R{mem},
+    \[
+      \<\ceu{\Mem(\Id)},n,\nil>\nst[1]\<\ceu{\Nop},n,\nil>_\Hnst\,.
+    \]
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{\EmitInt(e)}$.
+    %%
+    Then, by~\R{emit-int},
+    \[
+      \<\ceu{\EmitInt(e)},n,\nil>\nst[1]\<\ceu{\CanRun(n)},n,e>_\Hnst\,.
+    \]
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{\CanRun(n)}$.
+    %%
+    Then, by~\R{can-run},
+    \[
+      \<\ceu{\CanRun(n)},n,\nil>\nst[1]\<\ceu{\Nop},n,\nil>_\Hnst\,.
+    \]
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{\Ifelse{\Mem(\Id)}{p'}{p''}}$.
+    %%
+    There are two subcases.
+    \begin{subcase}
+      $\eval(\ceu{\Mem(\Id)})$~is true.
+      %%
+      Then, by~\R{if-true} and by the induction hypothesis, there is
+      a~$\delta'$ such that
+      \begin{align*}
+        \<\ceu{\Ifelse{\Mem(\Id)}{p'}{p''}},n,\nil>
+        &\nst[1]\<p',n,e>\\
+        &\nst[*]\delta'_\Hnst\,.
+      \end{align*}
+    \end{subcase}
+    \begin{subcase}
+      $\eval(\ceu{\Mem(\Id)})$~is false.
+      %%
+      Then, by~\R{if-false} and by the induction hypothesis, there is
+      a~$\delta'$ such that
+      \begin{align*}
+        \<\ceu{\Ifelse{\Mem(\Id)}{p'}{p''}},n,\nil>
+        &\nst[1]\<p'',n,e>\\
+        &\nst[*]\delta'_\Hnst\,.
+      \end{align*}
+    \end{subcase}
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{p';\,p''}$.
+    %%
+    There are three subcases.
+    \begin{subcase}
+      \label{thm.term-nst-*.seq-nop}
+      $p'=\ceu{\Nop}$.
+      %%
+      Then, by~\R{seq-nop} and by the induction hypothesis, there is
+      a~$\delta'$ such that
+      \[
+        \<\ceu{\Nop;\,p''},n,\nil>
+        \nst[1]\<p'',n,e>\nst[*]\delta'_\Hnst\,.
+      \]
+    \end{subcase}
+    \begin{subcase}
+      \label{thm.term-nst-*.seq-brk}
+      $p'=\ceu{\Break}$.
+      %%
+      Then, by~\R{seq-brk} and by the induction hypothesis, there is
+      a~$\delta'$ such that
+      \[
+        \<\ceu{\Break;\,p''},n,\nil>
+        \nst[1]\<p'',n,e>
+        \nst[*]\delta'_\Hnst\,.
+      \]
+    \end{subcase}
+    \begin{subcase}
+      $p'\ne\ceu{\Nop},\ceu{\Break}$.
+      %%
+      By induction hypothesis, there are~$p_1'$ and~$e$ such that
+      \[
+        \<p',n,\nil>\nst[*]\<p_1',n,e>_\Hnst\,.
+      \]
+      And by item~\eqref{lem.props-nst-n.a} of Lemma~\ref{lem.props-nst-n},
+      \begin{equation}
+        \label{thm.term-nst-*.seq-adv.eq1}
+        \<\ceu{p';\,p''},n,\nil>\nst[*]\<\ceu{p_1';\,p''},n,e>.
+      \end{equation}
+      It remains to be shown that~$\<\ceu{p_1';\,p''},n,e>$ is indeed
+      nested-irreducible.  There are four possibilities following from the
+      fact that the simpler~$\<p_1',n,e>$ is nested-irreducible.
+      %%
+      \begin{subsubcase}
+        $e\ne\nil$.  Then, by the definition of~$\Hnst$,
+        description~$\<\ceu{p_1';\,p''},n,e>$ is also nested-irreducible.
+      \end{subsubcase}
+      \begin{subsubcase}
+        $p_1'=\ceu{\Nop}$.
+        %%
+        From~\eqref{thm.term-nst-*.seq-adv.eq1},
+        \[
+          \<\ceu{p';\,p''},n,\nil>\nst[*]\<\ceu{\Nop;\,p''},n,e>\,.
+        \]
+        From this point on, this case is similar to
+        Case~\ref{thm.term-nst-*.seq-nop}.
+      \end{subsubcase}
+      \begin{subsubcase}
+        $p_1'=\ceu{\Break}$.
+        %%
+        From~\eqref{thm.term-nst-*.seq-adv.eq1},
+        \[
+          \<\ceu{p';\,p''},n,\nil>\nst[*]\<\ceu{\Break;\,p''},n,e>\,.
+        \]
+        From this point on, this case is similar to
+        Case~\ref{thm.term-nst-*.seq-brk}.
+      \end{subsubcase}
+      \begin{subsubcase}
+        $\isblocked(p_1',n)$ is true.
+        %%
+        Then, by definition,
+        \[
+          \isblocked(\ceu{p_1';p''},n)=\isblocked(p_1',n)=\mathit{true}\,.
+        \]
+        Thus, from~\eqref{thm.term-nst-*.seq-adv.eq1} and by the
+        definition~$\Hnst$, $\<\ceu{p_1';\,p''},n,e>$~is also
+        nested-irreducible.
+      \end{subsubcase}
+    \end{subcase}
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{\Loop{p'}}$.
+    %%
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{p'\AtLoop{p''}}$.
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{{p'}\And{p''}}$.
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{{p'}\AtAnd{p''}}$.
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{{p'}\Or{p''}}$.
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{{p'}\AtOr{p''}}$.
+  \end{case}
+\end{proof}
+
 % LocalWords:  subcase subcases

### 2018-02-19 06-appendix.tex
Adiciona ao apêndice prova (incompl.) da terminação NST

@@
 
-% \makeatletter
-% \newcommand\case{\@startsection{paragraph}{4}{\z@}%
-%    {-.5\baselineskip \@plus -2\p@ \@minus -.2\p@}%
-%    {-3.5\p@}%
-%    {\@parfont}}
-%  \def\@parfont{\itshape}
-% \makeatother
+\subsection*{Determinism}
 
@@
   than once in a row and cannot occur after a~$\outpop$ transition.)
-
-  Suppose
-  \[
-    \delta\out[1]\delta_1'\out[n-1]\delta_1
-    \quad\text{and}\quad
-    \delta\out[1]\delta_2'\out[n-1]\delta_2,
-  \]
-  for some~$n>1$ and~$\delta_1'$, $\delta_2'\in\Delta$.
   %%
-  By Lemma~\ref{lem.det-out}, $\delta_1'=\delta_2'$.  By induction
+  Suppose~$\delta\out[1]\delta_1'\out[n-1]\delta_1$
+  and~$\delta\out[1]\delta_2'\out[n-1]\delta_2$, for some~$n>1$
+  and~$\delta_1'$, $\delta_2'\in\Delta$.
+  %%
+  By Lemma~\ref{lem.det-out}, $\delta_1'=\delta_2'$.  By the induction
   hypothesis, $\delta_1=\delta_2$.\qedhere
@@
     \begin{subcase}
-      $\eval(\ceu{\Mem(\Id)})$ is true.
+      $\eval(\ceu{\Mem(\Id)})$~is true.
       %%
@@
       \end{align*}
-      for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By induction
+      for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By the induction
       hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
@@
     Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-expd}.
-    Hence~$p_1=p_2=\ceu{{p'}\And{(\CanRun(n);\,p'')}}$ and~$e_1=e_2=\nil$.
+    Hence~$p_1=p_2=\ceu{{p'}\AtAnd{(\CanRun(n);\,p'')}}$ and~$e_1=e_2=\nil$.
   \end{case}
@@
         \end{align*}
-        for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By induction
+        for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By the induction
         hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
@@
         \end{align*}
-        for some~$p_1'',p_2''\in\P$ and~$e_1'',e_2''\in\E$.  By induction
-        hypothesis, $p_1''=p_2''$ and~$e_1''=e_2''$.
+        for some~$p_1'',p_2''\in\P$ and~$e_1'',e_2''\in\E$.  By the
+        induction hypothesis, $p_1''=p_2''$ and~$e_1''=e_2''$.
         Hence~$p_1=\ceu{{p'}\And{p_1''}}=\ceu{{p'}\And{p_2''}}=p_2$
@@
     Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-expd}.
-    Hence~$p_1=p_2=\ceu{{p'}\Or{(\CanRun(n);\,p'')}}$ and~$e_1=e_2=\nil$.
+    Hence~$p_1=p_2=\ceu{{p'}\AtOr{(\CanRun(n);\,p'')}}$ and~$e_1=e_2=\nil$.
   \end{case}
@@
         Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-nop1}.
-        Hence~$p_1=p_2=\clear(p'')$ and~$e_1=e_2\nil$.
+        Hence~$p_1=p_2=\clear(p'')$ and~$e_1=e_2=\nil$.
       \end{subsubcase}
@@
         Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-nop1}.
-        Hence~$p_1=p_2=\clear(p')$ and~$e_1=e_2\nil$.
+        Hence~$p_1=p_2=\clear(p')$ and~$e_1=e_2=\nil$.
       \end{subsubcase}
@@
 
+\subsection*{Termination}
+
 \begin{definition}\label{def.Hnst}
@@
   transitions: they embody the result of an exhaustive number of~$\nst$
-  transitions.  We will write~$\delta_\Hnst$ to indicate that
+  applications.  We will write~$\delta_\Hnst$ to indicate that
   description~$\delta$ is nested-irreducible.
@@
     %%
-  \item\label{lem.props-net-n.c}
+  \item\label{lem.props-nst-n.c}
     $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[n]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$;
     %%
-  \item\label{lem.props-net-n.d}
+  \item\label{lem.props-nst-n.d}
     $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[n]\<\ceu{{p_1}'\AtOr{p_2}},n,e'>$.
@@
       \label{lem.props-nst-n.a.eq1}
-      \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>,
+      \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>\,,
     \end{equation}
     for some~$n>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible, i.e.,
-    $e=\nil$ and~$p\ne{\ceu{\Nop},\ceu{\Break}}$ and~$\isblocked(p_1,n)$ is
-    false.  By~\eqref{lem.props-nst-n.a.eq1} and by~\R{seq-adv},
+    $e=\nil$ and~$p\ne{\ceu{\Nop},\ceu{\Break}}$ and~$\isblocked(p_1'',n)$
+    is false.  By~\eqref{lem.props-nst-n.a.eq1} and by~\R{seq-adv},
     \begin{equation}
       \label{lem.props-nst-n.a.eq2}
-      \<\ceu{p_1;\,p_2},n,e>\nst[1]\<\ceu{p_1'';\,p_2},n,e''>.
+      \<\ceu{p_1;\,p_2},n,e>\nst[1]\<\ceu{p_1'';\,p_2},n,e''>\,.
     \end{equation}
-    From~\eqref{lem.props-nst-n.a.eq1}, by induction hypothesis,
+    From~\eqref{lem.props-nst-n.a.eq1}, by the induction hypothesis,
     \begin{equation}
       \label{lem.props-nst-n.a.eq3}
-      \<\ceu{p_1'';\,p_2},n,e''>\nst[n-1]\<\ceu{p_1';\,p_2},n,e'>.
+      \<\ceu{p_1'';\,p_2},n,e''>\nst[n-1]\<\ceu{p_1';\,p_2},n,e'>\,.
     \end{equation}
@@
     \[
-      \<\ceu{p_1;\,p_2},n,e>\nst[n]\<\ceu{p_1';\,p_2},n,e'>.
+      \<\ceu{p_1;\,p_2},n,e>\nst[n]\<\ceu{p_1';\,p_2},n,e'>\,.
     \]
+
+  \item Similar to Case~\eqref{lem.props-nst-n.a}.
+    %%
+    % The lemma is trivially true for~$n=0$, as~$p_1=p_1'$, and follows
+    % directly from~\R{loop-adv} for~$n=1$.  Suppose
+    % \begin{equation}
+    %   \label{lem.props-nst-n.b.eq1}
+    %   \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>\,,
+    % \end{equation}
+    % for some~$n>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible.
+    % By~\eqref{lem.props-nst-n.b.eq1} and by~\R{loop-adv},
+    % \begin{equation}
+    %   \label{lem.props-nst-n.b.eq2}
+    %   \<\ceu{p_1\AtLoop{p_2}},n,e>\nst[1]\<\ceu{p_1''\Loop{p_2}},n,e''>\,.
+    % \end{equation}
+    % From~\eqref{lem.props-nst-n.b.eq1}, by the induction hypothesis,
+    % \begin{equation}
+    %   \label{lem.props-nst-n.b.eq3}
+    %   \<\ceu{p_1''\AtLoop{p_2}},n,e''>
+    %   \nst[n-1]\<\ceu{p_1'\AtLoop{p_2}},n,e'>\,.
+    % \end{equation}
+    % From~\eqref{lem.props-nst-n.b.eq2} and~\eqref{lem.props-nst-n.b.eq3},
+    % \[
+    %   \<\ceu{p_1\AtLoop{p_2}},n,e>\nst[n]\<\ceu{p_1'\AtLoop{p_2}},n,e'>\,.
+    % \]
+    %%
+
+  \item Similar to Case~\eqref{lem.props-nst-n.a}.
+    %%
+    % The lemma is trivially true for~$n=0$, as~$p_1=p_1'$, and follows
+    % directly from~\R{and-adv1} for~$n=1$.  Suppose
+    % \begin{equation}
+    %   \label{lem.props-nst-n.c.eq1}
+    %   \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>\,,
+    % \end{equation}
+    % for some~$n>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible.
+    % By~\eqref{lem.props-nst-n.c.eq1} and by~\R{and-adv1},
+    % \begin{equation}
+    %   \label{lem.props-nst-n.c.eq2}
+    %   \<\ceu{{p_1}\AtAnd{p_2}},n,e>
+    %   \nst[1]\<\ceu{{p_1}''\AtAnd{p_2}},n,e''>\,.
+    % \end{equation}
+    % From~\eqref{lem.props-nst-n.c.eq1}, by the induction hypothesis,
+    % \begin{equation}
+    %   \label{lem.props-nst-n.a.eq3}
+    %   \<\ceu{{p_1''}\AtAnd{p_2}},n,e''>
+    %   \nst[n-1]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>\,.
+    % \end{equation}
+    % From~\eqref{lem.props-nst-n.c.eq2} and~\eqref{lem.props-nst-n.c.eq3},
+    % \[
+    %   \<\ceu{{p_1}\AtAnd{p_2}},n,e>
+    %   \nst[n]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>\,.
+    % \]
+    %%
+
+  \item Similar to Case~\eqref{lem.props-nst-n.a}.
+    %%
+    % The lemma is trivially true for~$n=0$, as~$p_1=p_1'$, and follows
+    % directly from~\R{or-adv1} for~$n=1$.  Suppose
+    % \begin{equation}
+    %   \label{lem.props-nst-n.d.eq1}
+    %   \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>\,,
+    % \end{equation}
+    % for some~$n>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible.
+    % By~\eqref{lem.props-nst-n.d.eq1} and by~\R{or-adv1},
+    % \begin{equation}
+    %   \label{lem.props-nst-n.d.eq2}
+    %   \<\ceu{{p_1}\AtOr{p_2}},n,e>
+    %   \nst[1]\<\ceu{{p_1}''\AtOr{p_2}},n,e''>\,.
+    % \end{equation}
+    % From~\eqref{lem.props-nst-n.d.eq1}, by the induction hypothesis,
+    % \begin{equation}
+    %   \label{lem.props-nst-n.a.eq3}
+    %   \<\ceu{{p_1''}\AtOr{p_2}},n,e''>
+    %   \nst[n-1]\<\ceu{{p_1'}\AtOr{p_2}},n,e'>\,.
+    % \end{equation}
+    % From~\eqref{lem.props-nst-n.d.eq2} and~\eqref{lem.props-nst-n.d.eq3},
+    % \[
+    %   \<\ceu{{p_1}\AtOr{p_2}},n,e>
+    %   \nst[n]\<\ceu{{p_1'}\AtOr{p_2}},n,e'>\,.
+    % \]
+    %%
   \end{enumerate}
 \end{proof}
+
+The syntactic restrictions discussed in Section~\ref{?}, regarding the body
+of~$\ceu{\Fin}$ and~$\ceu{\Loop}$ expressions, is formalized by the
+following assumptions.
+
+\begin{assumption}\label{ass.term-nst-fin}
+  If~$\delta=\<\ceu{\clear(p)},n,\nil>$ then there is a
+  description~$\delta'=\<\ceu{\Nop},n,\nil>$ such
+  that~$\delta\nst[*]\delta'$.
+\end{assumption}
+
+\begin{assumption}\label{ass.term-nst-loop}
+  If~$\delta=\<\ceu{\Loop{p}},n,\nil>$ then there is a
+  description~$\delta'=\<p',n,e>$ such that~$\delta\nst[*]\delta'$ where
+  either~$p'=\ceu{{\Break}\AtLoop{p}}$ or~$\isblockedext(p',n)$.
+\end{assumption}
 
@@
   Let~$\delta=\<p,n,\nil>$.  The theorem is trivially true if~$\delta$ is
-  nested-irreducible itself, as by definition~$\delta\nst[0]\delta_\Hnst$.
-  Suppose that is not the case.  Then, depending on the structure of~$p$,
-  there are~11 possibilities.  In each one of them, we show that
+  nested-irreducible, as by definition~$\delta\nst[0]\delta_\Hnst$.  Suppose
+  that is not the case.  Then, depending on the structure of~$p$, there
+  are~11 possibilities.  In each one of them, we show that
   such~$\delta'_\Hnst$ indeed exists.
@@
     \begin{subcase}
+      \label{thm.term-nst-*.seq-adv}
       $p'\ne\ceu{\Nop},\ceu{\Break}$.
       %%
-      By induction hypothesis, there are~$p_1'$ and~$e$ such that
+      Then, by the induction hypothesis, there are~$p_1'$ and~$e$ such that
       \[
@@
       \]
-      And by item~\eqref{lem.props-nst-n.a} of Lemma~\ref{lem.props-nst-n},
+      By item~\eqref{lem.props-nst-n.a} of Lemma~\ref{lem.props-nst-n},
       \begin{equation}
         \label{thm.term-nst-*.seq-adv.eq1}
-        \<\ceu{p';\,p''},n,\nil>\nst[*]\<\ceu{p_1';\,p''},n,e>.
+        \<\ceu{p';\,p''},n,\nil>\nst[*]\<\ceu{p_1';\,p''},n,e>\,.
       \end{equation}
@@
         $e\ne\nil$.  Then, by the definition of~$\Hnst$,
-        description~$\<\ceu{p_1';\,p''},n,e>$ is also nested-irreducible.
+        description~$\<\ceu{p_1';\,p''},n,e>$ is nested-irreducible.
       \end{subsubcase}
@@
       \begin{subsubcase}
-        $\isblocked(p_1',n)$ is true.
+        $\isblocked(p_1',n)$~is true.
         %%
@@
         \]
-        Thus, from~\eqref{thm.term-nst-*.seq-adv.eq1} and by the
-        definition~$\Hnst$, $\<\ceu{p_1';\,p''},n,e>$~is also
+        Hence, from~\eqref{thm.term-nst-*.seq-adv.eq1} and by the
+        definition~$\Hnst$, description~$\<\ceu{p_1';\,p''},n,e>$ is
         nested-irreducible.
@@
     %%
+    Then, by Assumption~\ref{ass.term-nst-loop},
+    \begin{equation}\label{thm.term-nst-*.loop-expd.eq1}
+      \<\ceu{\Loop{p'}},n,\nil>\nst[*]\<p_1',n,e>\,,
+    \end{equation}
+    for some~$e$ and~$p_1'$ such that either~$p_1'=\ceu{\Break\AtLoop{p'}}$
+    or~$\isblockedext(p_1',n)$.
+    \begin{subcase}
+      $p_1'=\ceu{\Break\AtLoop{p'}}$.
+      %%
+      From~\eqref{thm.term-nst-*.loop-expd.eq1}, by~\R{loop-brk},
+      \begin{align*}
+        \<\ceu{\Loop{p'}},n,\nil>
+        &\nst[*]\<\ceu{\Break\AtLoop{p'}},n,e>\\
+        &\nst[1]\<\ceu{\Nop},n,e>_\Hnst\,.
+      \end{align*}
+    \end{subcase}
+    \begin{subcase}
+      $\isblockedext(p_1',n)$ is true.  Then, by definition,
+      $\isblockedext(p_1',n)$ implies~$\isblocked(p_1',n)$.  Hence,
+      from~\eqref{thm.term-nst-*.loop-expd.eq1} and by the definition
+      of~$\Hnst$, $\<p_1',n,e>_\Hnst$.
+    \end{subcase}
   \end{case}
@@
   \begin{case}
-    $p=\ceu{p'\AtLoop{p''}}$.
+    $p=\ceu{p'\AtLoop{p''}}$.  There are three subcases.
+    \begin{subcase}
+      $p'=\ceu{\Nop}$.
+      %%
+      Then, by~\R{loop-nop} and by Assumption~\ref{ass.term-nst-loop},
+      \begin{align}
+        \label{thm.term-nst-*.loop-adv.eq1}
+        \<\ceu{\Nop\AtLoop{p''}},n,\nil>
+        &\nst[1]\<\ceu{\Loop{p''}},n,\nil>\notag\\
+        &\nst[*]\<p_1'',n,e>\,,
+      \end{align}
+      for some~$e$ and~$p_1''$ such that
+      either~$p_1''=\ceu{\Break\AtLoop{p''}}$ or~$\isblockedext(p_1'',n)$.
+      \begin{subsubcase}
+        $p_1''=\ceu{\Break\AtLoop{p''}}$.
+        From~\eqref{thm.term-nst-*.loop-adv.eq1}, by~\R{loop-brk},
+        \begin{align*}
+        \<\ceu{\Nop\AtLoop{p''}},n,\nil>
+        &\nst[1]\<\ceu{\Loop{p''}},n,\nil>\\
+        &\nst[*]\<\ceu{\Break\AtLoop{p''}},n,e>\\
+        &\nst[1]\<\ceu{\Nop},n,e>_\Hnst\,.
+        \end{align*}
+      \end{subsubcase}
+      \begin{subsubcase}
+        $\isblockedext(p_1'',n)$~is true.  Then, by definition,
+        $\isblockedext(p_1'',n)$ implies~$\isblocked(p_1'',n)$.  Hence,
+        from~\eqref{thm.term-nst-*.loop-adv.eq1} and by the definition
+        of~$\Hnst$, $\<p_1'',n,e>_\Hnst$.
+      \end{subsubcase}
+    \end{subcase}
+    \begin{subcase}
+      $p'=\ceu{\Break}$.  Then, by~\R{loop-brk},
+      \[
+        \<\ceu{\Break\AtLoop{p''}},n,\nil>
+        \nst[1]\<\ceu{\Nop},n,\nil>_\Hnst\,.
+      \]
+    \end{subcase}
+    \begin{subcase}
+      $p'\ne\ceu{\Nop},\ceu{\Break}$.  Then, by the induction hypothesis,
+      there are~$p_1'$ and~$e$ such that
+      \[
+        \<p',n,\nil>\nst[*]\<p'_1,n,e>_\Hnst\,.
+      \]
+      By item~\eqref{lem.props-nst-n.b} of Lemma~\ref{lem.props-nst-n},
+      \[
+        \<\ceu{p'\AtLoop{p''}},n,\nil>
+        \nst[*]\<\ceu{p_1'\AtLoop{p''}},n,e>\,.
+      \]
+      It remains to be show that~$\<\ceu{p_1'\AtLoop{p''}},n,e>$ is indeed
+      nested-irreducible.  The rest of this proof is similar to that of
+      Case~\ref{thm.term-nst-*.seq-adv}.
+    \end{subcase}
   \end{case}
@@
     $p=\ceu{{p'}\And{p''}}$.
+    %%
+    Then, by~\R{and-expd},
+    \[
+      \<\ceu{{p'}\And{p''}},n,\nil>
+      \nst[1]\<\ceu{{p'}\AtAnd{(\CanRun(n);\,p'')}},n,\nil>\,.
+    \]
+    From this point on, this case is similar to
+    Case~\ref{thm.term-nst-*.and}.
   \end{case}
 
-  \begin{case}
+  \begin{case}\label{thm.term-nst-*.and}
     $p=\ceu{{p'}\AtAnd{p''}}$.
+    %%
+    There are four subcases.
+    \begin{subcase}
+      $\isblocked(p',n)$ is false and~$\isblocked(p'',n)$ is false.
+      %%
+      There are three subcases.
+      \begin{subsubcase}
+        \label{thm.term-nst-*.and.nop1}
+        $p'=\ceu{\Nop}$.
+        %%
+        Then, by~\R{and-nop1} and by the induction hypothesis, there
+        is a~$\delta'$ such that
+        \[
+          \<\ceu{{\Nop}\AtAnd{p''}},n,\nil>
+          \nst[1]\<p'',n,\nil>
+          \nst[*]\delta'_\Hnst\,.
+        \]
+      \end{subsubcase}
+      \begin{subsubcase}
+        \label{thm.term-nst-*.and.brk1}
+        $p'=\ceu{\Break}$.
+        %%
+        Then, by~\R{and-brk1},
+        \begin{equation}
+          \label{thm.term-nst-*.and.brk1.eq1}
+          \<\ceu{{\Break}\AtAnd{p''}},n,\nil>
+          \nst[1]\<\ceu{\clear(p'');\,\Break},n,\nil>\,.
+        \end{equation}
+        From~\eqref{thm.term-nst-*.and.brk1.eq1}, by
+        Assumption~\ref{ass.term-nst-fin} and~\R{seq-nop},
+        \begin{align*}
+          \<\ceu{\clear(p'');\,\Break},n,\nil>
+          &\nst[*]\<\ceu{\Nop;\,\Break},n,\nil>\\
+          &\nst[1]\<\ceu{\Break},n,\nil>_\Hnst\,.
+        \end{align*}
+      \end{subsubcase}
+      \begin{subsubcase}
+        $p'\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Then, by the induction hypothesis, there are~$p_1'$ and~$e$ such
+        that
+        \[
+          \<p',n,\nil>\nst[*]\<p_1',n,e>_\Hnst\,.
+        \]
+        By item~\eqref{lem.props-nst-n.c} of Lemma~\ref{lem.props-nst-n},
+        \[
+          \<\ceu{{p'}\AtAnd{p''}},n,\nil>
+          \nst[*]\<\ceu{{p_1'}\AtAnd{p''}},n,e>\,.
+        \]
+        It remains to be show that~$\<\ceu{{p_1'}\AtAnd{p''}},n,e>$ leads to
+        an nested-irreducible description.  There are four possibilities
+        following from the fact that the simpler~$\<p_1',n,e>$ is
+        nested-irreducible.
+        \begin{enumerate}
+        \item If~$e\ne\nil$ then~$\<\ceu{{p_1'}\AtAnd{p''}},n,e>_\Hnst$.
+        \item If~$p_1'=\ceu{\Nop}$ this case becomes
+          Case~\ref{thm.term-nst-*.and.nop1}.
+        \item If~$p_1'=\ceu{\Break}$ this case becomes
+          Case~\ref{thm.term-nst-*.and.brk1}.
+        \item Otherwise, $\isblocked(p_1',n)$, and as~$\isblocked(p'',n)$ is
+          false, this case becomes\dots\ TODO.
+        \end{enumerate}
+      \end{subsubcase}
+    \end{subcase}
+    \begin{subcase}
+      $\isblocked(p',n)$ is false and~$\isblocked(p'',n)$ is true.
+      %%
+      TODO.
+    \end{subcase}
+    \begin{subcase}
+      $\isblocked(p',n)$ is true and~$\isblocked(p'',n)$ is false.
+      %%
+      TODO.
+    \end{subcase}
+    \begin{subcase}
+      $\isblocked(p',n)$ is true and~$\isblocked(p'',n)$ is true.
+      %%
+      TODO.
+    \end{subcase}
   \end{case}
@@
     $p=\ceu{{p'}\Or{p''}}$.
+    %%
+    $p=\ceu{{p'}\Or{p''}}$.  Then, by~\R{or-expd},
+    \[
+      \<\ceu{{p'}\Or{p''}},n,\nil>
+      \nst[1]\<\ceu{{p'}\AtOr{(\CanRun(n);\,p'')}},n,\nil>\,.
+    \]
+    From this point on, this case is similar to
+    Case~\ref{thm.term-nst-*.or}.
   \end{case}
 
-  \begin{case}
+  \begin{case}\label{thm.term-nst-*.or}
     $p=\ceu{{p'}\AtOr{p''}}$.
+    %%
+    TODO.
   \end{case}

### 2018-02-19 06-appendix.tex
Conserta erro de digitação no apêndice

@@
 The syntactic restrictions discussed in Section~\ref{?}, regarding the body
-of~$\ceu{\Fin}$ and~$\ceu{\Loop}$ expressions, is formalized by the
+of~$\ceu{\Fin}$ and~$\ceu{\Loop}$ expressions, are formalized by the
 following assumptions.

### 2018-02-19 01-intro.tex
Pequenos ajustes no fim da Introdução, Figura ilegível, e Título.

@@
 
-In this work, we present a formal semantics for a subset of \CEU that focus on
-its peculiarities in comparison to other synchronous languages.
-\begin{itemize}
-    \item qual a abordagem / operational semantics / dois passos
-    \item quais os resultados / provas
-    \item quais os desafios e limitações
-    \item \gl{TODO}
-\end{itemize}
+In this work, we present a formal small-step structural operational semantics
+for \CEU and proofs that it enforces memory-bounded, deterministic, and
+terminating reactions to the environment, i.e., that for a given arbitrary
+timeline of input events, multiple executions of the same program always react
+in bounded time and arrive at the same final finite memory state.
 
-\fs{Descrever seções.}
+% TODO: \fs{Descrever seções.}

### 2018-02-19 02-ceu.tex
Revisão da semântica.
Pequenos ajustes no fim da Introdução, Figura ilegível, e Título.

@@
 \centering
-\includegraphics[width=\columnwidth]{tick}
+\includegraphics[width=\columnwidth]{tick_min}
 \caption{The discrete notion of time in \CEU.}
@@
 %
-Event~\code{D} occurs during an idle period and can start immediately at
-\code{D-4}.
+Finally, event~\code{D} occurs during an idle period and can start immediately
+at \code{D-4}.
 %
-Finally, two instances of event~\code{E} occur during~\code{D-4}; they are
-handled in the subsequent reactions~\code{E-5} and~\code{E-6}.
+%Finally, two instances of event~\code{E} occur during~\code{D-4}; they are
+%handled in the subsequent reactions~\code{E-5} and~\code{E-6}.
 
@@
 \end{lstlisting}
-\vskip-\baselineskip
+%\vskip-\baselineskip
 

### 2018-02-19 index.tex
Adição do apêndice.
Adiciona ao apêndice prova (incompl.) da terminação NST
Pequenos ajustes no fim da Introdução, Figura ilegível, e Título.

@@
-\documentclass[sigplan, anonymous, review]{acmart}
+\documentclass[sigplan,anonymous,review,protrusion=true,expansion]{acmart}
 
@@
 
-\title[A Memory-Bounded, Deterministic, and Terminating Semantics for \CEU]
-{A Deterministic and Terminating Semantics
-  for the Synchronous Programming Language~\CEU}
+\title{A Memory-Bounded, Deterministic and Terminating Semantics for the
+       Synchronous Programming Language~\CEU}
 
@@
 
-\appendix
-%\include{proofs}
+\input{06-appendix}
 

### 2018-02-20 03-semantics.tex
Corrige a definição de and,or-expd na Seção 3.2
Completa a prova do teorema da terminação de NST

@@
 In this section, we introduce a reduced syntax for \CEU and propose an 
-operational semantics to formally describe the language.
+operational semantics to formally describe the behavior of its programs.
 We describe a small synchronous kernel highlighting the peculiarities of \CEU, 
@@
 We distinguish between two types of transition rules:
-    \emph{outermost} transitions $\out$ and
-    \emph{nested} transitions $\nst$\,.
+    \emph{outermost transitions} $\out$ and
+    \emph{nested transitions} $\nst$\,.
 
@@
 \begin{gather*}
-  \hskip-.5em
+  \hskip-.9em
   \<\ceu{p\And{q}},n,\nil>
-  \nst\<\ceu{p\And(\CanRun(n);q)},n,\nil>
+  \nst\<\ceu{p\AtAnd(\CanRun(n);q)},n,\nil>
   \Rtag{and-expd}\\[2\jot]
@@
   \<\ceu{p\Or{q}},n,\nil>
-  \nst\<\ceu{p\Or(\CanRun(n);q)},n,\nil>
+  \nst\<\ceu{p\AtOr(\CanRun(n);q)},n,\nil>
   \Rtag{or-expd}\\[2\jot]
@@
 
-Rules \R{and-expd} and \R{or-expd} insert a~$\ceu{\CanRun(n)}$ at the
+Rules~\R{and-expd} and~\R{or-expd} insert a~$\ceu{\CanRun(n)}$ at the
 beginning of the right branch.

### 2018-02-20 06-appendix.tex
Completa a prova do teorema da terminação de NST

@@
   The theorem is trivially true for~$n=0$ and follows directly from
-  Lemma~\ref{lem.det-out} for~$n=1$.  (Note that for~$n>1$ the theorem is
-  vacuously true for~$\outpush$ transitions---by the format of
-  rules~\R{push} and~\R{pop}, transitions~$\outpush$ cannot be applied more
-  than once in a row and cannot occur after a~$\outpop$ transition.)
+  Lemma~\ref{lem.det-out} for~$n=1$.  (The theorem is vacuously true
+  for~$\outpush$ transitions if~$n>1$ since, by definition,
+  $\outpush$~transitions cannot be applied more than once in a row and
+  cannot occur after a~$\outpop$ transition.)
   %%
@@
 
-  By the definition of transition~$\nst$, we have that~$e=\nil$ and
-  $n_1=n_2=n$.  It remains to be shown that~$p_1=p_2$ and~$e_1=e_2$.
+  By definition of~$\nst$, we have that~$e=\nil$ and $n_1=n_2=n$.  It
+  remains to be shown that~$p_1=p_2$ and~$e_1=e_2$.
 
@@
     \end{equation}
-    Since~$\delta'=\<p',n,e'>$ is nested-irreducible,
-    $p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p',n)$.  In either case, by
-    the definition of~$\nst$, there is no~$\delta_1'$ such
+    Since~$\delta'=\<p',n,e'>$ is nested-irreducible, $e'=\nil$
+    or~$p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p',n)$.  In any of these
+    cases, by the definition of~$\nst$, there is no~$\delta_1'$ such
     that~$\delta'\nst[1]\delta_1'$, which
@@
   \begin{case}
-    Suppose there are~$0\le{i}<n$ and~$\delta''_\Hnst$ such
-    that~$\delta\nst[i]\delta''$.  Then since~$n>i$, by
-    Case~\ref{lem.irr-nst-n-case1}, $\delta'$~could not exist, which is
-    absurd.  Therefore, the assumption that there is such~$i$ is
-    false.\qedhere
+    Suppose there are~$i<n$ and~$\delta''_\Hnst$ such
+    that~$\delta\nst[i]\delta''$.  Then, since~$n>i$, by
+    Case~\ref{lem.irr-nst-n-case1}, $\delta'$~cannot exist, which is absurd.
+    Therefore, the assumption that there is such~$i$ is false.\qedhere
   \end{case}
@@
     $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[n]\<\ceu{{p_1}'\AtOr{p_2}},n,e'>$.
+  \end{enumerate}
+  If~$\<p_2,n,e>\nst[n]\<p_2',n,e'>$, for any~$p_1$ such
+  that~$\isblocked(p_1,n)$:
+  \begin{enumerate}[(a)]
+    \setcounter{enumi}{4}
+  \item\label{lem.props-nst-n.e}
+    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[n]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>$;
+    %%
+  \item\label{lem.props-nst-n.f}
+    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[n]\<\ceu{{p_1}\AtOr{p_2'}},n,e'>$.
   \end{enumerate}
@@
     %%
+
+  \item The lemma is trivially true for~$n=0$, as~$p_2=p_2'$, and follows
+    directly from~\R{and-adv2} for~$n=1$.  Suppose
+    \begin{equation}
+      \label{lem.props-nst-n.e.eq1}
+      \<p_2,n,e>\nst[1]\<p_2'',n,e''>\nst[n-1]\<p_2',n,e'>\,,
+    \end{equation}
+    for some~$n>1$.  Then~$\<p_2'',n,e''>$ is not nested-irreducible.
+    By~\eqref{lem.props-nst-n.e.eq1} and by~\R{and-adv2},
+    \begin{equation}
+      \label{lem.props-nst-n.e.eq2}
+      \<\ceu{{p_1}\AtAnd{p_2}},n,e>
+      \nst[1]\<\ceu{{p_1}\AtAnd{p_2''}},n,e''>\,.
+    \end{equation}
+    From~\eqref{lem.props-nst-n.e.eq1}, by the induction hypothesis,
+    \begin{equation}
+      \label{lem.props-nst-n.e.eq3}
+      \<\ceu{{p_1}\AtAnd{p_2''}},n,e''>
+      \nst[n-1]\<\ceu{{p_1}\AtOr{p_2'}},n,e'>\,.
+    \end{equation}
+    From~\eqref{lem.props-nst-n.e.eq2} and~\eqref{lem.props-nst-n.e.eq3},
+    \[
+      \<\ceu{{p_1}\AtAnd{p_2}},n,e>
+      \nst[n]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>\,.
+    \]
+
+  \item Similar to Case~\eqref{lem.props-nst-n.e}.\qedhere
   \end{enumerate}
@@
       %%
-      Then, by~\R{seq-brk} and by the induction hypothesis, there is
-      a~$\delta'$ such that
+      Then, by~\R{seq-brk},
       \[
-        \<\ceu{\Break;\,p''},n,\nil>
-        \nst[1]\<p'',n,e>
-        \nst[*]\delta'_\Hnst\,.
+        \<\ceu{\Break;\,p''},n,\nil>\nst[1]\<\ceu{\Break},n,\nil>_\Hnst\,.
       \]
@@
       \end{equation}
-      It remains to be shown that~$\<\ceu{p_1';\,p''},n,e>$ is indeed
+      It remains to be shown that~$\<\ceu{p_1';\,p''},n,e>$ is
       nested-irreducible.  There are four possibilities following from the
@@
   \begin{case}
+    \label{thm.term-nst-*.loop}
     $p=\ceu{\Loop{p'}}$.
@@
       %%
-      Then, by~\R{loop-nop} and by Assumption~\ref{ass.term-nst-loop},
-      \begin{align}
-        \label{thm.term-nst-*.loop-adv.eq1}
+      Then, by~\R{loop-nop},
+      \[
         \<\ceu{\Nop\AtLoop{p''}},n,\nil>
-        &\nst[1]\<\ceu{\Loop{p''}},n,\nil>\notag\\
-        &\nst[*]\<p_1'',n,e>\,,
-      \end{align}
-      for some~$e$ and~$p_1''$ such that
-      either~$p_1''=\ceu{\Break\AtLoop{p''}}$ or~$\isblockedext(p_1'',n)$.
-      \begin{subsubcase}
-        $p_1''=\ceu{\Break\AtLoop{p''}}$.
-        From~\eqref{thm.term-nst-*.loop-adv.eq1}, by~\R{loop-brk},
-        \begin{align*}
-        \<\ceu{\Nop\AtLoop{p''}},n,\nil>
-        &\nst[1]\<\ceu{\Loop{p''}},n,\nil>\\
-        &\nst[*]\<\ceu{\Break\AtLoop{p''}},n,e>\\
-        &\nst[1]\<\ceu{\Nop},n,e>_\Hnst\,.
-        \end{align*}
-      \end{subsubcase}
-      \begin{subsubcase}
-        $\isblockedext(p_1'',n)$~is true.  Then, by definition,
-        $\isblockedext(p_1'',n)$ implies~$\isblocked(p_1'',n)$.  Hence,
-        from~\eqref{thm.term-nst-*.loop-adv.eq1} and by the definition
-        of~$\Hnst$, $\<p_1'',n,e>_\Hnst$.
-      \end{subsubcase}
+        \nst[1]\<\ceu{\Loop{p''}},n,\nil>\,.
+      \]
+      From this point on, this case is similar to
+      Case~\ref{thm.term-nst-*.loop}.
     \end{subcase}
@@
       \]
-      It remains to be show that~$\<\ceu{p_1'\AtLoop{p''}},n,e>$ is indeed
+      It remains to be show that~$\<\ceu{p_1'\AtLoop{p''}},n,e>$ is
       nested-irreducible.  The rest of this proof is similar to that of
@@
     %%
-    There are four subcases.
+    There are two subcases.
     \begin{subcase}
-      $\isblocked(p',n)$ is false and~$\isblocked(p'',n)$ is false.
+      $\isblocked(p',n)$ is false.
       %%
@@
       \begin{subsubcase}
+        \label{thm.term-nst-*.and.adv1}
         $p'\ne\ceu{\Nop},\ceu{\Break}$.
@@
         \begin{enumerate}
-        \item If~$e\ne\nil$ then~$\<\ceu{{p_1'}\AtAnd{p''}},n,e>_\Hnst$.
-        \item If~$p_1'=\ceu{\Nop}$ this case becomes
+        \item If~$e\ne\nil$ then, by
+          definition,~$\<\ceu{{p_1'}\AtAnd{p''}},n,e>_\Hnst$.
+        \item If~$p_1'=\ceu{\Nop}$, this case is similar to
           Case~\ref{thm.term-nst-*.and.nop1}.
-        \item If~$p_1'=\ceu{\Break}$ this case becomes
+        \item If~$p_1'=\ceu{\Break}$, this case is similar to
           Case~\ref{thm.term-nst-*.and.brk1}.
-        \item Otherwise, $\isblocked(p_1',n)$, and as~$\isblocked(p'',n)$ is
-          false, this case becomes\dots\ TODO.
+        \item If~$\isblocked(p_1',n)$, this case is similar to
+          Case~\ref{thm.term-nst-*.and2}.
         \end{enumerate}
@@
     \begin{subcase}
-      $\isblocked(p',n)$ is false and~$\isblocked(p'',n)$ is true.
+      \label{thm.term-nst-*.and2}
+      $\isblocked(p',n)$ is true.
       %%
-      TODO.
-    \end{subcase}
-    \begin{subcase}
-      $\isblocked(p',n)$ is true and~$\isblocked(p'',n)$ is false.
+      There are three subcases.
+      \begin{subsubcase}
+        \label{thm.term-nst-*.and.nop2}
+        $p''=\ceu{\Nop}$.
         %%
-      TODO.
-    \end{subcase}
-    \begin{subcase}
-      $\isblocked(p',n)$ is true and~$\isblocked(p'',n)$ is true.
+        Then, by~\R{and-nop2},
+        \[
+          \<\ceu{{p'}\AtAnd{\Nop}},n,\nil>\nst[1]\<p',n,\nil>_\Hnst\,.
+        \]
+      \end{subsubcase}
+      \begin{subsubcase}
+        \label{thm.term-nst-*.and.brk2}
+        $p''=\ceu{\Break}$.
         %%
-      TODO.
+        Then, by~\R{and-brk2},
+        \[
+          \<\ceu{{p'}\AtAnd{\Break}},n,\nil>
+          \nst[1]\<\ceu{\clear(p');\,\Break},n,\nil>\,.
+        \]
+        From this point on, this case is similar to
+        Case~\ref{thm.term-nst-*.and.brk1}.
+      \end{subsubcase}
+      \begin{subsubcase}
+        \label{thm.term-nst-*.and.adv2}
+        $p''\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Then, by the induction hypothesis, there are~$p_1''$ and~$e$ such
+        that
+        \[
+          \<p'',n,\nil>\nst[*]\<p_1'',n,e>_\Hnst\,.
+        \]
+        By item~\eqref{lem.props-nst-n.e} of Lemma~\ref{lem.props-nst-n},
+        \[
+          \<\ceu{{p'}\AtAnd{p''}},n,\nil>
+          \nst[*]\<\ceu{{p'}\AtAnd{p_1''}},n,e>\,.
+        \]
+        It remains to be show that~$\<\ceu{{p'}\AtAnd{p_1''}},n,e>$ leads to
+        an nested-irreducible description.  There are four possibilities
+        following from the fact that the simpler~$\<p_1'',n,e>$ is
+        nested-irreducible.
+        \begin{enumerate}
+        \item If~$e\ne\nil$ then, by definition,
+          $\<\ceu{{p'}\AtAnd{p_1''}},n,e>_\Hnst$.
+        \item If~$p_1''=\ceu{\Nop}$, this case is similar to
+          Case~\ref{thm.term-nst-*.and.nop2}.
+        \item If~$p_1''=\ceu{\Break}$, this case is similar to
+          Case~\ref{thm.term-nst-*.and.brk2}.
+        \item If~$\isblocked(p_1'',n)$ then, as both sides are blocked, by
+          definition, $\<\ceu{{p'}\AtAnd{p_1''}},n,e>_\Hnst$.
+        \end{enumerate}
+      \end{subsubcase}
     \end{subcase}
@@
     %%
-    $p=\ceu{{p'}\Or{p''}}$.  Then, by~\R{or-expd},
+    Then, by~\R{or-expd},
     \[
@@
 
-  \begin{case}\label{thm.term-nst-*.or}
+  \begin{case}
+    \label{thm.term-nst-*.or}
     $p=\ceu{{p'}\AtOr{p''}}$.
     %%
-    TODO.
+    There are two subcases.
+    \begin{subcase}
+      $\isblocked(p',n)$~is false.
+      %%
+      There are three subcases.
+      \begin{subsubcase}
+        \label{thm.term-nst-*.or.nop1}
+        $p'=\ceu{\Nop}$.  Then, by~\R{or-nop1},
+        \begin{equation}
+          \label{thm.term-nst-*.or.nop1.eq1}
+          \<\ceu{{\Nop}\AtOr{p''}},n,\nil>
+          \nst[1]\<\ceu{\clear(p'')},n,\nil>\,.
+        \end{equation}
+        From~\eqref{thm.term-nst-*.or.nop1.eq1}, by
+        Assumption~\ref{ass.term-nst-fin},
+        \[
+          \<\ceu{\clear(p'')},n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>_\Hnst\,.
+        \]
+      \end{subsubcase}
+      \begin{subsubcase}
+        \label{thm.term-nst-*.or.brk1}
+        $p'=\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{thm.term-nst-*.and.brk1}.
+      \end{subsubcase}
+      \begin{subsubcase}
+        $p'\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{thm.term-nst-*.and.adv1}.
+      \end{subsubcase}
+    \end{subcase}
+    \begin{subcase}
+      \label{thm.term-nst-*.or.adv1}
+      $\isblocked(p',n)$~is true.
+      %%
+      There are three subcases.
+      \begin{subsubcase}
+        $p''=\ceu{\Nop}$.
+        %%
+        Then, by~\R{or-nop2},
+        \begin{equation}
+          \label{thm.term-nst-*.or.nop2.eq1}
+          \<\ceu{p'\AtOr{\Nop}},n,\nil>
+          \nst[1]\<\clear(p'),n,\nil>\,.
+        \end{equation}
+        From~\eqref{thm.term-nst-*.or.nop2.eq1}, by
+        Assumption~\ref{ass.term-nst-fin},
+        \[
+          \<\ceu{\clear(p')},n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>_\Hnst\,.
+        \]
+      \end{subsubcase}
+      \begin{subsubcase}
+        $p''=\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{thm.term-nst-*.and.brk2}.
+      \end{subsubcase}
+      \begin{subsubcase}
+        $p''\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{thm.term-nst-*.and.adv2}.\qedhere
+      \end{subsubcase}
+    \end{subcase}
   \end{case}

### 2018-02-20 06-appendix.tex
Adiciona descrição do problema com bcast ao apêndice

@@
 
+\clearpage
+\section*{PROBLEMA}
+A reação abaixo não está definida pois~$\bcast$ não está definido
+para~$\ceu{\And}$:
+\[
+  \<\ceu{{\AwaitExt(e)}\And{\Mem(\Id);\AwaitExt(e)}},0,e>
+  \outpush\bot
+\]
+
+Mesmo que~$\bcast$ estivesse definido, como é o caso do~$\ceu{\AtAnd}$, não
+funcionaria:
+\[
+  \begin{split}
+    &\<\ceu{{\AwaitExt(e)}\AtAnd{\Mem(\Id);\AwaitExt(e)}},0,e>\\
+    &\quad\outpush\<\ceu{{\Nop(e)}\AtAnd{\Mem(\Id);\AwaitExt(e)}},1,\nil>.
+  \end{split}
+\]
+Repare que só o lado esquerdo consumiu o evento.
+
+Mesmo se assumirmos que inicialmente todas as trilhas do programa vão estar
+esperando também não funcionaria.  Por exemplo,
+\begin{align*}
+  &\<\ceu{\AwaitExt{E};({\EmitInt(e)}\And{\AwaitInt(e)}}),0,E>\\
+  &\outpush\<\ceu{\Nop;({\EmitInt(e)}\And{\AwaitInt(e)}}),1,\nil>\\
+  &\nst\<\ceu{{\EmitInt(e)}\And{\AwaitInt(e)}},1,\nil>\\
+  &\nst\<\ceu{{\EmitInt(e)}\AtAnd(\CanRun(1);\AwaitInt(e))},1,\nil>\\
+  &\nst\<\ceu{{\CanRun(1)}\AtAnd(\CanRun(1);\AwaitInt(e))},1,e>\\
+  &\outpush\<\ceu{{\CanRun(1)}\AtAnd(\CanRun(1);\AwaitInt(e))},2,\nil>\\
+  &\outpop\<\ceu{{\CanRun(1)}\AtAnd(\CanRun(1);\AwaitInt(e))},1,\nil>\\
+  &\nst\<\ceu{\Nop\AtAnd(\CanRun(1);\AwaitInt(e))},1,\nil>\\
+  &\nst\<\ceu{\CanRun(1);\AwaitInt(e)},0,\nil>\\
+  &\nst\<\ceu{\AwaitInt(e)},1,\nil>\\
+  &\outpop\<\ceu{\AwaitInt(e)},0,\nil>
+\end{align*}
+(Este~$\ceu{\AwaitInt(e)}$ que sobrou deveria ter sido acordado
+pelo~$\ceu{\EmitInt(e)}$ em paralelo, certo?)
+
+O problema é que a função~$\bcast$ só acorda as trilhas que estão esperando
+no momento do \R{push}, sem desconsiderar as instruções instantâneas que
+ocorrem antes da espera propriamente dita.  Não tem como resolver isso
+estaticamente.  Acho que a solução é voltar com a pilha de eventos emitidos.
+
 % LocalWords:  subcase subcases

### 2018-02-20 03-semantics.tex
Pequenas correções e alguns comentários

@@
 
-We assume that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, and~$\ceu{\AwaitInt}$ and~$\ceu{\EmitInt}$ expressions do not
+We assume that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$ and~$\ceu{\EmitInt}$ expressions do not
 share identifiers: any identifier is either a variable, an external event,
@@
 \]
-where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are
+where~
+%
+\rc{Na BNF o $p$ não deveria ser $\P$~(maiúsculo) tbm?}
+%
+$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are
 nonnegative integers representing the current stack level,
-and~$e,e'\in\E\cup\{\nil\}$ are the events emitted before and after the
-transition (both possibly the empty event~$\nil$).
+and~
+%
+\rc{Checar: Acho que o conjunto $\E$~não foi previamente definido em
+  lugar nenhum}
+%
+$e,e'\in\E\cup\{\nil\}$
+are the events emitted before and after the
+transition (both being possibly the empty event~$\nil$).
 
@@
         Figure~\ref{fig.bcast}),
-    (b)~creating a nested reaction by increasing the stack level, and, at the same time, and
+    (b)~creating a nested reaction by increasing the stack level, and, at the same time,
     (c)~consuming the event ($e$ becomes~$\nil$).
@@
 eventually resume in the current reaction (see rule \R{pop}).
+%
+\rc{Não entendi essa discussão}
 
@@
 
-Rule \textbf{pop} either has $p=@nop$ or $isblocked(n,p)$ as preconditions.
+Rule \textbf{pop} either has $p=@nop$ or $isblocked(p,n)$ as preconditions.
 %
@@
 For the second case, we need to show that given $\LL p,n,\epsilon \RR$, no
-$\NST$ transitions apply with $isblocked(n,p)$ and vice versa.
+$\NST$ transitions apply with $isblocked(p,n)$ and vice versa.
 Except for $@canrun$, there are no $\NST$ transitions for the other blocking
@@
 However, considering the precondition $\LL p,n,\epsilon \RR$,
-$isblocked(n,@canrun(n))$ is false.
+$isblocked(@canrun(n),n)$ is false.
 Hence, given the preconditions for rule \textbf{pop}, no $\NST$ transitions can
 occur.
-Conversely, if a $\NST$ transition is possible, then $isblocked(n,p)$ must be
+Conversely, if a $\NST$ transition is possible, then $isblocked(p,n)$ must be
 false.
 Again, except for $@canrun$, all other transitions do not involve blocking
-expressions, hence, for these transitions, $isblocked(n,p)$ must be false.
-For rule \textbf{canrun}, a transition can only occur if the current stack
+expressions, hence, for these transitions, $isblocked(p,n)$ must be false.
+For rule \R{can-run}, a transition can only occur if the current stack
 level matches $@canrun(n)$.
-In this case, $isblocked(n,@canrun(n))$ is false.
+In this case, $isblocked(@canrun(n),n)$ is false.
 
@@
 cannot be blocked and vice-versa, i.e., that
-$\LL p,n,\epsilon \RR \NST \LL p',n,e \RR$ and $isblocked(n,p)$ are mutually
+$\LL p,n,\epsilon \RR \NST \LL p',n,e \RR$ and $isblocked(p,n)$ are mutually
 exclusive, which is exactly the same reasoning for rule \textbf{pop} above.

### 2018-02-21 06-appendix.tex
Remove a seção PROBLEMA do apêndice
Adiciona definição da função \pot ao apêndice

--- 2018-02-20 06-appendix.tex
@@
   \begin{case}
-    $p=\ceu{\Ifelse{\Mem(\Id)}{p'}{p''}}$.  There are two subcases.
+    $p=\ceu{\IfElse{\Mem(\Id)}{p'}{p''}}$.  There are two subcases.
     \begin{subcase}
@@
   \begin{case}
-    $p=\ceu{\Ifelse{\Mem(\Id)}{p'}{p''}}$.
+    $p=\ceu{\IfElse{\Mem(\Id)}{p'}{p''}}$.
     %%
@@
       \begin{align*}
-        \<\ceu{\Ifelse{\Mem(\Id)}{p'}{p''}},n,\nil>
+        \<\ceu{\IfElse{\Mem(\Id)}{p'}{p''}},n,\nil>
         &\nst[1]\<p',n,e>\\
@@
       \begin{align*}
-        \<\ceu{\Ifelse{\Mem(\Id)}{p'}{p''}},n,\nil>
+        \<\ceu{\IfElse{\Mem(\Id)}{p'}{p''}},n,\nil>
         &\nst[1]\<p'',n,e>\\
@@
 
-% LocalWords:  subcase subcases
+\begin{definition}
+  The \emph{potency} of a program~$p$, denoted~$\pot(p)$, is defined as
+  follows:
+  \begin{alignat*}{2}
+    \pot(p)&=1
+    &\quad &\text{if~$p=\ceu{\EmitInt(e)}$}\\
+    %%
+    \pot(p)&=\pot(p_1)+\pot(p_2)
+    &\quad &
+    \begin{aligned}[t]
+      &\text{if~$p=p_1;\,p_2$ or~$p=\ceu{p_1\AtLoop{p_2}}$}\\
+      &\quad\text{and~$p_1\ne{\ceu{\AwaitExt{e}}}$}
+    \end{aligned}\\
+    %%
+    \pot(p)&=\max\{\pot(p_1),\pot(p_2)\}
+    &\quad &
+    \begin{aligned}[t]
+      &\text{if~$p=\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}}$}\\
+      &\quad\text{or~$p=\ceu{{p_1}\And{p_2}}$
+        or~$p=\ceu{{p_1}\Or{p_2}}$}\\
+      &\quad\text{or~$p=\ceu{{p_1}\AtAnd{p_2}}$
+        or~$p=\ceu{{p_1}\AtOr{p_2}}$}
+    \end{aligned}\\
+    %%
+    \pot(p)&=0
+    &\quad &\text{otherwise}\,.
+  \end{alignat*}
+  % \item$\pot(p)=\pot(p_1)$,\\
+  %   \strut\quad if~$p=\ceu{\Loop{p_1}}$ or~$p=\ceu{\Every{e}\,{p_1}}$;
+  %   %%
+  % \item$\pot(p)=\pot(p_1)+\pot(p_2)$,\\
+  %   \strut\quad if~$p=p_1;\,p_2$ or~$p=\ceu{p_1\AtLoop{p_2}}$
+  %   and~$p_1\ne{\ceu{\AwaitExt{e}}}$;
+  %   %%
+  % \item$\pot(p)=\max\{\pot(p_1),\pot(p_2)\}$,\\
+  %   \strut\quad if~$p=\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}}$\\
+  %   \strut\qquad or~$p=\ceu{{p_1}\And{p_2}}$ or~$p=\ceu{{p_1}\Or{p_2}}$\\
+  %   \strut\qquad or~$p=\ceu{{p_1}\AtAnd{p_2}}$ or~$p=\ceu{{p_1}\AtOr{p_2}}$;
+  %   %%
+  % \item$\pot(p)=0$,\enspace otherwise.
+  % \end{enumerate}
+  % \begin{align*}
+  %   &\pot(p)=\\
+  %   &\enspace\begin{cases}
+  %     1
+  %     &\text{if~$p=\ceu{\AwaitExt(e)}$}\\[2\jot]
+  %     %%
+  %     \pot(p_1)
+  %     &\text{if~$p=\ceu{\Loop{p1}}$ or~$p=\ceu{\Every{e}{p_1}}$}\\[2\jot]
+  %     %%
+  %     \pot(p_1)+\pot(p_2)
+  %     &\parbox[t]{.6\hsize}{%
+  %       if~$p=p_1;\,p_2$ or~$p=\ceu{p_1\AtLoop{p_2}}$\\
+  %       \strut\quad and~$p_1\ne{\ceu{\EmitInt{e}}}$}\strut\\[6\jot]
+  %     %%
+  %     \max\{\pot(p_1),\pot(p_2)\}
+  %     &\parbox[t]{.6\hsize}{%
+  %       if~$p=\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}}$\\
+  %       \strut\quad or~$p=\ceu{{p_1}\And{p_2}}$\\
+  %       \strut\quad or~$p=\ceu{{p_1}\AtAnd{p_2}}$\\
+  %       \strut\quad or~$p=\ceu{{p_1}\Or{p_2}}$\\
+  %       \strut\quad or~$p=\ceu{{p_1}\AtOr{p_2}}$}
+  %   \end{cases}
+  % \end{align*}
+\end{definition}

### 2018-02-21 04-related.tex
Pequenas correções e alguns comentários
Trabalhos Relacionados

@@
 
-\CEU was strongly influenced by Esterel but they differ in the most fundamental 
-aspect of the notion of time~\cite{ceu.sensys13}.
-%
-Esterel defines time as a discrete sequence of logical unit instants or 
-``ticks''.
-At each tick, the program reacts to an arbitrary number of simultaneous input 
-events from the environment.
+\CEU follows the lineage of imperative synchronous languages initiated by
+Esterel~\cite{esterel.ieee91}.
+These languages typically define time as a discrete sequence of logical
+``ticks'' in which multiple simultaneous input events can be
+active~\cite{ceu.tecs17}.
 %
@@
 reject programs with \emph{causality cycles} and \emph{schizophrenia problems}%
-~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
+~\cite{esterel.constructive}.
 %
 In contrast, \CEU defines time as a discrete sequence of reactions to
-unique input events.
+unique input events, which is a prerequisite for the concurrency checks that
+enable safe shared-memory concurrency.
+
+Also, in most synchronous languages, the behavior of internal and external
+events is equivalent.
+In \CEU, internal events introduce stack-based micro reactions within external
+reactions, providing more fine-grained control for intra-reaction execution.
 %
-In the formal semantics, ... \fs{como isso aparece na semantica} ...
-\CEU also rejects some syntactically correct programs to avoid infinite
-execution, but with simple restrictions in the abstract syntax tree.
+This allows for memory-bounded subroutines that execute multiple times in the
+same external reaction.
+%
+The synchronous languages Statecharts~\cite{statecharts.variants} and
+Statemate~\cite{statecharts.statemate} also distinguish internal from external
+events.
+In the former, \emph{``reactions to external and internal events (...) can be
+sensed only after completion of the step''}.
+In the latter, \emph{``the receiving state (of the internal event) acts here as
+a function''}.
+Although the descriptions suggest a stack-based semantics, we are not aware of
+formalizations for these ideas for a deeper comparison with \CEU.
 
-Another distinction is that, in Esterel, the behavior of internal and external
-events is equivalent, while in \CEU internal events introduce stack-based micro
-reactions within external reactions, providing more fine-grained control for
-intra-reaction execution.
+Like \CEU, many synchronous languages%
+~\cite{rp.rc,wsn.protothreads,wsn.sol,rp.synchc,rp.pretc}
+also rely on lexical scheduling to preserve determinism.
 %
-In the formal semantics, ... \fs{como isso aparece na semantica} ...
-%
-Some variants of the Statecharts synchronous visual language also distinguish
-internal from external events~\cite{statecharts.variants}.
-In Statemate~\cite{statecharts.statemate}, \emph{``reactions to external and
-internal events (...) can be sensed only after completion of the step''},
-implying queue-based execution.
-In Stateflow~\cite{statecharts.stateflow}, \emph{``the receiving state (of the
-event) acts here as a function''}, which is similar to \CEU's
-stack-based execution.
-We are not aware of formalizations for these ideas for a deeper comparison with
-\CEU.
-
-Like other synchronous languages
-(%
-    \emph{Reactive C}~\cite{rp.rc},                 % 1991
-    \emph{Protothreads}~\cite{wsn.protothreads},    % 2006
-    \emph{SOL}~\cite{wsn.sol},                      % 2007
-    \emph{SC}~\cite{rp.synchc},                     % 2009
-    and
-    \emph{PRET-C}~\cite{rp.pretc}%                  % 2010
-),
-\CEU relies on deterministic scheduling to preserve intra-reaction determinism.
-%
-In addition, it also performs concurrency checks to detect trails that, when
-reordered, change the observable behavior of the program, i.e., trails that
-actually rely on deterministic scheduling~\cite{ceu.sensys13}.
-%
-Esterel is only deterministic with respect to external behavior: \emph{``the
-same sequence of inputs always produces the same sequence of 
-outputs''}~\cite{esterel.primer}.
-%
-However, the execution order for operations within a reaction is 
-non-deterministic: \emph{``if there is no control dependency, as in
+In contrast, in Esterel, the execution order for operations within a reaction
+is non-deterministic: \emph{``if there is no control dependency, as in
 \code{(call~f1()~||~call~f2())},
@@
 
-Esterel describes a finalization mechanism in a standardization
-proposal~\cite{esterel.v7} that is similar to \CEU's.
-However, we are not aware of an open implementation or a formal semantics for a
-deeper comparison.
-
-\fs{outras linguagens sincronas}
-\fs{outras linguagens deterministicas}
-\fs{outras linguagens com terminacao}
+Regarding the integration with the $C$ language-based environment, \CEU
+supports a finalization mechanism for external resources.
+%
+In addition, \CEU also tracks pointers representing resources that cross $C$
+boundaries and forces the programmer to provide associated finalizers.

### 2018-02-21 06-appendix.tex
Correção Rodrigo e remoção do "Problema"

--- 2018-02-20 06-appendix.tex
@@
   \begin{case}
-    $p=\ceu{\Ifelse{\Mem(\Id)}{p'}{p''}}$.  There are two subcases.
+    $p=\ceu{\IfElse{\Mem(\Id)}{p'}{p''}}$.
+    %%
+    There are two subcases.
     \begin{subcase}
@@
   \begin{case}
-    $p=\ceu{p';\,p''}$.  There are three subcases.
+    $p=\ceu{p';\,p''}$.
+    %%
+    There are three subcases.
     \begin{subcase}
@@
   Let~$\delta=\<p,n,\nil>$.  The theorem is trivially true if~$\delta$ is
-  nested-irreducible, as by definition~$\delta\nst[0]\delta_\Hnst$.  Suppose
-  that is not the case.  Then, depending on the structure of~$p$, there
-  are~11 possibilities.  In each one of them, we show that
-  such~$\delta'_\Hnst$ indeed exists.
+  nested-irreducible, as by definition~$\delta\nst[0]\delta$.  Suppose that
+  is not the case.  Then, depending on the structure of~$p$, there are~11
+  possibilities.  In each one of them, we show that such~$\delta'_\Hnst$
+  indeed exists.
   \begin{case}
@@
 
-% LocalWords:  subcase subcases
+\begin{definition}\label{def.pot}
+  The potency of a program~$p$ in reaction to event~$e$,
+  denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ expressions
+  that can be reached during a reaction of~$p$ to~$e$.  Formally,
+  $\pot(p,e)$ is defined as follows:
+  \[
+    \pot(p,e)=\pot'(\bcast(p,e))\,,
+  \]
+  where
+  \begin{enumerate}[(a)]
+  \item$\pot'(p)=1$, if~$p=\ceu{\EmitInt(e)}$;
+    %%
+  \item$\pot'(p)=\max\{\pot'(p_1),\pot'(p_2)\}$,\\
+    if~$p=\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}}$;
+    %%
+  \item$\pot'(p)=\pot'(p_1)$, if~$p=\ceu{\Loop{p_1}}$
+    or~$p=\ceu{\Every{e}\,{p_1}}$;
+    %%
+  \item$\pot'(p)=\pot'(p_1)+\pot'(p_2)$, if~$p=p_1;\,p_2$
+    or~$p=\ceu{p_1\AtLoop{p_2}}$
+    with~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
+    or~$p=\ceu{{p_1}\And{p_2}}$ or~$p=\ceu{{p_1}\Or{p_2}}$
+    or~$p=\ceu{{p_1}\AtAnd{p_2}}$ or~$p=\ceu{{p_1}\AtOr{p_2}}$;
+    %%
+  \item$\pot'(p)=0$, otherwise.
+  \end{enumerate}
+  Note that~$\pot'$ only considers the~$\ceu{\EmitInt}$ expressions that
+  occur before the first~$\ceu{\AwaitExt}$ in a given execution path.  This
+  is because such~$\ceu{\AwaitExt}$ act as barriers that cannot be consumed
+  in the reaction to the original event~$e$.
+\end{definition}
+
+\begin{definition}\label{def.rank}
+  The \emph{rank} of a description~$\delta=\<p,n,e>$,
+  denoted~$\rank(\delta)$, is a pair of nonnegative integers~$\<i,j>$ such
+  that~$i=\pot(p,e)$, and $j=n$~if~$e=\nil$ or~$j=n+1$ otherwise.
+\end{definition}
+
+The next two lemmas establish that a single application of an~$\out$
+or~$\nst$ transition either preserves or decreases the rank of the input
+description.  All rank comparisons assume lexicographical order.
+
+\begin{lemma}\strut
+  \begin{enumerate}[(a)]
+  \item If~$\delta\outpush\delta'$ then~$\rank(\delta)=\rank(\delta')$.
+  \item If~$\delta\outpop\delta'$ then~$\rank(\delta)>\rank(\delta')$.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}\strut
+  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$,
+  $\rank(\delta)=\<i,j>$, and~$\rank(\delta')=\<i',j'>$.
+  \begin{enumerate}[(a)]
+  \item Suppose
+    \[
+      \<p,n,e>\outpush\<p',n',e'>\,.
+    \]
+    By~\R{push}, $e\ne\nil$, $p'=\bcast(p,e)$, $n'=n+1$, and~$e'=\nil$.  By
+    Definition~\ref{def.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
+    as~$e'=\nil$ and~$n'=n+1$; hence~$j=j'$.
+
+    It remains to be shown that~$i=i'$:
+    \begin{align*}
+      i&=\pot(p,e)
+         \tag*{by Definition~\ref{def.rank}}\\
+       &=\pot'(\bcast(p,e))
+         \tag*{by Definition~\ref{def.pot}}\\
+       &=\pot'(p')
+         \tag*{since~$p'=\bcast(p,e)$}\\
+       &=\pot'(\bcast(p',\nil))
+         \tag*{by definition of~$\bcast$}\\
+       &=\pot'(\bcast(p',e'))
+         \tag*{since~$e'=\nil$}\\
+       &=\pot(p',e')
+         \tag*{by Definition~\ref{def.pot}}\\
+       &=i'
+         \tag*{by Definition~\ref{def.rank}}
+    \end{align*}
+
+  \item Suppose
+    \[
+      \<p,n,e>\outpop\<p',n',e'>\,.
+    \]
+    By~\R{pop}, $p=p'$, $n>0$, $n'=n-1$, and~$e=e'=\nil$.
+    %%
+    By Definition~\ref{def.pot}, $\pot(\bcast(p,e))=\pot(\bcast(p',e')$;
+    hence~$i=i'$.  And by Definition~\ref{def.rank}, $j=n$, as~$e=\nil$,
+    and~$j'=n-1$, as~$e'=\nil$ and~$n'=n-1$; hence~$j>j'$.
+    Therefore,~${\<i,j>}>{\<i',j'>}$.\qedhere
+  \end{enumerate}
+\end{proof}
+
+\begin{lemma}\label{lem.rank-nst}
+If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
+\end{lemma}
+\begin{proof}
+  By induction on the structure of~$\nst$ derivations.
+  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$, $\rank(\delta)=\<i,j>$,
+  and~$\rank(\delta')=\<i',j'>$.  Suppose there is a derivation~$\pi$
+  such that
+  \[
+    \pi\Vdash\<p,n,e>\nst\<p',n',e'>\,.
+  \]
+  By definition of~$\nst$, we have that~$e=\nil$ and $n=n'$.  Depending on
+  the structure of program~$p$, there are~11 possibilities.  In each one of
+  them, we show that~$\rank(\delta)\ge\rank(\delta')$.
+
+  \begin{case}
+    $p=\ceu{\Mem(id)}$.
+    %%
+    Then~$\pi$ is an instance of~\R{mem}.  Hence~$p'=\ceu{\Nop}$
+    and~$e'=\nil$.  Thus~$\rank(\delta)=\rank(\delta')=\<0,n>$.
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{\EmitInt(e_1)}$.
+    %%
+    Then~$\pi$ is an instance of~\R{emit-int}.  Hence~$p'=\ceu{\Nop}$
+    and~$e'=e_1$.
+    Thus~${\rank(\delta)={\<1,n>}}>{\<0,n+1>=\rank(\delta')}$.
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{\CanRun(n)}$.
+    %%
+    Then~$\pi$ is an instance of~\R{can-run}.  Hence~$p'=\ceu{\Nop}$
+    and~$e'=\nil$.  Thus~$\rank(\delta)=\rank(\delta')=\<0,n>$.
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{\IfElse{p}{p_1}{p_2}}$.
+    %%
+    There are two subcases.
+    \begin{subcase}
+      \label{lem.rank-nst.if-true}
+      $\eval(\ceu{\Mem(\Id)})$~is true.
+      %%
+      Then~$\pi$ is an instance of~\R{if-true}.  Hence~$p'=\ceu{p_1}$
+      and~$e'=\nil$.  Thus~$\rank(\delta)=\<i,n>$
+      and~$\rank(\delta')=\<j,n>$, where~$i=\max\{pot'(p_1),pot'(p_2)\}$
+      and~$j=\pot'(p_1)$.  Therefore~$i>j$.
+    \end{subcase}
+    \begin{subcase}
+      $\eval(\ceu{\Mem(\Id)})$~is false.
+      %%
+      Similar to Case~\ref{lem.rank-nst.if-true}.
+    \end{subcase}
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{p_1;\,p_2}$.
+    %%
+    There are three subcases.
+    \begin{subcase}
+      $p_1=\ceu{\Nop}$.
+      %%
+      Then~$\pi$ is an instance of~\R{seq-nop}.  TODO.
+    \end{subcase}
+    \begin{subcase}
+      $p_1=\ceu{\Break}$.
+      %%
+      Then~$\pi$ is an instance of~\R{seq-brk}.  TODO.
+    \end{subcase}
+    \begin{subcase}
+      \label{lem.rank-nst.seq-adv}
+      $p_1\ne\ceu{\Nop},\ceu{\Break}$.
+      %%
+      Then~$\pi$ is an instance of~\R{seq-adv}.
+      Thus there is a derivation~$\pi'$ such that
+      \[
+        \pi'\Vdash\<p_1,n,\nil>\nst\<p_1',n,e_1'>\,,
+      \]
+      for some~$p_1'\in\P$ and~$e_1'\in\E$.  Thus~$p'=p_1';p_2$
+      and~$e'=e_1'$.  By the induction hypothesis,
+      \begin{equation}
+        \label{lem.rank-nst.seq-adv.eq1}
+        \rank(\<p_1,n,\nil>)\ge\rank(\<p_1',n,e_1'>)\,.
+      \end{equation}
+      There are two subcases.
+      \begin{subsubcase}
+        $e'=\nil$.
+        %%
+        Then
+        \begin{align*}
+          \rank(\delta)&=\<\pot'(p_1)+\pot'(p_2),n>\,,\enspace\text{and}\\
+          \rank(\delta')&=\<\pot'(p_1')+\pot'(p_2),n>\,.
+        \end{align*}
+        By~\eqref{lem.rank-nst.seq-adv.eq1} and from the fact
+        that~$e_1'=\nil$, $\pot'(p_1)\ge\pot'(p_1')$.  Therefore,
+        $\rank(\delta)\ge\rank(\delta')$.
+      \end{subsubcase}
+      \begin{subsubcase}
+        $e'\ne\nil$.
+        %%
+        Then~$\pi'$ is an instance of~\R{emit-int}, as no other rule affects
+        the third component of the output description.
+        \[
+          \rank(\delta)=\<i,n>
+          >\<i-1,n+1>
+          =\rank(\delta')\,.
+        \]
+      \end{subsubcase}
+    \end{subcase}
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{\Loop{p_1}}$.
+  \end{case}
+
+  TODO\dots
+\end{proof}
+
+\begin{theorem}[Reaction termination]
+  For any~$\delta$ there is a~$\delta'_\Hnst$ such
+  that~$\delta\trans[*]\delta'$ and~$\rank(\delta')=\<0,0>$.
+\end{theorem}
+\begin{proof}
+  By lexicographic induction on~$rank(\delta)$.
+\end{proof}
+%   \begin{basis}
+%     If~$\<i,j>=\<0,0>$ then, by~\ref{?}, $\delta$ cannot be advanced
+%     by~$\out$ transitions.  There are two possibilities\dots\ (it can or
+%     cannot be advanced by nst transitions).
+%   \end{basis}
+%   \begin{induction}
+%     Let~$\<i,j>\ne\<0,0>$ and suppose that the theorem holds for
+%     all~$\delta''$ such that~$\rank(\delta'')<\rank(\delta)$.
+%     \begin{case}
+%       $\delta$~is nested-irreducible.
+%       %%
+%       \begin{subcase}
+%         $\delta\outpush\delta_1$, for some~$\delta_1$.
+%       \end{subcase}
+%       \begin{subcase}
+%         $\delta\outpop\delta_1$, for some~$\delta_1$.
+%         %%
+%         Then, by definition of~\R{pop}, $\rank(\delta_1)=\<i,j-1>$, and by
+%         the induction hypothesis\dots
+%       \end{subcase}
+%     \end{case}
+%     \begin{case}
+%       $\delta$~is not nested-irreducible.
+%       %%
+%     \end{case}
+%   \end{induction}

### 2018-02-22 05-conclusion.tex
Conclusão

@@
 
-\fs{TODO}
+The programming language \CEU aims to offer a concurrent, safe, and realistic
+alternative to C for embedded soft real-time systems, such as sensor networks
+and multimedia applications.
+%
+%Its synchronous semantics enables safe shared-memory concurrency and
+%safe abortion of lines of execution, while enforcing memory-bounded,
+%deterministic, and terminating reactions to the environment.
+%
+\CEU inherits the synchronous and imperative mindset of Esterel but adopts a
+simpler semantics with fine-grained execution control, which makes the
+language fully deterministic.
+%
+In addition, its stack-based execution for internal events provides a limited
+but memory-bounded form of subroutines.
+%
+\CEU also provides a finalization mechanism for resources when interacting with
+the external environment.
+
+We present a small-step structural operational semantics for
+\CEU and a proof that reactions are deterministic, terminate in finite time,
+and use bounded memory, i.e., that for a given arbitrary timeline of input
+events, multiple executions of the same program always react in bounded time
+and arrive at the same final finite memory state.

### 2018-02-22 06-appendix.tex
Conserta a definição de \pot' no apêndice

@@
       %%
-      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-break}.
+      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-brk}.
       Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
@@
       \end{subsubcase}
-      \begin{subsubcase}\label{lem.det-nst.and-brk1}
+      \begin{subsubcase}
+        \label{lem.det-nst.and-brk1}
         $p'=\ceu{\Break}$.
@@
 
-\begin{definition}\label{def.pot}
+\begin{definition}
+  \label{def.pot}
+  %%
   The potency of a program~$p$ in reaction to event~$e$,
   denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ expressions
-  that can be reached during a reaction of~$p$ to~$e$.  Formally,
-  $\pot(p,e)$ is defined as follows:
+  that can be reached during a reaction of~$p$ to~$e$.
+
+  More formally,
   \[
@@
   \]
-  where
+  where~$\pot'$ is an auxiliary function that counts the number of
+  reachable~$\ceu{\EmitInt}$ in the program resulting from the broadcast of
+  event~$e$ to~$p$.
+
+  The auxiliary function~$\pot'$ is defined by the following clauses:
   \begin{enumerate}[(a)]
-  \item$\pot'(p)=1$, if~$p=\ceu{\EmitInt(e)}$;
-    %%
-  \item$\pot'(p)=\max\{\pot'(p_1),\pot'(p_2)\}$,\\
-    if~$p=\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}}$;
-    %%
-  \item$\pot'(p)=\pot'(p_1)$, if~$p=\ceu{\Loop{p_1}}$
-    or~$p=\ceu{\Every{e}\,{p_1}}$;
-    %%
-  \item$\pot'(p)=\pot'(p_1)+\pot'(p_2)$, if~$p=p_1;\,p_2$
-    or~$p=\ceu{p_1\AtLoop{p_2}}$
-    with~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
-    or~$p=\ceu{{p_1}\And{p_2}}$ or~$p=\ceu{{p_1}\Or{p_2}}$
-    or~$p=\ceu{{p_1}\AtAnd{p_2}}$ or~$p=\ceu{{p_1}\AtOr{p_2}}$;
-    %%
-  \item$\pot'(p)=0$, otherwise.
+    \item$\pot'(\ceu{\EmitInt}(e))=1$.
+    \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})
+      =\max\{\pot'(p_1),\pot'(p_2)\}$.
+    \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$.
+    \item If~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
+      \begin{align*}
+        \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\
+        %%
+        \pot'(\ceu{p_1\AtLoop{p_2}})&=
+        \begin{cases}
+          \pot'(p_1)              &\text{if~(\dag)}\\
+          \pot'(p_1)+\pot'(p_2)   &\text{otherwise}\\
+        \end{cases}\\
+        %%
+        \pot'(\ceu{{p_1}\And{p_2}})&=\pot'(p_1)+\pot'(p_2)\\
+        %%
+        \pot'(\ceu{{p_1}\AtAnd{p_2}})&=\pot'(p_1)+\pot'(p_2)\\
+        %%
+        \pot'(\ceu{{p_1}\Or{p_2}})&=\pot'(p_1)+\pot'(p_2)\\
+        %%
+        \pot'(\ceu{{p_1}\AtOr{p_2}})&=\pot'(p_1)+\pot'(p_2)\,,
+      \end{align*}
+      where~(\dag) stands for: ``A~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$
+      occurs in all execution paths of~$p_1$.''
+      \item Otherwise, if none of~(a)--(d) applies, $\pot(\ast)=0$.
     \end{enumerate}
-  Note that~$\pot'$ only considers the~$\ceu{\EmitInt}$ expressions that
-  occur before the first~$\ceu{\AwaitExt}$ in a given execution path.  This
-  is because such~$\ceu{\AwaitExt}$ act as barriers that cannot be consumed
-  in the reaction to the original event~$e$.
+
+  TODO: Define~$\pot'(\ceu{\Every{e}\,{p_1}})$ above.
 \end{definition}
@@
   denoted~$\rank(\delta)$, is a pair of nonnegative integers~$\<i,j>$ such
-  that~$i=\pot(p,e)$, and $j=n$~if~$e=\nil$ or~$j=n+1$ otherwise.
+  that
+  \begin{alignat*}{2}
+    i&=\pot(p,e) &\quad\text{and}\quad
+    j&=
+       \begin{cases}
+         n  &\text{if~$e=\nil$}\\
+         n+1&\text{otherwise\,.}
+       \end{cases}
+  \end{alignat*}
 \end{definition}
@@
   \begin{enumerate}[(a)]
-  \item Suppose
-    \[
-      \<p,n,e>\outpush\<p',n',e'>\,.
-    \]
-    By~\R{push}, $e\ne\nil$, $p'=\bcast(p,e)$, $n'=n+1$, and~$e'=\nil$.  By
+  \item Suppose~$\<p,n,e>\outpush\<p',n',e'>$.  Then, by~\R{push},
+    $e\ne\nil$, $p'=\bcast(p,e)$, $n'=n+1$, and~$e'=\nil$.  By
     Definition~\ref{def.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
     as~$e'=\nil$ and~$n'=n+1$; hence~$j=j'$.
-
+    %%
     It remains to be shown that~$i=i'$:
@@
     \end{align*}
-
-  \item Suppose
-    \[
-      \<p,n,e>\outpop\<p',n',e'>\,.
-    \]
-    By~\R{pop}, $p=p'$, $n>0$, $n'=n-1$, and~$e=e'=\nil$.
-    %%
-    By Definition~\ref{def.pot}, $\pot(\bcast(p,e))=\pot(\bcast(p',e')$;
+    Therefore, $\<i,j>=\<i',j'>$.
+
+  \item Suppose~$\<p,n,e>\outpop\<p',n',e'>$.  Then, by~\R{pop}, $p=p'$,
+    $n>0$, $n'=n-1$, and~$e=e'=\nil$.
+    %%
+    By Definition~\ref{def.pot}, $\pot(\bcast(p,e))=\pot(\bcast(p',e'))$;
     hence~$i=i'$.  And by Definition~\ref{def.rank}, $j=n$, as~$e=\nil$,
@@
 
-\begin{lemma}\label{lem.rank-nst}
+\begin{lemma}
+  \label{lem.rank-nst}
+  %%
   If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
@@
 \begin{proof}
-  By induction on the structure of~$\nst$ derivations.
+  We proceed by induction on the structure of~$\nst$ derivations.
   Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$, $\rank(\delta)=\<i,j>$,
-  and~$\rank(\delta')=\<i',j'>$.  Suppose there is a derivation~$\pi$
-  such that
+  and~$\rank(\delta')=\<i',j'>$.  By the theorem hypothesis, there is a
+  derivation~$\pi$ such that
   \[
@@
   \]
-  By definition of~$\nst$, we have that~$e=\nil$ and $n=n'$.  Depending on
-  the structure of program~$p$, there are~11 possibilities.  In each one of
-  them, we show that~$\rank(\delta)\ge\rank(\delta')$.
+  By definition of~$\nst$, $e=\nil$ and $n=n'$.  Depending on the structure
+  of program~$p$, there are~11 possibilities.  In each one of them, we show
+  that~$\rank(\delta)\ge\rank(\delta')$.
 
@@
     Then~$\pi$ is an instance of~\R{mem}.  Hence~$p'=\ceu{\Nop}$
-    and~$e'=\nil$.  Thus~$\rank(\delta)=\rank(\delta')=\<0,n>$.
+    and~$e'=\nil$.  Thus
+    \[
+      \rank(\delta)=\rank(\delta')=\<0,n>\,.
+    \]
   \end{case}
@@
     Then~$\pi$ is an instance of~\R{emit-int}.  Hence~$p'=\ceu{\Nop}$
-    and~$e'=e_1$.
-    Thus~${\rank(\delta)={\<1,n>}}>{\<0,n+1>=\rank(\delta')}$.
+    and~$e'=e_1\ne\nil$.
+    Thus
+    \[
+      {\rank(\delta)={\<1,n>}}>{\<0,n+1>=\rank(\delta')}\,.
+    \]
   \end{case}
@@
     Then~$\pi$ is an instance of~\R{can-run}.  Hence~$p'=\ceu{\Nop}$
-    and~$e'=\nil$.  Thus~$\rank(\delta)=\rank(\delta')=\<0,n>$.
+    and~$e'=\nil$.  Thus
+    \[
+      \rank(\delta)=\rank(\delta')=\<0,n>\,.
+    \]
   \end{case}
@@
       Then~$\pi$ is an instance of~\R{if-true}.  Hence~$p'=\ceu{p_1}$
-      and~$e'=\nil$.  Thus~$\rank(\delta)=\<i,n>$
-      and~$\rank(\delta')=\<j,n>$, where~$i=\max\{pot'(p_1),pot'(p_2)\}$
-      and~$j=\pot'(p_1)$.  Therefore~$i>j$.
+      and~$e'=\nil$.  Thus
+      \begin{align*}
+        \rank(\delta)&=\<\max\{pot'(p_1),pot'(p_2)\},n>\\
+                     &\ge\<\pot'(p_1),n>=\rank(\delta')\,.
+      \end{align*}
     \end{subcase}
@@
     \begin{subcase}
+      \label{lem.rank-nst.seq-nop}
       $p_1=\ceu{\Nop}$.
       %%
-      Then~$\pi$ is an instance of~\R{seq-nop}.  TODO.
-    \end{subcase}
-    \begin{subcase}
+      Then~$\pi$ is an instance of~\R{seq-nop}.
+      Hence~$p'=p_2$ and~$e'=\nil$.  Thus
+      \begin{align*}
+        \rank(\delta)&=\<\pot'(p_1)+\pot'(p_2),n>\\
+                     &\ge\<pot'(p_2),n>=\rank(\delta')\,.
+      \end{align*}
+    \end{subcase}
+    \begin{subcase}
+      \label{lem.rank-nst.seq-brk}
       $p_1=\ceu{\Break}$.
       %%
-      Then~$\pi$ is an instance of~\R{seq-brk}.  TODO.
+      Then~$\pi$ is an instance of~\R{seq-brk}.
+      Hence~$p'=p_1$ and~$e'=\nil$.  Thus
+      \[
+        \rank(\delta)=\<0,n>=\rank(\delta')\,.
+      \]
     \end{subcase}
@@
       %%
-      Then~$\pi$ is an instance of~\R{seq-adv}.
-      Thus there is a derivation~$\pi'$ such that
+      Then~$\pi$ is an instance of~\R{seq-adv}.  Hence there is a
+      derivation~$\pi'$ such that
       \[
@@
       \]
-      for some~$p_1'\in\P$ and~$e_1'\in\E$.  Thus~$p'=p_1';p_2$
-      and~$e'=e_1'$.  By the induction hypothesis,
+      for some~$p_1'$ and~$e_1'$.  Thus~$p'=p_1';p_2$ and~$e'=e_1'$.  By the
+      induction hypothesis,
       \begin{equation}
@@
         \begin{align*}
-          \rank(\delta)&=\<\pot'(p_1)+\pot'(p_2),n>\,,\enspace\text{and}\\
+          \rank(\delta)&=\<\pot'(p_1)+\pot'(p_2),n>\enspace\text{and}\\
           \rank(\delta')&=\<\pot'(p_1')+\pot'(p_2),n>\,.
         \end{align*}
-        By~\eqref{lem.rank-nst.seq-adv.eq1} and from the fact
-        that~$e_1'=\nil$, $\pot'(p_1)\ge\pot'(p_1')$.  Therefore,
-        $\rank(\delta)\ge\rank(\delta')$.
+        By~\eqref{lem.rank-nst.seq-adv.eq1}, $\pot'(p_1)\ge\pot'(p_1')$.
+        Thus
+        \[
+          \rank(\delta)\ge\rank(\delta')\,.
+        \]
       \end{subsubcase}
@@
         %%
-        Then~$\pi'$ is an instance of~\R{emit-int}, as no other rule affects
-        the third component of the output description.
-        \[
-          \rank(\delta)=\<i,n>
-          >\<i-1,n+1>
-          =\rank(\delta')\,.
-        \]
-      \end{subsubcase}
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
+        Then~$\pi'$ contain one application of~\R{emit-int}, which consumes
+        one~$\ceu{\EmitInt(e')}$ expression from~$p_1$ and implies
+        $\pot'(p_1)>\pot'(p_1')$.  Thus
+        \begin{align*}
+          \rank(\delta)&=\<\pot'(p_1)+\pot'(p_2),n>\\
+                       &>\<\pot'(p_1')+\pot'(p_2),n+1>=\rank(\delta')\,.
+        \end{align*}
+      \end{subsubcase}
+    \end{subcase}
+  \end{case}
+
+  \begin{case}
+    \label{lem.rank-nst.loop-expd}
     $p=\ceu{\Loop{p_1}}$.
-  \end{case}
-
-  TODO\dots
+    %%
+    Then~$\pi$ is an instance of~\R{loop-expd}.
+    Hence~$p'=\ceu{p_1\AtLoop{p_1}}$ and~$e'=\nil$.
+    %%
+    TODO: Use condition~(\dag) in definition of~$\pot'$.
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{{p_1}\AtLoop{p_2}}$.
+    %%
+    There are three cases.
+    \begin{subcase}
+      $p_1=\ceu{\Nop}$.
+      %%
+      Similar to Case~\ref{lem.rank-nst.seq-nop}.
+    \end{subcase}
+    \begin{subcase}
+      $p_1=\ceu{\Break}$.
+      %%
+      Similar to Case~\ref{lem.rank-nst.seq-brk}.
+    \end{subcase}
+    \begin{subcase}
+      \label{lem.rank-nst.loop-adv}
+      $p_1\ne\ceu{\Nop},\ceu{\Break}$.
+      %%
+      Similar to Case~\ref{lem.rank-nst.seq-adv}.
+    \end{subcase}
+  \end{case}
+
+  \begin{case}
+    \label{lem.rank-nst.and.expd}
+    $p=\ceu{{p_1}\And{p_2}}$.
+    %%
+    Then~$\pi$ is an instance of~\R{and-expd}.
+    Hence~$p'=\ceu{{p_1}\AtAnd(\CanRun(n);\,p_2)}$ and~$e'=\nil$.
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{{p_1}\AtAnd{p_2}}$.
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{{p_1}\Or{p_2}}$.
+  \end{case}
+
+  \begin{case}
+    $p=\ceu{{p_1}\AtOr{p_2}}$.
+  \end{case}
 \end{proof}

### 2018-02-22 06-appendix.tex
Typo no apêndice

@@
   denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ expressions
-  that can be reached during a reaction of~$p$ to~$e$.
+  that can be executed during a reaction of~$p$ to~$e$.
 

### 2018-02-22 01-intro.tex
Revisão Seção 1-2   - tirei "par do .. end" (não é usado no paper)

@@
 reduction in source code size (around 25\%) with a small increase in memory
-usage (around 5--10\% for \emph{text} and \emph{data})~\cite{ceu.sensys13}.
+usage (around 5--10\%)~\cite{ceu.sensys13}.
 %

### 2018-02-22 02-ceu.tex
Revisão Seção 1-2   - tirei "par do .. end" (não é usado no paper)

@@
 par/and do $\T<stmts>$ with $\T<stmts>$ end | \C{.45}{terminates when all sides ends} |
-par     do $\T<stmts>$ with $\T<stmts>$ end | \C{.45}{never terminates}               |
 
@@
 Then, it uses a \code{par/or} composition to run two activities in parallel:
-a single statement that waits for a button press before terminating (ln.~4),
-and an endless loop that toggles the LED on and off on radio receives
+a single-statement trail that waits for a button press before terminating
+(ln.~4), and an endless loop that toggles the LED on and off on radio receives
 (ln.~9--14).
@@
   caption={A program in \CEU that toggles a LED on every radio receive,
-           terminating on a button press in a consistent state.},
+           terminating on a button press always with the LED off.},
   label={lst.ceu},
@@
 environment should only provide access to~C operations that can be assumed to
-be instantaneous, such as non-blocking~I/O and simple accesses to data
-structures.
+be instantaneous, such as non-blocking~I/O and simple data structure accessors.
 
@@
 \CEU defines time as a discrete sequence of reactions to unique external
-input events.
+input events received from the environment.
 %
-External input events are received from the environment, and each delimits a
-new logical unit of time that triggers an associated reaction.
+Each input event delimits a new logical unit of time that triggers an
+associated reaction.
 %
@@
 %
-Considering the context soft real-time systems,
+Considering the context of soft real-time systems,
 %targeted by \CEU (e.g., sensor networks, multimedia systems, interactive games, etc.)
-such delays and postponed reactions might be tolerated as long as they are
-infrequent and the application does not take too long to catch up with real
-time.
+postponed reactions might be tolerated as long as they are infrequent and the
+application does not take too long to catch up with real time.
 %
-Note that the synchronous semantics is the norm in typical event-driven
+Note that the synchronous semantics is also the norm in typical event-driven
 systems, such as event dispatching in UI toolkits, game loops in game engines,
@@
 
-This form of subroutines has a significant limitation though: it cannot
-express recursion, since an \code{emit} to itself is always ignored as a
-running trail cannot be waiting on itself.
+This form of subroutines has a significant limitation that it cannot express
+recursion, since an \code{emit} to itself is always ignored as a running trail
+cannot be waiting on itself.
 %
@@
 However, the \code{loop} variation would not compile because it does not
-contain a path to an external input \code{await} (\code{e} is an internal
+contain a path to an external input \code{await} (since \code{e} is an internal
 event).
 %
-However, event iterators enforce syntactic restrictions and cannot contain
-\code{await} or \code{break} statements.
+However, event iterators enforce other syntactic restrictions and cannot
+contain \code{await} or \code{break} statements.
 The absence of \code{break} guarantees that an iterator never terminates from
@@
 \end{lstlisting}
-\centering\small{\bx Accesses to \code{y} are concurrent but deterministic.}
+\centering\small{\bx Accesses to \code{y} are concurrent but still deterministic.}
 \end{minipage}
@@
 In the example \ax, the two assignments to \code{x} can only execute in
-reactions to different events \code{A} and \code{B}, which cannot occur
-simultaneously by definition (Section~\ref{sec.ceu.evts}).
+reactions to different events \code{A} and \code{B} (ln. 5,8), which cannot
+occur simultaneously by definition.
 Hence, for the sequence of events \code{A->B}, \code{x} becomes \code{4}

### 2018-02-22 03-semantics.tex
Conserta if-else na gramática da Seção 3.2
Adiciona teorema da terminação no fim do apêndice
Correção Rodrigo e remoção do "Problema"
Revisão Seção 3-5

@@
 
-In this section, we introduce a reduced syntax for \CEU and propose an
-operational semantics to formally describe the behavior of its programs.
+In this section, we introduce a reduced syntax of \CEU and propose an
+operational semantics to formally describe the behavior of programs.
 We describe a small synchronous kernel highlighting the peculiarities of \CEU,
@@
 \bgroup
-\def\lbl#1{\qquad\text{\emph{#1}}}%
+\def\lbl#1{\quad\text{\emph{#1}}}%
 \newdimen\X
@@
       %%
-  \mid&\enspace\ceu{\Ifelse{b}{p_1}{p_2}}
+  \mid&\enspace\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}}
       &&\lbl{conditional}\\[\X]
       %%
-  \mid&\enspace\ceu{p1\,;\,p2}
+  \mid&\enspace\ceu{p_1\,;\,p_2}
       &&\lbl{sequence}\\[\X]
@@
 
-The~$\ceu{\Mem(id)}$ primitive represents all accesses, assignments, and system calls
-that affect a memory location identified by~$id$.
+The~$\ceu{\Mem(id)}$ primitive represents all accesses, assignments, and system
+calls that affect a memory location identified by~$id$.
 %
@@
 
-We assume that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$ and~$\ceu{\EmitInt}$ expressions do not
-share identifiers: any identifier is either a variable, an external event,
-or an internal event.
+We assume that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$
+and~$\ceu{\EmitInt}$ expressions do not share identifiers:
+any identifier is either a variable, an external event, or an internal event.
 
-Most expressions in the abstract language are mapped to
-their counterparts in the concrete language.  The exceptions are the
-finalization block~$\ceu{\Fin{p}}$ and the \texttt{@}-expressions, which are
-internal expressions that result from the
-expansion of awaits, emits, and loops by the transition rules to be discussed.
+Most expressions in the abstract language are mapped to their counterparts in
+the concrete language.
+The exceptions are the finalization block~$\ceu{\Fin{p}}$ and the
+\texttt{@}-expressions which result from expansions in the transition rules to
+be presented.
 
-Regarding other mismatches between the concrete and abstract languages, the concrete
+Regarding mismatches between the concrete and abstract languages, the concrete
 \code{await} and \code{emit} primitives support communication of values between
@@
 variable to hold the value of an $\ceu{\EmitInt}$ and access it after the
-corresponding $\ceu{\AwaitInt}$.
+corresponding $\ceu{\AwaitInt}$ awakes.
 %
@@
 %
-We use a set of small-step operational rules, which are built in such a way
+We use a set of small-step operational rules, which are designed in such a way
 that at most one transition is possible at any time, resulting in deterministic
@@
 \]
-where~
-%
-\rc{Na BNF o $p$ não deveria ser $\P$~(maiúsculo) tbm?}
-%
-$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are
-nonnegative integers representing the current stack level,
-and~
-%
-\rc{Checar: Acho que o conjunto $\E$~não foi previamente definido em
-  lugar nenhum}
-%
-$e,e'\in\E\cup\{\nil\}$
-are the events emitted before and after the
-transition (both being possibly the empty event~$\nil$).
+where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are non-negative
+integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
+the events emitted before and after the transition (both being possibly the
+empty event~$\nil$).
 
@@
 At the beginning of a reaction to an input event~$id$, the input description is
-initialized with stack level~0 ($n=0$) and with the emitted event
+initialized with stack level~0 ($n=0$) and with the externally emitted event
 ($e=id$).
@@
 
-The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions
-that only apply to the program as a whole and manipulate the stack level:
+The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions which only
+apply to the program as a whole, and are the only to manipulate the stack
+level:
 \begin{align*}
@@
 This condition ensures that an $\ceu{\EmitInt}$ only resumes after its internal
-reaction completes and blocks, as discussed in Section~\ref{sec.ceu.evts}.
+reaction completes and blocks in the current stack level.
 
@@
 same program at stack level~0.
-(Rule \R{pop} is the only rule that decreases the stack level.)
 
@@
 
-The~$\nst$ rules are recursive definitions with the following general
-format:
+The~$\nst$ rules are recursive definitions with the following general format:
 \[
@@
 
-A complete reaction consists of the following series of transitions:
+A complete reaction consists of a series of transitions:
 \begin{align*}
@@
 
-A $\ceu{\Mem}$ operation becomes a $\ceu{\Nop}$ which indicates the memory access (rule
-\R{mem}).
-An $\ceu{\EmitInt(id)}$ generates an event $\ceu{\Id}$ and transits to a $\ceu{\CanRun(n)}$ which
-can only resume at level~$n$ (rule \R{emit-int}).
-Since all~$\nst$ rules can only transit with $e=\nil$, an $\ceu{\EmitInt}$ causes
-rule \R{push} to execute at the outer level, creating a new level~$n+1$ on
-the stack.
+A $\ceu{\Mem}$ operation becomes a $\ceu{\Nop}$ which indicates the memory
+access (rule \R{mem}).
+An $\ceu{\EmitInt(id)}$ generates an event $\ceu{\Id}$ and transits to a
+$\ceu{\CanRun(n)}$ which can only resume at level~$n$ (rule \R{emit-int}).
+Since all~$\nst$ rules can only transit with $e=\nil$, an $\ceu{\EmitInt}$
+inevitably causes rule \R{push} to execute at the outer level, creating a new
+level~$n+1$ on the stack.
 Also, with the new stack level, the resulting $\ceu{\CanRun}(n)$ itself cannot
-transit (rule~\R{can-run}), providing the desired stack-based semantics for internal events.
+transit yet (rule~\R{can-run}), providing the desired stack-based semantics for
+internal events.
 
@@
   \AxiomC{$\eval(\ceu{\Mem(\Id)})$}
-  \UnaryInfC{$\<\ceu{\Ifelse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<p,n,\nil>$}
+  \UnaryInfC{$\<\ceu{\IfElse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<p,n,\nil>$}
   \DisplayProof
@@
   \AxiomC{$\lnot\eval(\ceu{\Mem(\Id)})$}
-  \UnaryInfC{$\<\ceu{\Ifelse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<q,n,\nil>$}
+  \UnaryInfC{$\<\ceu{\IfElse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<q,n,\nil>$}
   \DisplayProof
@@
 %
-Function~$\eval$ evaluates a given~$\ceu{\Mem}$ expression.
+Function~$\eval$ evaluates a given~$\ceu{\Mem}$ expression to a boolean value.
 %
@@
 
-Rules~\R{and-expd} and~\R{or-expd} insert a~$\ceu{\CanRun(n)}$ at the
-beginning of the right branch.
-This ensures that~any $\ceu{\EmitInt}$ on the left branch, which transits to a~$\ceu{\CanRun(n)}$, still resumes before the right branch starts.
+Rules~\R{and-expd} and~\R{or-expd} insert a~$\ceu{\CanRun(n)}$ at the beginning
+of the right branch.
+This ensures that~any $\ceu{\EmitInt}$ on the left branch, which transits to
+a~$\ceu{\CanRun(n)}$, still resumes before the right branch starts.
 %
@@
 Finally, a~$\ceu{\Break}$ in one of the sides in parallel escapes the closest
-enclosing~$\ceu{\Loop}$, properly aborting the other side by applying the~$\clear$
+enclosing~$\ceu{\Loop}$, properly aborting the other side with the~$\clear$
 function:
@@
 
-A reaction eventually blocks in~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$, $\ceu{\Fin}$,
-and~$\ceu{\CanRun}$ expressions in parallel trails.
+A reaction eventually blocks in~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$,
+$\ceu{\Every}$, $\ceu{\Fin}$, and~$\ceu{\CanRun}$ expressions in parallel
+trails.
 %
-If no trails are blocked in~$\ceu{\CanRun}$ expressions, it means that the program
-cannot advance in the current reaction.
+Then, if none of the trails is blocked in~$\ceu{\CanRun}$ expressions, it means
+that the program cannot advance in the current reaction.
 %
-However, $\ceu{\CanRun}$ expressions can still resume in lower stack indexes and will
-eventually resume in the current reaction (see rule \R{pop}).
-%
-\rc{Não entendi essa discussão}
+However, $\ceu{\CanRun}$ expressions can still resume at lower stack indexes
+and will eventually resume in the current reaction (see rule \R{pop}).
 
@@
       bcast(\_,e)
-      &=\bot\enspace
+      &=\_\enspace
         (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
-      &\quad\ceu{\Ifelse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
+      &\quad\ceu{\IfElse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
       \\[1\jot]
@@
         (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
-      &\quad\ceu{\Ifelse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
+      &\quad\ceu{\IfElse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
         \\[1\jot]
@@
         (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
-      &\quad\ceu{\Ifelse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
+      &\quad\ceu{\IfElse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
     \end{align*}}
@@
 
+- reference the appendix A
+
 \subsubsection*{Determinism}

### 2018-02-22 04-related.tex
Revisão Seção 3-5

@@
 unique input events, which is a prerequisite for the concurrency checks that
-enable safe shared-memory concurrency.
+enable safe shared-memory concurrency, as discussed in
+Section~\ref{sec.ceu.shared}.
 
-Also, in most synchronous languages, the behavior of internal and external
-events is equivalent.
-In \CEU, internal events introduce stack-based micro reactions within external
-reactions, providing more fine-grained control for intra-reaction execution.
+In most synchronous languages, the behavior of external and internal events is
+equivalent.
+However, in \CEU, internal events introduce stack-based micro reactions within
+external reactions, providing more fine-grained control for intra-reaction
+execution.
 %
-This allows for memory-bounded subroutines that execute multiple times in the
-same external reaction.
+This allows for memory-bounded subroutines that can execute multiple times
+during the same external reaction.
 %
@@
 
-Like \CEU, many synchronous languages%
+Like \CEU, many other synchronous languages%
 ~\cite{rp.rc,wsn.protothreads,wsn.sol,rp.synchc,rp.pretc}
@@
 
-Regarding the integration with the $C$ language-based environment, \CEU
+Regarding the integration with $C$ language-based environments, \CEU
 supports a finalization mechanism for external resources.
@@
 boundaries and forces the programmer to provide associated finalizers.
+%
+As far as we know, this extra safety level is unique to \CEU.

### 2018-02-22 05-conclusion.tex
Revisão Seção 3-5

@@
 alternative to C for embedded soft real-time systems, such as sensor networks
-and multimedia applications.
+and multimedia systems.
 %
@@
 
-We present a small-step structural operational semantics for
-\CEU and a proof that reactions are deterministic, terminate in finite time,
-and use bounded memory, i.e., that for a given arbitrary timeline of input
-events, multiple executions of the same program always react in bounded time
-and arrive at the same final finite memory state.
+We propose a small-step structural operational semantics for \CEU and a proof
+that reactions are deterministic, terminate in finite time, and use bounded
+memory, i.e., that for a given arbitrary timeline of input events, multiple
+executions of the same program always react in bounded time and arrive at the
+same final finite memory state.

### 2018-02-23 06-appendix.tex
Conserta typo no apêndice
Corrige a prova da terminação do \nst[1]

@@
 
-\begin{lemma}\label{lem.det-out}
+\begin{lemma}
+  \label{lem.det-out}
+  %%
   If~$\delta\out\delta_1$ and~$\delta\out\delta_2$ then~$\delta_1=\delta_2$.
@@
 
-\begin{theorem}\label{thm.det-out-pop-n}
+\begin{theorem}
+  \label{thm.det-out-pop-n}
+  %%
   If~$\delta\out[n]\delta_1$ and~$\delta\out[n]\delta_2$
@@
   and~$\delta\out[1]\delta_2'\out[n-1]\delta_2$, for some~$n>1$
-  and~$\delta_1'$, $\delta_2'\in\Delta$.
+  and~$\delta_1'$ and~$\delta_2'$.
   %%
@@
 
-\begin{lemma}\label{lem.det-nst}
+\begin{lemma}
+  \label{lem.det-nst}
+  %%
   If~$\delta\nst\delta_1$ and~$\delta\nst\delta_2$ then~$\delta_1=\delta_2$.
@@
       \end{align*}
-      for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By the induction
+      for some~$p_1'$, $p_2'$, $e_1'$, and~$e_2'$.  By the induction
       hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
@@
   \begin{case}
-    $p=\ceu{p'\AtLoop{p''}}$.  There are three subcases.
+    $p=\ceu{p'\AtLoop{p''}}$.
+    %%
+    There are three subcases.
     \begin{subcase}
@@
   \begin{case}
-    $p=\ceu{p'\AtAnd{p''}}$.  There are two subcases.
-    \begin{subcase}
-      $\isblocked(p',n)$~is false.  There are three subcases.
+    $p=\ceu{p'\AtAnd{p''}}$.
+    %%
+    There are two subcases.
+    \begin{subcase}
+      $\isblocked(p',n)$~is false.
+      %%
+      There are three subcases.
       \begin{subsubcase}
@@
         Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop1}.
-        Hence~$p_1=p_2=p''$ and~$e_1=e_2\nil$.
+        Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
       \end{subsubcase}
@@
         Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-brk1}.
-        Hence~$p_1=p_2=\ceu{\clear(p'');\Break}$ and~$e_1=e_2\nil$.
-      \end{subsubcase}
-      \begin{subsubcase}\label{lem.det-nst.and-adv1}
+        Hence~$p_1=p_2=\ceu{\clear(p'');\Break}$ and~$e_1=e_2=\nil$.
+      \end{subsubcase}
+      \begin{subsubcase}
+        \label{lem.det-nst.and-adv1}
         $p'\ne\ceu{\Nop},\ceu{\Break}$.
@@
         \end{align*}
-        for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By the induction
+        for some~$p_1'$, $p_2'$, $e_1'$, $e_2'$.  By the induction
         hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
@@
     \begin{subcase}
-      $\isblocked(p',n)$~is true.  There are three subcases.
+      $\isblocked(p',n)$~is true.
+      %%
+      There are three subcases.
       \begin{subsubcase}
@@
         Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop2}.
-        Hence~$p_1=p_2=p'$ and~$e_1=e_2\nil$.
-      \end{subsubcase}
-      \begin{subsubcase}\label{lem.det-nst.and-brk2}
+        Hence~$p_1=p_2=p'$ and~$e_1=e_2=\nil$.
+      \end{subsubcase}
+      \begin{subsubcase}
+        \label{lem.det-nst.and-brk2}
         $p''=\ceu{\Break}$.
@@
       \end{subsubcase}
-      \begin{subsubcase}\label{lem.det-nst.and-adv2}
+      \begin{subsubcase}
+        \label{lem.det-nst.and-adv2}
         $p''\ne\ceu{\Nop},\ceu{\Break}$.
@@
         \end{align*}
-        for some~$p_1'',p_2''\in\P$ and~$e_1'',e_2''\in\E$.  By the
-        induction hypothesis, $p_1''=p_2''$ and~$e_1''=e_2''$.
+        for some~$p_1''$, $p_2''$, $e_1''$, and~$e_2''$.  By the induction
+        hypothesis, $p_1''=p_2''$ and~$e_1''=e_2''$.
         Hence~$p_1=\ceu{{p'}\And{p_1''}}=\ceu{{p'}\And{p_2''}}=p_2$
@@
   \begin{case}
-    $p=\ceu{p'\AtOr{p''}}$.  There are two subcases.
-    \begin{subcase}
-      $\isblocked(p',n)$~is false.  There are three subcases.
+    $p=\ceu{p'\AtOr{p''}}$.
+    %%
+    There are two subcases.
+    \begin{subcase}
+      $\isblocked(p',n)$~is false.
+      %%
+      There are three subcases.
       \begin{subsubcase}
@@
 
-\begin{theorem}\label{thm.det-nst-n}
+\begin{theorem}
+  \label{thm.det-nst-n}
+  %%
   If~$\delta\nst[n]\delta_1$ and~$\delta\nst[n]\delta_2$
@@
 
-\begin{definition}\label{def.Hnst}
+\begin{definition}
+  \label{def.Hnst}
+  %%
   A description~$\delta=\<p,n,e>$ is \emph{nested-irre\-ducible}
@@
 
-\begin{lemma}\label{lem.irr-nst-n}
+\begin{lemma}
+  \label{lem.irr-nst-n}
+  %%
   If~$\delta\nst[n]\delta_\Hnst'$ then, for all~$i\ne{n}$, there is
@@
   Let~$\delta\nst[n]\delta'_\Hnst$, for some~$n\ge0$.
+  %%
   There are two cases.
-  \begin{case}\label{lem.irr-nst-n-case1}
+  \begin{case}
+    \label{lem.irr-nst-n-case1}
+    %%
     Suppose there are~$i>n$ and~$\delta''_\Hnst$ such
@@
     Then, by definition of~$\nst[i]$,
-    \begin{equation}\label{lem.irr-nst-n-eq1}
+    \begin{equation}
+      \label{lem.irr-nst-n-eq1}
+      %%
       \delta\nst[n]\delta'\nst[n+1]\delta_1'\nst[n+2]\cdots\nst[i]\delta''.
@@
 
-\begin{lemma}\label{lem.props-nst-n}
+\begin{lemma}
+  \label{lem.props-nst-n}
+  %%
   If~$\<p_1,n,e>\nst[n]\<p_1',n,e'>$ then, for any~$p_2$:
@@
 
-The syntactic restrictions discussed in Section~\ref{?}, regarding the body
-of~$\ceu{\Fin}$ and~$\ceu{\Loop}$ expressions, are formalized by the
-following assumptions.
-
-\begin{assumption}\label{ass.term-nst-fin}
-  If~$\delta=\<\ceu{\clear(p)},n,\nil>$ then there is a
-  description~$\delta'=\<\ceu{\Nop},n,\nil>$ such
-  that~$\delta\nst[*]\delta'$.
+The syntactic restrictions discussed in Section~\ref{sec.sem}, regarding the
+body of~$\ceu{\Fin}$ and~$\ceu{\Loop}$ expressions, are formalized by the
+next assumption.
+
+\begin{assumption}[Syntactical restrictions]\strut
+  \label{ass.syn-rest}
+  %%
+  \begin{enumerate}[(a)]
+  \item\label{ass.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contain no
+    occurrences of~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$,
+    or~$\ceu{\Fin}$ expressions.  Consequently, for any~$n$,
+    \[
+      \<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>\,.
+    \]
+    %%
+  \item\label{ass.syn-rest.loop} If~$p=\ceu{\Loop{p_1}}$ then all execution
+    paths of~$p_1$ contain at least one occurrence of a
+    matching~$\ceu{\Break}$ or an~$\ceu{\AwaitExt}$ expression.
+    Consequently, for all~$n$, there are~$p_1'$ and~$e$ such that
+    \[
+      \<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>\,,
+    \]
+    where~$p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$.
+  \end{enumerate}
 \end{assumption}
 
-\begin{assumption}\label{ass.term-nst-loop}
-  If~$\delta=\<\ceu{\Loop{p}},n,\nil>$ then there is a
-  description~$\delta'=\<p',n,e>$ such that~$\delta\nst[*]\delta'$ where
-  either~$p'=\ceu{{\Break}\AtLoop{p}}$ or~$\isblockedext(p',n)$.
-\end{assumption}
-
-\begin{theorem}\label{thm.term-nst-*}
+\begin{theorem}
+  \label{thm.term-nst-*}
+  %%
   For any~$\delta$ there is a~$\delta'_\Hnst$ such
@@
     %%
-    Then, by Assumption~\ref{ass.term-nst-loop},
+    Then, by item~\eqref{ass.syn-rest.loop} of
+    Assumption~\ref{ass.syn-rest},
     \begin{equation}\label{thm.term-nst-*.loop-expd.eq1}
@@
     for some~$e$ and~$p_1'$ such that either~$p_1'=\ceu{\Break\AtLoop{p'}}$
-    or~$\isblockedext(p_1',n)$.
+    or~$\isblocked(p_1',n)$.
     \begin{subcase}
@@
     \begin{subcase}
-      $\isblockedext(p_1',n)$ is true.  Then, by definition,
-      $\isblockedext(p_1',n)$ implies~$\isblocked(p_1',n)$.  Hence,
+      $\isblocked(p_1',n)$ is true.  Hence,
       from~\eqref{thm.term-nst-*.loop-expd.eq1} and by the definition
@@
 
-  \begin{case}\label{thm.term-nst-*.and}
+  \begin{case}
+    \label{thm.term-nst-*.and}
+    %%
     $p=\ceu{{p'}\AtAnd{p''}}$.
@@
       \begin{subsubcase}
-        \label{thm.term-nst-*.and.nop1}
+        \label{thm.term-nst-*.and-nop1}
         $p'=\ceu{\Nop}$.
@@
       \begin{subsubcase}
-        \label{thm.term-nst-*.and.brk1}
+        \label{thm.term-nst-*.and-brk1}
         $p'=\ceu{\Break}$.
@@
         Then, by~\R{and-brk1},
-        \begin{equation}
-          \label{thm.term-nst-*.and.brk1.eq1}
-          \<\ceu{{\Break}\AtAnd{p''}},n,\nil>
-          \nst[1]\<\ceu{\clear(p'');\,\Break},n,\nil>\,.
-        \end{equation}
-        From~\eqref{thm.term-nst-*.and.brk1.eq1}, by
-        Assumption~\ref{ass.term-nst-fin} and~\R{seq-nop},
+        \begin{align}
+          \label{thm.term-nst-*.and-brk1.eq1}
+          &\<\ceu{{\Break}\AtAnd{p''}},n,\nil>\\
+          &\qquad\nst[1]\<\ceu{\clear(p'');\,\Break},n,\nil>\,.\notag
+        \end{align}
+        From~\eqref{thm.term-nst-*.and-brk1.eq1}, by
+        item~\eqref{ass.syn-rest.fin} of Assumption~\ref{ass.syn-rest} and
+        by~\R{seq-nop},
         \begin{align*}
@@
       \begin{subsubcase}
-        \label{thm.term-nst-*.and.adv1}
+        \label{thm.term-nst-*.and-adv1}
         $p'\ne\ceu{\Nop},\ceu{\Break}$.
@@
         \item If~$p_1'=\ceu{\Nop}$, this case is similar to
-          Case~\ref{thm.term-nst-*.and.nop1}.
+          Case~\ref{thm.term-nst-*.and-nop1}.
         \item If~$p_1'=\ceu{\Break}$, this case is similar to
-          Case~\ref{thm.term-nst-*.and.brk1}.
+          Case~\ref{thm.term-nst-*.and-brk1}.
         \item If~$\isblocked(p_1',n)$, this case is similar to
@@
       \begin{subsubcase}
-        \label{thm.term-nst-*.and.nop2}
+        \label{thm.term-nst-*.and-nop2}
         $p''=\ceu{\Nop}$.
@@
       \begin{subsubcase}
-        \label{thm.term-nst-*.and.brk2}
+        \label{thm.term-nst-*.and-brk2}
         $p''=\ceu{\Break}$.
@@
         From this point on, this case is similar to
-        Case~\ref{thm.term-nst-*.and.brk1}.
-      \end{subsubcase}
-      \begin{subsubcase}
-        \label{thm.term-nst-*.and.adv2}
+        Case~\ref{thm.term-nst-*.and-brk1}.
+      \end{subsubcase}
+      \begin{subsubcase}
+        \label{thm.term-nst-*.and-adv2}
         $p''\ne\ceu{\Nop},\ceu{\Break}$.
@@
         \item If~$p_1''=\ceu{\Nop}$, this case is similar to
-          Case~\ref{thm.term-nst-*.and.nop2}.
+          Case~\ref{thm.term-nst-*.and-nop2}.
         \item If~$p_1''=\ceu{\Break}$, this case is similar to
-          Case~\ref{thm.term-nst-*.and.brk2}.
+          Case~\ref{thm.term-nst-*.and-brk2}.
         \item If~$\isblocked(p_1'',n)$ then, as both sides are blocked, by
@@
       \begin{subsubcase}
-        \label{thm.term-nst-*.or.nop1}
+        \label{thm.term-nst-*.or-nop1}
         $p'=\ceu{\Nop}$.  Then, by~\R{or-nop1},
         \begin{equation}
-          \label{thm.term-nst-*.or.nop1.eq1}
+          \label{thm.term-nst-*.or-nop1.eq1}
           \<\ceu{{\Nop}\AtOr{p''}},n,\nil>
@@
         \end{equation}
-        From~\eqref{thm.term-nst-*.or.nop1.eq1}, by
-        Assumption~\ref{ass.term-nst-fin},
+        From~\eqref{thm.term-nst-*.or-nop1.eq1}, by
+        item~\eqref{ass.syn-rest.fin} Assumption~\ref{ass.syn-rest},
         \[
@@
       \begin{subsubcase}
-        \label{thm.term-nst-*.or.brk1}
+        \label{thm.term-nst-*.or-brk1}
         $p'=\ceu{\Break}$.
         %%
-        Similar to Case~\ref{thm.term-nst-*.and.brk1}.
+        Similar to Case~\ref{thm.term-nst-*.and-brk1}.
       \end{subsubcase}
@@
         %%
-        Similar to Case~\ref{thm.term-nst-*.and.adv1}.
-      \end{subsubcase}
-    \end{subcase}
-    \begin{subcase}
-      \label{thm.term-nst-*.or.adv1}
+        Similar to Case~\ref{thm.term-nst-*.and-adv1}.
+      \end{subsubcase}
+    \end{subcase}
+    \begin{subcase}
+      \label{thm.term-nst-*.or-adv1}
       $\isblocked(p',n)$~is true.
@@
         \begin{equation}
-          \label{thm.term-nst-*.or.nop2.eq1}
+          \label{thm.term-nst-*.or-nop2.eq1}
           \<\ceu{p'\AtOr{\Nop}},n,\nil>
@@
         \end{equation}
-        From~\eqref{thm.term-nst-*.or.nop2.eq1}, by
-        Assumption~\ref{ass.term-nst-fin},
+        From~\eqref{thm.term-nst-*.or-nop2.eq1}, by
+        item~\eqref{ass.syn-rest.fin} of Assumption~\ref{ass.syn-rest} and
+        by definition of~$\clear$,
         \[
@@
         %%
-        Similar to Case~\ref{thm.term-nst-*.and.brk2}.
+        Similar to Case~\ref{thm.term-nst-*.and-brk2}.
       \end{subsubcase}
@@
         %%
-        Similar to Case~\ref{thm.term-nst-*.and.adv2}.\qedhere
+        Similar to Case~\ref{thm.term-nst-*.and-adv2}.\qedhere
       \end{subsubcase}
@@
   where~$\pot'$ is an auxiliary function that counts the number of
-  reachable~$\ceu{\EmitInt}$ in the program resulting from the broadcast of
-  event~$e$ to~$p$.
+  reachable~$\ceu{\EmitInt}$ expressions in the program resulting from the
+  broadcast of event~$e$ to~$p$.
 
@@
   \begin{enumerate}[(a)]
-    \item$\pot'(\ceu{\EmitInt}(e))=1$.
+  \item\label{def.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$;
+    %%
   \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})
-      =\max\{\pot'(p_1),\pot'(p_2)\}$.
-    \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$.
+    =\max\{\pot'(p_1),\pot'(p_2)\}$;
+    %%
+  \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$;
+    %%
+  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(p_1)+\pot'(p_2)$;
+    %%
+  \item$\pot'(\ceu{{p_1}\Or{p_2}})=\pot'(p_1)+\pot'(p_2)$;
+    %%
   \item If~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
@@
       \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\
-        %%
       \pot'(\ceu{p_1\AtLoop{p_2}})&=
       \begin{cases}
-          \pot'(p_1)              &\text{if~(\dag)}\\
+        \pot'(p_1)              &\text{if\enspace(\dag)}\\
         \pot'(p_1)+\pot'(p_2)   &\text{otherwise}\\
-        \end{cases}\\
-        %%
-        \pot'(\ceu{{p_1}\And{p_2}})&=\pot'(p_1)+\pot'(p_2)\\
-        %%
-        \pot'(\ceu{{p_1}\AtAnd{p_2}})&=\pot'(p_1)+\pot'(p_2)\\
-        %%
-        \pot'(\ceu{{p_1}\Or{p_2}})&=\pot'(p_1)+\pot'(p_2)\\
-        %%
-        \pot'(\ceu{{p_1}\AtOr{p_2}})&=\pot'(p_1)+\pot'(p_2)\,,
+      \end{cases}
     \end{align*}
-      where~(\dag) stands for: ``A~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$
-      occurs in all execution paths of~$p_1$.''
-      \item Otherwise, if none of~(a)--(d) applies, $\pot(\ast)=0$.
+    where~(\dag) stands for: ``a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ occurs
+    in all execution paths of~$p_1$'';
+    %%
+  \item If~$p_1,p_2\ne\ceu{\Break}$,
+    \[
+      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
+    \]
+    %%
+  \item\label{def.pot.last} If~$p_1,p_2\ne\ceu{\Break}$
+    and~$p_1,p_2\ne\ceu{\Nop}$,
+    \[
+      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
+    \]
+    %%
+  \item Otherwise, if none of~\eqref{def.pot.first}--\eqref{def.pot.last}
+    applies, $\pot(\_)=0$.
   \end{enumerate}
 
-  TODO: Define~$\pot'(\ceu{\Every{e}\,{p_1}})$ above.
+  \gl{Should~$\pot'(\ceu{\protect\Every{e}\,{p_1}})\ne0$ above?}
 \end{definition}
 
-\begin{definition}\label{def.rank}
+\begin{definition}
+  \label{def.rank}
+  %%
   The \emph{rank} of a description~$\delta=\<p,n,e>$,
@@
       \[
-        \rank(\delta)=\<0,n>=\rank(\delta')\,.
+        \rank(\delta)=\rank(\delta')=\<0,n>\,.
       \]
@@
     %%
-    TODO: Use condition~(\dag) in definition of~$\pot'$.
+    By item~\eqref{ass.syn-rest.loop} of Assumption~\ref{ass.syn-rest}, all
+    execution paths of~$p_1$ contain at least one occurrence
+    of~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$.  Thus, by condition~(\dag) in
+    Definition~\ref{def.pot},
+    \[
+      \rank(\delta)=\rank(\delta')=\<\pot'(p_1),n>\,.
+    \]
   \end{case}
@@
       %%
-      Similar to Case~\ref{lem.rank-nst.seq-adv}.
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    \label{lem.rank-nst.and.expd}
+      Then~$\pi$ is an instance of~\R{loop-adv}.  Hence there is a
+      derivation~$\pi'$ such that
+      \[
+        \pi'\Vdash\<p_1,n,\nil>\nst\<p_1',n,e_1'>\,,
+      \]
+      for some~$p_1'$ and~$e_1'$.  Thus~$p'=p_1';\,p_2$ and~$e'=e_1'$.
+      %%
+      % By the induction hypothesis,
+      % \begin{equation}
+      %   \label{lem.rank-nst.loop-adv-eq1}
+      %   \rank(\<p_1,n,\nil>)\ge\rank(\<p_1',n,e_1'>)\,.
+      % \end{equation}
+      %%
+      There are two subcases.
+      \begin{subsubcase}
+        $\pot'(p)=\pot'(p_1)$.
+        %%
+        Then all execution paths of~$p_1$ contain a~$\ceu{\Break}$
+        or~$\ceu{\AwaitExt}$ expression.  A single~$\out$ cannot terminate
+        the loop, since~$p_1\ne\ceu{\Break}$, nor can it consume
+        an~$\ceu{\AwaitExt}$, which means that all execution paths in~$p_1'$
+        still contain a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$.
+        Hence~$\pot'(p')=\pot'(p_1')$.  The rest of this proof is similar to
+        that of Case~\ref{lem.rank-nst.seq-adv}.
+      \end{subsubcase}
+      \begin{subsubcase}
+        $\pot'(p)=\pot'(p_1)+\pot'(p_2)$.
+        %%
+        Then some execution path in~$p_1$ does not contain a~$\ceu{\Break}$
+        or~$\ceu{\AwaitExt}$ expression.  Since~$p_1\ne\ceu{\Nop}$, a
+        single~$\out$ cannot restart the loop, which means that~$p_1'$ still
+        contain some execution path in which a~$\ceu{\Break}$
+        or~$\ceu{\AwaitExt}$ does not occur.
+        Hence~$\pot'(p')=\pot'(p_1')+\pot'(p_2)$.  The rest of this proof is
+        similar to that of Case~\ref{lem.rank-nst.seq-adv}.
+      \end{subsubcase}
+    \end{subcase}
+  \end{case}
+
+  \begin{case}
+    \label{lem.rank-nst.and-expd}
     $p=\ceu{{p_1}\And{p_2}}$.
@@
     Hence~$p'=\ceu{{p_1}\AtAnd(\CanRun(n);\,p_2)}$ and~$e'=\nil$.
-  \end{case}
-
-  \begin{case}
+    Thus
+    \[
+      \rank(\delta)=\rank(\delta')=\<\pot'(p_1)+\pot'(p_2),n>\,.
+    \]
+  \end{case}
+
+  \begin{case}
+    \label{lem.rank-nst.and}
     $p=\ceu{{p_1}\AtAnd{p_2}}$.
+    %%
+    There are two subcases.
+    \begin{subcase}
+      \label{lem.rank-nst.and1}
+      $\isblocked(p_1,n)$~is false.
+      %%
+      There are three subcases.
+      \begin{subsubcase}
+        \label{lem.rank-nst.and-nop1}
+        $p_1=\ceu{\Nop}$.
+        %%
+        Then~$\pi$ is an instance of~\R{and-nop1}.  Hence~$p'=p_2$
+        and~$e'=\nil$.  Thus
+        \[
+          \rank(\delta)=\rank(\delta')=\<0+\pot'(p_2),n>\,.
+        \]
+      \end{subsubcase}
+      \begin{subsubcase}
+        \label{lem.rank-nst.and-brk1}
+        $p_1=\ceu{\Break}$.
+        %%
+        Then~$\pi$ is an instance of~\R{and-brk1}.
+        Hence~$p'=\ceu{\clear(p_2);\,\Break}$ and~$e'=\nil$.  By item
+        \eqref{ass.syn-rest.fin} of Assumption~\ref{ass.syn-rest} and by the
+        definition of~$\clear$, $\clear(p_2)$ does not
+        contain~$\ceu{\EmitInt}$ expressions.  Thus
+        \[
+          \rank(\delta)=\rank(\delta')=\<0,n>\,.
+        \]
+      \end{subsubcase}
+      \begin{subsubcase}
+        \label{lem.rank-nst.and-adv1}
+        $p_1\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Then~$\pi$ is an instance of~\R{and-adv1}.  As~$p_1\ne\ceu{\Break}$
+        and~$p_2\ne\ceu{\Break}$ (otherwise~\R{and-brk2} would have taken
+        precedence), the rest of this proof is similar to that of
+        Case~\ref{lem.rank-nst.seq-adv}.
+      \end{subsubcase}
+    \end{subcase}
+    \begin{subcase}
+      $\isblocked(p_1,n)$~is true.
+      %%
+      Similar to Case~\ref{lem.rank-nst.and1}
+    \end{subcase}
   \end{case}
@@
     $p=\ceu{{p_1}\Or{p_2}}$.
+    %%
+    Then~$\pi$ is an instance of~\R{or-expd}.
+    Hence~$p'=\ceu{{p_1}\AtOr(\CanRun(n);\,p_2)}$ and~$e'=\nil$.
+    Thus
+    \[
+      \rank(\delta)=\rank(\delta')=\<\pot'(p_1)+\pot'(p_2),n>\,.
+    \]
   \end{case}
@@
     $p=\ceu{{p_1}\AtOr{p_2}}$.
+    %%
+    There are two subcases.
+    \begin{subcase}
+      \label{lem.rank-nst.or1}
+      $\isblocked(p_1,n)$ is false.
+      %%
+      There are three subcases.
+      \begin{subsubcase}
+        $p_1=\ceu{\Nop}$.
+        %%
+        Then~$\pi$ is an instance of~\R{or-nop1}.  Hence~$p'=\clear(p_2)$
+        and~$e'=\nil$.  By item \eqref{ass.syn-rest.fin} of
+        Assumption~\ref{ass.syn-rest} and by the definition of~$\clear$,
+        $p'$ does not contain~$\ceu{\EmitInt}$ expressions.  Thus
+        \[
+          \rank(\delta)=\rank(\delta')=\<0,n>\,.
+        \]
+      \end{subsubcase}
+      \begin{subsubcase}
+        $p_1=\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{lem.rank-nst.and-brk1}.
+      \end{subsubcase}
+      \begin{subsubcase}
+        $p_1\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{lem.rank-nst.and-adv1}.
+      \end{subsubcase}
+    \end{subcase}
+    \begin{subcase}
+      $\isblocked(p_1,n)$ is true.
+      %%
+      Similar to Case~\ref{lem.rank-nst.or1}.\qedhere
+    \end{subcase}
   \end{case}
@@
 
-\begin{theorem}[Reaction termination]
-  For any~$\delta$ there is a~$\delta'_\Hnst$ such
-  that~$\delta\trans[*]\delta'$ and~$\rank(\delta')=\<0,0>$.
-\end{theorem}
-\begin{proof}
-  By lexicographic induction on~$rank(\delta)$.
-\end{proof}
+% \begin{theorem}[Reaction termination]
+%   For any~$\delta$ there is a~$\delta'_\Hnst$ such
+%   that~$\delta\trans[*]\delta'$ and~$\rank(\delta')=\<0,0>$.
+% \end{theorem}
+% \begin{proof}
+%   By lexicographic induction on~$rank(\delta)$.
+% \end{proof}
 %   \begin{basis}

### 2018-02-23 06-appendix.tex
Teorema da terminação geral

@@
   iff~$e\ne\nil$ or~$p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p,n)$~is
-  true.  Nested-irreducible descriptions serve as normal forms for~$\nst$
+  true.
+
+  Nested-irreducible descriptions serve as normal forms for~$\nst$
   transitions: they embody the result of an exhaustive number of~$\nst$
@@
 \begin{lemma}\strut
+  \label{lem.rank-out}
   \begin{enumerate}[(a)]
-  \item If~$\delta\outpush\delta'$ then~$\rank(\delta)=\rank(\delta')$.
-  \item If~$\delta\outpop\delta'$ then~$\rank(\delta)>\rank(\delta')$.
+  \item\label{lem.rank-out-push} If~$\delta\outpush\delta'$
+    then~$\rank(\delta)=\rank(\delta')$.
+  \item\label{lem.rank-out-pop} If~$\delta\outpop\delta'$
+    then~$\rank(\delta)>\rank(\delta')$.
   \end{enumerate}
@@
 
-% \begin{theorem}[Reaction termination]
-%   For any~$\delta$ there is a~$\delta'_\Hnst$ such
-%   that~$\delta\trans[*]\delta'$ and~$\rank(\delta')=\<0,0>$.
-% \end{theorem}
-% \begin{proof}
-%   By lexicographic induction on~$rank(\delta)$.
-% \end{proof}
-%   \begin{basis}
-%     If~$\<i,j>=\<0,0>$ then, by~\ref{?}, $\delta$ cannot be advanced
-%     by~$\out$ transitions.  There are two possibilities\dots\ (it can or
-%     cannot be advanced by nst transitions).
-%   \end{basis}
-%   \begin{induction}
-%     Let~$\<i,j>\ne\<0,0>$ and suppose that the theorem holds for
-%     all~$\delta''$ such that~$\rank(\delta'')<\rank(\delta)$.
-%     \begin{case}
-%       $\delta$~is nested-irreducible.
-%       %%
-%       \begin{subcase}
-%         $\delta\outpush\delta_1$, for some~$\delta_1$.
-%       \end{subcase}
-%       \begin{subcase}
-%         $\delta\outpop\delta_1$, for some~$\delta_1$.
-%         %%
-%         Then, by definition of~\R{pop}, $\rank(\delta_1)=\<i,j-1>$, and by
-%         the induction hypothesis\dots
-%       \end{subcase}
-%     \end{case}
-%     \begin{case}
-%       $\delta$~is not nested-irreducible.
-%       %%
-%     \end{case}
-%   \end{induction}
+\begin{theorem}
+  \label{thm.rank-nst-*}
+  %%
+  If~$\delta\nst[*]\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
+\end{theorem}
+\begin{proof}
+  If~$\delta\nst[*]\delta'$ then there is an~$n$ such that
+  $\delta\nst[n]\delta'$.  We proceed by induction on~$n$.
+  %%
+  The theorem is trivially true for~$n=0$ and follows directly from
+  Lemma~\ref{lem.rank-nst} for~$n=1$.  Suppose
+  $\delta\nst[1]\delta_1'\nst[n-1]\delta'$, for some~$n>1$ and~$\delta_1'$.
+  Thus, by Lemma~\ref{lem.rank-nst} and by the induction hypothesis,
+  \[
+    \rank(\delta)\ge\rank(\delta_1')\ge\rank(\delta')\,.\qedhere
+  \]
+\end{proof}
+
+\begin{definition}
+  \label{def.H}
+  %%
+  A description~$\delta$ is \emph{irreducible} (in symbols, $\delta_\#$) iff
+  it is nested-irreducible and its~$\rank(\delta)$ is~$\<i,0>$, for
+  some~$i\ge0$.
+
+  An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form
+  for transitions in general.  It cannot be advanced by~$\nst$, as it is
+  nested-irreducible, and neither by~$\outpush$ nor~$\outpop$, as the second
+  coordinate of its rank is~0, which implies~$e=\nil$ and~$n=0$.
+\end{definition}
+
+\begin{theorem}
+  \label{thm.term}
+  %%
+  For any~$\delta$, there is a~$\delta'_\#$ such
+  that~$\delta\trans[*]\delta'_\#$.
+\end{theorem}
+\begin{proof}
+  By lexicographic induction on~$rank(\delta)$.  Let~$\delta=\<p,n,e>$ and
+  $\rank(\delta)=\<i,j>$.
+  %%
+  \begin{basis}
+    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$
+    transitions, as~$j=0$ implies~$e=\nil$ and~$n=0$.  Thus neither~\R{push}
+    nor~\R{pop} can be applied.  There are two possibilities:
+    either~$\delta$ is nested-irreducible or it is not.  In the first case,
+    the theorem is trivially true, as~$\delta\nst[0]\delta_\Hnst$.
+
+    Suppose~$\delta$ is not nested-irreducible.  Then, by
+    Theorem~\ref{thm.term-nst-*}, $\delta\nst[*]\delta'_\Hnst$ for
+    some~$\delta'_\Hnst$.  By Theorem~\ref{thm.rank-nst-*},
+    \[
+      \<i,j>=\<0,0>\ge\rank(\delta')\,,
+    \]
+    which implies~$\rank(\delta')=\<0,0>$.
+  \end{basis}
+  \begin{induction}
+    Let~$\<i,j>\ne\<0,0>$.
+    %%
+    There are two subcases.
+    \begin{case}
+      \label{thm.term.Hnst}
+      $\delta$~is nested-irreducible.
+      %%
+      There are two cases.
+      \begin{subcase}
+        \label{thm.term.Hnst-j0}
+        $j=0$.  By Definition~\ref{def.H}, $\delta_\#$.
+        Thus~$\delta\trans[0]\delta_\#$.
+      \end{subcase}
+      \begin{subcase}
+        \label{thm.term.Hnst-j>0}
+        $j>0$.
+        %%
+        There are two subcases.
+        \begin{subsubcase}
+          \label{thm.term.Hnst-j>0-nonnil}
+          $e\ne\nil$.
+          %%
+          Then, by~\R{out} and by Theorem~\ref{thm.term-nst-*}, there
+          are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
+          \[
+            \delta\outpush\delta_1'\nst[*]\delta'_\Hnst\,.
+          \]
+          Thus, by Lemma by item~\eqref{lem.rank-out-push} of
+          Lemma~\ref{lem.rank-out} and by Theorem~\ref{thm.rank-nst-*},
+          \begin{align*}
+            \rank(\delta)=\rank(\delta_1')&=\<i,j>\\
+                                          &\ge\rank(\delta')=\<i',j'>\,.
+          \end{align*}
+          If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof
+          is similar to that of Case~\ref{thm.term.Hnst-j>0-nil}.
+          Otherwise, if~$e'\ne\nil$ then $i>i'$, since
+          an~$\ceu{\EmitInt(e')}$ was consumed by the nested transitions.
+          Thus,
+          \[
+            \rank(\delta)>\rank(\delta')\,.
+          \]
+          By the induction hypothesis, $\delta'\trans[*]\delta''_\#$, for
+          some~$\delta''_\#$.  Thus~$\delta\trans[*]\delta''_\#$.
+        \end{subsubcase}
+        \begin{subsubcase}
+          \label{thm.term.Hnst-j>0-nil}
+          $e=\nil$.
+          %%
+          Then, since~$j>0$, $\delta\outpop\delta'$, for
+          some~$\delta''$.  By item~\eqref{lem.rank-out-pop} of
+          Lemma~\ref{lem.rank-out},
+          \[
+            \rank(\delta)>\rank(\delta')\,.
+          \]
+          Hence, by the induction hypothesis, there is a~$\delta''_\#$ such
+          that~$\delta'\trans[*]\delta''_\#$.
+          Thus~$\delta\trans[*]\delta''_\#$.
+        \end{subsubcase}
+      \end{subcase}
+    \end{case}
+    \begin{case}
+      $\delta$~is not nested-irreducible.
+      %%
+      There are two subcases.
+      \begin{subcase}
+        $e\ne\nil$.  Similar to Case~\ref{thm.term.Hnst-j>0-nonnil}.
+      \end{subcase}
+      \begin{subcase}
+        $e=\nil$.  Then, by Theorems~\ref{thm.rank-nst-*}
+        and~\ref{thm.rank-nst-*} there is a~$\delta'_\Hnst$ such
+        that~$\delta\nst[*]\delta'_\Hnst$
+        with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
+        is similar to that of Case~\ref{thm.term.Hnst}.\qedhere
+      \end{subcase}
+    \end{case}
+  \end{induction}
+\end{proof}

### 2018-02-23 06-appendix.tex
Conserta confusão "n" pilha/passos no apêndice

@@
+\clearpage
 \appendix
@@
 
-\begin{theorem}
-  \label{thm.det-out-pop-n}
-  %%
-  If~$\delta\out[n]\delta_1$ and~$\delta\out[n]\delta_2$
-  then~$\delta_1=\delta_2$.
-\end{theorem}
-\begin{proof}
-  By induction on~$n$.
-  %%
-  The theorem is trivially true for~$n=0$ and follows directly from
-  Lemma~\ref{lem.det-out} for~$n=1$.  (The theorem is vacuously true
-  for~$\outpush$ transitions if~$n>1$ since, by definition,
-  $\outpush$~transitions cannot be applied more than once in a row and
-  cannot occur after a~$\outpop$ transition.)
-  %%
-  Suppose~$\delta\out[1]\delta_1'\out[n-1]\delta_1$
-  and~$\delta\out[1]\delta_2'\out[n-1]\delta_2$, for some~$n>1$
-  and~$\delta_1'$ and~$\delta_2'$.
-  %%
-  By Lemma~\ref{lem.det-out}, $\delta_1'=\delta_2'$.  By the induction
-  hypothesis, $\delta_1=\delta_2$.\qedhere
-\end{proof}
+% \begin{theorem}
+%   \label{thm.det-out-i}
+%   %%
+%   If~$\delta\out[i]\delta_1$ and~$\delta\out[i]\delta_2$
+%   then~$\delta_1=\delta_2$.
+% \end{theorem}
+% \begin{proof}
+%   By induction on~$i$.
+%   %%
+%   The theorem is trivially true if~$i=0$ and follows directly from
+%   Lemma~\ref{lem.det-out} if~$i=1$.  (The theorem is vacuously true
+%   for~$\outpush$ transitions if~$i>1$ since they cannot be applied more
+%   than once in a row and cannot occur after a~$\outpop$ transition.)
+%   %%
+%   Suppose
+%   \[
+%     \delta\out[1]\delta_1'\out[i-1]\delta_1
+%     \quad\text{and}\quad
+%     \delta\out[1]\delta_2'\out[i-1]\delta_2\,,
+%   \]
+%   for some~$i>1$, $\delta_1'$ and~$\delta_2'$.
+%   %%
+%   By Lemma~\ref{lem.det-out}, $\delta_1'=\delta_2'$.  By the induction
+%   hypothesis, $\delta_1=\delta_2$.\qedhere
+% \end{proof}
 
@@
 
-\begin{theorem}
-  \label{thm.det-nst-n}
-  %%
-  If~$\delta\nst[n]\delta_1$ and~$\delta\nst[n]\delta_2$
+% \begin{theorem}
+%   \label{thm.det-nst-i}
+%   %%
+%   If~$\delta\nst[i]\delta_1$ and~$\delta\nst[i]\delta_2$
+%   then~$\delta_1=\delta_2$.
+% \end{theorem}
+% \begin{proof}
+%   By induction on~$i$.  Similar to the proof of
+%   Theorem~\ref{thm.det-out-pop-i}.
+% \end{proof}
+
+The next theorem establishes that transitions are deterministic, i.e., that
+the transition relation~$\trans$ between descriptions is a function.
+
+\begin{theorem}[Determinism]\strut
+  \label{thm.det}
+  %%
+  If~$\delta\trans[i]\delta_1$ and~$\delta\trans[i]\delta_2$
   then~$\delta_1=\delta_2$.
@@
 \begin{proof}
-  Similar to the proof of Theorem~\ref{thm.det-out-pop-n}.
+  By induction on~$i$.  The theorem is trivially true if~$i=0$ and follows
+  directly from Lemmas~\ref{lem.det-out} and~\ref{lem.det-nst} for~$i=1$.
+  %%
+  Suppose
+  \[
+    \delta\trans[1]\delta_1'\trans[i-1]\delta_1
+    \quad\text{and}\quad
+    \delta\trans[1]\delta_2'\trans[i-1]\delta_2\,,
+  \]
+  for some~$i>1$, $\delta_1'$ and~$\delta_2'$.
+  There are two possibilities.
+  \begin{case}
+    $\delta\out[1]\delta_1'$ and~$\delta\out[1]\delta_2'$.  Then, by
+    Lemma~\ref{lem.det-out}, $\delta_1'=\delta_2'$, and by the induction
+    hypothesis, $\delta_1=\delta_2$.
+  \end{case}
+  \begin{case}
+    $\delta\nst[1]\delta_1'$ and~$\delta\nst[1]\delta_2'$.  Then, by
+    Lemma~\ref{lem.det-nst}, $\delta_1'=\delta_2'$, and by the induction
+    hypothesis, $\delta_1=\delta_2$.\qedhere
+  \end{case}
 \end{proof}
@@
   true.
+\end{definition}
 
@@
 description~$\delta$ is nested-irreducible.
-\end{definition}
 
@@
 \begin{lemma}
-  \label{lem.irr-nst-n}
-  %%
-  If~$\delta\nst[n]\delta_\Hnst'$ then, for all~$i\ne{n}$, there is
-  no~$\delta_\Hnst''$ such that~$\delta\nst[i]\delta''_\Hnst$.
+  \label{lem.irr-nst-i}
+  %%
+  If~$\delta\nst[i]\delta_\Hnst'$ then, for all~$k\ne{i}$, there is
+  no~$\delta_\Hnst''$ such that~$\delta\nst[k]\delta''_\Hnst$.
 \end{lemma}
 \begin{proof}
-  By contradiction on the hypothesis that there is such~$i$.
-  %%
-  Let~$\delta\nst[n]\delta'_\Hnst$, for some~$n\ge0$.
+  By contradiction on the hypothesis that there is such~$k$.
+  %%
+  Let~$\delta\nst[i]\delta'_\Hnst$, for some~$i\ge0$.
   %%
@@
   \begin{case}
-    \label{lem.irr-nst-n-case1}
-    %%
-    Suppose there are~$i>n$ and~$\delta''_\Hnst$ such
-    that~$\delta\nst[i]\delta''$.
-    %%
-    Then, by definition of~$\nst[i]$,
+    \label{lem.irr-nst-i-case1}
+    %%
+    Suppose there are~$k>i$ and~$\delta''_\Hnst$ such
+    that~$\delta\nst[k]\delta''$.
+    %%
+    Then, by definition of~$\nst[k]$,
     \begin{equation}
-      \label{lem.irr-nst-n-eq1}
-      %%
-      \delta\nst[n]\delta'\nst[n+1]\delta_1'\nst[n+2]\cdots\nst[i]\delta''.
+      \label{lem.irr-nst-i-eq1}
+      %%
+      \delta\nst[i]\delta'\nst[i+1]\delta_1'\nst[i+2]\cdots\nst[k]\delta''.
     \end{equation}
@@
     that~$\delta'\nst[1]\delta_1'$, which
-    contradicts~\eqref{lem.irr-nst-n-eq1}.  Therefore, no such~$i$ can
+    contradicts~\eqref{lem.irr-nst-i-eq1}.  Therefore, no such~$k$ can
     exist.
@@
   \begin{case}
-    Suppose there are~$i<n$ and~$\delta''_\Hnst$ such
-    that~$\delta\nst[i]\delta''$.  Then, since~$n>i$, by
-    Case~\ref{lem.irr-nst-n-case1}, $\delta'$~cannot exist, which is absurd.
-    Therefore, the assumption that there is such~$i$ is false.\qedhere
+    Suppose there are~$k<i$ and~$\delta''_\Hnst$ such
+    that~$\delta\nst[k]\delta''$.  Then, since~$i>k$, by
+    Case~\ref{lem.irr-nst-i-case1}, $\delta'$~cannot exist, which is absurd.
+    Therefore, the assumption that there is such~$k$ is false.\qedhere
   \end{case}
@@
 \begin{lemma}
-  \label{lem.props-nst-n}
-  %%
-  If~$\<p_1,n,e>\nst[n]\<p_1',n,e'>$ then, for any~$p_2$:
+  \label{lem.props-nst-i}\strut\\
+  %%
+  If~$\<p_1,n,e>\nst[i]\<p_1',n,e'>$ then, for any~$p_2$:
   \begin{enumerate}[(a)]
-  \item\label{lem.props-nst-n.a}
-    $\<\ceu{p_1;\,p_2},n,e>\nst[n]\<p_1';p_2,n,e'>$;
-    %%
-  \item\label{lem.props-nst-n.b}
-    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[n]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$;
-    %%
-  \item\label{lem.props-nst-n.c}
-    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[n]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$;
-    %%
-  \item\label{lem.props-nst-n.d}
-    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[n]\<\ceu{{p_1}'\AtOr{p_2}},n,e'>$.
+  \item\label{lem.props-nst-i.a}
+    $\<\ceu{p_1;\,p_2},n,e>\nst[i]\<p_1';p_2,n,e'>$;
+    %%
+  \item\label{lem.props-nst-i.b}
+    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[i]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$;
+    %%
+  \item\label{lem.props-nst-i.c}
+    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$;
+    %%
+  \item\label{lem.props-nst-i.d}
+    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}'\AtOr{p_2}},n,e'>$.
   \end{enumerate}
-  If~$\<p_2,n,e>\nst[n]\<p_2',n,e'>$, for any~$p_1$ such
+  If~$\<p_2,n,e>\nst[i]\<p_2',n,e'>$, for any~$p_1$ such
   that~$\isblocked(p_1,n)$:
@@
     \setcounter{enumi}{4}
-  \item\label{lem.props-nst-n.e}
-    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[n]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>$;
-    %%
-  \item\label{lem.props-nst-n.f}
-    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[n]\<\ceu{{p_1}\AtOr{p_2'}},n,e'>$.
+  \item\label{lem.props-nst-i.e}
+    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>$;
+    %%
+  \item\label{lem.props-nst-i.f}
+    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtOr{p_2'}},n,e'>$.
   \end{enumerate}
@@
 \begin{proof}
-  By induction on~$n$.
+  By induction on~$i$.
   %%
   \begin{enumerate}[(a)]
-  \item The lemma is trivially true for~$n=0$, as~$p_1=p_1'$, and follows
-    directly from~\R{seq-adv} for~$n=1$.  Suppose
+  \item The lemma is trivially true for~$i=0$, as~$p_1=p_1'$, and follows
+    directly from~\R{seq-adv} for~$i=1$.  Suppose
     \begin{equation}
-      \label{lem.props-nst-n.a.eq1}
-      \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>\,,
+      \label{lem.props-nst-i.a.eq1}
+      \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[i-1]\<p_1',n,e'>\,,
     \end{equation}
-    for some~$n>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible, i.e.,
+    for some~$i>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible, i.e.,
     $e=\nil$ and~$p\ne{\ceu{\Nop},\ceu{\Break}}$ and~$\isblocked(p_1'',n)$
-    is false.  By~\eqref{lem.props-nst-n.a.eq1} and by~\R{seq-adv},
+    is false.  By~\eqref{lem.props-nst-i.a.eq1} and by~\R{seq-adv},
     \begin{equation}
-      \label{lem.props-nst-n.a.eq2}
+      \label{lem.props-nst-i.a.eq2}
       \<\ceu{p_1;\,p_2},n,e>\nst[1]\<\ceu{p_1'';\,p_2},n,e''>\,.
     \end{equation}
-    From~\eqref{lem.props-nst-n.a.eq1}, by the induction hypothesis,
+    From~\eqref{lem.props-nst-i.a.eq1}, by the induction hypothesis,
     \begin{equation}
-      \label{lem.props-nst-n.a.eq3}
-      \<\ceu{p_1'';\,p_2},n,e''>\nst[n-1]\<\ceu{p_1';\,p_2},n,e'>\,.
+      \label{lem.props-nst-i.a.eq3}
+      \<\ceu{p_1'';\,p_2},n,e''>\nst[i-1]\<\ceu{p_1';\,p_2},n,e'>\,.
     \end{equation}
-    From~\eqref{lem.props-nst-n.a.eq2} and~\eqref{lem.props-nst-n.a.eq3},
-    \[
-      \<\ceu{p_1;\,p_2},n,e>\nst[n]\<\ceu{p_1';\,p_2},n,e'>\,.
-    \]
-
-  \item Similar to Case~\eqref{lem.props-nst-n.a}.
+    From~\eqref{lem.props-nst-i.a.eq2} and~\eqref{lem.props-nst-i.a.eq3},
+    \[
+      \<\ceu{p_1;\,p_2},n,e>\nst[i]\<\ceu{p_1';\,p_2},n,e'>\,.
+    \]
+
+  \item Similar to Case~\eqref{lem.props-nst-i.a}.
     %%
@@
     % \begin{equation}
-    %   \label{lem.props-nst-n.b.eq1}
+    %   \label{lem.props-nst-i.b.eq1}
     %   \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>\,,
@@
     % for some~$n>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible.
-    % By~\eqref{lem.props-nst-n.b.eq1} and by~\R{loop-adv},
+    % By~\eqref{lem.props-nst-i.b.eq1} and by~\R{loop-adv},
     % \begin{equation}
-    %   \label{lem.props-nst-n.b.eq2}
+    %   \label{lem.props-nst-i.b.eq2}
     %   \<\ceu{p_1\AtLoop{p_2}},n,e>\nst[1]\<\ceu{p_1''\Loop{p_2}},n,e''>\,.
     % \end{equation}
-    % From~\eqref{lem.props-nst-n.b.eq1}, by the induction hypothesis,
+    % From~\eqref{lem.props-nst-i.b.eq1}, by the induction hypothesis,
     % \begin{equation}
-    %   \label{lem.props-nst-n.b.eq3}
+    %   \label{lem.props-nst-i.b.eq3}
     %   \<\ceu{p_1''\AtLoop{p_2}},n,e''>
@@
     % \end{equation}
-    % From~\eqref{lem.props-nst-n.b.eq2} and~\eqref{lem.props-nst-n.b.eq3},
+    % From~\eqref{lem.props-nst-i.b.eq2} and~\eqref{lem.props-nst-i.b.eq3},
     % \[
@@
 
-  \item Similar to Case~\eqref{lem.props-nst-n.a}.
+  \item Similar to Case~\eqref{lem.props-nst-i.a}.
     %%
@@
     % \begin{equation}
-    %   \label{lem.props-nst-n.c.eq1}
+    %   \label{lem.props-nst-i.c.eq1}
     %   \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>\,,
@@
     % for some~$n>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible.
-    % By~\eqref{lem.props-nst-n.c.eq1} and by~\R{and-adv1},
+    % By~\eqref{lem.props-nst-i.c.eq1} and by~\R{and-adv1},
     % \begin{equation}
-    %   \label{lem.props-nst-n.c.eq2}
+    %   \label{lem.props-nst-i.c.eq2}
     %   \<\ceu{{p_1}\AtAnd{p_2}},n,e>
@@
     % \end{equation}
-    % From~\eqref{lem.props-nst-n.c.eq1}, by the induction hypothesis,
+    % From~\eqref{lem.props-nst-i.c.eq1}, by the induction hypothesis,
     % \begin{equation}
-    %   \label{lem.props-nst-n.a.eq3}
+    %   \label{lem.props-nst-i.a.eq3}
     %   \<\ceu{{p_1''}\AtAnd{p_2}},n,e''>
@@
     % \end{equation}
-    % From~\eqref{lem.props-nst-n.c.eq2} and~\eqref{lem.props-nst-n.c.eq3},
+    % From~\eqref{lem.props-nst-i.c.eq2} and~\eqref{lem.props-nst-i.c.eq3},
     % \[
@@
     % \begin{equation}
-    %   \label{lem.props-nst-n.d.eq1}
+    %   \label{lem.props-nst-i.d.eq1}
     %   \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>\,,
@@
     % for some~$n>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible.
-    % By~\eqref{lem.props-nst-n.d.eq1} and by~\R{or-adv1},
+    % By~\eqref{lem.props-nst-i.d.eq1} and by~\R{or-adv1},
     % \begin{equation}
-    %   \label{lem.props-nst-n.d.eq2}
+    %   \label{lem.props-nst-i.d.eq2}
     %   \<\ceu{{p_1}\AtOr{p_2}},n,e>
@@
     % \end{equation}
-    % From~\eqref{lem.props-nst-n.d.eq1}, by the induction hypothesis,
+    % From~\eqref{lem.props-nst-i.d.eq1}, by the induction hypothesis,
     % \begin{equation}
-    %   \label{lem.props-nst-n.a.eq3}
+    %   \label{lem.props-nst-i.a.eq3}
     %   \<\ceu{{p_1''}\AtOr{p_2}},n,e''>
@@
     % \end{equation}
-    % From~\eqref{lem.props-nst-n.d.eq2} and~\eqref{lem.props-nst-n.d.eq3},
+    % From~\eqref{lem.props-nst-i.d.eq2} and~\eqref{lem.props-nst-i.d.eq3},
     % \[
@@
 
-  \item The lemma is trivially true for~$n=0$, as~$p_2=p_2'$, and follows
-    directly from~\R{and-adv2} for~$n=1$.  Suppose
+  \item The lemma is trivially true for~$i=0$, as~$p_2=p_2'$, and follows
+    directly from~\R{and-adv2} for~$i=1$.  Suppose
     \begin{equation}
-      \label{lem.props-nst-n.e.eq1}
-      \<p_2,n,e>\nst[1]\<p_2'',n,e''>\nst[n-1]\<p_2',n,e'>\,,
+      \label{lem.props-nst-i.e.eq1}
+      \<p_2,n,e>\nst[1]\<p_2'',n,e''>\nst[i-1]\<p_2',n,e'>\,,
     \end{equation}
-    for some~$n>1$.  Then~$\<p_2'',n,e''>$ is not nested-irreducible.
-    By~\eqref{lem.props-nst-n.e.eq1} and by~\R{and-adv2},
+    for some~$i>1$.  Then~$\<p_2'',n,e''>$ is not nested-irreducible.
+    By~\eqref{lem.props-nst-i.e.eq1} and by~\R{and-adv2},
     \begin{equation}
-      \label{lem.props-nst-n.e.eq2}
+      \label{lem.props-nst-i.e.eq2}
       \<\ceu{{p_1}\AtAnd{p_2}},n,e>
@@
     \end{equation}
-    From~\eqref{lem.props-nst-n.e.eq1}, by the induction hypothesis,
+    From~\eqref{lem.props-nst-i.e.eq1}, by the induction hypothesis,
     \begin{equation}
-      \label{lem.props-nst-n.e.eq3}
+      \label{lem.props-nst-i.e.eq3}
       \<\ceu{{p_1}\AtAnd{p_2''}},n,e''>
-      \nst[n-1]\<\ceu{{p_1}\AtOr{p_2'}},n,e'>\,.
+      \nst[i-1]\<\ceu{{p_1}\AtOr{p_2'}},n,e'>\,.
     \end{equation}
-    From~\eqref{lem.props-nst-n.e.eq2} and~\eqref{lem.props-nst-n.e.eq3},
+    From~\eqref{lem.props-nst-i.e.eq2} and~\eqref{lem.props-nst-i.e.eq3},
     \[
       \<\ceu{{p_1}\AtAnd{p_2}},n,e>
-      \nst[n]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>\,.
-    \]
-
-  \item Similar to Case~\eqref{lem.props-nst-n.e}.\qedhere
+      \nst[i]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>\,.
+    \]
+
+  \item Similar to Case~\eqref{lem.props-nst-i.e}.\qedhere
   \end{enumerate}
@@
     \]
-    where~$p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$.
+    where~$p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
   \end{enumerate}
@@
     \begin{subcase}
+      \label{thm.term-nst-*.if-true}
       $\eval(\ceu{\Mem(\Id)})$~is true.
@@
       %%
-      Then, by~\R{if-false} and by the induction hypothesis, there is
-      a~$\delta'$ such that
-      \begin{align*}
-        \<\ceu{\IfElse{\Mem(\Id)}{p'}{p''}},n,\nil>
-        &\nst[1]\<p'',n,e>\\
-        &\nst[*]\delta'_\Hnst\,.
-      \end{align*}
+      Similar to Case~\ref{thm.term-nst-*.if-true}.
+      %%
+      % Then, by~\R{if-false} and by the induction hypothesis, there is
+      % a~$\delta'$ such that
+      % \begin{align*}
+      %   \<\ceu{\IfElse{\Mem(\Id)}{p'}{p''}},n,\nil>
+      %   &\nst[1]\<p'',n,e>\\
+      %   &\nst[*]\delta'_\Hnst\,.
+      % \end{align*}
+      %%
     \end{subcase}
@@
       \]
-      By item~\eqref{lem.props-nst-n.a} of Lemma~\ref{lem.props-nst-n},
+      By item~\eqref{lem.props-nst-i.a} of Lemma~\ref{lem.props-nst-i},
       \begin{equation}
@@
         Case~\ref{thm.term-nst-*.seq-nop}.
+        %%
       \end{subsubcase}
@@
       \]
-      By item~\eqref{lem.props-nst-n.b} of Lemma~\ref{lem.props-nst-n},
+      By item~\eqref{lem.props-nst-i.b} of Lemma~\ref{lem.props-nst-i},
       \[
@@
         \]
-        By item~\eqref{lem.props-nst-n.c} of Lemma~\ref{lem.props-nst-n},
+        By item~\eqref{lem.props-nst-i.c} of Lemma~\ref{lem.props-nst-i},
         \[
@@
         \]
-        By item~\eqref{lem.props-nst-n.e} of Lemma~\ref{lem.props-nst-n},
+        By item~\eqref{lem.props-nst-i.e} of Lemma~\ref{lem.props-nst-i},
         \[
@@
     \[
-      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
+      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,;
     \]
@@
     \[
-      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
+      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,;
     \]
@@
   \end{enumerate}
-
-  \gl{Should~$\pot'(\ceu{\protect\Every{e}\,{p_1}})\ne0$ above?}
+  %%
+  %%\gl{Should~$\pot'(\ceu{\protect\Every{e}\,{p_1}})\ne0$ above?}
+  %%
 \end{definition}
@@
 
-The next two lemmas establish that a single application of an~$\out$
-or~$\nst$ transition either preserves or decreases the rank of the input
-description.  All rank comparisons assume lexicographical order.
+The next two lemmas establish that a single application of~$\out$ or~$\nst$
+either preserves or decreases the rank of the input description.  All rank
+comparisons assume lexicographical order.
 
@@
   Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$, $\rank(\delta)=\<i,j>$,
-  and~$\rank(\delta')=\<i',j'>$.  By the theorem hypothesis, there is a
-  derivation~$\pi$ such that
+  and~$\rank(\delta')=\<i',j'>$.  By the hypothesis of the theorem, there is
+  a derivation~$\pi$ such that
   \[
@@
     Then~$\pi$ is an instance of~\R{mem}.  Hence~$p'=\ceu{\Nop}$
-    and~$e'=\nil$.  Thus
-    \[
-      \rank(\delta)=\rank(\delta')=\<0,n>\,.
-    \]
+    and~$e'=\nil$.  Thus $\rank(\delta)=\rank(\delta')=\<0,n>$.
   \end{case}
@@
 \begin{proof}
-  If~$\delta\nst[*]\delta'$ then there is an~$n$ such that
-  $\delta\nst[n]\delta'$.  We proceed by induction on~$n$.
-  %%
-  The theorem is trivially true for~$n=0$ and follows directly from
-  Lemma~\ref{lem.rank-nst} for~$n=1$.  Suppose
-  $\delta\nst[1]\delta_1'\nst[n-1]\delta'$, for some~$n>1$ and~$\delta_1'$.
+  If~$\delta\nst[*]\delta'$ then~$\delta\nst[n]\delta'$, for some~$i$.  We
+  proceed by induction on~$i$.
+  %%
+  The theorem is trivially true for~$i=0$ and follows directly from
+  Lemma~\ref{lem.rank-nst} for~$i=1$.  Suppose
+  $\delta\nst[1]\delta_1'\nst[i-1]\delta'$, for some~$i>1$ and~$\delta_1'$.
   Thus, by Lemma~\ref{lem.rank-nst} and by the induction hypothesis,
@@
   some~$i\ge0$.
-
-  An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form
-  for transitions in general.  It cannot be advanced by~$\nst$, as it is
-  nested-irreducible, and neither by~$\outpush$ nor~$\outpop$, as the second
-  coordinate of its rank is~0, which implies~$e=\nil$ and~$n=0$.
 \end{definition}
 
-\begin{theorem}
-  \label{thm.term}
+An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for
+general transitions~$\trans$.  Such descriptions cannot be advanced
+by~$\nst$, as it is nested-irreducible, and neither by~$\outpush$
+nor~$\outpop$, as the second coordinate of its rank is~0, which
+implies~$e=\nil$ and~$n=0$.
+
+\begin{theorem}[Termination]\label{thm.term}\strut\\
   %%
@@
   \begin{basis}
-    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$
-    transitions, as~$j=0$ implies~$e=\nil$ and~$n=0$.  Thus neither~\R{push}
-    nor~\R{pop} can be applied.  There are two possibilities:
-    either~$\delta$ is nested-irreducible or it is not.  In the first case,
-    the theorem is trivially true, as~$\delta\nst[0]\delta_\Hnst$.
-
-    Suppose~$\delta$ is not nested-irreducible.  Then, by
-    Theorem~\ref{thm.term-nst-*}, $\delta\nst[*]\delta'_\Hnst$ for
-    some~$\delta'_\Hnst$.  By Theorem~\ref{thm.rank-nst-*},
+    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$, as~$j=0$
+    implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be
+    applied).  There are two possibilities:  either~$\delta$ is
+    nested-irreducible or it is not.  In the first case, the theorem is
+    trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
+    nested-irreducible.  Then, by Theorem~\ref{thm.term-nst-*},
+    $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
+    Theorem~\ref{thm.rank-nst-*},
     \[
@@
           By the induction hypothesis, $\delta'\trans[*]\delta''_\#$, for
-          some~$\delta''_\#$.  Thus~$\delta\trans[*]\delta''_\#$.
+          some~$\delta''_\#$.  Therefore, $\delta\trans[*]\delta''_\#$.
         \end{subsubcase}
@@
           that~$\delta'\trans[*]\delta''_\#$.
-          Thus~$\delta\trans[*]\delta''_\#$.
+          Therefore, $\delta\trans[*]\delta''_\#$.
         \end{subsubcase}

### 2018-02-23 03-semantics.tex
Seção 3.3.1, determinismo

@@
 
-- informal discussion
+\subsubsection{Determinism}
 
-- reference the appendix A
+Transitions~$\out$ and~$\nst$ are defined in such a way that given an input
+description either no rule is applicable or exactly one of them can be
+applied.  This means that the resulting relation~$\trans$ is in fact a
+partial function and that, consequently, the same number of applications
+of~$\trans$ will always yield the same result.
 
-\subsubsection*{Determinism}
+The next two lemmas establish the determinism of a single application
+of~$\out$ and~$\nst$.  Lemma~\ref{lem.x.det-out} follows from a simple
+inspection of rules~\R{push} and~\R{pop}.  The proof of
+Lemma~\ref{lem.x.det-out} proceeds by induction on the structure of the
+derivation trees produced by applications of the rules for~$\nst$.  Both
+lemmas are used in the proof of Theorem~\ref{thm.x.det}, the main result of
+this section, which establishes that any given number of applications
+of~$\trans$ starting from the same input description will always lead to the
+same output description.
 
-The proof for determinism relies on the fact all semantic rules are mutually
-exclusive, i.e., their preconditions are unique in the set of rules.
-This can be verified by direct inspection of rules.
+\begin{lemma}
+  \label{lem.x.det-out}
+  %%
+  If~$\delta\out\delta_1$ and~$\delta\out\delta_2$ then~$\delta_1=\delta_2$.
+\end{lemma}
 
-Rule \textbf{push} is the only one with $e \neq \epsilon$ as a precondition,
-and is trivially mutually exclusive with all other rules.
+\begin{lemma}
+  \label{lem.x.det-nst}
+  %%
+  If~$\delta\nst\delta_1$ and~$\delta\nst\delta_2$ then~$\delta_1=\delta_2$.
+\end{lemma}
 
-Rule \textbf{pop} either has $p=@nop$ or $isblocked(p,n)$ as preconditions.
-%
-Note that rule \textbf{pop} only applies syntactically to top-level
-transitions.
-For instance, it can never match $\NST$ rules for subprograms as in rule
-\textbf{seq-adv}.
-%
-Hence, for the first case, rule \textbf{pop} only applies, and is the only one
-to apply, to $nop$ as the whole program (i.e., a $nop$ not surrounded by other
-expressions, such as in rule \textbf{seq-nop}).
-%
-For the second case, we need to show that given $\LL p,n,\epsilon \RR$, no
-$\NST$ transitions apply with $isblocked(p,n)$ and vice versa.
-Except for $@canrun$, there are no $\NST$ transitions for the other blocking
-expressions ($awaitExt$, $awaitInt$, $every$, and $fin$).
-However, considering the precondition $\LL p,n,\epsilon \RR$,
-$isblocked(@canrun(n),n)$ is false.
-Hence, given the preconditions for rule \textbf{pop}, no $\NST$ transitions can
-occur.
-Conversely, if a $\NST$ transition is possible, then $isblocked(p,n)$ must be
-false.
-Again, except for $@canrun$, all other transitions do not involve blocking
-expressions, hence, for these transitions, $isblocked(p,n)$ must be false.
-For rule \R{can-run}, a transition can only occur if the current stack
-level matches $@canrun(n)$.
-In this case, $isblocked(@canrun(n),n)$ is false.
+\begin{theorem}[Determinism]
+  \label{thm.x.det}\strut\\
+  %%
+  If~$\delta\trans[i]\delta_1$ and~$\delta\trans[i]\delta_2$
+  then~$\delta_1=\delta_2$.
+\end{theorem}
+\begin{proof}
+  By induction on~$i$.  The theorem is trivially true if~$i=0$ and follows
+  directly from the lemmas if~$i=1$.  Suppose
+  \[
+    \delta\trans[1]\delta_1'\trans[i-1]\delta_1
+    \quad\text{and}\quad
+    \delta\trans[1]\delta_2'\trans[i-1]\delta_2\,,
+  \]
+  for some~$i>1$, $\delta_1'$ and~$\delta_2'$.
+  %%
+  Then, by Lemma~\ref{lem.det-out} or~\ref{lem.det-nst}, depending on
+  whether the first transition is~$\out$ or~$\nst$ (it cannot be both),
+  $\delta_1'=\delta_2'$, and by the induction hypothesis,
+  $\delta_1=\delta_2$.
+\end{proof}
 
-Finally, we need to show that $\NST$ transitions are mutually exclusive among
-themselves.
-%
-Note that most rules have unique syntactic prefixes, e.g., $(@nop~@and~q)$
-(rule \textbf{and-nop1}) is trivially mutually exclusive with $(@nop~@loop~p)$
-(rule \textbf{or-nop1}).
-%
-The only exceptions are rules \textbf{and-adv1} vs. \textbf{and-adv2}, and
-\textbf{or-adv1} vs. \textbf{or-adv2}.
-In both cases, we need to show that if the left branch can advance, then it
-cannot be blocked and vice-versa, i.e., that
-$\LL p,n,\epsilon \RR \NST \LL p',n,e \RR$ and $isblocked(p,n)$ are mutually
-exclusive, which is exactly the same reasoning for rule \textbf{pop} above.
+% The proof for determinism relies on the fact all semantic rules are
+% mutually exclusive, i.e., their preconditions are unique in the set of
+% rules.  This can be verified by direct inspection of rules.
+
+% Rule \textbf{push} is the only one with $e \neq \epsilon$ as a
+% precondition, and is trivially mutually exclusive with all other rules.
+
+% Rule \textbf{pop} either has $p=@nop$ or $isblocked(p,n)$ as
+% preconditions.
+% %
+% Note that rule \textbf{pop} only applies syntactically to top-level
+% transitions.  For instance, it can never match $\NST$ rules for
+% subprograms as in rule \textbf{seq-adv}.
+% %
+% Hence, for the first case, rule \textbf{pop} only applies, and is the only
+% one to apply, to $nop$ as the whole program (i.e., a $nop$ not surrounded
+% by other expressions, such as in rule \textbf{seq-nop}).
+% %
+% For the second case, we need to show that given $\LL p,n,\epsilon \RR$, no
+% $\NST$ transitions apply with $isblocked(p,n)$ and vice versa.  Except for
+% $@canrun$, there are no $\NST$ transitions for the other blocking
+% expressions ($awaitExt$, $awaitInt$, $every$, and $fin$).  However,
+% considering the precondition $\LL p,n,\epsilon \RR$,
+% $isblocked(@canrun(n),n)$ is false.  Hence, given the preconditions for
+% rule \textbf{pop}, no $\NST$ transitions can occur.  Conversely, if a
+% $\NST$ transition is possible, then $isblocked(p,n)$ must be false.
+% Again, except for $@canrun$, all other transitions do not involve blocking
+% expressions, hence, for these transitions, $isblocked(p,n)$ must be false.
+% For rule \R{can-run}, a transition can only occur if the current stack
+% level matches $@canrun(n)$.  In this case, $isblocked(@canrun(n),n)$ is
+% false.
+
+% Finally, we need to show that $\NST$ transitions are mutually exclusive
+% among themselves.
+% %
+% Note that most rules have unique syntactic prefixes, e.g., $(@nop~@and~q)$
+% (rule \textbf{and-nop1}) is trivially mutually exclusive with
+% $(@nop~@loop~p)$ (rule \textbf{or-nop1}).
+% %
+% The only exceptions are rules \textbf{and-adv1} vs. \textbf{and-adv2}, and
+% \textbf{or-adv1} vs. \textbf{or-adv2}.  In both cases, we need to show
+% that if the left branch can advance, then it cannot be blocked and
+% vice-versa, i.e., that $\LL p,n,\epsilon \RR \NST \LL p',n,e \RR$ and
+% $isblocked(p,n)$ are mutually exclusive, which is exactly the same
+% reasoning for rule \textbf{pop} above.
 

### 2018-02-23 03-semantics.tex
Seção 3.3, terminação (início)

@@
 
-\subsection{Determinism, Termination, and Memory Bounds}
+\subsection{Properties}
 
@@
 applied.  This means that the resulting relation~$\trans$ is in fact a
-partial function and that, consequently, the same number of applications
-of~$\trans$ will always yield the same result.
+partial function.
 
@@
 inspection of rules~\R{push} and~\R{pop}.  The proof of
-Lemma~\ref{lem.x.det-out} proceeds by induction on the structure of the
-derivation trees produced by applications of the rules for~$\nst$.  Both
-lemmas are used in the proof of Theorem~\ref{thm.x.det}, the main result of
-this section, which establishes that any given number of applications
-of~$\trans$ starting from the same input description will always lead to the
-same output description.
+Lemma~\ref{lem.x.det-out}, however, requires an induction on the structure
+of the derivation trees produced by the rules for~$\nst$.  Both lemmas are
+used in the proof of Theorem~\ref{thm.x.det}, the main result of this
+section.  Theorem~\ref{thm.x.det} establishes that any given number of
+applications of~$\trans$ starting from the same input description will
+always lead to the same output description.
 
@@
 
-\subsubsection*{Termination}
-
-- there is always a possible transition until n=0
-
-every cannot restart itself
-    - break disallowed
-    - emit ignored
+\subsubsection{Termination}
+
+In this section, we prove that a sufficiently long sequence of applications
+of~$\trans$, and consequently~$\out$ and~$\nst$, will eventually lead to an
+irreducible description, viz., one that cannot be modified by further
+transitions.
+
+\begin{definition}
+  \label{def.x.Hnst}
+  %%
+  A description~$\delta=\<p,n,e>$ is \emph{nested-irre\-ducible}
+  iff~$e\ne\nil$ or~$p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p,n)$~is
+  true.
+\end{definition}
+
+Nested-irreducible descriptions serve as normal forms for~$\nst$
+transitions: they embody the result of an exhaustive number of~$\nst$
+applications.  We will write~$\delta_\Hnst$ to indicate that
+description~$\delta$ is nested-irreducible.
+
+To characterize irreducible descriptions in general, we will need to define
+the notions of potency of a program and the rank of a description.
+
+\begin{definition}
+  \label{def.x.pot}
+  %%
+  The potency of a program~$p$ in reaction to event~$e$,
+  denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ expressions
+  that can be executed during a reaction of~$p$ to~$e$.
+
+  More formally,
+  \[
+    \pot(p,e)=\pot'(\bcast(p,e))\,,
+  \]
+  where~$\pot'$ is an auxiliary function that counts the number of
+  reachable~$\ceu{\EmitInt}$ expressions in the program resulting from the
+  broadcast of event~$e$ to~$p$.
+
+  The auxiliary function~$\pot'$ is defined by the following clauses:
+  \begin{enumerate}[(a)]
+  \item\label{def.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$;
+    %%
+  \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})
+    =\max\{\pot'(p_1),\pot'(p_2)\}$;
+    %%
+  \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$;
+    %%
+  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(p_1)+\pot'(p_2)$;
+    %%
+  \item$\pot'(\ceu{{p_1}\Or{p_2}})=\pot'(p_1)+\pot'(p_2)$;
+    %%
+  \item If~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
+    \begin{align*}
+      \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\
+      \pot'(\ceu{p_1\AtLoop{p_2}})&=
+      \begin{cases}
+        \pot'(p_1)              &\text{if\enspace(\dag)}\\
+        \pot'(p_1)+\pot'(p_2)   &\text{otherwise}\\
+      \end{cases}
+    \end{align*}
+    where~(\dag) stands for: ``a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ occurs
+    in all execution paths of~$p_1$'';
+    %%
+  \item If~$p_1,p_2\ne\ceu{\Break}$,
+    \[
+      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,;
+    \]
+    %%
+  \item\label{def.pot.last} If~$p_1,p_2\ne\ceu{\Break}$
+    and~$p_1,p_2\ne\ceu{\Nop}$,
+    \[
+      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,;
+    \]
+    %%
+  \item Otherwise, if none of~\eqref{def.pot.first}--\eqref{def.pot.last}
+    applies, $\pot(\_)=0$.
+  \end{enumerate}
+\end{definition}
+
+\begin{definition}
+  \label{def.x.rank}
+  %%
+  The \emph{rank} of a description~$\delta=\<p,n,e>$,
+  denoted~$\rank(\delta)$, is a pair of nonnegative integers~$\<i,j>$ such
+  that
+  \begin{alignat*}{2}
+    i&=\pot(p,e) &\quad\text{and}\quad
+    j&=
+       \begin{cases}
+         n  &\text{if~$e=\nil$}\\
+         n+1&\text{otherwise\,.}
+       \end{cases}
+  \end{alignat*}
+\end{definition}
+
+\begin{definition}
+  \label{def.x.H}
+  %%
+  A description~$\delta$ is \emph{irreducible} (in symbols, $\delta_\#$) iff
+  it is nested-irreducible and its~$\rank(\delta)$ is~$\<i,0>$, for
+  some~$i\ge0$.
+\end{definition}
+
+An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for
+transitions~$\trans$ in general.  Such descriptions cannot be advanced
+by~$\nst$, as it is nested-irreducible, and neither by~$\outpush$
+nor~$\outpop$, as the second coordinate of its rank is~0, which
+implies~$e=\nil$ and~$n=0$.
+
+TODO: Lemas e teorema da terminação.  Sketch da prova.  Mencionar que as
+restrições sintáticas são fundamentais.
+
+% - there is always a possible transition until n=0
+%
+% every cannot restart itself
+%     - break disallowed
+%     - emit ignored
 
 \subsubsection*{Memory Bounds}
+
+TODO: Acho que sai direto do~$\pot(p)$.
 

### 2018-02-23 02-ceu.tex
Revisão Seção 3-5
Sugestões do Hermann

@@
 
-\CEU is a synchronous reactive language in which programs advance in a sequence
+\CEU is a synchronous reactive language in which programs evolve in a sequence
 of discrete reactions to external events.
@@
 \CEU provides an \code{await} statement that blocks the current running trail
-allowing the program to advance its other trails; when all trails are blocked,
+allowing the program to execute its other trails; when all trails are blocked,
 the reaction terminates and control returns to the environment.
@@
 \subsection{Shared-Memory Concurrency}
+\label{sec.ceu.shared}
 

### 2018-02-23 03-semantics.tex
Sugestões do Hermann
(Faltou definicao de E)
Comprime os o conteúdo da Seção 3.2

@@
-\newcommand{\NST}{\1\xrightarrow[\mathit{nst}]\1}
-\newcommand{\OUT}{\1\xrightarrow[\mathit{out}]\1}
-\newcommand{\LL}{\langle}
-\newcommand{\RR}{\rangle}
-\newcommand{\DS}{\displaystyle}
-
-\newcommand{\1}{\;}
-\newcommand{\2}{\;\;}
-\newcommand{\3}{\;\;\;}
-\newcommand{\5}{\;\;\;\;\;}
+% \newcommand{\NST}{\1\xrightarrow[\mathit{nst}]\1}
+% \newcommand{\OUT}{\1\xrightarrow[\mathit{out}]\1}
+% \newcommand{\LL}{\langle}
+% \newcommand{\RR}{\rangle}
+% \newcommand{\DS}{\displaystyle}
+
+% \newcommand{\1}{\;}
+% \newcommand{\2}{\;\;}
+% \newcommand{\3}{\;\;\;}
+% \newcommand{\5}{\;\;\;\;\;}
+
+\def\JOT{1.5\jot}
 
@@
 \begin{alignat*}{2}
-  p\Coloneqq
+  P\Coloneqq
       &\enspace\ceu{\Mem(\Id)}
@@
       %%
-  \mid&\enspace\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}}
+  \mid&\enspace\ceu{\IfElse{\Mem(\Id)}{P_1}{P_2}}
       &&\lbl{conditional}\\[\X]
       %%
-  \mid&\enspace\ceu{p_1\,;\,p_2}
+  \mid&\enspace\ceu{P_1\,;\,P_2}
       &&\lbl{sequence}\\[\X]
       %%
-  \mid&\enspace\ceu{\Loop p_1}
+  \mid&\enspace\ceu{\Loop P_1}
       &&\lbl{repetition}\\[\X]
       %%
-  \mid&\enspace\ceu{\Every{\Id}\ p_1}
+  \mid&\enspace\ceu{\Every{\Id}\ P_1}
       &&\lbl{event iteration}\\[\X]
       %%
-  \mid&\enspace\ceu{p_1\And p_2}
+  \mid&\enspace\ceu{P_1\And P_2}
       &&\lbl{par/and}\\[\X]
       %%
-  \mid&\enspace\ceu{p_2\Or p_2}
+  \mid&\enspace\ceu{P_2\Or P_2}
       &&\lbl{par/or}\\[\X]
       %%
-  \mid&\enspace\ceu{\Fin p}
+  \mid&\enspace\ceu{\Fin P}
       &&\lbl{finalization}\\[\X]
       %%
-  \mid&\enspace\ceu{p_1\AtLoop p_2}
+  \mid&\enspace\ceu{P_1\AtLoop P_2}
       &&\lbl{unwinded loop}\\[\X]
       %%
-  \mid&\enspace\ceu{p_1\AtAnd\ p_2}
+  \mid&\enspace\ceu{P_1\AtAnd\ P_2}
       &&\lbl{unwinded par/and}\\[\X]
       %%
-  \mid&\enspace\ceu{p_1\AtOr\ p_2}
+  \mid&\enspace\ceu{P_1\AtOr\ P_2}
       &&\lbl{unwinded par/or}\\[\X]
@@
 the concrete language.
-The exceptions are the finalization block~$\ceu{\Fin{p}}$ and the
+The exceptions are the finalization block~$\ceu{\Fin{P}}$ and the
 \texttt{@}-expressions which result from expansions in the transition rules to
@@
 empty event~$\nil$).
+$\E$ is a set of primitive events.
 
@@
   \DisplayProof
-  \Rtag{push}\\[2\jot]
+  \Rtag{push}\\[\JOT]
   %%
@@
 
-The~$\nst$ rules are recursive definitions with the following general format:
+The~$\nst$ rules are recursive definitions of the form
 \[
@@
 
-A complete reaction consists of a series of transitions:
+A complete reaction consists of a series of transitions of the form
 \begin{align*}
@@
 
-The~$\nst$ transition rules for atomic expressions are defined as follows:
+The~$\nst$ rules for atoms are defined as follows:
 \begin{align*}
   \<\ceu{\Mem(\Id)},n,\nil>
-  &\nst\<\ceu{\Nop},n,\nil>\Rtag{mem}\\
+  &\nst\<\ceu{\Nop},n,\nil>\Rtag{mem}\\[\JOT]
   %%
   \<\ceu{\EmitInt(\Id)},n,\nil>
-  &\nst\<\ceu{\CanRun(n)},n,\ceu{\Id}>\Rtag{emit-int}\\
+  &\nst\<\ceu{\CanRun(n)},n,\ceu{\Id}>\Rtag{emit-int}\\[\JOT]
   %%
@@
 
-Proceeding to compound expressions, the rules for conditionals and sequences
-are straightforward:
+The rules for conditionals and sequences are straightforward:
 %
@@
   \DisplayProof
-  \Rtag{if-true}\\[2\jot]
+  \Rtag{if-true}\\[\JOT]
   %%
@@
   \DisplayProof
-  \Rtag{if-false}\\[2\jot]
+  \Rtag{if-false}\\[\JOT]
   %%
@@
 \vskip-\abovedisplayskip
-\vskip2\jot
+\vskip\JOT
 \begin{align*}
-  \<\ceu{\Nop;\,q},n,\nil>&\nst\<q,n,\nil>\Rtag{seq-nop}\\[2\jot]
+  \<\ceu{\Nop;\,q},n,\nil>&\nst\<q,n,\nil>\Rtag{seq-nop}\\[\JOT]
   %%
@@
 
-The rules for loops are analogous to sequences, but use ``\code{@}'' as
-separators to properly bind breaks to their enclosing loops:
+The rules for loops are similar to those for sequences, but use ``\code{@}''
+as separators to bind breaks to their enclosing loops:
 \begin{align*}
   \<\ceu{\Loop{p}},n,\nil>
-  &\nst\<\ceu{p\AtLoop{p}},n,\nil>\Rtag{loop-expd}\\[2\jot]
+  &\nst\<\ceu{p\AtLoop{p}},n,\nil>\Rtag{loop-expd}\\[\JOT]
   %%
@@
   \DisplayProof
-  \Rtag{loop-adv}\\[2\jot]
+  \Rtag{loop-adv}\\[\JOT]
   %%
   \<\ceu{\Nop\AtLoop{p}},n,\nil>
-  &\nst\<\ceu{\Loop{p}},n,\nil>\Rtag{loop-nop}\\[2\jot]
+  &\nst\<\ceu{\Loop{p}},n,\nil>\Rtag{loop-nop}\\[\JOT]
   %%
@@
 
-The semantic rules for~$\ceu{\And}$ and~$\ceu{\Or}$ compositions
-force transitions on their left branches~$p$ to occur before their right
-branches~$q$:
+The rules for~$\ceu{\And}$ and~$\ceu{\Or}$ compositions make sure that their
+left branches always transition before their right branches:
+%%
 \begin{gather*}
@@
   \nst\<\ceu{p\AtAnd(\CanRun(n);q)},n,\nil>
-  \Rtag{and-expd}\\[2\jot]
+  \Rtag{and-expd}\\[\JOT]
   %%
@@
   \DisplayProof
-  \Rtag{and-adv1}\\[2\jot]
+  \Rtag{and-adv1}\\[\JOT]
   %%
@@
   \DisplayProof
-  \Rtag{and-adv2}\\[2\jot]
-  %%
+  \Rtag{and-adv2}
+\end{gather*}
+
+\begin{gather*}
   \<\ceu{p\Or{q}},n,\nil>
   \nst\<\ceu{p\AtOr(\CanRun(n);q)},n,\nil>
-  \Rtag{or-expd}\\[2\jot]
+  \Rtag{or-expd}\\[\JOT]
   %%
@@
   \DisplayProof
-  \Rtag{or-adv1}\\[2\jot]
+  \Rtag{or-adv1}\\[\JOT]
   %%
@@
 %
-In a parallel~$\ceu{\AtOr}$, however, if one of the sides terminates, the whole composition
+In a parallel~$\ceu{\AtOr}$, however, if one of the sides terminates, the
+whole composition
 terminates and function~$\clear$ is used to properly finalize the aborted
@@
 \begin{gather*}
-  \<\ceu{{\Nop}\AtAnd{q}},n,\nil>\nst\<q,n,\nil>\Rtag{and-nop1}\\[2\jot]
+  \<\ceu{{\Nop}\AtAnd{q}},n,\nil>\nst\<q,n,\nil>\Rtag{and-nop1}\\[\JOT]
   %%
@@
   \DisplayProof
-  \Rtag{and-nop2}\\[2\jot]
-  %%
-  \<\ceu{{\Nop}\AtOr{q}},n,\nil>\nst\<\clear(q),n,\nil>\Rtag{or-nop1}\\[2\jot]
+  \Rtag{and-nop2}\\[\JOT]
+  %%
+  \<\ceu{{\Nop}\AtOr{q}},n,\nil>\nst\<\clear(q),n,\nil>\Rtag{or-nop1}\\[\JOT]
   %%
@@
   \<\ceu{{\Break}\AtAnd{q}},n,\nil>\nst\<\ceu{\clear(q);\Break},n,\nil>
-  \Rtag{and-brk1}\\[2\jot]
+  \Rtag{and-brk1}\\[\JOT]
   %%
@@
   \DisplayProof
-  \Rtag{and-brk2}\\[2\jot]
+  \Rtag{and-brk2}\\[\JOT]
   %%
   \<\ceu{{\Break}\AtOr{q}},n,\nil>\nst\<\ceu{\clear(q);\Break},n,\nil>
-  \Rtag{or-brk1}\\[2\jot]
+  \Rtag{or-brk1}\\[\JOT]
   %%
@@
 
-In this section, we prove that a sufficiently long sequence of applications
-of~$\trans$, and consequently~$\out$ and~$\nst$, will eventually lead to an
-irreducible description, viz., one that cannot be modified by further
-transitions.
+We proceed to prove that a sufficiently long sequence of applications
+of~$\trans$ will eventually lead to an irreducible description, i.e., one
+that cannot be modified by further transitions.  Before that, however, we
+need to introduce some definitions and basic results regarding~$\nst$
+and~$\out$.
 

### 2018-02-23 02-ceu.tex
Adiciona mais enunciados à Seção 3.3.2 (terminação)

@@
 \centering\small{\ax Accesses to \code{x} are never concurrent.}
-\end{minipage}
+\end{minipage}%
 %
@@
 \centering\small{\bx Accesses to \code{y} are concurrent but still deterministic.}
-\end{minipage}
+\end{minipage}%
 %\rule{8.4cm}{0.37pt}
-\caption{ Shared-memory concurrency in \CEU:
+\caption{Shared-memory concurrency in \CEU:
 example \ax is safe because the trails access \code{x} atomically in different
@@
 example \bx is unsafe because both trails access \code{y} in the same reaction.
-\label{lst.shared}
+\label{lst.shared}%
 }
@@
 
-In \CEU, when multiple trails are active during the same reaction, they are
+In \CEU, when multiple trails are active in the same reaction, they are
 scheduled in lexical order, i.e., in the order they appear in the program
@@
 %
-For instance, consider the two examples in Figure~\ref{lst.shared}, both
-defining shared variables (ln. 3), and assigning to them in parallel trails
-(ln. 6, 9).
+For instance, consider the examples in Figure~\ref{lst.shared}, both
+defining shared variables (ln.~3), and assigning to them in parallel trails
+(ln.~6, 9).
 
-In the example \ax, the two assignments to \code{x} can only execute in
-reactions to different events \code{A} and \code{B} (ln. 5,8), which cannot
+In example \ax, the two assignments to \code{x} can only execute in
+reactions to different events \code{A} and \code{B} (ln.~5, 8), which cannot
 occur simultaneously by definition.
@@
 
-In the example \bx, the two assignments to \code{y} are simultaneous because
+In example \bx, the two assignments to \code{y} are simultaneous because
 they execute in reaction to the same event \code{A}.
@@
 if a variable is written in a trail segment, then a concurrent trail segment
-cannot read or write to that variable~\cite{ceu.sensys13}.
+cannot access that variable~\cite{ceu.sensys13}.
 %
@@
 
-\begin{figure}
+\begin{figure}[h]
 \begin{minipage}[t]{0.45\linewidth}
@@
 \end{lstlisting}
-\centering\small{\ax Local resource finalization}
-\end{minipage}
+\small{\ax\!Local resource finalization}
+\end{minipage}%
 %
@@
 \centering\small{\bx External resource finalization}
-\end{minipage}
+\end{minipage}%
 %\rule{8.4cm}{0.37pt}
-\caption{
+\caption{%
 \CEU enforces the use of finalization to prevent \emph{dangling pointers} for
-local resources and \emph{memory leaks} for external resources.
-\label{lst.fin.ceu}
-}
+local resources and \emph{memory leaks} for external resources.%
+\label{lst.fin.ceu}}
 \end{figure}
@@
 The \code{par/or} of \CEU is an orthogonal abortion mechanism because the two
-sides in the composition need not be tweaked with synchronization primitives or
-state variables in order to affect each other.
+sides in the composition need not be tweaked with synchronization primitives nor
+state variables to affect each other.
 %
@@
 \begin{itemize}
-    \item If \CEU \textbf{passes} a pointer to a system call (ln. \ax:5), the
-          pointer represents a \textbf{local} resource (ln. \ax:2) that
-          requires finalization (ln. \ax:7).
+    \item If \CEU \textbf{passes} a pointer to a system call (ln.~\ax:5), the
+          pointer represents a \textbf{local} resource (ln.~\ax:2) that
+          requires finalization (ln.~\ax:7).
     \item If \CEU \textbf{receives} a pointer from a system call return
-          (ln.  \bx:4), the pointer represents an \textbf{external} resource
-          (ln. \bx:2) that requires finalization (ln. \bx:6).
+          (ln.~ \bx:4), the pointer represents an \textbf{external} resource
+          (ln.~\bx:2) that requires finalization (ln.~\bx:6).
 \end{itemize}
@@
 %
-In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
-(ln. 2) is an internal resource passed as a pointer to \code{\_send} (ln. 5),
+In Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
+(ln.~2) is an internal resource passed as a pointer to \code{\_send} (ln.~5),
 which is an asynchronous call that transmits the buffer in the background.
-If the block aborts (ln. 11) before receiving an acknowledge from the
-environment (ln. 9), the local \code{msg} goes out of scope and the external
+If the block aborts (ln.~11) before receiving an acknowledge from the
+environment (ln.~9), the local \code{msg} goes out of scope and the external
 transmission now holds a \emph{dangling pointer}.
-The finalization ensures that the transmission also aborts (ln. 7).
+The finalization ensures that the transmission also aborts (ln.~7).
 %
-In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
+In Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
 4) returns an external file resource as a pointer.
-If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file
+If the block aborts (ln.~12) during the \code{await A} (ln.~9), the file
 remains open as a \emph{memory leak}.
-The finalization ensures that the file closes properly (ln. 6).
+The finalization ensures that the file closes properly (ln.~6).
 %
@@
 construct.%
-\footnote{
+\footnote{%
 The compiler only forces the programmer to write the finalization clause, but

### 2018-02-23 03-semantics.tex
Adiciona mais enunciados à Seção 3.3.2 (terminação)

@@
 of~$\trans$ will eventually lead to an irreducible description, i.e., one
-that cannot be modified by further transitions.  Before that, however, we
-need to introduce some definitions and basic results regarding~$\nst$
-and~$\out$.
+that cannot be modified by further transitions.  Before doing that, however,
+we need to introduce some notation and establish some basic properties
+of~$\nst$ and~$\out$.
 
@@
 description~$\delta$ is nested-irreducible.
+
+The use of qualifier ``irreducible'' in Definition~\ref{def.x.Hnst} is
+justified by Proposition~\ref{prop.x.irr-nst-i}.  The proposition
+establishes that if a finite number of applications of~$\nst$ results in an
+irreducible description, then that can only occur exactly once, at some
+specific number~$i$.  The proof of the lemma follows directly from the
+definition of~$\nst$ by contradiction on the hypothesis that there is
+such~$k\ne{i}$.
+
+\begin{proposition}
+  \label{prop.x.irr-nst-i}
+  %%
+  If~$\delta\nst[i]\delta_\Hnst'$ then, for all~$k\ne{i}$, there is
+  no~$\delta_\Hnst''$ such that~$\delta\nst[k]\delta''_\Hnst$.
+\end{proposition}
+
+The next lemma establishes that sequences of~$\nst$ transitions are
+well-behaved regarding the order of evaluation of composition members.  Its
+proof follows by induction on~$i$.
+
+\begin{lemma}
+  \label{lem.x.props-nst-i}\strut\\
+  %%
+  If~$\<p_1,n,e>\nst[i]\<p_1',n,e'>$ then, for any~$p_2$:
+  \begin{enumerate}[(a)]
+  \item\label{lem.x.props-nst-i.a}
+    $\<\ceu{p_1;\,p_2},n,e>\nst[i]\<p_1';p_2,n,e'>$;
+    %%
+  \item\label{lem.x.props-nst-i.b}
+    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[i]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$;
+    %%
+  \item\label{lem.x.props-nst-i.c}
+    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$;
+    %%
+  \item\label{lem.x.props-nst-i.d}
+    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}'\AtOr{p_2}},n,e'>$.
+  \end{enumerate}
+  If~$\<p_2,n,e>\nst[i]\<p_2',n,e'>$, for any~$p_1$ such
+  that~$\isblocked(p_1,n)$:
+  \begin{enumerate}[(a)]
+    \setcounter{enumi}{4}
+  \item\label{lem.x.props-nst-i.e}
+    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>$;
+    %%
+  \item\label{lem.x.props-nst-i.f}
+    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtOr{p_2'}},n,e'>$.
+  \end{enumerate}
+\end{lemma}
+%%
+% \begin{proof}[Proof of~(a)]
+%   By induction on~$i$.  The lemma is trivially true for~$i=0$,
+%   as~$p_1=p_1'$, and follows directly from~\R{seq-adv} for~$i=1$.  Suppose
+%   \begin{equation}
+%     \label{lem.x.props-nst-i.a.eq1}
+%     \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[i-1]\<p_1',n,e'>\,,
+%   \end{equation}
+%   for some~$i>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible, i.e.,
+%   $e=\nil$ and~$p\ne{\ceu{\Nop},\ceu{\Break}}$ and~$\isblocked(p_1'',n)$
+%   is false.  By~\eqref{lem.x.props-nst-i.a.eq1} and by~\R{seq-adv},
+%   \begin{equation}
+%     \label{lem.x.props-nst-i.a.eq2}
+%     \<\ceu{p_1;\,p_2},n,e>\nst[1]\<\ceu{p_1'';\,p_2},n,e''>\,.
+%   \end{equation}
+%   From~\eqref{lem.x.props-nst-i.a.eq1}, by the induction hypothesis,
+%   \begin{equation}
+%     \label{lem.x.props-nst-i.a.eq3}
+%     \<\ceu{p_1'';\,p_2},n,e''>\nst[i-1]\<\ceu{p_1';\,p_2},n,e'>\,.
+%   \end{equation}
+%   From~\eqref{lem.x.props-nst-i.a.eq2} and~\eqref{lem.x.props-nst-i.a.eq3},
+%   \[
+%     \<\ceu{p_1;\,p_2},n,e>\nst[i]\<\ceu{p_1';\,p_2},n,e'>\,.
+%   \]
+%   The proofs of the other cases are similar.
+% \end{proof}
+%%
+
+The syntactic restrictions discussed in Section~\ref{sec.sem}, regarding the
+body of~$\ceu{\Fin}$ and~$\ceu{\Loop}$ expressions, and their consequences,
+are formalized in the next assumption.
+
+\begin{assumption}[Syntactic restrictions]\strut
+  \label{ass.x.syn-rest}
+  %%
+  \begin{enumerate}[(a)]
+  \item\label{ass.x.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contain
+    no occurrences of~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$,
+    or~$\ceu{\Fin}$ expressions.  Consequently, for any~$n$,
+    $\<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>$.
+    %%
+  \item\label{ass.x.syn-rest.loop} If~$p=\ceu{\Loop{p_1}}$ then all
+    execution paths of~$p_1$ contain at a matching~$\ceu{\Break}$ or
+    an~$\ceu{\AwaitExt}$.  Consequently, for all~$n$, there are~$p_1'$
+    and~$e$ such that $\<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>$, where
+    $p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
+  \end{enumerate}
+\end{assumption}
+
+Theorem~\ref{thm.x.term-nst-*} establishes that a finite (possibly zero)
+number of~$\nst$ transitions eventually lead to a nested-irreducible
+description.  The proof follows by induction on the structure of programs
+(set~$P$) and depends on Lemma~\ref{lem.props-nst-i} and
+Assumption~\ref{ass.x.syn-rest}.
+
+\begin{theorem}
+  \label{thm.x.term-nst-*}
+  %%
+  For any~$\delta$ there is a~$\delta'_\Hnst$ such
+  that~$\delta\nst[*]\delta'_\Hnst$.
+\end{theorem}
+
+% -------------------------------------------------------------------------
 

### 2018-02-23 06-appendix.tex
Adiciona mais enunciados à Seção 3.3.2 (terminação)

@@
 
-The next theorem establishes that transitions are deterministic, i.e., that
-the transition relation~$\trans$ between descriptions is a function.
-
 \begin{theorem}[Determinism]\strut
@@
 
-Nested-irreducible descriptions serve as normal forms for~$\nst$
-transitions: they embody the result of an exhaustive number of~$\nst$
-applications.  We will write~$\delta_\Hnst$ to indicate that
-description~$\delta$ is nested-irreducible.
-
-The next lemma justifies the use of qualifier ``irreducible'' in
-Definition~\ref{def.Hnst}.
-
-\begin{lemma}
-  \label{lem.irr-nst-i}
+\begin{proposition}
+  \label{prop.irr-nst-i}
   %%
@@
   no~$\delta_\Hnst''$ such that~$\delta\nst[k]\delta''_\Hnst$.
-\end{lemma}
+\end{proposition}
 \begin{proof}
@@
   \begin{case}
-    \label{lem.irr-nst-i-case1}
+    \label{prop.irr-nst-i-case1}
     %%
@@
     \begin{equation}
-      \label{lem.irr-nst-i-eq1}
+      \label{prop.irr-nst-i-eq1}
       %%
@@
     that~$\delta'\nst[1]\delta_1'$, which
-    contradicts~\eqref{lem.irr-nst-i-eq1}.  Therefore, no such~$k$ can
+    contradicts~\eqref{prop.irr-nst-i-eq1}.  Therefore, no such~$k$ can
     exist.
@@
     that~$\delta\nst[k]\delta''$.  Then, since~$i>k$, by
-    Case~\ref{lem.irr-nst-i-case1}, $\delta'$~cannot exist, which is absurd.
-    Therefore, the assumption that there is such~$k$ is false.\qedhere
+    Case~\ref{prop.irr-nst-i-case1}, $\delta'$~cannot exist, which is
+    absurd.  Therefore, the assumption that there is such~$k$ is
+    false.\qedhere
   \end{case}
 \end{proof}
-
-The next lemma establishes some basic properties of sequences of~$\nst$
-transitions.
 
@@
 
-The syntactic restrictions discussed in Section~\ref{sec.sem}, regarding the
-body of~$\ceu{\Fin}$ and~$\ceu{\Loop}$ expressions, are formalized by the
-next assumption.
-
 \begin{assumption}[Syntactical restrictions]\strut

### 2018-02-23 index.tex
Completa a prova do teorema da terminação de NST
Header page: title, keywords, conference
Adiciona mais enunciados à Seção 3.3.2 (terminação)

@@
 %Conference
-\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El
-  Paso, Texas USA}
-\acmYear{1997}
-\copyrightyear{2016}
+\acmConference[LCTES'18]{ACM SIGPLAN/SIGBED}{June 2018}{Philadelphia, USA}
+%\acmYear{1997}
+%\copyrightyear{2016}
 
-\acmPrice{15.00}
+%\acmPrice{15.00}
 
@@
 
-\title{A Memory-Bounded, Deterministic and Terminating Semantics for the
+\title[A Memory-Bounded, Deterministic and Terminating Semantics for~\CEU]
+{A Memory-Bounded, Deterministic and Terminating Semantics for the
        Synchronous Programming Language~\CEU}
@@
 \terms{Languages, design, theory}
-\keywords{Operational semantics, \CEU, Synchronous languages}
+\keywords{Determinism, Termination, Operational semantics, Synchronous languages}
 

### 2018-02-23 03-semantics.tex
Atualiza mais coisa na Seção 3.3.2

@@
 %
-We use a set of small-step operational rules, which are designed in such a way
-that at most one transition is possible at any time, resulting in deterministic
-reactions.
+We use a set of small-step operational rules~\cite{Plotkin-G-D-1981}, which
+are designed in such a way that at most one transition is possible at any
+time, resulting in deterministic reactions.
 %
@@
   \item\label{lem.x.props-nst-i.a}
-    $\<\ceu{p_1;\,p_2},n,e>\nst[i]\<p_1';p_2,n,e'>$;
+    $\<\ceu{p_1;\,p_2},n,e>\nst[i]\<p_1';p_2,n,e'>$.
     %%
   \item\label{lem.x.props-nst-i.b}
-    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[i]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$;
+    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[i]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$.
     %%
   \item\label{lem.x.props-nst-i.c}
-    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$;
+    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$.
     %%
@@
   \item\label{lem.x.props-nst-i.e}
-    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>$;
+    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>$.
     %%
@@
 number of~$\nst$ transitions eventually lead to a nested-irreducible
-description.  The proof follows by induction on the structure of programs
-(set~$P$) and depends on Lemma~\ref{lem.props-nst-i} and
+description.  The proof is by induction on the structure of programs
+(members of set~$P$) and uses Lemma~\ref{lem.props-nst-i} and
 Assumption~\ref{ass.x.syn-rest}.
@@
 
-% -------------------------------------------------------------------------
-
-To characterize irreducible descriptions in general, we will need to define
-the notions of potency of a program and the rank of a description.
+The main result of this section, Theorem~\ref{?}, is similar to
+Theorem~\ref{thm.x.term-nst-*} but applies to transitions~$\trans$ in
+general.  Therefore, before stating and proving it, we need to characterize
+irreducible descriptions in general.
 
@@
   %%
-  The potency of a program~$p$ in reaction to event~$e$,
+  The \emph{potency} of a program~$p$ in reaction to event~$e$,
   denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ expressions
@@
 
-  The auxiliary function~$\pot'$ is defined by the following clauses:
+  The auxiliary function~$\pot'$ is defined by
+  clauses~\eqref{def.x.pot.first}--\eqref{def.x.pot.last}:
   \begin{enumerate}[(a)]
-  \item\label{def.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$;
+  \item\label{def.x.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$.
     %%
   \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})
-    =\max\{\pot'(p_1),\pot'(p_2)\}$;
-    %%
-  \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$;
-    %%
-  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(p_1)+\pot'(p_2)$;
-    %%
-  \item$\pot'(\ceu{{p_1}\Or{p_2}})=\pot'(p_1)+\pot'(p_2)$;
+    =\max\{\pot'(p_1),\pot'(p_2)\}$.
+    %%
+  \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$.
+    %%
+  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(p_1)+\pot'(p_2)$.
+    %%
+  \item$\pot'(\ceu{{p_1}\Or{p_2}})=\pot'(p_1)+\pot'(p_2)$.
     %%
@@
     where~(\dag) stands for: ``a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ occurs
-    in all execution paths of~$p_1$'';
-    %%
-  \item If~$p_1,p_2\ne\ceu{\Break}$,
+    in all execution paths of~$p_1$''.
+    %%
+  \item If~$p_1,p_2\ne\ceu{\Break}$.
     \[
-      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,;
+      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
     \]
     %%
-  \item\label{def.pot.last} If~$p_1,p_2\ne\ceu{\Break}$
+  \item\label{def.x.pot.before-last} If~$p_1,p_2\ne\ceu{\Break}$
     and~$p_1,p_2\ne\ceu{\Nop}$,
     \[
-      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,;
+      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
     \]
     %%
-  \item Otherwise, if none of~\eqref{def.pot.first}--\eqref{def.pot.last}
-    applies, $\pot(\_)=0$.
+  \item\label{def.x.pot.last} Otherwise, if none
+    of~\eqref{def.pot.first}--\eqref{def.pot.before-last} applies,
+    $\pot(\_)=0$.
   \end{enumerate}

### 2018-02-23 03-semantics.tex
Refatora definições de \pot e \rank na 3.3.2

@@
 Theorem~\ref{thm.x.term-nst-*} but applies to transitions~$\trans$ in
-general.  Therefore, before stating and proving it, we need to characterize
-irreducible descriptions in general.
+general.  Before stating and proving it, we need to characterize irreducible
+descriptions in general, which depend on the notions of potency of a program
+and rank of a description.
 
@@
 
-  More formally,
-  \[
-    \pot(p,e)=\pot'(\bcast(p,e))\,,
-  \]
-  where~$\pot'$ is an auxiliary function that counts the number of
-  reachable~$\ceu{\EmitInt}$ expressions in the program resulting from the
-  broadcast of event~$e$ to~$p$.
+  More formally, $\pot(p,e)=\pot'(\bcast(p,e))$, where~$\pot'$ is an
+  auxiliary function that counts the number of reachable~$\ceu{\EmitInt}$
+  expressions in the program resulting from the broadcast of event~$e$
+  to~$p$.
 
@@
 
+The next two lemmas establish that a single application of~$\out$ or~$\nst$
+either preserves or decreases the rank of the input description.  All rank
+comparisons assume lexicographical order.
+
+\begin{lemma}\strut
+  \label{lem.rank-out}
+  \begin{enumerate}[(a)]
+  \item\label{lem.rank-out-push} If~$\delta\outpush\delta'$
+    then~$\rank(\delta)=\rank(\delta')$.
+  \item\label{lem.rank-out-pop} If~$\delta\outpop\delta'$
+    then~$\rank(\delta)>\rank(\delta')$.
+  \end{enumerate}
+\end{lemma}
+
+\begin{lemma}
+  \label{lem.rank-nst}
+  %%
+  If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
+\end{lemma}
+
 An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for

### 2018-02-23 03-semantics.tex
Seção 3.3.2, lemas de preservação do \rank

@@
   \item\label{def.x.pot.last} Otherwise, if none
-    of~\eqref{def.pot.first}--\eqref{def.pot.before-last} applies,
+    of~\eqref{def.x.pot.first}--\eqref{def.x.pot.before-last} applies,
     $\pot(\_)=0$.
@@
 
+We can now characterize irreducible descriptions in general.
+
 \begin{definition}
@@
 
+An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for
+arbitrary transitions~$\trans$.  Such descriptions cannot be advanced
+by~$\nst$, as it is nested-irreducible, and neither by~$\outpush$
+nor~$\outpop$, as the second coordinate of its rank is~0, which
+implies~$e=\nil$ and~$n=0$.
+
 The next two lemmas establish that a single application of~$\out$ or~$\nst$
 either preserves or decreases the rank of the input description.  All rank
-comparisons assume lexicographical order.
+comparisons assume lexicographical order, i.e., if~$\rank(\delta)=\<i,j>$
+and~$\rank(\delta')=\<i',j'>$ then~$\rank(\delta)>\rank(\delta')$ iff~$i>i'$
+or~$i=i$ and~$j>j'$.
+%%
+The proof of Lemma~\ref{lem.x.rank-out} follows directly from~\R{push}
+and~\R{pop} and from Definitions~\ref{def.x.pot} and~\ref{def.x.rank}.  The
+proof of Lemma~\ref{lem.x.rank-nst}, however, is by induction on the
+structure of~$\nst$ derivations.
 
 \begin{lemma}\strut
-  \label{lem.rank-out}
+  \label{lem.x.rank-out}
+  %%
   \begin{enumerate}[(a)]
@@
 \begin{lemma}
-  \label{lem.rank-nst}
+  \label{lem.x.rank-nst}
   %%
@@
 \end{lemma}
-
-An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for
-transitions~$\trans$ in general.  Such descriptions cannot be advanced
-by~$\nst$, as it is nested-irreducible, and neither by~$\outpush$
-nor~$\outpop$, as the second coordinate of its rank is~0, which
-implies~$e=\nil$ and~$n=0$.
-
-TODO: Lemas e teorema da terminação.  Sketch da prova.  Mencionar que as
-restrições sintáticas são fundamentais.
-
-% - there is always a possible transition until n=0
-%
-% every cannot restart itself
-%     - break disallowed
-%     - emit ignored
 

### 2018-02-23 06-appendix.tex
Seção 3.3.2, lemas de preservação do \rank

@@
   \item\label{lem.props-nst-i.a}
-    $\<\ceu{p_1;\,p_2},n,e>\nst[i]\<p_1';p_2,n,e'>$;
+    $\<\ceu{p_1;\,p_2},n,e>\nst[i]\<p_1';p_2,n,e'>$.
     %%
   \item\label{lem.props-nst-i.b}
-    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[i]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$;
+    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[i]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$.
     %%
   \item\label{lem.props-nst-i.c}
-    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$;
+    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$.
     %%
@@
   \item\label{lem.props-nst-i.e}
-    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>$;
+    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>$.
     %%
@@
 
-The next two lemmas establish that a single application of~$\out$ or~$\nst$
-either preserves or decreases the rank of the input description.  All rank
-comparisons assume lexicographical order.
+\begin{definition}
+  \label{def.H}
+  %%
+  A description~$\delta$ is \emph{irreducible} (in symbols, $\delta_\#$) iff
+  it is nested-irreducible and its~$\rank(\delta)$ is~$\<i,0>$, for
+  some~$i\ge0$.
+\end{definition}
 
@@
   \label{lem.rank-out}
+  %%
   \begin{enumerate}[(a)]
@@
 
-\begin{definition}
-  \label{def.H}
-  %%
-  A description~$\delta$ is \emph{irreducible} (in symbols, $\delta_\#$) iff
-  it is nested-irreducible and its~$\rank(\delta)$ is~$\<i,0>$, for
-  some~$i\ge0$.
-\end{definition}
-
-An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for
-general transitions~$\trans$.  Such descriptions cannot be advanced
-by~$\nst$, as it is nested-irreducible, and neither by~$\outpush$
-nor~$\outpop$, as the second coordinate of its rank is~0, which
-implies~$e=\nil$ and~$n=0$.
-
 \begin{theorem}[Termination]\label{thm.term}\strut\\

### 2018-02-23 03-semantics.tex
Minor changes
* 03-semantics.tex

@@
   %%
-  Then, by Lemma~\ref{lem.det-out} or~\ref{lem.det-nst}, depending on
+  Then, by Lemma~\ref{lem.x.det-out} or~\ref{lem.x.det-nst}, depending on
   whether the first transition is~$\out$ or~$\nst$ (it cannot be both),
@@
 
-The next lemma establishes that sequences of~$\nst$ transitions are
-well-behaved regarding the order of evaluation of composition members.  Its
+The next lemma establishes that sequences of~$\nst$ transitions behave
+as expected regarding the order of evaluation of composition members.  Its
 proof follows by induction on~$i$.
@@
   \begin{enumerate}[(a)]
-  \item\label{ass.x.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contain
+  \item\label{ass.x.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contains
     no occurrences of~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$,
@@
   \item\label{ass.x.syn-rest.loop} If~$p=\ceu{\Loop{p_1}}$ then all
-    execution paths of~$p_1$ contain at a matching~$\ceu{\Break}$ or
+    execution paths of~$p_1$ contain a matching~$\ceu{\Break}$ or
     an~$\ceu{\AwaitExt}$.  Consequently, for all~$n$, there are~$p_1'$
@@
 description.  The proof is by induction on the structure of programs
-(members of set~$P$) and uses Lemma~\ref{lem.props-nst-i} and
+(members of set~$P$) and uses Lemma~\ref{lem.x.props-nst-i} and
 Assumption~\ref{ass.x.syn-rest}.

### 2018-02-24 03-semantics.tex
Prova do Lema 3.13

@@
 
-We proceed to prove that a sufficiently long sequence of applications
-of~$\trans$ will eventually lead to an irreducible description, i.e., one
+In this section, we prove that any sufficiently long sequence of
+applications of~$\trans$ will lead to an irreducible description, i.e., one
 that cannot be modified by further transitions.  Before doing that, however,
@@
   %%
-  If~$\<p_1,n,e>\nst[i]\<p_1',n,e'>$ then, for any~$p_2$:
+  If~$\<p_1,n,e>\nst[i]\<p_1',n,e'>$, for any~$p_2$:
   \begin{enumerate}[(a)]
@@
   \end{enumerate}
+  \medskip
   If~$\<p_2,n,e>\nst[i]\<p_2',n,e'>$, for any~$p_1$ such
@@
     or~$\ceu{\Fin}$ expressions.  Consequently, for any~$n$,
-    $\<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>$.
+    \[
+      \<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>\,.
+    \]
     %%
@@
     an~$\ceu{\AwaitExt}$.  Consequently, for all~$n$, there are~$p_1'$
-    and~$e$ such that $\<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>$, where
-    $p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
+    and~$e$ such that
+    \[
+      \<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>\,,
+    \]
+    where $p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
   \end{enumerate}
@@
         \pot'(p_1)              &\text{if\enspace(\dag)}\\
-        \pot'(p_1)+\pot'(p_2)   &\text{otherwise}\\
+        \pot'(p_1)+\pot'(p_2)   &\text{otherwise,}\\
       \end{cases}
@@
 
-We can now characterize irreducible descriptions in general.
-
 \begin{definition}
@@
 An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for
-arbitrary transitions~$\trans$.  Such descriptions cannot be advanced
+transitions~$\trans$ in general.  Such descriptions cannot be advanced
 by~$\nst$, as it is nested-irreducible, and neither by~$\outpush$
@@
 proof of Lemma~\ref{lem.x.rank-nst}, however, is by induction on the
-structure of~$\nst$ derivations.
+structure of~$\nst$ derivations.  Since Lemma~\ref{lem.x.rank-nst} is of
+fundamental importance for the termination result, we detail the proof of
+one of its cases.
 
@@
 \end{lemma}
+\begin{proof}
+  By induction on the structure of~$\nst$ derivations.  Let
+  \[
+    \delta=\<p,n,e>
+    \quad\text{and}\quad
+    \delta'=\<p',n',e'>
+  \]
+  be descriptions with~$\rank(\delta)=\<i,j>$ and~$\rank(\delta')=\<i',j'>$
+  such that~$\delta\nst\delta'$.  Then there is a derivation~$\pi$ whose
+  conclusion is~$\delta\nst\delta'$, i.e.,
+  \[
+    \pi\Vdash\<p,n,e>\nst\<p',n',e'>\,.
+  \]
+  By definition of~$\nst$, $e=\nil$ and~$n=n'$.  Depending on the structure
+  of program~$p$, there are~11 distinct cases to be considered.  In each one
+  of them, we need to show that~$\rank(\delta)\ge\rank(\delta')$.
+
+  We proceed to prove the case where~$p=\ceu{{p_1}\AtLoop{p_2}}$, which
+  itself consists of three subcases.
+  \begin{case}
+    $p_1=\ceu{\Nop}$.
+    %%
+    Then~$\pi$ is an instance of~\R{loop-nop}.  i.e., its conclusion is
+    obtained by an application of this rule.  Hence~$p'=p_2$ and~$e'=\nil$.
+    Thus
+    \begin{align*}
+        \rank(\delta)&=\<\pot'(p_1)+\pot'(p_2),n>\\
+                     &\ge\<pot'(p_2),n>=\rank(\delta')\,.
+    \end{align*}
+  \end{case}
+  \begin{case}
+    $p_1=\ceu{\Break}$.
+    %%
+    Then~$\pi$ is an instance of~\R{loop-brk}.
+    Hence~$p'=p_1$ and~$e'=\nil$.  Thus
+    \[
+      \rank(\delta)=\rank(\delta')=\<0,n>\,.
+    \]
+  \end{case}
+  \begin{case}
+    $p_1\ne\ceu{\Nop},\ceu{\Break}$.
+    %%
+    Then~$\pi$ is an instance of~\R{loop-adv}.  Hence there is a
+    derivation~$\pi'$ such that
+    \begin{equation}
+      \label{lem.x.rank-nst-eq1}
+      \pi'\Vdash\<p_1,n,\nil>\nst\<p_1',n,e_1'>\,,
+    \end{equation}
+    for some~$p_1'$ and~$e_1'$.  Thus~$p'=\ceu{p_1'\AtLoop{\,p_2}}$
+    and~$e'=e_1'$.  There are two subcases.
+    \begin{subcase}
+      \label{lem.x.rank-nst-case1}
+      $\pot'(p)=\pot'(p_1)$.
+      %%
+      Then, by Definition~\ref{def.pot}, every execution path of~$p_1$
+      contains a (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
+      A single~$\nst$ cannot terminate the loop, since~$p_1\ne\ceu{\Break}$,
+      nor can it consume an~$\ceu{\AwaitExt}$, which means that all
+      execution paths in~$p_1'$ still contain a~$\ceu{\Break}$
+      or~$\ceu{\AwaitExt}$.  Hence~$\pot'(p')=\pot'(p_1')$.
+
+      There are two further subcases depending on event~$e'$.
+      \begin{subsubcase}
+        $e'=\nil$.
+        %%
+        Then $\rank(\delta)=\<pot'(p_1),n>$ and
+        $\rank(\delta')=\<pot'(p_1'),n>$.  From~\eqref{lem.x.rank-nst-eq1},
+        by the induction hypothesis, $\pot'(p_1)\ge\pot(p_1')$.  Thus
+        \[
+          \rank(\delta)\ge\rank(\delta')\,.
+        \]
+      \end{subsubcase}
+      \begin{subsubcase}
+        $e'\ne\nil$.
+        %%
+        Then~$\pi'$ contains one application of~\R{emit-int} which consumes
+        one~$\ceu{\EmitInt(e')}$ expression from~$p_1$ and which implies
+        $\pot'(p_1)>\pot'(p_1')$.  Thus
+        \begin{align*}
+          \rank(\delta)&=\<\pot'(p_1),n>\\
+                       &>\<\pot'(p_1'),n+1>=\rank(\delta')\,.
+        \end{align*}
+      \end{subsubcase}
+    \end{subcase}
+    \begin{subsubcase}
+      $\pot'(p)=\pot'(p_1)+\pot'(p_2)$.
+      %%
+      Then some execution path in~$p_1$ does not contain a
+      (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
+      Since~$p_1\ne\ceu{\Nop}$, a single~$\nst$ cannot restart the loop,
+      which means that~$p_1'$ still contain some execution path in which
+      a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ does not occur.
+      Hence~$\pot'(p')=\pot'(p_1')+\pot'(p_2)$.  The rest of this proof is
+      similar to that of Case~\ref{lem.x.rank-nst-case1}.\qedhere
+    \end{subsubcase}
+  \end{case}
+\end{proof}
 

### 2018-02-24 06-appendix.tex
Prova do Lema 3.13

@@
   Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$, $\rank(\delta)=\<i,j>$,
-  and~$\rank(\delta')=\<i',j'>$.  By the hypothesis of the theorem, there is
+  and~$\rank(\delta')=\<i',j'>$.  By the hypothesis of the lemma, there is
   a derivation~$\pi$ such that
@@
         %%
-        Then~$\pi'$ contain one application of~\R{emit-int}, which consumes
+        Then~$\pi'$ contains one application of~\R{emit-int}, which consumes
         one~$\ceu{\EmitInt(e')}$ expression from~$p_1$ and implies
@@
       \]
-      for some~$p_1'$ and~$e_1'$.  Thus~$p'=p_1';\,p_2$ and~$e'=e_1'$.
+      for some~$p_1'$ and~$e_1'$.  Thus~$p'=\ceu{p_1'\AtLoop{\,p_2}}$
+      and~$e'=e_1'$.
       %%
@@
         %%
-        Then all execution paths of~$p_1$ contain a~$\ceu{\Break}$
-        or~$\ceu{\AwaitExt}$ expression.  A single~$\out$ cannot terminate
+        Then every execution path of~$p_1$ contains a~$\ceu{\Break}$
+        or~$\ceu{\AwaitExt}$ expression.  A single~$\nst$ cannot terminate
         the loop, since~$p_1\ne\ceu{\Break}$, nor can it consume
@@
         or~$\ceu{\AwaitExt}$ expression.  Since~$p_1\ne\ceu{\Nop}$, a
-        single~$\out$ cannot restart the loop, which means that~$p_1'$ still
+        single~$\nst$ cannot restart the loop, which means that~$p_1'$ still
         contain some execution path in which a~$\ceu{\Break}$

### 2018-02-24 03-semantics.tex
Corrije e Comprime prova do lema da terminação em 3.3.2

@@
   \end{enumerate}
-  \medskip
+  \smallskip
   If~$\<p_2,n,e>\nst[i]\<p_2',n,e'>$, for any~$p_1$ such
@@
     or~$\ceu{\Fin}$ expressions.  Consequently, for any~$n$,
-    \[
-      \<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>\,.
-    \]
+    $\<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>$.
+    %%
+    % \[
+    %   \<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>\,.
+    % \]
     %%
@@
     and~$e$ such that
-    \[
-      \<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>\,,
-    \]
+    $\<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>$,
+    %%
+    % \[
+    %   \<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>\,,
+    % \]
+    %%
     where $p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
@@
 
-The main result of this section, Theorem~\ref{?}, is similar to
+The main result of this section, Theorem~\ref{thm.x.term}, is similar to
 Theorem~\ref{thm.x.term-nst-*} but applies to transitions~$\trans$ in
 general.  Before stating and proving it, we need to characterize irreducible
-descriptions in general, which depend on the notions of potency of a program
-and rank of a description.
+descriptions in general.  This characterization depends on the notions of
+potency and rank.
 
@@
   conclusion is~$\delta\nst\delta'$, i.e.,
-  \[
-    \pi\Vdash\<p,n,e>\nst\<p',n',e'>\,.
-  \]
+  $\pi\Vdash\<p,n,e>\nst\<p',n',e'>$.
+  %%
+  % \[
+  %   \pi\Vdash\<p,n,e>\nst\<p',n',e'>\,.
+  % \]
+  %%
   By definition of~$\nst$, $e=\nil$ and~$n=n'$.  Depending on the structure
@@
     \begin{align*}
-        \rank(\delta)&=\<\pot'(p_1)+\pot'(p_2),n>\\
-                     &\ge\<pot'(p_2),n>=\rank(\delta')\,.
+        \rank(\delta)=\<\pot'(p_1)+\pot'(p_2),n>
+                     \ge\<pot'(p_2),n>=\rank(\delta')\,.
     \end{align*}
@@
     Hence~$p'=p_1$ and~$e'=\nil$.  Thus
-    \[
-      \rank(\delta)=\rank(\delta')=\<0,n>\,.
-    \]
+    $\rank(\delta)=\rank(\delta')=\<0,n>$.
+    %%
+    % \[
+    %   \rank(\delta)=\rank(\delta')=\<0,n>\,.
+    % \]
+    %%
   \end{case}
@@
         by the induction hypothesis, $\pot'(p_1)\ge\pot(p_1')$.  Thus
-        \[
-          \rank(\delta)\ge\rank(\delta')\,.
-        \]
+        $\rank(\delta)\ge\rank(\delta')$.
+        %%
+        % \[
+        %   \rank(\delta)\ge\rank(\delta')\,.
+        % \]
+        %%
       \end{subsubcase}
@@
         $\pot'(p_1)>\pot'(p_1')$.  Thus
-        \begin{align*}
-          \rank(\delta)&=\<\pot'(p_1),n>\\
-                       &>\<\pot'(p_1'),n+1>=\rank(\delta')\,.
-        \end{align*}
+        \[
+          \hskip1em\rank(\delta)\!=\!\<\pot'(p_1),n>
+          >\<\pot'(p_1'),n+1>\!=\!\rank(\delta')\,.
+        \]
+        %%
+        % \begin{align*}
+        %   \rank(\delta)&=\<\pot'(p_1),n>\\
+        %                &>\<\pot'(p_1'),n+1>=\rank(\delta')\,.
+        % \end{align*}
+        %%
       \end{subsubcase}
@@
 
-\subsubsection*{Memory Bounds}
+The next theorem is a generalization of Lemma~\ref{lem.x.rank-nst}
+for~$\nst[*]$ transitions.  Its proof is by induction on~$i$.
+
+\begin{theorem}
+  \label{thm.x.rank-nst-*}
+  %%
+  If~$\delta\nst[*]\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
+\end{theorem}
+
+We now state and prove the main result of this section,
+Theorem~\ref{thm.x.term}, viz., the termination theorem for~$\trans[*]$.
+The idea of the proof is that a series of~$\nst$ or~$\out$ transitions
+eventually decrease the rank of the current description until an irreducible
+description is reached.  This irreducible description is the final result of
+the reaction.
+%%
+\begin{theorem}[Termination]\label{thm.x.term}\strut\\
+  %%
+  For any~$\delta$, there is a~$\delta'_\#$ such
+  that~$\delta\trans[*]\delta'_\#$.
+\end{theorem}
+\begin{proof}
+  By lexicographic induction on~$\rank(\delta)$.  Let~$\delta=\<p,n,e>$ and
+  $\rank(\delta)=\<i,j>$.
+  \begin{basis}
+    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$, as~$j=0$
+    implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be
+    applied).  There are two possibilities:  either~$\delta$ is
+    nested-irreducible or it is not.  In the first case, the theorem is
+    trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
+    nested-irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
+    $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
+    Theorem~\ref{thm.x.rank-nst-*},
+    \[
+      \<i,j>=\<0,0>\ge\rank(\delta')\,,
+    \]
+    which implies~$\rank(\delta')=\<0,0>$.
+  \end{basis}
+  \begin{induction}
+    Let~$\<i,j>\ne\<0,0>$.
+    %%
+    There are two subcases depending on whether or not $\delta$~is
+    nested-irreducible.
+    \begin{case}
+      \label{thm.x.term.Hnst}
+      $\delta$~is nested-irreducible.
+      %%
+      If~$j=0$ then, by Definition~\ref{def.x.H}, $\delta_\#$, and
+      thus~$\delta\trans[0]\delta_\#$.  Suppose that is not the case, i.e.,
+      that $j>0$.  Then there are two subcases.
+      \begin{subcase}
+        \label{thm.x.term.Hnst-j>0-nonnil}
+        $e\ne\nil$.
+        %%
+        Then, by~\R{out} and by Theorem~\ref{thm.x.term-nst-*}, there
+        are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
+        $\delta\outpush\delta_1'\nst[*]\delta'_\Hnst$.  Thus, by
+        Lemma~\ref{lem.x.rank-out} and by Theorem~\ref{thm.x.rank-nst-*},
+        \begin{align*}
+          \rank(\delta)=\rank(\delta_1')=\<i,j>
+                                        \ge\rank(\delta')=\<i',j'>\,.
+        \end{align*}
+        If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof is
+        similar to that of Case~\ref{thm.x.term.Hnst-j>0-nil} below.
+        Otherwise, if~$e'\ne\nil$ then $i>i'$, since an~$\ceu{\EmitInt(e')}$
+        was consumed by the nested transitions.  Thus,
+        $\rank(\delta)>\rank(\delta')$.  By the induction hypothesis,
+        $\delta'\trans[*]\delta''_\#$, for some~$\delta''_\#$.  Therefore,
+        $\delta\trans[*]\delta''_\#$.
+      \end{subcase}
+      \begin{subcase}
+        \label{thm.x.term.Hnst-j>0-nil}
+        $e=\nil$.
+        %%
+        Then, since~$j>0$, $\delta\outpop\delta'$, for some~$\delta''$.  By
+        item~\eqref{lem.rank-out-pop} of Lemma~\ref{lem.rank-out},
+        $\rank(\delta)>\rank(\delta')$.  Hence, by the induction hypothesis,
+        there is a~$\delta''_\#$ such that~$\delta'\trans[*]\delta''_\#$.
+        Therefore, $\delta\trans[*]\delta''_\#$.
+      \end{subcase}
+    \end{case}
+    \begin{case}
+      $\delta$~is not nested-irreducible.
+      %%
+      Then~$e=\nil$ and, by Theorems~\ref{thm.x.rank-nst-*}
+      and~\ref{thm.x.rank-nst-*}, there is a~$\delta'_\Hnst$ such
+      that~$\delta\nst[*]\delta'_\Hnst$
+      with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
+      is similar to that of Case~\ref{thm.x.term.Hnst} above.\qedhere
+    \end{case}
+  \end{induction}
+\end{proof}
+
+\subsubsection*{Memory bounds}
 

### 2018-02-24 06-appendix.tex
Corrije e Comprime prova do lema da terminação em 3.3.2

@@
           \]
-          Thus, by Lemma by item~\eqref{lem.rank-out-push} of
+          Thus, by item~\eqref{lem.rank-out-push} of
           Lemma~\ref{lem.rank-out} and by Theorem~\ref{thm.rank-nst-*},
@@
       %%
-      There are two subcases.
-      \begin{subcase}
-        $e\ne\nil$.  Similar to Case~\ref{thm.term.Hnst-j>0-nonnil}.
-      \end{subcase}
-      \begin{subcase}
-        $e=\nil$.  Then, by Theorems~\ref{thm.rank-nst-*}
-        and~\ref{thm.rank-nst-*} there is a~$\delta'_\Hnst$ such
+      Then~$e=\nil$ and, by Theorems~\ref{thm.rank-nst-*}
+      and~\ref{thm.rank-nst-*}, there is a~$\delta'_\Hnst$ such
       that~$\delta\nst[*]\delta'_\Hnst$
@@
       is similar to that of Case~\ref{thm.term.Hnst}.\qedhere
-      \end{subcase}
     \end{case}

### 2018-02-24 03-semantics.tex
Conserta a legenda da Figura 4 na Seção 3

@@
       \clear(\_)
-      &=\bot\enspace
+      &=\xi\enspace
         (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
@@
   converting~$\ceu{\protect\AwaitExt}$ and~$\ceu{\protect\AwaitInt}$
-  to~$\ceu{\protect\Nop}$ expressions, and by unwinding $\ceu{\protect\Every}$
+  to~$\ceu{\protect\Nop}$, and by unwinding $\ceu{\protect\Every}$
   expressions.
   %%
-  \enspace(ii)~Predicate~$\isblocked$ is true only if all branches in parallel
-  are blocked waiting for events, for finalization clauses, or for certain
+  \space(ii)~Predicate~$\isblocked$ is true only if all branches in parallel
+  are blocked waiting for events, finalization clauses, or certain
   stack levels.
   %%
-  \enspace(iii)~Function~$\clear$ extracts~$\ceu{\protect\Fin}$ expressions in
+  \space(iii)~Function~$\clear$ extracts~$\ceu{\protect\Fin}$ expressions in
   parallel and put their bodies in sequence.
+  %%
+  In~(i), (ii), and~(iii),~``$\_$'' denotes the omitted cases and~``$\xi$''
+  denotes the empty string.
+  %%
 }

### 2018-02-24 03-semantics.tex
Renomeia apêndice
Adiciona parágrafo sobre memory bound, Seção 3.3.3

@@
   &\AxiomC{$n>0$}
-  \AxiomC{$\ceu{p=\Nop}\vee\isblocked(p,n)$}
+  \AxiomC{$p=\ceu{\Nop,\Break}\vee\isblocked(p,n)$}
   \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
@@
     Theorem~\ref{thm.x.rank-nst-*},
-    \[
-      \<i,j>=\<0,0>\ge\rank(\delta')\,,
-    \]
+    $\<i,j>=\<0,0>\ge\rank(\delta')$,
+    %%
+    % \[
+    %   \<i,j>=\<0,0>\ge\rank(\delta')\,,
+    % \]
+    %%
     which implies~$\rank(\delta')=\<0,0>$.
@@
 
-\subsubsection*{Memory bounds}
-
-TODO: Acho que sai direto do~$\pot(p)$.
-
-- program is finite
-- lexical scope
-    - no heap allocation
-- no code reentrancy
-    - reexecution only due to loops
-    - loop reuse nested vars
--
+\subsubsection{Memory bound}
+
+As \CEU has no mechanism for heap allocation, unbounded iteration, or
+general recursion, the maximum memory usage of a given \CEU program is
+determined solely by the length of its code, the number of variables it
+uses, and the size of the event stack that it requires to run.  The code
+length and the number of variables used are easily determined by code
+inspection.  The maximum size of the event stack during a reaction of
+program~$p$ to external event~$e$ corresponds to~$\pot(p,e)$, i.e., to the
+maximum number of internal events that~$p$ may emit in a reaction to~$e$.
+If~$p$ may react to external events~$e_1$, \dots,~$e_n$ then, in the worst
+case, its event stack will need to
+store~$\max\{\pot(p,e_1),\dots,\pot(p,e_n)\}$ events.
+
+% - program is finite
+% - lexical scope
+%     - no heap allocation
+% - no code reentrancy
+%     - reexecution only due to loops
+%     - loop reuse nested vars

### 2018-02-24 01-intro.tex
Typo na introdução

@@
     \item Finalization mechanism for lines of execution, which makes abortion
-          safe with regards to external resources.
+          safe with regard to external resources.
 \end{itemize}
 
-In this work, we present a formal small-step structural operational semantics
-for \CEU and proofs that it enforces memory-bounded, deterministic, and
-terminating reactions to the environment, i.e., that for a given arbitrary
-timeline of input events, multiple executions of the same program always react
-in bounded time and arrive at the same final finite memory state.
+In this work, we present a formal small-step structural operational
+semantics for \CEU and prove that it enforces memory-bounded, deterministic,
+and terminating reactions to the environment, i.e., that for a given
+arbitrary timeline of input events, multiple executions of the same program
+always react in bounded time and arrive at the same final finite memory
+state.
 

### 2018-02-24 02-ceu.tex
Consertos pequenos na Seção 3

@@
 
-Listing~\ref{lst.syntax} shows a compact reference of~\CEU:
+Listing~\ref{lst.syntax} shows a compact reference of~\CEU.
 
@@
 stack-based processing, can be used to implement a limited form of subroutines,
-as illustrated in Listing~\ref{lst.sub}:
+as illustrated in Listing~\ref{lst.sub}.
 

### 2018-02-24 03-semantics.tex
Consertos pequenos na Seção 3

@@
 
-In this section, we introduce a reduced syntax of \CEU and propose an
-operational semantics to formally describe the behavior of programs.
+In this section, we introduce a reduced syntax for \CEU and propose an
+operational semantics to formally describe the behavior of its programs.
 We describe a small synchronous kernel highlighting the peculiarities of \CEU,
@@
 %
-We use a set of small-step operational rules~\cite{Plotkin-G-D-1981}, which
+We use a set of small-step operational rules~\cite{Plotkin-G-D-1981} which
 are designed in such a way that at most one transition is possible at any
@@
 \]
-where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are non-negative
+where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are nonnegative
 integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
-the events emitted before and after the transition (both being possibly the
+the events emitted before and after the transition (both possibly being the
 empty event~$\nil$).
@@
 We will refer to the triples on the left-hand and right-hand sides of
-symbol~$\to$ as \emph{descriptions} (denoted~$\delta$).  The triple on the
-left-hand side of symbol~$\to$ is called the \emph{input description}, and
+symbol~$\trans$ as \emph{descriptions} (denoted~$\delta$).  The triple on the
+left-hand side of symbol~$\trans$ is called the \emph{input description}, and
 the triple on its right-hand side is called the \emph{output description}.
@@
 
-The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions which only
-apply to the program as a whole, and are the only to manipulate the stack
-level:
+The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions that
+apply to the program as a whole.  These are the only rules that manipulate
+the stack level.
 \begin{align*}
@@
 
-The rules for~$\ceu{\And}$ and~$\ceu{\Or}$ compositions make sure that their
-left branches always transition before their right branches:
+The rules for~$\ceu{\And}$ and~$\ceu{\Or}$ compositions ensure that the
+left-hand side of compositions always transition before their right-hand
+side:
 %%
@@
 
-  More formally, $\pot(p,e)=\pot'(\bcast(p,e))$, where~$\pot'$ is an
+  More formally, $\pot(p,e)=\pot'(\bcast(p,e))$ where~$\pot'$ is an
   auxiliary function that counts the number of reachable~$\ceu{\EmitInt}$
@@
 An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for
-transitions~$\trans$ in general.  Such descriptions cannot be advanced
+transitions~$\trans$ in general.  Such description cannot be advanced
 by~$\nst$, as it is nested-irreducible, and neither by~$\outpush$
@@
 either preserves or decreases the rank of the input description.  All rank
-comparisons assume lexicographical order, i.e., if~$\rank(\delta)=\<i,j>$
+comparisons assume lexicographic order, i.e., if~$\rank(\delta)=\<i,j>$
 and~$\rank(\delta')=\<i',j'>$ then~$\rank(\delta)>\rank(\delta')$ iff~$i>i'$
@@
     for some~$p_1'$ and~$e_1'$.  Thus~$p'=\ceu{p_1'\AtLoop{\,p_2}}$
-    and~$e'=e_1'$.  There are two subcases.
+    and~$e'=e_1'$.
+
+    There are two subcases.
     \begin{subcase}
@@
       A single~$\nst$ cannot terminate the loop, since~$p_1\ne\ceu{\Break}$,
-      nor can it consume an~$\ceu{\AwaitExt}$, which means that all
+      nor it can consume an~$\ceu{\AwaitExt}$, which means that all
       execution paths in~$p_1'$ still contain a~$\ceu{\Break}$
@@
 
-      There are two further subcases depending on event~$e'$.
+      There are two subcases.
       \begin{subsubcase}
@@
         %%
-        Then, since~$j>0$, $\delta\outpop\delta'$, for some~$\delta''$.  By
+        Then, since~$j>0$, $\delta\outpop\delta'$, for some~$\delta'$.  By
         item~\eqref{lem.rank-out-pop} of Lemma~\ref{lem.rank-out},
@@
 program~$p$ to external event~$e$ corresponds to~$\pot(p,e)$, i.e., to the
-maximum number of internal events that~$p$ may emit in a reaction to~$e$.
+maximum number of internal events that~$p$ may emit in reaction to~$e$.
 If~$p$ may react to external events~$e_1$, \dots,~$e_n$ then, in the worst

### 2018-02-24 03-semantics.tex
Pequenos consertos Seção 3

@@
 Theorem~\ref{thm.x.term}, viz., the termination theorem for~$\trans[*]$.
-The idea of the proof is that a series of~$\nst$ or~$\out$ transitions
-eventually decrease the rank of the current description until an irreducible
-description is reached.  This irreducible description is the final result of
-the reaction.
+The idea of the proof is that a sufficiently large sequence of~$\nst$
+and~$\out$ transitions eventually decrease the rank of the current
+description until an irreducible description is reached.  This irreducible
+description is the final result of the reaction.
 %%

### 2018-02-24 05-conclusion.tex
Coloca a conclusão no passado

@@
 
-We propose a small-step structural operational semantics for \CEU and a proof
-that reactions are deterministic, terminate in finite time, and use bounded
-memory, i.e., that for a given arbitrary timeline of input events, multiple
-executions of the same program always react in bounded time and arrive at the
-same final finite memory state.
+In this paper, we proposed a small-step structural operational semantics for
+\CEU and proved that under it reactions are deterministic, terminate in
+finite time, and use bounded memory, i.e., that for a given arbitrary
+timeline of input events, multiple executions of the same program always
+react in bounded time and arrive at the same final finite memory state.

### 2018-02-24 06-appendix.tex
Renomeia apêndice
Fix Lemma A.13 Case 2
* 06-appendix.tex
Change n by i in Theorem A.14
* 06-appendix.tex
Uniformiza enunciados da Seção 3 e Apêndice

@@
 \appendix
-\section{Proofs}
+\section{Appendix: Detailed Proofs (optional)}
 \label{sec.proofs}
@@
 
-\begin{theorem}[Determinism]\strut
-  \label{thm.det}
+\begin{theorem}[Determinism]\label{thm.det}\strut\\
   %%
@@
   \end{enumerate}
+  \smallskip
   If~$\<p_2,n,e>\nst[i]\<p_2',n,e'>$, for any~$p_1$ such
@@
 
-\begin{assumption}[Syntactical restrictions]\strut
+\begin{assumption}[Syntactic restrictions]\strut
   \label{ass.syn-rest}
@@
 
-  The auxiliary function~$\pot'$ is defined by the following clauses:
+  The auxiliary function~$\pot'$ is defined by
+  clauses~\label{def.pot.first}--\label{def.pot.last}:
   \begin{enumerate}[(a)]
@@
     %%
-  \item If~$p_1,p_2\ne\ceu{\Break}$,
+  \item\label{def.pot.before-last} If~$p_1,p_2\ne\ceu{\Break}$,
     \[
@@
     %%
-  \item Otherwise, if none of~\eqref{def.pot.first}--\eqref{def.pot.last}
-    applies, $\pot(\_)=0$.
+  \item Otherwise, if none
+    of~\eqref{def.pot.first}--\eqref{def.pot.before-last} applies,
+    $\pot(\_)=0$.
   \end{enumerate}
@@
     %%
-    Then~$\pi$ is an instance of~\R{emit-int}.  Hence~$p'=\ceu{\Nop}$
+    Then~$\pi$ is an instance of~\R{emit-int}.  Hence~$p'=\ceu{\CanRun}$
     and~$e'=e_1\ne\nil$.
@@
 \begin{proof}
-  If~$\delta\nst[*]\delta'$ then~$\delta\nst[n]\delta'$, for some~$i$.  We
+  If~$\delta\nst[*]\delta'$ then~$\delta\nst[i]\delta'$, for some~$i$.  We
   proceed by induction on~$i$.
@@
       %%
-      Then~$e=\nil$ and, by Theorems~\ref{thm.rank-nst-*}
+      Then~$e=\nil$ and, by Theorems~\ref{thm.term-nst-*}
       and~\ref{thm.rank-nst-*}, there is a~$\delta'_\Hnst$ such

### 2018-03-19 03-semantics.tex
Typos levantados pelos revisores

@@
 
-At the beginning of the reaction, an external event is emitted, which
+At the beginning of a reaction, an external event is emitted, which
 will trigger rule \R{push}, which will immediately raise the stack level
@@
       %%
-      \bcast(\ceu{p;\,q})
+      \bcast(\ceu{p;\,q},e)
       &=\ceu{\bcast(p,e);\,q}\\[-1\jot]
@@
 inspection of rules~\R{push} and~\R{pop}.  The proof of
-Lemma~\ref{lem.x.det-out}, however, requires an induction on the structure
+Lemma~\ref{lem.x.det-nst}, however, requires an induction on the structure
 of the derivation trees produced by the rules for~$\nst$.  Both lemmas are
@@
       %%
-      Then~$e=\nil$ and, by Theorems~\ref{thm.x.rank-nst-*}
+      Then~$e=\nil$ and, by Theorem~\ref{thm.x.term-nst-*}
       and~\ref{thm.x.rank-nst-*}, there is a~$\delta'_\Hnst$ such

### 2018-03-19 index.tex
Typos levantados pelos revisores

@@
 %
-It focus on control-flow safety features, such as safe shared-memory
+It focuses on control-flow safety features, such as safe shared-memory
 concurrency and safe abortion of lines of execution, while enforcing

### 2018-03-19 lctes18-reviews-11-rebuttal.txt
Add reviews and rebuttal.

We would like to thank the reviewers for the thoughtful comments, questions,
and criticism.

In a final version, we will address the main concern among reviews regarding
the paper motivations.
We will discuss why having a formal semantics of the language is important, and
how it affected our own implementation.
We will also discuss why the main results of the paper (memory boundedness,
determinism, and termination) are particularly important in the context of
embedded systems.

We will accommodate all requested clarifications and corrections in a final
version of the paper (more intuitive/complete descriptions of the semantics,
compare with some specific work, etc.)
Some of the comments are also addressed individually as follows.

Review #11A
===========

> Concerning the language itself, it is clear indeed that the strong
> restrictions <...> help to ensure the properties which are proved.
> However, even if it is the case for other languages <...>, the fact of
> linking the semantics of the language to what could be a particular
> implementation (scheduling in lexical order) could be considered somewhat
> questionable. While the language uses, in principle, parallel constructions,
> the parallelism seems in any case very largely limited by such a choice.

Considering the constant and low-level interactions with the underlying
architecture in embedded systems (e.g., direct port manipulation), we believe
that lexical scheduling must be part of the language specification as a
pragmatic design decision.

However, Céu has compile-time mechanisms to detect trails in parallel that
never share resources (`[Section 2.2]`).
In such cases, an implementation could take advantage of real parallelism.

Review #11B
===========

> - There are no semantics given for certain language features, e.g. what
>   happens when a call to a C function is made using the underscore syntax?
>   Also, it’s mentioned that side effects and system calls behave like in
>   conventional imperative languages, and it would be good to have a reference
>   for what this behaviour is.

In this context, C calls and system calls are synonymous.
The behavior for calls and side effects, all encapsulated in the `mem(id)`
primitive, is to modify a memory location in "instantaneous" time.
They are opaque operations that have no effect on the application execution
flow (e.g., they cannot block or abort execution).

> Are there approaches to this distributed synchronous programming problem
> other than the event-driven model?

Most synchronous languages have a clock/tick-based model of time
(`[Section 4]`).
These models are duals since we can accommodate a single event in a single tick
and vice versa.

Review #11C
===========

> How does the focus on the control aspects of the language (and leaving out
> side-effects and system calls) affect the semantics and the conclusions which
> can be drawn from analysing it?

The focus on control does not affect the logical semantics, abstract program
behavior (independent of real-time considerations).
Any modeling of side effects would have to assume, or somehow guarantee, not to
break the synchronous model (e.g., that system calls are non-blocking).
Such assumptions or guarantees would lead us to the same result.

Review #11D
===========

> How is determinism ensured across different runs for the example in fig 2a,
> if events A and B can occur in any order?

The program behaves deterministically in the sense that the same inputs (event
history) will necessarily produce the same output.

> Overall, the paper gives the impression of proposing CEU, <...>. Is the
> contribution the actual formalization of these semantics?

Yes, the contribution is the actual formalization and proofs of memory
boundedness, determinism, and termination.

> Otherwise, how was synchronization between traits implemented previously in
> CEU to ensure correctness?

From the very beginning, the language has been designed (and implemented) with
these safety properties in mind.
The proposed formalization of this paper has helped us to validate the design
and has also influenced the current implementation.

> Can await-emit lead to a deadlock if called from two concurrent trails,
> waiting on each other?

No.
An `emit` remains pending until its triggered internal reaction completes.
If it is pending it cannot be in an `await`, so a new `emit` in the triggered
reaction cannot retrigger the original `emit`.
Stack-based internal events prevents code reentrancy that would occur with a
queue-based policy.

> How does the compiler ensure that no non-blocking operations are called from
> CEU? Is there a black list of library calls?

The actual implementation of Céu supports a command-line option that accepts a
white list of library calls.
If a program tries to use a function not in this list, the compiler raises an
error.

### 2018-03-19 lctes18-reviews-11-rebuttal.txt
Sugestões do Roberto para o rebuttal.

@@
 
-We will accommodate all requested clarifications and corrections in a final
-version of the paper (more intuitive/complete descriptions of the semantics,
-compare with some specific work, etc.)
+We will also accommodate all requested clarifications and corrections (e.g.,
+more intuitive/complete descriptions of the semantics, compare with some
+specific work, etc.)
 Some of the comments are also addressed individually as follows.
@@
 history) will necessarily produce the same output.
+So, a history where `A` occurs before `B` is different from a history where `B`
+occurs first.
+Different inputs can produce different outputs.
 
+> It is not clear how many of these semantics exist in the implementation of
+> CEU presented in TECS'17. <...>.
 > Overall, the paper gives the impression of proposing CEU, <...>. Is the
@@
 boundedness, determinism, and termination.
+Céu itself has already been presented in other publications such as TECS'17.
 

### 2018-03-22 01-intro.tex
Revisões do artigo.

@@
 \CEU inherits the synchronous and imperative mindset of Esterel but adopts a
-simpler semantics with fine-grained execution control~\cite{ceu.tecs17}.
+simpler semantics with fine-grained execution control.
 %
-The list that follows summarizes the semantic peculiarities of \CEU:
+The list that follows summarizes the semantic peculiarities of
+\CEU~\cite{ceu.tecs17}:
 %
@@
 state.
+%
+Conceiving a formal semantics for \CEU leads to a number of capabilities and
+outcomes as follows:
 
-% TODO: \fs{Descrever seções.}
+\begin{enumerate}
+\item
+    Understanding, finding corner cases, and setting compromises.
+    After the semantics was complete and discussed in extent, we found a bug in
+    the order of execution of statements in a specific case.
+\item
+    Explaining core aspects of the language in a reduced, precise, and
+    unambiguous way.
+    This is particularly important when comparing languages that are similar in
+    the surface (e.g., \CEU and Esterel).
+\item
+    Implementing the language.
+    Small-step operational semantics typically describe abstract machines that
+    are close to concrete implementations.
+    The current implementation of the \CEU scheduler is based on the formal
+    semantics.
+\item
+    Proving properties for particular or all programs in the language.
+    We prove that all programs in \CEU are memory bounded, deterministic, and
+    react in finite time.
+\end{enumerate}
+
+The last item is particularly important in the context of embedded systems:
+
+\begin{description}
+\item[Memory Boundedness:]
+At compile time, we can ensure that the program fits in the device's restricted
+memory and that it will not grow unexpectedly during runtime.
+\item[Deterministic Execution:]
+We can simulate an entire program execution providing an input history with the
+guarantee that it will always behave the same.
+This can be done in negligible time in a controlled development environment and
+with multiple users (e.g., multiple developers each on its desktop) before
+deploying the application in the actual device.
+\item[Terminating Reactions:]
+Real-time applications must guarantee responses within specified deadlines.
+A terminating semantics enforces upper bounds for all reactions and guarantees
+that programs always progress with the time.
+In addition, it also ensures that programs always reach an idle state amenable
+to entering in standby mode, which is can improve the autonomy of
+battery-powered devices.
+\end{description}

### 2018-03-22 03-semantics.tex
Revisões do artigo.

@@
 For the sake of simplicity, we focus on the control aspects of the language,
-leaving out side-effects and system calls (which behave like in conventional
-imperative languages).
+and encapsulate all side effects and system calls into a single abstract memory
+operation.
+Memory operations are opaque and behave like in conventional imperative
+languages (i.e., reads and writes), having no effect on the application
+execution flow (e.g., they cannot block or abort execution).
 

### 2018-03-22 lctes18-reviews-11-rebuttal.txt
Revisões oficiais.

@@
 more intuitive/complete descriptions of the semantics, compare with some
-specific work, etc.)
+specific work, etc.).
 Some of the comments are also addressed individually as follows.

### 2018-04-10 07-artifact.tex
Add artifact evaluation

\appendix
\section{Artifact appendix}

%Submission and reviewing guidelines and methodology:
%{\em http://cTuning.org/ae/submission.html}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Abstract}

Our artifact includes an open-source implementation of the programming language
\CEU.
The implementation is based on and should conform with the formal semantics
presented in this paper.
The artifact also includes an executable script with over 3500 test cases of
valid and invalid programs in \CEU.
The script is customizable and allows to create new tests providing inputs and
expected outputs.
The evaluation platform is a Linux/Intel with Lua-5.3 and GCC-7.2 installed.

\subsection{Artifact check-list (meta-information)}

{\em Obligatory. Use just a few informal keywords in all fields applicable to your artifacts
and remove the rest. This information is needed to find appropriate reviewers and gradually 
unify artifact meta information in Digital Libraries.}

{\small
\begin{itemize}
  %\item {\bf Algorithm: }
  %\item {\bf Program: }
  \item {\bf Compilation: } The output of the \CEU compiler is a C program that requires a C compiler (e.g., GCC-7.2).
  \item {\bf Transformations: } The compiler of \CEU is a Lua program (Lua-5.3) that generates a C program.
  %\item {\bf Binary: }
  \item {\bf Data set: } The data set is a set of program test cases included in the language distribution.
  \item {\bf Run-time environment: } Linux with Lua-5.3 and GCC-7.2. Root access is not required.
  \item {\bf Hardware: } Off-the-shelf Intel machine.
  %\item {\bf Run-time state: }
  \item {\bf Execution: } 5-10 minutes for the full test.
  \item {\bf Output: } Console output: Success (termination) or Fail (abortion).
  \item {\bf Experiments: } Manual steps by user.
  %\item {\bf Workflow frameworks used?: }
  %\item {\bf Publicly available?: } Yes: \url{https://github.com/fsantanna/ceu}
\end{itemize}

\begin{itemize}
  \item {\bf Artifacts publicly available?:} Yes.
  \item {\bf Artifacts functional?:} Yes.
  \item {\bf Artifacts reusable?:} No.
  \item {\bf Results validated?:} No.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Description}

\subsubsection{How delivered}

The compiler/distribution of \CEU is available on GitHub:
\url{https://github.com/fsantanna/ceu}

\subsubsection{Hardware dependencies}

Off-the-shelf Intel machine.

\subsubsection{Software dependencies}

Linux, Lua-5.3 (with lpeg-1.0.0), and GCC-7.2.

\subsubsection{Data sets}

A set of more than 3500 programs packed in a single script file which is
included with the compiler distribution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Installation}

Install all required software (assuming an Ubuntu-based distribution):

\begin{verbatim}
$ sudo apt-get install git lua5.3 lua-lpeg liblua5.3-0 \
                       liblua5.3-dev
\end{verbatim}

Clone the repository of \CEU:

\begin{verbatim}
$ git clone https://github.com/fsantanna/ceu
$ cd ceu/
$ git checkout v0.30
\end{verbatim}

Install \CEU:

\begin{verbatim}
$ make
$ sudo make install     # install as "/usr/local/bin/ceu"
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment workflow}

Run the experiment:

\begin{verbatim}
$ cd tst/
$ ./run.lua
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Evaluation and expected result}

The experiment will execute all test cases.
In about 5-10 minutes, a summary will be printed on screen:

\begin{verbatim}
$ cd tst/
$ ./run.lua
<...> # output with the test programs
stats = {
    count  = 3483,
    trails = 9094,
    bytes  = 52555128,
    bcasts = 0,
    visits = 4065937,
}
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment customization}

The file \code{tst/tests.lua} includes all test cases.
Each test case contains a program in \CEU as well as the expected result, e.g.:

\begin{verbatim}
Test { [[
    var int ret = 0;
    var int i;
    loop i in [0 -> 10[ do
        ret = ret + 1;
    end
    escape ret;
]],
    run = 10,
}
\end{verbatim}

This test case should compile and run successfully yielding \code{10}.

To customize the experiment, include a new test case at line \code{440}, after
the string ``\code{--~OK:~well~tested}''.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Notes}

### 2018-04-10 07-artifact.tex
Ajustes no Artifact evaluation

@@
-\appendix
 \section{Artifact appendix}
@@
 
-{\em Obligatory. Use just a few informal keywords in all fields applicable to your artifacts
-and remove the rest. This information is needed to find appropriate reviewers and gradually 
-unify artifact meta information in Digital Libraries.}
-
 {\small
@@
   \item {\bf Data set: } The data set is a set of program test cases included in the language distribution.
-  \item {\bf Run-time environment: } Linux with Lua-5.3 and GCC-7.2. Root access is not required.
-  \item {\bf Hardware: } Off-the-shelf Intel machine.
+  \item {\bf Run-time environment: } Linux with Lua-5.3 and GCC-7.2. Root access is required to install a single executable file.
+  \item {\bf Hardware: } An off-the-shelf Intel machine.
   %\item {\bf Run-time state: }
@@
   \item {\bf Output: } Console output: Success (termination) or Fail (abortion).
-  \item {\bf Experiments: } Manual steps by user.
+  \item {\bf Experiments: } Manual steps performed by the user in the command line.
   %\item {\bf Workflow frameworks used?: }
@@
 
-Off-the-shelf Intel machine.
+An off-the-shelf Intel machine.
 

### 2018-04-14 01-intro.tex
Revisão Seção 1

@@
 \item
-    Understanding, finding corner cases, and setting compromises.
-    After the semantics was complete and discussed in extent, we found a bug in
-    the order of execution of statements in a specific case.
+    Understanding, finding corner cases, and stabilizing the language.
+    After the semantics was complete and discussed in extent in our group, we
+    found a critical bug in the order of execution of statements.
 \item
@@
     Implementing the language.
-    Small-step operational semantics typically describe abstract machines that
-    are close to concrete implementations.
+    A small-step operational semantics describes an abstract machine that is
+    close to a concrete implementation.
     The current implementation of the \CEU scheduler is based on the formal
-    semantics.
+    semantics presented in this paper.
 \item
     Proving properties for particular or all programs in the language.
-    We prove that all programs in \CEU are memory bounded, deterministic, and
-    react in finite time.
+    For instance, in this work, we prove that all programs in \CEU are memory
+    bounded, deterministic, and react in finite time.
 \end{enumerate}
 
-The last item is particularly important in the context of embedded systems:
+The last item is particularly important in the context of constrained embedded
+systems:
 
@@
 We can simulate an entire program execution providing an input history with the
-guarantee that it will always behave the same.
-This can be done in negligible time in a controlled development environment and
-with multiple users (e.g., multiple developers each on its desktop) before
-deploying the application in the actual device.
+guarantee that it will always have the same behavior.
+This can be done in negligible time in a controlled development environment
+before deploying the application in the actual device (e.g., by multiple
+developers in standard desktops).
 \item[Terminating Reactions:]
@@
 that programs always progress with the time.
-In addition, it also ensures that programs always reach an idle state amenable
-to entering in standby mode, which is can improve the autonomy of
-battery-powered devices.
+%In addition, it also ensures that programs always reach an idle state amenable
+%to entering in standby mode, which can improve the autonomy of battery-powered
+%devices.
 \end{description}

### 2018-04-14 03-semantics.tex
Minor rephrasing in Section 3

--- 2018-03-19 03-semantics.tex
@@
 establishes that if a finite number of applications of~$\nst$ results in an
-irreducible description, then that can only occur exactly once, at some
-specific number~$i$.  The proof of the lemma follows directly from the
-definition of~$\nst$ by contradiction on the hypothesis that there is
-such~$k\ne{i}$.
+irreducible description, then that occurs exactly once, at some specific
+number~$i$.  The proof of the lemma follows directly from the definition
+of~$\nst$ by contradiction on the hypothesis that there is such~$k\ne{i}$.
 
@@
 Theorem~\ref{thm.x.term-nst-*} establishes that a finite (possibly zero)
-number of~$\nst$ transitions eventually lead to a nested-irreducible
+number of~$\nst$ transitions eventually leads to a nested-irreducible
 description.  The proof is by induction on the structure of programs
@@
   More formally, $\pot(p,e)=\pot'(\bcast(p,e))$ where~$\pot'$ is an
-  auxiliary function that counts the number of reachable~$\ceu{\EmitInt}$
-  expressions in the program resulting from the broadcast of event~$e$
-  to~$p$.
+  auxiliary function that counts the maximum number of
+  reachable~$\ceu{\EmitInt}$ expressions in the program resulting from the
+  broadcast of event~$e$ to~$p$.
 

### 2018-04-14 01-intro.tex

--- 2018-03-22 01-intro.tex
@@
 We can simulate an entire program execution providing an input history with the
-guarantee that it will always behave the same.
-This can be done in negligible time in a controlled development environment and
-with multiple users (e.g., multiple developers each on its desktop) before
-deploying the application in the actual device.
+guarantee that it will always have the same behavior.
+This can be done in negligible time in a controlled development environment by
+multiple developers before deploying the application in the actual device.
 \item[Terminating Reactions:]
@@
 In addition, it also ensures that programs always reach an idle state amenable
-to entering in standby mode, which is can improve the autonomy of
-battery-powered devices.
+to entering in standby mode, which can improve the autonomy of battery-powered
+devices.
 \end{description}

### 2018-04-14 02-ceu.tex

@@
 environment should only provide access to~C operations that can be assumed to
-be instantaneous, such as non-blocking~I/O and simple data structure accessors.
+be instantaneous, such as non-blocking~I/O and simple data structure accessors.%
+\footnote {
+The actual implementation of Céu supports a command-line option that accepts a
+white list of library calls.
+If a program tries to use a function not in this list, the compiler raises an
+error.
+}
 

### 2018-04-14 04-related.tex
Revisões do artigo.

@@
 nor be written by concurrent threads''}~\cite{esterel.primer}.
+%
+Considering the constant and low-level interactions with the underlying
+architecture in embedded systems (e.g., direct port manipulation), we believe
+that it is advantageous to embrace lexical scheduling as part of the language
+specification as a pragmatic design decision to enforce determinism.
+%
+However, since \CEU statically detects trails not sharing memory, an optimized
+scheduler could exploit real parallelism in such cases.
 

### 2018-04-14 01-intro.tex
Fix prepositions in Introduction

@@
 This can be done in negligible time in a controlled development environment by
-multiple developers before deploying the application in the actual device.
+multiple developers before deploying the application to the actual device.
 \item[Terminating Reactions:]
@@
 A terminating semantics enforces upper bounds for all reactions and guarantees
-that programs always progress with the time.
-In addition, it also ensures that programs always reach an idle state amenable
+that programs always progress with time.
+In addition, it ensures that programs always reach an idle state amenable
 to entering in standby mode, which can improve the autonomy of battery-powered

### 2018-04-14 02-ceu.tex
Revisão Seção 2

@@
 
-\CEU is a synchronous reactive language in which programs evolve in a sequence
-of discrete reactions to external events.
+\CEU~\cite{ceu.sensys13,ceu.tecs17} is a synchronous reactive language in which
+programs evolve in a sequence of discrete reactions to external events.
 %
@@
 white list of library calls.
-If a program tries to use a function not in this list, the compiler raises an
+If a program tries to use a function not in the list, the compiler raises an
 error.
@@
   float=h,
-  caption={A \CEU program with a ``subroutine''.},
+  caption={A ``subroutine'' that increments its argument.},
   label={lst.sub},
@@
 
-At first sight, event iteration such as in ``\code{every e do <...> end}'' seems
-to be equivalent to ``\code{loop do await e; <...> end}''.
-However, the \code{loop} variation would not compile because it does not
-contain a path to an external input \code{await} (since \code{e} is an internal
-event).
+At first sight, the constructions ``\code{every e do <...> end}'' and
+``\code{loop do await e; <...> end}'' seem to be equivalent.
+However, the \code{loop} variation would not compile since it does not contain
+an external input \code{await} (\code{e} is an internal event).
 %
-However, event iterators enforce other syntactic restrictions and cannot
-contain \code{await} or \code{break} statements.
-The absence of \code{break} guarantees that an iterator never terminates from
-itself, essentially behaving as a safe blocking point in the program.
+The \code{every} variation compiles because event iterators have an additional
+syntactic restriction that they cannot contain \code{break} or \code{await}
+statements.
+This restriction guarantees that an iterator never terminates from itself and,
+thus, always awaits its identifying event, essentially behaving as a safe
+blocking point in the program.
 %
@@
 In addition, abortion is \emph{immediate} in the sense that it executes
-atomically in the current micro reaction.
+atomically inside the current micro reaction.
 %
@@
 
-\CEU also enforces the use of \code{finalize} for system calls that deal with
-pointers representing resources, as illustrated in the two examples of
-Figure~\ref{lst.fin.ceu}:
+\CEU also enforces, at compile time, the use of \code{finalize} for system
+calls that deal with pointers representing resources, as illustrated in the two
+examples of Figure~\ref{lst.fin.ceu}:
 %

### 2018-04-15 03-semantics.tex
Revisão Seção 3

@@
 Memory operations are opaque and behave like in conventional imperative
-languages (i.e., reads and writes), having no effect on the application
+languages (i.e., reads and writes), having no direct effect on the application
 execution flow (e.g., they cannot block or abort execution).
@@
 
-We assume that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$
-and~$\ceu{\EmitInt}$ expressions do not share identifiers:
+We assume that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, and
+$\ceu{\AwaitInt}$/$\ceu{\EmitInt}$ expressions do not share identifiers:
 any identifier is either a variable, an external event, or an internal event.
@@
 \]
-where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are nonnegative
+where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are non-negative
 integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
@@
   &\AxiomC{$n>0$}
-  \AxiomC{$p=\ceu{\Nop,\Break}\vee\isblocked(p,n)$}
+  \AxiomC{$p=\ceu{\Nop} \vee p=\ceu{\Break} \vee \isblocked(p,n)$}
   \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
@@
 Rule \R{pop} only decreases the stack level, not affecting the
-program, and only applies if the program is blocked (see function~$\isblocked$ in
-Figure~\ref{fig.isblocked}).
+program, and only applies if the program is blocked (see function~$\isblocked$
+in Figure~\ref{fig.isblocked}) or terminated ($p=\ceu{\Nop}$ or $p=\ceu{\Break}$).
 This condition ensures that an $\ceu{\EmitInt}$ only resumes after its internal

### 2018-04-15 02-ceu.tex
Compress captions of figs and listings in Section 2

--- 2018-04-14 02-ceu.tex
@@
 \def\T<#1>{\langle\mathit{#1}\rangle}
-\def\C#1#2{\hfill\rmfamily\itshape\makebox[#1\columnwidth][l]{//~#2}}
+\def\C#1{\rmfamily\itshape//~#1}
+\def\CC#1#2{\hfill\makebox[#1][l]{\C{#2}}}
 \begin{lstlisting}[
@@
 ]
-|\C{1.}{Declarations}|
-input $\T<type>$ $\T<id>$; | \C{.6}{declares an external input event} |
-event $\T<type>$ $\T<id>$; | \C{.6}{declares an internal event}       |
-var   $\T<type>$ $\T<id>$; | \C{.6}{declares a variable}              |
+|\C{Declarations}|
+input $\T<type>$ $\T<id>$; | \CC{19.5em}{declares an external input event} |
+event $\T<type>$ $\T<id>$; | \CC{19.5em}{declares an internal event}       |
+var   $\T<type>$ $\T<id>$; | \CC{19.5em}{declares a variable}              |
 
-|\C{1.}{Event handling}|
-$\T<id>$ = await $\T<id>$;   | \C{.6}{awaits an event and assigns the received value} |
-emit $\T<id>$($\T<exp>$);    | \C{.6}{emits an event passing a value} |
+|\C{Event handling}|
+$\T<id>$ = await $\T<id>$; | \CC{19.5em}{awaits an event and assigns the received value} |
+emit $\T<id>$($\T<exp>$); | \CC{19.5em}{emits an event passing a value} |
 
-|\C{1.}{Control flow}|
-$\T<stmt>$ ; $\T<stmt>$                          | \C{.45}{sequence}     |
-if $\T<exp>$ then $\T<stmts>$ else $\T<stmts>$ end    | \C{.45}{conditional}  |
-loop do $\T<stmts>$ end                      | \C{.45}{repetition}   |
-every $\T<id>$ in $\T<id>$ do $\T<stmts>$ end         | \C{.45}{event iteration}   |
-finalize [$\T<stmts>$] with $\T<stmts>$ end       | \C{.45}{finalization} |
+|\C{Control flow}|
+$\T<stmt>$ ; $\T<stmt>$ | \CC{10em}{sequence}     |
+if $\T<exp>$ then $\T<stmts>$ else $\T<stmts>$ end | \CC{10em}{conditional}  |
+loop do $\T<stmts>$ end | \CC{10em}{repetition}   |
+every $\T<id>$ in $\T<id>$ do $\T<stmts>$ end | \CC{10em}{event iteration}   |
+finalize [$\T<stmts>$] with $\T<stmts>$ end | \CC{10em}{finalization} |
 
-|\C{1.}{Logical parallelism}|
-par/or  do $\T<stmts>$ with $\T<stmts>$ end | \C{.45}{aborts when any side ends}      |
-par/and do $\T<stmts>$ with $\T<stmts>$ end | \C{.45}{terminates when all sides ends} |
+|\C{Logical parallelism}|
+par/or  do $\T<stmts>$ with $\T<stmts>$ end | \CC{13em}{aborts when any side ends}      |
+par/and do $\T<stmts>$ with $\T<stmts>$ end | \CC{13em}{terminates when all sides end} |
 
-|\C{1.}{Assignment \& Integration with C}|
-$\T<id>$ = $\T<exp>$; | \C{.6}{assigns a value to a variable}              |
-_$\T<id>$($\T<exps>$) | \C{.6}{calls a C function (id starts with `\_'\,)} |
+|\C{Assignment \& Integration with C}|
+$\T<id>$ = $\T<exp>$; | \CC{19.5em}{assigns a value to a variable}              |
+_$\T<id>$($\T<exps>$) | \CC{19.5em}{calls a C function (id starts with `\_'\,)} |
 \end{lstlisting}
@@
 Listing~\ref{lst.ceu} shows a complete example in \CEU that toggles a LED
-whenever a radio packet is received, terminating with a button press always
+whenever a radio packet is received, terminating on a button press always
 with the LED off.
 %
-The implementation first declares the \code{BUTTON} and \code{RADIO\_RECV} as
+The program first declares the \code{BUTTON} and \code{RADIO\_RECV} as
 input events (ln.~1--2).
@@
 The \code{finalize} clause (ln.~6--8) ensures that, no matter how its enclosing
-trail terminates, the LED will be unconditionally turned off (ln.~7).
+trail terminates, the LED will be unconditionally turned off~(ln.~7).
 
@@
 
-\begin{figure}[b]
-\centering
-\includegraphics[width=\columnwidth]{tick_min}
-\caption{The discrete notion of time in \CEU.}
-\label{fig.ticks}
-\end{figure}
-
 The sequential processing of external input events induces a discrete notion of
@@
 %handled in the subsequent reactions~\code{E-5} and~\code{E-6}.
+
+\begingroup
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}[x=.9em,y=.9em,
+  font=\footnotesize\sffamily,>=stealth,
+  time/.style={font=\scriptsize\sffamily},
+  cross/.style={inner sep=0pt,outer sep=0pt},
+  evt/.style={fill=white,inner sep=1.85pt,outer sep=0pt},
+  rng/.style={fill=white,inner sep=1pt,outer sep=0pt,
+    font=\ttfamily\scriptsize}]
+  \draw[->]
+  (0,0)
+    node[align=right,anchor=east,xshift=-.5em]
+        {Continuous\\Timeline\\(Real World)}
+    node[anchor=north west,xshift=-.5em,time]
+        {0ms000us}
+    node[anchor=north west,xshift=8.75em,time]
+        {17ms521us}
+  -- (18.5,0)
+    node[anchor=north east,xshift=.1em,time]
+        {30ms000us};
+  \node[cross] at (0,0){x};
+  %%
+  \coordinate(Ext) at ($(0,0)!.5!(0,-5)$);
+  \coordinate(A) at (Ext-|4,0);
+  \coordinate(B) at (Ext-|5.5,0);
+  \coordinate(C) at (Ext-|9,0);
+  \coordinate(D) at (Ext-|15,0);
+  %%
+  \draw[semithick,dotted] (0,1-|A) -- (0,-6-|A);
+  \draw[semithick,dotted] (0,1-|B) -- (0,-6-|B);
+  \draw[semithick,dotted] (0,1-|C) -- (0,-6-|C);
+  \draw[semithick,dotted] (0,1-|D) -- (0,-5-|D);
+  \node[evt](Aevt) at (A) {\textbf{A}};
+  \node[evt](Bevt) at (B) {\textbf{B}};
+  \node[evt](Cevt) at (C) {\textbf{C}};
+  \node[evt](Devt) at (D) {\textbf{D}};
+  \node[anchor=east,xshift=-.5em] at (A){\emph{External Input}};
+  %%
+  \draw[-]
+  (0,-5)
+    node[align=right,anchor=east,xshift=-.5em]
+        {Discrete\\Timeline\\(\CEU)}
+  -- (18.5,-5);
+  \node[cross](P0) at (0,-5){x};
+  \node[cross](P1) at (4.75,-5){x};
+  \node[cross](P2) at (10,-5){x};
+  \node[cross](P3) at (12.5,-5){x};
+  \node[cross](P4) at (14.5,-5){x};
+  \node[cross](P5) at (15.5,-5){x};
+  \node[cross](P6) at (18.5,-5){x};
+  \node[rng] at ($(P0)!.5!(P1)$){boot-0};
+  \node[rng] at ($(P1)!.5!(P2)$){A-1};
+  \node[rng] at ($(P2)!.5!(P3)$){B-2};
+  \node[rng] at ($(P3)!.5!(P4)$){C-3};
+  \node[rng] at ($(P5)!.5!(P6)$){D-4};
+  \draw[decorate,decoration={brace,raise=-1pt,mirror,amplitude=2pt}]
+  ($(P4)+(0,-.5)$) -- node[below,time] {Idle period} ($(P5)+(0,-.5)$);
+  %%
+  \begin{scope}[shorten >=1.5pt]
+  \draw[->] (Aevt) -- (P1);
+  \draw[->] (Bevt) -- (P2);
+  \draw[->] (Cevt) -- (P3);
+  \draw[->] (Devt) -- (P5);
+  \end{scope}
+\end{tikzpicture}
+%\includegraphics[width=\columnwidth]{tick_min}
+\caption{The discrete notion of time in \CEU.}
+\label{fig.ticks}
+\end{figure}
+\endgroup
 
@@
 
-\begin{figure}[h]
+\begingroup
+\begin{figure}[t]
 \begin{minipage}[h]{0.45\linewidth}
-\begin{lstlisting}[numbers=right]
+\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\footnotesize]
 input void A;
@@
 \end{lstlisting}
-\centering\small{\ax Accesses to \code{x} are never concurrent.}
+\small{\ax Accesses to \code{x} are never\\\phantom{\ax}\relax concurrent.}
 \end{minipage}%
@@
 %\begin{lstlisting}
-\begin{lstlisting}[xleftmargin=2em]
+\begin{lstlisting}[xleftmargin=2em,basicstyle=\ttfamily\footnotesize]
 input void A;
@@
 \end{lstlisting}
-\centering\small{\bx Accesses to \code{y} are concurrent but still deterministic.}
+\small{\bx Accesses to \code{y} are concurrent\\
+\phantom{\bx}\relax but still deterministic.}
 \end{minipage}%
@@
 \caption{Shared-memory concurrency in \CEU:
-example \ax is safe because the trails access \code{x} atomically in different
+\ax is safe because the trails access \code{x} atomically in different
 reactions;
-example \bx is unsafe because both trails access \code{y} in the same reaction.
-\label{lst.shared}%
-}
+\bx is unsafe because both trails access \code{y} in the same reaction}
+\label{lst.shared}
 \end{figure}
+\endgroup
 
@@
 \begin{minipage}[t]{0.45\linewidth}
-\begin{lstlisting}[numbers=right]
+\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\footnotesize]
 par/or do
@@
 \end{lstlisting}
-\small{\ax\!Local resource finalization}
+\small{\ax\!Local resource finalization\rlap{.}}
 \end{minipage}%
@@
 \begin{minipage}[t]{0.53\linewidth}
-\begin{lstlisting}[xleftmargin=2em]
+\begin{lstlisting}[xleftmargin=2em,basicstyle=\ttfamily\footnotesize]
 par/or do
@@
 \end{lstlisting}
-\centering\small{\bx External resource finalization}
+\small{\quad\bx\!External resource finalization\rlap{.}}
 \end{minipage}%
@@
 \caption{%
-\CEU enforces the use of finalization to prevent \emph{dangling pointers} for
-local resources and \emph{memory leaks} for external resources.%
-\label{lst.fin.ceu}}
+\CEU enforces the use of finalization to prevent dangling pointers and memory leaks.}
+\label{lst.fin.ceu}
 \end{figure}
@@
 pointers representing resources, as illustrated in the two examples of
-Figure~\ref{lst.fin.ceu}:
+Figure~\ref{lst.fin.ceu}.
 %
-\begin{itemize}
-    \item If \CEU \textbf{passes} a pointer to a system call (ln.~\ax:5), the
+%\begin{itemize}
+ If \CEU \textbf{passes} a pointer to a system call (ln.~\ax:5), the
           pointer represents a \textbf{local} resource (ln.~\ax:2) that
           requires finalization (ln.~\ax:7).
-    \item If \CEU \textbf{receives} a pointer from a system call return
+ If \CEU \textbf{receives} a pointer from a system call return
           (ln.~ \bx:4), the pointer represents an \textbf{external} resource
           (ln.~\bx:2) that requires finalization (ln.~\bx:6).
-\end{itemize}
-%
+%\end{itemize}
+
 \CEU tracks the interaction of system calls with pointers and requires

### 2018-04-15 03-semantics.tex
Minor rewrites for brevity and clarity in the beginning of Section 3

--- 2018-04-14 03-semantics.tex
@@
 
-In this section, we introduce a reduced syntax for \CEU and propose an
-operational semantics to formally describe the behavior of its programs.
-We describe a small synchronous kernel highlighting the peculiarities of \CEU,
-in particular, the stack-based execution for internal events.
-%
-For the sake of simplicity, we focus on the control aspects of the language,
-leaving out side-effects and system calls (which behave like in conventional
-imperative languages).
+In this section, we introduce a reduced, abstract syntax for \CEU and
+present an operational semantics that formalizes the behavior of its
+programs.
+% We describe a small synchronous kernel highlighting the peculiarities of \CEU,
+% in particular, the stack-based execution for internal events.
+%
+The semantics deals only with the control aspects of \CEU.
+Side-effects and system calls are encapsulated in an abstract memory access
+primitive and are assumed to behave like in conventional imperative
+languages.
 
@@
 The grammar below defines the syntax of a subset of \CEU that is
-sufficient to describe all semantic peculiarities of the language.
+sufficient to describe all peculiarities of the language.
 \bgroup
@@
 \newdimen\X
-\X=-1\jot
+\X=-1.25\jot
 \begin{alignat*}{2}
@@
 
-The~$\ceu{\Mem(id)}$ primitive represents all accesses, assignments, and system
-calls that affect a memory location identified by~$id$.
-%
-According to the synchronous hypothesis of \CEU, $\ceu{\Mem}$ expressions are
+The~$\ceu{\Mem(id)}$ primitive represents any access, assignment, or system
+call that affects the memory location identified by~$id$.\footnotemark\
+%
+Following the synchronous hypothesis, $\ceu{\Mem}$ expressions are
 considered to be atomic and instantaneous.
 %
-As the challenging parts of \CEU reside on its control structures, we are not
-concerned here with a precise semantics for side effects, but only with their
-occurrences in programs.
+% As the challenging parts of \CEU reside on its control structures, we are not
+% concerned here with a precise semantics for side effects, but only with their
+% occurrences in programs.
 %
@@
 %unused identifier).
-
-We assume that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$
-and~$\ceu{\EmitInt}$ expressions do not share identifiers:
-any identifier is either a variable, an external event, or an internal event.
-
-Most expressions in the abstract language are mapped to their counterparts in
-the concrete language.
-The exceptions are the finalization block~$\ceu{\Fin{P}}$ and the
-\texttt{@}-expressions which result from expansions in the transition rules to
-be presented.
-
-Regarding mismatches between the concrete and abstract languages, the concrete
-\code{await} and \code{emit} primitives support communication of values between
-them, e.g., an ``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
-variable~\code{v} to~10.
-To reproduce this functionality in the formal semantics, we can use a shared
-variable to hold the value of an $\ceu{\EmitInt}$ and access it after the
-corresponding $\ceu{\AwaitInt}$ awakes.
-%
-Also, a ``\code{finalize $A$ with $B$ end; $C$}'' in the concrete language is
-equivalent to ``\ceu{A;\;((\Fin{B})\ \Or\ C)}'' in the abstract language.
-In the concrete language, $A$ and~$C$ execute in sequence, and
+%
+\footnotetext{Although the same symbol~$\ceu{\Id}$ is used in their
+  definition, the $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$
+  and~$\ceu{\EmitInt}$ expressions do not share identifiers: any identifier
+  is either a variable, an external event, or an internal event.}
+
+Most expressions in the abstract language are mapped to their counterparts
+in the concrete language.  The exceptions are the finalization
+block~$\ceu{\Fin{p}}$ and the \texttt{@}-expressions which do not exist in
+the concrete language.  (These expressions result from the expansion of the
+transition rules to be presented.)
+
+A further difference between the concrete and abstract languages regards the
+emit-await pair.  In the concrete language the \code{await} and \code{emit}
+expressions support communication of values between them, e.g., an
+``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
+variable~\code{v} to~10.  Although the abstract $\ceu{\AwaitInt}$
+and~$\ceu{\EmitInt}$ expression do not support such communication of values,
+it can be easily simulated: we can use a shared variable to hold the value
+of an $\ceu{\EmitInt}$ and access it after the corresponding
+$\ceu{\AwaitInt}$ awakes.
+
+Finally, a ``\code{finalize $A$ with $B$ end; $C$}'' in the concrete
+language is equivalent to ``\ceu{A;\;((\Fin{B})\ \Or\ C)}'' in the abstract
+language.  In the concrete language, $A$ and~$C$ execute in sequence, and
 the finalization code~$B$ is implicitly suspended waiting for~$C$
-termination.
-In the abstract language, ``$\ceu{\Fin B}$'' suspends forever when reached (it is
-an awaiting expression that never awakes).
-Hence, we need an explicit \code{or} to execute~$C$ in parallel, whose
-termination aborts ``$\ceu{\Fin B}$'', which finally causes~$B$ to
-execute (by the semantic rules to be discussed).
+termination.  In the abstract language, ``$\ceu{\Fin B}$'' suspends forever
+when reached (it is an awaiting expression that never awakes).  Hence, we
+need an explicit \code{or} to execute~$C$ in parallel, whose termination
+aborts ``$\ceu{\Fin B}$'', which finally causes~$B$ to execute (by the
+semantic rules below).
 
@@
 We use a set of small-step operational rules~\cite{Plotkin-G-D-1981} which
-are designed in such a way that at most one transition is possible at any
-time, resulting in deterministic reactions.
+ensure that at most one transition is possible at any time, resulting in
+deterministic reactions.
 %
@@
 empty event~$\nil$).
-$\E$ is a set of primitive events.
+%$\E$ is a set of primitive events.
 
 We will refer to the triples on the left-hand and right-hand sides of
-symbol~$\trans$ as \emph{descriptions} (denoted~$\delta$).  The triple on the
-left-hand side of symbol~$\trans$ is called the \emph{input description}, and
-the triple on its right-hand side is called the \emph{output description}.
+symbol~$\trans$ as \emph{descriptions} (denoted~$\delta$).  The description
+on the left-hand side of~$\trans$ is the \emph{input description},
+and the one on the right-hand side is the \emph{output description}.
 
@@
 At the beginning of a reaction to an input event~$id$, the input description is
-initialized with stack level~0 ($n=0$) and with the externally emitted event
-($e=id$).
+initialized with stack level~0 ($n=0$) and with the externally emitted event~$e=id$.
 %, but \code{emitInt} expressions can increase the stack level.
 At the end of a reaction, after an arbitrary but finite number of transitions,
-the last output description will block with a (possibly) modified program~$p'$, at stack
-level~0, and with no event emitted~($\nil$):
+the last output description will block with a (possibly) modified program~$p'$ at stack
+level~0 and with no event emitted:
 \[

### 2018-04-15 01-intro.tex

--- 2018-04-14 01-intro.tex
@@
 This can be done in negligible time in a controlled development environment
-before deploying the application in the actual device (e.g., by multiple
+before deploying the application to the actual device (e.g., by multiple
 developers in standard desktops).
@@
 that programs always progress with the time.
-%In addition, it also ensures that programs always reach an idle state amenable
-%to entering in standby mode, which can improve the autonomy of battery-powered
-%devices.
 \end{description}

### 2018-04-15 02-ceu.tex

@@
   float=h,
-  caption={A program in \CEU that toggles a LED on every radio receive,
-           terminating on a button press always with the LED off.},
+  caption={A \CEU program that toggles a LED on radio receive,
+           terminating on a button press always with the LED~off.},
   label={lst.ceu},
@@
 \begingroup
-\begin{figure}[h]
+\begin{figure}[t]
 \centering
@@
 
-In example \ax, the two assignments to \code{x} can only execute in
+In \ax, the two assignments to \code{x} can only execute in
 reactions to different events \code{A} and \code{B} (ln.~5, 8), which cannot
@@
 
-In example \bx, the two assignments to \code{y} are simultaneous because
+In \bx, the two assignments to \code{y} are simultaneous because
 they execute in reaction to the same event \code{A}.
@@
 
-\begin{figure}[h]
+The \code{par/or} of \CEU is an orthogonal abortion mechanism because the two
+sides in the composition need not be tweaked with synchronization primitives nor
+state variables to affect each other.
+%
+In addition, abortion is \emph{immediate} in the sense that it executes
+atomically inside the current micro reaction.
+%
+Immediate orthogonal abortion is a distinctive feature of synchronous languages
+and cannot be expressed effectively in traditional (asynchronous)
+multi-threaded languages~\cite{esterel.preemption,sync_async.threadsstop}.
+
+However, aborting lines of execution that deal with external resources may lead
+to inconsistencies.
+%
+For this reason, \CEU provides a \code{finalize} construct to unconditionally
+execute a series of statements even if the enclosing block is externally
+aborted.
+
+\CEU also enforces, at compile time, the use of \code{finalize} for system
+calls that deal with pointers representing resources, as illustrated in the two
+examples of Figure~\ref{lst.fin.ceu}.
+%
+%\begin{itemize}
+ If \CEU \textbf{passes} a pointer to a system call (ln.~\ax:5), the
+          pointer represents a \textbf{local} resource (ln.~\ax:2) that
+          requires finalization (ln.~\ax:7).
+ If \CEU \textbf{receives} a pointer from a system call return
+          (ln.~ \bx:4), the pointer represents an \textbf{external} resource
+          (ln.~\bx:2) that requires finalization (ln.~\bx:6).
+%\end{itemize}
+
+\CEU tracks the interaction of system calls with pointers and requires
+finalization clauses to accompany them.
+%
+In Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
+(ln.~2) is an internal resource passed as a pointer to \code{\_send} (ln.~5),
+which is an asynchronous call that transmits the buffer in the background.
+If the block aborts (ln.~11) before receiving an acknowledge from the
+environment (ln.~9), the local \code{msg} goes out of scope and the external
+transmission now holds a \emph{dangling pointer}.
+The finalization ensures that the transmission also aborts (ln.~7).
+%
+In Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
+4) returns an external file resource as a pointer.
+If the block aborts (ln.~12) during the \code{await A} (ln.~9), the file
+remains open as a \emph{memory leak}.
+The finalization ensures that the file closes properly (ln.~6).
+%
+In both cases, the code does not compile without the \code{finalize}
+construct.%
+\footnote{%
+The compiler only forces the programmer to write the finalization clause, but
+cannot check if it actually handles the resource properly.}
+
+\begin{figure}[t]
 \begin{minipage}[t]{0.45\linewidth}
@@
 
-The \code{par/or} of \CEU is an orthogonal abortion mechanism because the two
-sides in the composition need not be tweaked with synchronization primitives nor
-state variables to affect each other.
-%
-In addition, abortion is \emph{immediate} in the sense that it executes
-atomically in the current micro reaction.
-%
-Immediate orthogonal abortion is a distinctive feature of synchronous languages
-and cannot be expressed effectively in traditional (asynchronous)
-multi-threaded languages~\cite{esterel.preemption,sync_async.threadsstop}.
-
-However, aborting lines of execution that deal with external resources may lead
-to inconsistencies.
-%
-For this reason, \CEU provides a \code{finalize} construct to unconditionally
-execute a series of statements even if the enclosing block is externally
-aborted.
-
-\CEU also enforces the use of \code{finalize} for system calls that deal with
-pointers representing resources, as illustrated in the two examples of
-Figure~\ref{lst.fin.ceu}.
-%
-%\begin{itemize}
- If \CEU \textbf{passes} a pointer to a system call (ln.~\ax:5), the
-          pointer represents a \textbf{local} resource (ln.~\ax:2) that
-          requires finalization (ln.~\ax:7).
- If \CEU \textbf{receives} a pointer from a system call return
-          (ln.~ \bx:4), the pointer represents an \textbf{external} resource
-          (ln.~\bx:2) that requires finalization (ln.~\bx:6).
-%\end{itemize}
-
-\CEU tracks the interaction of system calls with pointers and requires
-finalization clauses to accompany them.
-%
-In Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
-(ln.~2) is an internal resource passed as a pointer to \code{\_send} (ln.~5),
-which is an asynchronous call that transmits the buffer in the background.
-If the block aborts (ln.~11) before receiving an acknowledge from the
-environment (ln.~9), the local \code{msg} goes out of scope and the external
-transmission now holds a \emph{dangling pointer}.
-The finalization ensures that the transmission also aborts (ln.~7).
-%
-In Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
-4) returns an external file resource as a pointer.
-If the block aborts (ln.~12) during the \code{await A} (ln.~9), the file
-remains open as a \emph{memory leak}.
-The finalization ensures that the file closes properly (ln.~6).
-%
-In both cases, the code does not compile without the \code{finalize}
-construct.%
-\footnote{%
-The compiler only forces the programmer to write the finalization clause, but
-cannot check if it actually handles the resource properly.
-}
-
 The finalization mechanism of \CEU is fundamental to preserve the orthogonality

### 2018-04-15 02-ceu.tex
\small no listings da Seção 2 (ao invés de \footnotesize)
Usar \footnotesize nesses listings não ganha quase nada de espaço e deixa o
código mais difícil de ler.  Também tirei o \small do \code, para ficar
uniforme com a Seção 3.  O ganho do texttt-small é muito pequeno, e fica
feio quando misturado com matemática (como na seção 3).

@@
 
-\bgroup
-\def\T<#1>{\langle\mathit{#1}\rangle}
 \def\C#1{\rmfamily\itshape//~#1}
 \def\CC#1#2{\hfill\makebox[#1][l]{\C{#2}}}
+\def\T<#1>{\langle\mathit{#1}\rangle}
+\bgroup
 \begin{lstlisting}[
-  basicstyle=\ttfamily\footnotesize,
+  basicstyle=\ttfamily\small,
   caption={The concrete syntax of \CEU.},
@@
 |\C{Declarations}|
-input $\T<type>$ $\T<id>$; | \CC{19.5em}{declares an external input event} |
-event $\T<type>$ $\T<id>$; | \CC{19.5em}{declares an internal event}       |
-var   $\T<type>$ $\T<id>$; | \CC{19.5em}{declares a variable}              |
+input $\T<type>$ $\T<id>$; |\CC{19.5em}{declares an external input event} |
+event $\T<type>$ $\T<id>$; |\CC{19.5em}{declares an internal event}       |
+var   $\T<type>$ $\T<id>$; |\CC{19.5em}{declares a variable}              |
 
 |\C{Event handling}|
-$\T<id>$ = await $\T<id>$; | \CC{19.5em}{awaits an event and assigns the received value} |
-emit $\T<id>$($\T<exp>$); | \CC{19.5em}{emits an event passing a value} |
+$\T<id>$ = await $\T<id>$; |\CC{19.5em}{awaits event and assigns the received value} |
+emit $\T<id>$($\T<expr>$); |\CC{19.5em}{emits event passing a value} |
 
 |\C{Control flow}|
-$\T<stmt>$ ; $\T<stmt>$ | \CC{10em}{sequence}     |
-if $\T<exp>$ then $\T<stmts>$ else $\T<stmts>$ end | \CC{10em}{conditional}  |
-loop do $\T<stmts>$ end | \CC{10em}{repetition}   |
-every $\T<id>$ in $\T<id>$ do $\T<stmts>$ end | \CC{10em}{event iteration}   |
-finalize [$\T<stmts>$] with $\T<stmts>$ end | \CC{10em}{finalization} |
+$\T<stmt>$ ; $\T<stmt>$ | \CC{8em}{sequence}     |
+if $\T<expr>$ then $\T<stmt>$ else $\T<stmt>$ end | \CC{8em}{conditional}  |
+loop do $\T<stmt>$ end | \CC{8em}{repetition}   |
+every $\T<id>$ in $\T<id>$ do $\T<stmt>$ end | \CC{8em}{event iteration}   |
+finalize [$\T<stmt>$] with $\T<stmt>$ end | \CC{8em}{finalization} |
 
 |\C{Logical parallelism}|
-par/or  do $\T<stmts>$ with $\T<stmts>$ end | \CC{13em}{aborts when any side ends}      |
-par/and do $\T<stmts>$ with $\T<stmts>$ end | \CC{13em}{terminates when all sides end} |
+par/or  do $\!\T<stmt>\!$ with $\!\T<stmt>\!$ end |\CC{13.5em}{aborts if any side ends}      |
+par/and do $\!\T<stmt>\!$ with $\!\T<stmt>\!$ end |\CC{13.5em}{terminates if all sides end} |
 
 |\C{Assignment \& Integration with C}|
-$\T<id>$ = $\T<exp>$; | \CC{19.5em}{assigns a value to a variable}              |
-_$\T<id>$($\T<exps>$) | \CC{19.5em}{calls a C function (id starts with `\_'\,)} |
+$\T<id>$ = $\T<expr>$; | \CC{19.5em}{assigns a value to a variable}              |
+_$\T<id>$($\T<exprs>$) | \CC{19.5em}{calls a C function (id starts with `\_'\,)} |
 \end{lstlisting}
@@
   numbers=left,
-  basicstyle=\ttfamily\footnotesize,
+  basicstyle=\ttfamily\small,
   float=h,
@@
 %
-\begin{enumerate}[i]
+\begin{enumerate}[i.]
 \item The program initiates a ``boot reaction'' in a single trail (parallel
@@
 timestamps'' for the event occurrences (e.g., event~\code{C} occurs at
-$17ms521us$).
+17ms521us).
 %
@@
 \begingroup
-\begin{figure}[t]
+\begin{figure}[h]
 \centering
-\begin{tikzpicture}[x=.9em,y=.9em,
+\begin{tikzpicture}[x=1em,y=1em,
   font=\footnotesize\sffamily,>=stealth,
@@
   rng/.style={fill=white,inner sep=1pt,outer sep=0pt,
-    font=\ttfamily\scriptsize}]
+    font=\ttfamily\footnotesize}]
   \draw[->]
@@
   \draw[semithick,dotted] (0,1-|D) -- (0,-5-|D);
-  \node[evt](Aevt) at (A) {\textbf{A}};
-  \node[evt](Bevt) at (B) {\textbf{B}};
-  \node[evt](Cevt) at (C) {\textbf{C}};
-  \node[evt](Devt) at (D) {\textbf{D}};
+  \node[evt](Aevt) at (A) {\texttt{\textbf{A}}};
+  \node[evt](Bevt) at (B) {\texttt{\textbf{B}}};
+  \node[evt](Cevt) at (C) {\texttt{\textbf{C}}};
+  \node[evt](Devt) at (D) {\texttt{\textbf{D}}};
   \node[anchor=east,xshift=-.5em] at (A){\emph{External Input}};
@@
 \end{tikzpicture}
+\vskip-.6em
 %\includegraphics[width=\columnwidth]{tick_min}
@@
 ]
-event int* inc;         // declares subroutine "inc"
+event int* inc;       |\C{declares subroutine "inc"}|
 par/or do
     var int* p;
-    every p in inc do   // implements "inc" through an event iterator
+    every p in inc do |\C{implements "inc" through event iterator}|
         *p = *p + 1;
@@
     var int v = 1;
-    emit inc(&v);       // calls "inc"
-    emit inc(&v);       // calls "inc"
-    _assert(v==3);      // asserts result after the two returns
+    emit inc(&v);     |\C{calls "inc"}|
+    emit inc(&v);     |\C{calls "inc"}|
+    _assert(v==3);    |\C{asserts result after the two returns}|
 end
@@
 
-At first sight, the constructions ``\code{every e do <...> end}'' and
-``\code{loop do await e; <...> end}'' seem to be equivalent.
+At first sight, the constructions ``\code{every e do $\T<\cdots>$ end}'' and
+``\code{loop do await e; $\T<\cdots>$ end}'' seem to be equivalent.
 However, the \code{loop} variation would not compile since it does not contain
@@
 \begin{figure}[t]
-\begin{minipage}[h]{0.45\linewidth}
-\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\footnotesize]
+\begin{minipage}[t]{0.45\linewidth}
+\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\small]
 input void A;
@@
 \end{lstlisting}
-\small{\ax Accesses to \code{x} are never\\\phantom{\ax}\relax concurrent.}
+\small{\ax\ Accesses to \code{x} are never concurrent.}
 \end{minipage}%
 %
-\begin{minipage}[h]{0.53\linewidth}
+\begin{minipage}[t]{0.53\linewidth}
 %\begin{lstlisting}
-\begin{lstlisting}[xleftmargin=2em,basicstyle=\ttfamily\footnotesize]
+\begin{lstlisting}[xleftmargin=1.75em,basicstyle=\ttfamily\small]
 input void A;
-// (empty line)
+|\C{(empty line)}|
 var int y = 1;
@@
 \end{lstlisting}
-\small{\bx Accesses to \code{y} are concurrent\\
-\phantom{\bx}\relax but still deterministic.}
+\hskip1.5em
+\begin{minipage}[t]{.9\columnwidth}
+\small{\bx\ Accesses to \code{y} are concurrent but still deterministic.}
+\end{minipage}%
 \end{minipage}%
@@
 \caption{Shared-memory concurrency in \CEU:
-\ax is safe because the trails access \code{x} atomically in different
+\ax\ is safe because the trails access \code{x} atomically in different
 reactions;
-\bx is unsafe because both trails access \code{y} in the same reaction}
+\bx\ is unsafe because both trails access \code{y} in the same reaction}
 \label{lst.shared}
@@
 %\begin{itemize}
- If \CEU \textbf{passes} a pointer to a system call (ln.~\ax:5), the
-          pointer represents a \textbf{local} resource (ln.~\ax:2) that
+ If \CEU \emph{passes} a pointer to a system call (ln.~\ax:5), the
+          pointer represents a \emph{local} resource (ln.~\ax:2) that
           requires finalization (ln.~\ax:7).
- If \CEU \textbf{receives} a pointer from a system call return
-          (ln.~ \bx:4), the pointer represents an \textbf{external} resource
+ If \CEU \emph{receives} a pointer from a system call return
+          (ln.~ \bx:4), the pointer represents an \emph{external} resource
           (ln.~\bx:2) that requires finalization (ln.~\bx:6).
@@
 \begin{minipage}[t]{0.45\linewidth}
-\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\footnotesize]
+\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\small]
 par/or do
    var _msg_t msg;
-   <...> // prepare msg
+   $\T<\cdots>$   |\C{prepare msg}|
    finalize
@@
 with
-   <...>
+   $\T<\cdots>$
 end
-//
+|\C{(empty line)}|
 \end{lstlisting}
-\small{\ax\!Local resource finalization\rlap{.}}
+\small{\ax\ Local resource finalization.}
 \end{minipage}%
@@
 \begin{minipage}[t]{0.53\linewidth}
-\begin{lstlisting}[xleftmargin=2em,basicstyle=\ttfamily\footnotesize]
+\begin{lstlisting}[xleftmargin=1.75em,basicstyle=\ttfamily\small]
 par/or do
@@
    finalize
-      f = _fopen(...);
+      f = _fopen($\dots$);
    with
@@
    end
-   _fwrite(..., f);
+   _fwrite($\dots$, f);
    await A;
-   _fwrite(..., f);
+   _fwrite($\dots$, f);
 with
-   <...>
+   $\T<\cdots>$
 end
 \end{lstlisting}
-\small{\quad\bx\!External resource finalization\rlap{.}}
+\hskip1em
+\begin{minipage}[t]{.94\columnwidth}
+\small{\bx\ External resource finalization.}
+\end{minipage}%
 \end{minipage}%

### 2018-04-15 03-semantics.tex
Cleanup and compress stuff in Section 3 up to Determinism

@@
 
-\def\JOT{1.5\jot}
+\def\JOT{.8\jot}
 
@@
 The semantics deals only with the control aspects of \CEU.
-Side-effects and system calls are encapsulated in an abstract memory access
+Side-effects and system calls are encapsulated in a memory access
 primitive and are assumed to behave like in conventional imperative
@@
 \bgroup
-\def\lbl#1{\quad\text{\emph{#1}}}%
+\def\lbl#1{\enspace\text{\emph{#1}}}%
 \newdimen\X
-\X=-1.25\jot
+\X=-1.3\jot
 \begin{alignat*}{2}
@@
 
-The~$\ceu{\Mem(id)}$ primitive represents any access, assignment, or system
+The~$\ceu{\Mem(id)}$ primitive represents an assignment or system
 call that affects the memory location identified by~$id$.\footnotemark\
 %
-Following the synchronous hypothesis, $\ceu{\Mem}$ expressions are
-considered to be atomic and instantaneous.
+Following the synchronous hypothesis, $\ceu{\Mem}$ statements and
+expressions are considered to be atomic and instantaneous.
 %
@@
 \footnotetext{Although the same symbol~$\ceu{\Id}$ is used in their
-  definition, the $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$
-  and~$\ceu{\EmitInt}$ expressions do not share identifiers: any identifier
-  is either a variable, an external event, or an internal event.}
-
-Most expressions in the abstract language are mapped to their counterparts
-in the concrete language.  The exceptions are the finalization
-block~$\ceu{\Fin{p}}$ and the \texttt{@}-expressions which do not exist in
-the concrete language.  (These expressions result from the expansion of the
-transition rules to be presented.)
+  definition, $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$
+  and~$\ceu{\EmitInt}$ do not share identifiers: any identifier is either a
+  variable, an external event, or an internal event.}
+
+Most statements in the abstract language are mapped to their counterparts in
+the concrete language.  The exceptions are the finalization
+block~$\ceu{\Fin{p}}$ and the \texttt{@}-statements which do not exist in
+the concrete language.  (These result from the expansion of the transition
+rules to be presented.)
 
 A further difference between the concrete and abstract languages regards the
-emit-await pair.  In the concrete language the \code{await} and \code{emit}
-expressions support communication of values between them, e.g., an
+emit-await pair.  In the concrete language, the \code{await} can be used as
+an expression which evaluates to the value stored in the emitted
+event, e.g., an
 ``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
 variable~\code{v} to~10.  Although the abstract $\ceu{\AwaitInt}$
-and~$\ceu{\EmitInt}$ expression do not support such communication of values,
-it can be easily simulated: we can use a shared variable to hold the value
-of an $\ceu{\EmitInt}$ and access it after the corresponding
-$\ceu{\AwaitInt}$ awakes.
+and~$\ceu{\EmitInt}$ do not support such communication of values, it can be
+easily simulated: one can use a shared variable to hold the value of an
+$\ceu{\EmitInt}$ and access it after the corresponding $\ceu{\AwaitInt}$
+awakes.
 
@@
 the finalization code~$B$ is implicitly suspended waiting for~$C$
-termination.  In the abstract language, ``$\ceu{\Fin B}$'' suspends forever
-when reached (it is an awaiting expression that never awakes).  Hence, we
+to terminate.  In the abstract language, ``$\ceu{\Fin B}$'' suspends forever
+when reached (it is an awaiting statement that never awakes).  Hence, we
 need an explicit \code{or} to execute~$C$ in parallel, whose termination
@@
 
-The core of our semantics describes how a program reacts to a single external
-input event, i.e., starting from an input event, how the program behaves and
-becomes idle again to proceed to a subsequent reaction.
-%
-We use a set of small-step operational rules~\cite{Plotkin-G-D-1981} which
-ensure that at most one transition is possible at any time, resulting in
-deterministic reactions.
-%
-The transition rules map a triple with a program~$p$, a stack level~$n$, and an
-emitted event~$e$ to a modified triple as follows:
+The operational semantics is a mathematical model that describes how an
+abstract \CEU\ programs reacts to a single external input event, i.e., how
+starting from this input event the program advances its state until all its
+trails are blocked waiting for a another input event.
+%%
+The formalism we use here is that of small-step operational
+semantics~\cite{Plotkin-G-D-1981} in which the meaning of programs is
+defined in terms of transitions of an abstract machine.  Each transition
+transforms a triple consisting of a program~$p$, a stack level~$n$, and an
+emitted event~$e$ into a modified triple as follows:
 \[
@@
 symbol~$\trans$ as \emph{descriptions} (denoted~$\delta$).  The description
-on the left-hand side of~$\trans$ is the \emph{input description},
-and the one on the right-hand side is the \emph{output description}.
+on the left of~$\trans$ is called the \emph{input description},
+the one on its right is called the \emph{output description}.
 
@@
 At the end of a reaction, after an arbitrary but finite number of transitions,
-the last output description will block with a (possibly) modified program~$p'$ at stack
+the last output description will block with a possibly modified program~$p'$ at stack
 level~0 and with no event emitted:
@@
 
-We distinguish between two types of transition rules:
-    \emph{outermost transitions} $\out$ and
-    \emph{nested transitions} $\nst$\,.
+We now proceed to give the rules for possible the transitions.  We
+distinguish between two types of transitions:  \emph{outermost transitions}
+$\out$ and \emph{nested transitions} $\nst$\,.
 
@@
 
-The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions that
-apply to the program as a whole.  These are the only rules that manipulate
-the stack level.
+The rules \R{push} and \R{pop} for~$\out$ transitions are non-recursive
+definitions that apply to the program as a whole.  These are the only rules
+that manipulate the stack level.
 \begin{align*}
@@
   \DisplayProof
-  \Rtag{push}\\[\JOT]
+  \Rtag{push}\\[1.2\jot]
   %%
@@
 
-Rule \R{push} matches whenever there is an emitted event in the input
+Rule \R{push} can be applied whenever there is a nonempty event in the input
 description,
 and instantly broadcasts the event to the program, which means
-    (a)~awaking active $\ceu{\AwaitExt}$ or $\ceu{\AwaitInt}$ expressions altogether (see function~$\bcast$ in
+    (a)~awaking any active $\ceu{\AwaitExt}$ or $\ceu{\AwaitInt}$ statements
+    (see $\bcast$ in
         Figure~\ref{fig.bcast}),
@@
 %
-Rule \R{push} is the only rule in the semantics that matches an
+Rule \R{push} is the only rule that matches an
 emitted event and also immediately consumes it.
 
-Rule \R{pop} only decreases the stack level, not affecting the
-program, and only applies if the program is blocked (see function~$\isblocked$
+Rule \R{pop} decreases the stack level by one and can only be applied if the
+program is blocked (see $\isblocked$
 in Figure~\ref{fig.isblocked}) or terminated ($p=\ceu{\Nop}$ or $p=\ceu{\Break}$).
@@
 At the beginning of a reaction, an external event is emitted, which
-will trigger rule \R{push}, which will immediately raise the stack level
+triggers rule \R{push}, immediately raising the stack level
 to~1.
@@
 successive applications of
-rule~\R{pop} will eventually lead to a description containing this
+rule~\R{pop} will lead to a description with this
 same program at stack level~0.
@@
   \Big[\null\nst[*]\null\out\null\Big]\!\!\ast
-  \null\nst[*]\null\outpop\<p',0,\nil>\,.
+  \null\nst[*]\null\outpop\<p',0,\nil>\,,
 \end{align*}
@@
 %
+where the~$\ast$'s stand for a finite (possibly zero) number of repetitions.
 First, a~$\outpush$ starts a nested reaction at level~1.
@@
 access (rule \R{mem}).
-An $\ceu{\EmitInt(id)}$ generates an event $\ceu{\Id}$ and transits to a
+An $\ceu{\EmitInt(id)}$ generates an event $\ceu{\Id}$ and becomes a
 $\ceu{\CanRun(n)}$ which can only resume at level~$n$ (rule \R{emit-int}).
-Since all~$\nst$ rules can only transit with $e=\nil$, an $\ceu{\EmitInt}$
+Since all~$\nst$ rules can only be applied if $e=\nil$, an $\ceu{\EmitInt}$
 inevitably causes rule \R{push} to execute at the outer level, creating a new
@@
 Also, with the new stack level, the resulting $\ceu{\CanRun}(n)$ itself cannot
-transit yet (rule~\R{can-run}), providing the desired stack-based semantics for
+transition yet (rule~\R{can-run}), providing the desired stack-based semantics for
 internal events.
 
-The rules for conditionals and sequences are straightforward:
-%
+The rules for conditionals and sequences are the following:
+\vskip-.6\baselineskip
+\begingroup
+\def\JOT{.5\jot}
 \begin{gather*}
@@
   \DisplayProof
-  \Rtag{seq-adv}
+  \Rtag{seq-adv}\\[\JOT]
+  \<\ceu{\Nop;\,q},n,\nil>\nst\<q,n,\nil>\Rtag{seq-nop}\\[\JOT]
+  %%
+  \<\ceu{\Break;\,q},n,\nil>\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
 \end{gather*}
-\vskip-\belowdisplayskip
-\vskip-\abovedisplayskip
-\vskip\JOT
-\begin{align*}
-  \<\ceu{\Nop;\,q},n,\nil>&\nst\<q,n,\nil>\Rtag{seq-nop}\\[\JOT]
-  %%
-  \<\ceu{\Break;\,q},n,\nil>&\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
-\end{align*}
+\endgroup
+% \begin{align*}
+%   \<\ceu{\Nop;\,q},n,\nil>&\nst\<q,n,\nil>\Rtag{seq-nop}\\[\JOT]
+%   %%
+%   \<\ceu{\Break;\,q},n,\nil>&\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
+% \end{align*}
 
@@
 Given that our semantics focuses on control, rules \R{if-true} and
-\R{if-false} are the only to query~$\ceu{\Mem}$ expressions.
-%
-Function~$\eval$ evaluates a given~$\ceu{\Mem}$ expression to a boolean value.
+\R{if-false} are the only to query~$\ceu{\Mem}$ as expressions.
+%
+Function~$\eval$ evaluates a~$\ceu{\Mem}$ expression to a boolean value.
 %
@@
 
-The rules for~$\ceu{\And}$ and~$\ceu{\Or}$ compositions ensure that the
-left-hand side of compositions always transition before their right-hand
-side:
+The rules for~$\ceu{\And}$ and~$\ceu{\Or}$ compositions ensure that their
+left branch always transition before their right
+branch:
 %%
@@
   \DisplayProof
-  \Rtag{and-adv2}
-\end{gather*}
-
-\begin{gather*}
+  \Rtag{and-adv2}\\[1.2\jot]
+% \end{gather*}
+% \begin{gather*}
   \<\ceu{p\Or{q}},n,\nil>
@@
 of the right branch.
-This ensures that~any $\ceu{\EmitInt}$ on the left branch, which transits to
-a~$\ceu{\CanRun(n)}$, still resumes before the right branch starts.
+This ensures that~any $\ceu{\EmitInt}$ on the left branch, which eventually becomes
+a~$\ceu{\CanRun(n)}$, resumes before the right branch starts.
 %
 The deterministic behavior of the semantics relies on the \emph{isblocked}
-predicate (see Figure~\ref{fig.isblocked}) which appears in rules
+predicate (see Figure~\ref{fig.isblocked}) which is used in rules
 \R{and-adv2} and \R{or-adv2}.
@@
 
-In a parallel~$\ceu{\AtAnd}$, if one of the sides terminates, the composition is
-simply substituted by the other side (rules \R{and-nop1} and
+In a parallel~$\ceu{\AtAnd}$, if one branch terminates, the composition becomes the other branch (rules \R{and-nop1} and
 \R{and-nop2} below).
 %
-In a parallel~$\ceu{\AtOr}$, however, if one of the sides terminates, the
+In a parallel~$\ceu{\AtOr}$, however, if one branch terminates, the
 whole composition
-terminates and function~$\clear$ is used to properly finalize the aborted
-side (rules \R{or-nop1} and \R{or-nop2}).
+terminates and~$\clear$ is called to finalize the aborted
+branch (rules \R{or-nop1} and \R{or-nop2}).
+
+\begingroup
 \begin{gather*}
@@
 \end{gather*}
+\endgroup
 
@@
 The~$\clear$ function (see Figure~\ref{fig.clear}) concatenates all
-active~$\ceu{\Fin}$ bodies of the side being aborted, so that they execute before the
+active~$\ceu{\Fin}$ bodies of the branch being aborted, so that they execute before the
 composition rejoins.
-Note that there are no transition rules for~$\ceu{\Fin}$ expressions.
-This is because once reached, a $\ceu{\Fin}$ expression halts and will only execute
-when it is aborted by a parallel trail and is expanded by the~$\clear$
-function.
-%In Section~\ref{sec.formal.fins}, we show how to map a finalization block in
-%the concrete language to a $fin$ in the formal semantics.
-%
-Note also that there is a syntactic restriction that postulates that~$\ceu{\Fin}$ bodies cannot
-contain awaiting expressions ($\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$,
-$\ceu{\Every}$, or $\ceu{\Fin}$),
-i.e., the result of a~$\clear$ call is guaranteed to execute entirely within a reaction.
-
-Finally, a~$\ceu{\Break}$ in one of the sides in parallel escapes the closest
-enclosing~$\ceu{\Loop}$, properly aborting the other side with the~$\clear$
+
+As there are no transition rules for~$\ceu{\Fin}$ statements,
+ once reached, a $\ceu{\Fin}$ halts and will only be consumed
+if its trail is aborted.  At this point, its body will
+execute as a result of the~$\clear$ call.  The body of a~$\ceu{\Fin}$
+statement always execute withing a reaction.  This is due to a syntactic
+restriction: $\ceu{\Fin}$ bodies cannot
+contain awaiting statements (namely, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$,
+$\ceu{\Every}$, or $\ceu{\Fin}$).
+
+Finally, a~$\ceu{\Break}$ in one branch of a parallel escapes the closest
+enclosing~$\ceu{\Loop}$, properly aborting the other branch with the~$\clear$
 function:
+%
+\begingroup
 \begin{gather*}
@@
 A reaction eventually blocks in~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$,
-$\ceu{\Every}$, $\ceu{\Fin}$, and~$\ceu{\CanRun}$ expressions in parallel
+$\ceu{\Every}$, $\ceu{\Fin}$, and~$\ceu{\CanRun}$ statements in parallel
 trails.
 %
-Then, if none of the trails is blocked in~$\ceu{\CanRun}$ expressions, it means
+Then, if none of the trails is blocked in~$\ceu{\CanRun}$ statements, it means
 that the program cannot advance in the current reaction.
 %
-However, $\ceu{\CanRun}$ expressions can still resume at lower stack indexes
+However, $\ceu{\CanRun}$ statements can still resume at lower stack indexes
 and will eventually resume in the current reaction (see rule \R{pop}).
@@
       %%-
-      \intertext{\llap{(i)~}Function~$\bcast$:}
+      \shortintertext{\llap{(i)~}Function~$\bcast$:}
       %%-
@@
       %%-
-      \intertext{\llap{(ii)~}Predicate~$\isblocked$:}
+      \shortintertext{\llap{(ii)~}Predicate~$\isblocked$:}
       %%-
@@
       %%-
-      \intertext{\llap{(iii)~}Function~$\clear$:}
+      \shortintertext{\llap{(iii)~}Function~$\clear$:}
       %%-
@@
 description either no rule is applicable or exactly one of them can be
-applied.  This means that the resulting relation~$\trans$ is in fact a
-partial function.
+applied.  This means that the resulting transition relation~$\trans$ is in
+fact a partial function.
 
@@
 inspection of rules~\R{push} and~\R{pop}.  The proof of
-Lemma~\ref{lem.x.det-nst}, however, requires an induction on the structure
+Lemma~\ref{lem.x.det-nst} follows by induction on the structure
 of the derivation trees produced by the rules for~$\nst$.  Both lemmas are
-used in the proof of Theorem~\ref{thm.x.det}, the main result of this
-section.  Theorem~\ref{thm.x.det} establishes that any given number of
-applications of~$\trans$ starting from the same input description will
-always lead to the same output description.
-
-\begin{lemma}
+used in the proof of the next theorem.
+
+\begin{restatable}{lemma}{lemxdetout}
   \label{lem.x.det-out}
@@
   If~$\delta\out\delta_1$ and~$\delta\out\delta_2$ then~$\delta_1=\delta_2$.
-\end{lemma}
-
-\begin{lemma}
+\end{restatable}
+
+\begin{restatable}{lemma}{lemxdetnst}
   \label{lem.x.det-nst}
@@
   If~$\delta\nst\delta_1$ and~$\delta\nst\delta_2$ then~$\delta_1=\delta_2$.
-\end{lemma}
-
-\begin{theorem}[Determinism]
-  \label{thm.x.det}\strut\\
-  %%
-  If~$\delta\trans[i]\delta_1$ and~$\delta\trans[i]\delta_2$
-  then~$\delta_1=\delta_2$.
-\end{theorem}
+\end{restatable}
+
+The main result of this section, Theorem~\ref{thm.x.det}, establishes that
+any given number~$i\ge0$ of applications of arbitrary transition rules,
+starting from the same input description, will always lead to the same output
+description.  In other words, any finite sequence of transitions behave
+deterministically.
+
+\begin{restatable}[Determinism]{theorem}{thmxdet}
+  \label{thm.x.det}
+  %%
+  $\delta\trans[i]\delta_1$ and~$\delta\trans[i]\delta_2$
+  implies~$\delta_1=\delta_2$.
+\end{restatable}
+
 \begin{proof}
   By induction on~$i$.  The theorem is trivially true if~$i=0$ and follows
-  directly from the lemmas if~$i=1$.  Suppose
+  directly from the previous lemmas if~$i=1$.  Suppose
   \[
@@
 
-In this section, we prove that any sufficiently long sequence of
-applications of~$\trans$ will lead to an irreducible description, i.e., one
-that cannot be modified by further transitions.  Before doing that, however,
-we need to introduce some notation and establish some basic properties
-of~$\nst$ and~$\out$.
+We now turn to the problem of termination.  We want to show that any
+sufficiently long sequence of applications of arbitrary transition rules
+will eventually lead to an irreducible description, i.e., one that cannot be
+modified by further transitions.  Before doing that, however, we need to
+introduce some notation and establish some basic properties of
+transitions~$\nst$ and~$\out$.
 
@@
   If~$\<p_1,n,e>\nst[i]\<p_1',n,e'>$, for any~$p_2$:
-  \begin{enumerate}[(a)]
+  \begin{enumerate:a}
   \item\label{lem.x.props-nst-i.a}
@@
     $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}'\AtOr{p_2}},n,e'>$.
-  \end{enumerate}
+  \end{enumerate:a}
   \smallskip
@@
   that~$\isblocked(p_1,n)$:
-  \begin{enumerate}[(a)]
+  \begin{enumerate:a}
     \setcounter{enumi}{4}
@@
     $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtOr{p_2'}},n,e'>$.
-  \end{enumerate}
+  \end{enumerate:a}
 \end{lemma}
@@
   %%
-  \begin{enumerate}[(a)]
+  \begin{enumerate:a}
   \item\label{ass.x.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contains
@@
     where $p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
-  \end{enumerate}
+  \end{enumerate:a}
 \end{assumption}
@@
   clauses~\eqref{def.x.pot.first}--\eqref{def.x.pot.last}:
-  \begin{enumerate}[(a)]
+  \begin{enumerate:a}
   \item\label{def.x.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$.
@@
     $\pot(\_)=0$.
-  \end{enumerate}
+  \end{enumerate:a}
 \end{definition}
@@
   %%
-  \begin{enumerate}[(a)]
+  \begin{enumerate:a}
   \item\label{lem.rank-out-push} If~$\delta\outpush\delta'$
@@
     then~$\rank(\delta)>\rank(\delta')$.
-  \end{enumerate}
+  \end{enumerate:a}
 \end{lemma}

### 2018-04-15 06-appendix.tex
Add artifact evaluation
Cleanup and compress stuff in Section 3 up to Determinism

--- 2018-02-24 06-appendix.tex
@@
-\clearpage
-\appendix
-\section{Appendix: Detailed Proofs (optional)}
+\section{Detailed Proofs}
 \label{sec.proofs}
 
-\subsection*{Determinism}
-
-\begin{lemma}
-  \label{lem.det-out}
-  %%
-  If~$\delta\out\delta_1$ and~$\delta\out\delta_2$ then~$\delta_1=\delta_2$.
-\end{lemma}
+\lemxdetout*
 \begin{proof}
@@
   two possibilities.
-  \begin{case}
-    $e\ne\nil$.  Both transitions are applications of~\R{push}.
+  \begin{casex}
+  \item $e\ne\nil$.  Both transitions are applications of~\R{push}.
     Hence~$p_1=p_2=\bcast(p,e)$, $n_1=n_2=n+1$, and~$e_1=e_2=\nil$.
-  \end{case}
-  \begin{case}
-    $e=\nil$.  Both transitions are applications of~\R{pop}.
-    Hence~$p_1=p_2=p$, $n_1=n_2=n-1$, and~$e_1=e_2=\nil$.\qedhere
-  \end{case}
+    %%
+  \item $e=\nil$.  Both transitions are applications of~\R{pop}.
+    Hence $p_1=p_2=p$, $n_1=n_2=n-1$, and~$e_1=e_2=\nil$.\qedhere
+  \end{casex}
 \end{proof}
 
-% \begin{theorem}
-%   \label{thm.det-out-i}
-%   %%
-%   If~$\delta\out[i]\delta_1$ and~$\delta\out[i]\delta_2$
-%   then~$\delta_1=\delta_2$.
-% \end{theorem}
-% \begin{proof}
-%   By induction on~$i$.
-%   %%
-%   The theorem is trivially true if~$i=0$ and follows directly from
-%   Lemma~\ref{lem.det-out} if~$i=1$.  (The theorem is vacuously true
-%   for~$\outpush$ transitions if~$i>1$ since they cannot be applied more
-%   than once in a row and cannot occur after a~$\outpop$ transition.)
-%   %%
-%   Suppose
-%   \[
-%     \delta\out[1]\delta_1'\out[i-1]\delta_1
-%     \quad\text{and}\quad
-%     \delta\out[1]\delta_2'\out[i-1]\delta_2\,,
-%   \]
-%   for some~$i>1$, $\delta_1'$ and~$\delta_2'$.
-%   %%
-%   By Lemma~\ref{lem.det-out}, $\delta_1'=\delta_2'$.  By the induction
-%   hypothesis, $\delta_1=\delta_2$.\qedhere
-% \end{proof}
-
-\begin{lemma}
-  \label{lem.det-nst}
-  %%
-  If~$\delta\nst\delta_1$ and~$\delta\nst\delta_2$ then~$\delta_1=\delta_2$.
-\end{lemma}
+\lemxdetnst*
 \begin{proof}
@@
   $\ceu{\AwaitInt}$, $\ceu{\Break}$, $\ceu{\Every}$, $\ceu{\Fin}$,
-  or~$\ceu{\Nop}$ expression as there are no~$\nst$ rules to transition such
+  or~$\ceu{\Nop}$ statement as there are no~$\nst$ rules to transition such
   programs.)
 
-  \begin{case}
-    $p=\ceu{\Mem(\Id)}$.
+  \begin{casex}
+  \item $p=\ceu{\Mem(\Id)}$.
     %%
@@
     Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{\EmitInt(e')}$.
+    %%
+  \item $p=\ceu{\EmitInt(e')}$.
     %%
@@
     Hence~$p_1=p_2=\ceu{\CanRun(n)}$ and~$e_1=e_2=e'$.
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{\CanRun(n)}$.
-    %%
+    %%
+  \item $p=\ceu{\CanRun(n)}$.
+    %% \
     Then~$\pi_1$ and~$\pi_2$ are instances of~\R{can-run}.
     Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{\IfElse{\Mem(\Id)}{p'}{p''}}$.
+    %%
+  \item $p=\ceu{\IfElse{\Mem(\Id)}{p'}{p''}}$.
     %%
     There are two subcases.
-    \begin{subcase}
-      $\eval(\ceu{\Mem(\Id)})$~is true.
+    \begin{casex}
+    \item$\eval(\ceu{\Mem(\Id)})$.
       %%
@@
       Hence~$p_1=p_2=p'$ and~$e_1=e_2=\nil$.
-    \end{subcase}
-    \begin{subcase}
-      $\eval(\ceu{\Mem(\Id)})$ is false.
+      %%
+    \item$\lnot\eval(\ceu{\Mem(\Id)})$.
       %%
@@
       Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{p';\,p''}$.
+    \end{casex}
+    %%
+  \item$p=\ceu{p';\,p''}$.
     %%
     There are three subcases.
-    \begin{subcase}
-      $p'=\ceu{\Nop}$.
+    \begin{casex}
+    \item $p'=\ceu{\Nop}$.
       %%
@@
       Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
-    \end{subcase}
-    \begin{subcase}
-      $p'=\ceu{\Break}$.
+      %%
+    \item $p'=\ceu{\Break}$.
       %%
@@
       Hence~$p_1=p_2=\ceu{\Break}$ and~$e_1=e_2=\nil$.
-    \end{subcase}
-    \begin{subcase}
-      $p'\ne\ceu{\Nop},\ceu{\Break}$.
-      %%
-      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-adv}.
-      Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
+      %%
+    \item $p'\ne\ceu{\Nop},\ceu{\Break}$.
+      %%
+      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-adv}.  Thus there
+      are derivations~$\pi_1'$ and~$\pi_2'$ such that
       \begin{align*}
@@
       hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
-      Hence~$p_1=\ceu{p_1';p''}=\ceu{p_2';p''}=p_2$ and~$e_1=e_1'=e_2'=e_2$.
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{\Loop{p'}}$.
+      Hence~$p_1=\ceu{p_1';p''}=\ceu{p_2';p''}=p_2$
+      and~$e_1=e_1'=e_2'=e_2$.
+    \end{casex}
+    %%
+  \item $p=\ceu{\Loop{p'}}$.
     %%
@@
     Hence~$p_1=p_2=\ceu{p'\AtLoop{p'}}$ and~$e_1=e_2=\nil$.
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{p'\AtLoop{p''}}$.
+    %%
+  \item $p=\ceu{p'\AtLoop{p''}}$.
     %%
     There are three subcases.
-    \begin{subcase}
-      $p'=\ceu{\Nop}$.
+    \begin{casex}
+    \item $p'=\ceu{\Nop}$.
       %%
@@
       Hence~$p_1=p_2=\ceu{\Loop{p''}}$ and~$e_1=e_2=\nil$.
-    \end{subcase}
-    \begin{subcase}
-      $p'=\ceu{\Break}$.
+      %%
+    \item $p'=\ceu{\Break}$.
       %%
@@
       Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
-    \end{subcase}
-    \begin{subcase}
-      $p'\ne\ceu{\Nop},\ceu{\Break}$.
-      %%
-      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-adv}.
-      Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
+      %%
+    \item $p'\ne\ceu{\Nop},\ceu{\Break}$.
+      %%
+      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-adv}.  Thus there
+      are derivations~$\pi_1'$ and~$\pi_2'$ such that
       \begin{align*}
@@
       and~$e_1=e_1'=e_2'=e_2$.
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{p'\And{p''}}$.
+    \end{casex}
+  \item $p=\ceu{p'\And{p''}}$.
     %%
     Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-expd}.
-    Hence~$p_1=p_2=\ceu{{p'}\AtAnd{(\CanRun(n);\,p'')}}$ and~$e_1=e_2=\nil$.
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{p'\AtAnd{p''}}$.
+    Hence~$p_1=p_2=\ceu{{p'}\AtAnd{(\CanRun(n);\,p'')}}$
+    and~$e_1=e_2=\nil$.
+    %%
+  \item $p=\ceu{p'\AtAnd{p''}}$.
     %%
     There are two subcases.
-    \begin{subcase}
-      $\isblocked(p',n)$~is false.
+    \begin{casex}
+    \item$\lnot\isblocked(p',n)$.
       %%
       There are three subcases.
-      \begin{subsubcase}
-        $p'=\ceu{\Nop}$.
+      \begin{casex}
+      \item $p'=\ceu{\Nop}$.
         %%
@@
         Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
-      \end{subsubcase}
-      \begin{subsubcase}
-        \label{lem.det-nst.and-brk1}
+        %%
+      \item\label{lem.x.det-nst.and-brk1}
         $p'=\ceu{\Break}$.
@@
         Hence~$p_1=p_2=\ceu{\clear(p'');\Break}$ and~$e_1=e_2=\nil$.
-      \end{subsubcase}
-      \begin{subsubcase}
-        \label{lem.det-nst.and-adv1}
-        $p'\ne\ceu{\Nop},\ceu{\Break}$.
-        %%
-        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv1}.
-        Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
+        %%
+      \item\label{lem.x.det-nst.and-adv1}$p'\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv1}.  Thus
+        there are derivations~$\pi_1'$ and~$\pi_2'$ such that
         \begin{align*}
@@
         and~$e_1=e_1'=e_2'=e_2$.
-      \end{subsubcase}
-    \end{subcase}
-    \begin{subcase}
-      $\isblocked(p',n)$~is true.
+      \end{casex}
+      %%
+    \item$\isblocked(p',n)$.
       %%
       There are three subcases.
-      \begin{subsubcase}
-        $p''=\ceu{\Nop}$.
+      \begin{casex}
+      \item $p''=\ceu{\Nop}$.
         %%
@@
         Hence~$p_1=p_2=p'$ and~$e_1=e_2=\nil$.
-      \end{subsubcase}
-      \begin{subsubcase}
-        \label{lem.det-nst.and-brk2}
-        $p''=\ceu{\Break}$.
+        %%
+      \item\label{lem.x.det-nst.and-brk2}$p''=\ceu{\Break}$.
         %%
@@
         Hence~$p_1=p_2=\ceu{\clear(p');\Break}$ and~$e_1=e_2=\nil$.
-      \end{subsubcase}
-      \begin{subsubcase}
-        \label{lem.det-nst.and-adv2}
-        $p''\ne\ceu{\Nop},\ceu{\Break}$.
-        %%
-        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv2}.  Thus there
-        are derivations~$\pi_1''$ and~$\pi_2''$ such that
+        %%
+      \item\label{lem.x.det-nst.and-adv2}$p''\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-adv2}.  Thus
+        there are derivations~$\pi_1''$ and~$\pi_2''$ such that
         \begin{align*}
@@
         and~$e_1=e_1''=e_2''=e_2$.
-      \end{subsubcase}
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{p'\Or{p''}}$.
+      \end{casex}
+    \end{casex}
+    %%
+  \item$p=\ceu{p'\Or{p''}}$.
     %%
     Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-expd}.
-    Hence~$p_1=p_2=\ceu{{p'}\AtOr{(\CanRun(n);\,p'')}}$ and~$e_1=e_2=\nil$.
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{p'\AtOr{p''}}$.
+    Hence~$p_1=p_2=\ceu{{p'}\AtOr{(\CanRun(n);\,p'')}}$
+    and~$e_1=e_2=\nil$.
+    %%
+  \item$p=\ceu{p'\AtOr{p''}}$.
     %%
     There are two subcases.
-    \begin{subcase}
-      $\isblocked(p',n)$~is false.
+    \begin{casex}
+    \item$\lnot\isblocked(p',n)$.
       %%
       There are three subcases.
-      \begin{subsubcase}
-        $p'=\ceu{\Nop}$.
+      \begin{casex}
+      \item $p'=\ceu{\Nop}$.
         %%
@@
         Hence~$p_1=p_2=\clear(p'')$ and~$e_1=e_2=\nil$.
-      \end{subsubcase}
-      \begin{subsubcase}
-        $p'=\ceu{\Break}$.
-        %%
-        Similar to Case~\ref{lem.det-nst.and-brk1}.
-      \end{subsubcase}
-      \begin{subsubcase}
-        $p'\ne\ceu{\Nop},\ceu{\Break}$.
-        %%
-        Similar to Case~\ref{lem.det-nst.and-adv1}.
-      \end{subsubcase}
-    \end{subcase}
-    \begin{subcase}
-      $\isblocked(p',n)$~is true.
+        %%
+      \item $p'=\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{lem.x.det-nst.and-brk1}.
+        %%
+      \item$p'\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{lem.x.det-nst.and-adv1}.
+      \end{casex}
+      %%
+    \item$\isblocked(p',n)$.
       %%
       There are three subcases.
-      \begin{subsubcase}
-        $p''=\ceu{\Nop}$.
+      \begin{casex}
+      \item $p''=\ceu{\Nop}$.
         %%
@@
         Hence~$p_1=p_2=\clear(p')$ and~$e_1=e_2=\nil$.
-      \end{subsubcase}
-      \begin{subsubcase}
-        $p''=\ceu{\Break}$.
-        %%
-        Similar to Case~\ref{lem.det-nst.and-brk2}.
-      \end{subsubcase}
-      \begin{subsubcase}
-        $p''\ne\ceu{\Nop},\ceu{\Break}$.
-        %%
-        Similar to Case~\ref{lem.det-nst.and-adv2}.
-        %%
-        \qedhere
-      \end{subsubcase}
-    \end{subcase}
-  \end{case}
+        %%
+      \item $p''=\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{lem.x.det-nst.and-brk2}.
+        %%
+      \item $p''\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{lem.x.det-nst.and-adv2}.
+        %%
+      \end{casex}
+    \end{casex}
+  \end{casex}
 \end{proof}
 
-% \begin{theorem}
-%   \label{thm.det-nst-i}
-%   %%
-%   If~$\delta\nst[i]\delta_1$ and~$\delta\nst[i]\delta_2$
-%   then~$\delta_1=\delta_2$.
-% \end{theorem}
-% \begin{proof}
-%   By induction on~$i$.  Similar to the proof of
-%   Theorem~\ref{thm.det-out-pop-i}.
-% \end{proof}
-
-\begin{theorem}[Determinism]\label{thm.det}\strut\\
-  %%
-  If~$\delta\trans[i]\delta_1$ and~$\delta\trans[i]\delta_2$
-  then~$\delta_1=\delta_2$.
-\end{theorem}
+\thmxdet*
 \begin{proof}
   By induction on~$i$.  The theorem is trivially true if~$i=0$ and follows
-  directly from Lemmas~\ref{lem.det-out} and~\ref{lem.det-nst} for~$i=1$.
+  directly from Lemmas~\ref{lem.x.det-out} and~\ref{lem.x.det-nst} for~$i=1$.
   %%
@@
   There are two possibilities.
-  \begin{case}
-    $\delta\out[1]\delta_1'$ and~$\delta\out[1]\delta_2'$.  Then, by
-    Lemma~\ref{lem.det-out}, $\delta_1'=\delta_2'$, and by the induction
+  \begin{casex}
+  \item $\delta\out[1]\delta_1'$ and~$\delta\out[1]\delta_2'$.  Then, by
+    Lemma~\ref{lem.x.det-out}, $\delta_1'=\delta_2'$, and by the induction
     hypothesis, $\delta_1=\delta_2$.
-  \end{case}
-  \begin{case}
-    $\delta\nst[1]\delta_1'$ and~$\delta\nst[1]\delta_2'$.  Then, by
-    Lemma~\ref{lem.det-nst}, $\delta_1'=\delta_2'$, and by the induction
+    %%
+  \item $\delta\nst[1]\delta_1'$ and~$\delta\nst[1]\delta_2'$.  Then, by
+    Lemma~\ref{lem.x.det-nst}, $\delta_1'=\delta_2'$, and by the induction
     hypothesis, $\delta_1=\delta_2$.\qedhere
-  \end{case}
+  \end{casex}
 \end{proof}
-
-\subsection*{Termination}
 
@@
   If~$\<p_1,n,e>\nst[i]\<p_1',n,e'>$, for any~$p_2$:
-  \begin{enumerate}[(a)]
+  \begin{enumerate:a}
   \item\label{lem.props-nst-i.a}
@@
   %%
-  \begin{enumerate}[(a)]
+  \begin{enumerate:a}
   \item The lemma is trivially true for~$i=0$, as~$p_1=p_1'$, and follows
@@
   \item Similar to Case~\eqref{lem.props-nst-i.e}.\qedhere
-  \end{enumerate}
+  \end{enumerate:a}
 \end{proof}
@@
   %%
-  \begin{enumerate}[(a)]
+  \begin{enumerate:a}
   \item\label{ass.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contain no
@@
     where~$p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
-  \end{enumerate}
+  \end{enumerate:a}
 \end{assumption}
@@
   clauses~\label{def.pot.first}--\label{def.pot.last}:
-  \begin{enumerate}[(a)]
+  \begin{enumerate:a}
   \item\label{def.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$;
@@
     $\pot(\_)=0$.
-  \end{enumerate}
+  \end{enumerate:a}
   %%
@@
   $\rank(\delta)=\<i,j>$, and~$\rank(\delta')=\<i',j'>$.
-  \begin{enumerate}[(a)]
+  \begin{enumerate:a}
   \item Suppose~$\<p,n,e>\outpush\<p',n',e'>$.  Then, by~\R{push},
@@
     Therefore,~${\<i,j>}>{\<i',j'>}$.\qedhere
-  \end{enumerate}
+  \end{enumerate:a}
 \end{proof}

### 2018-04-15 03-semantics.tex
Typo in Section 3
Cleanup Section 3 and Appendix A up to Theorem 3.8

@@
 execute as a result of the~$\clear$ call.  The body of a~$\ceu{\Fin}$
-statement always execute withing a reaction.  This is due to a syntactic
+statement always execute within a reaction.  This is due to a syntactic
 restriction: $\ceu{\Fin}$ bodies cannot
@@
   A description~$\delta=\<p,n,e>$ is \emph{nested-irre\-ducible}
-  iff~$e\ne\nil$ or~$p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p,n)$~is
-  true.
+  iff~$e\ne\nil$ or~$p=\ceu{\Nop}$ or~$p=\ceu{\Break}$
+  or~$\isblocked(p,n)$.\footnote{We often abbreviate ``$p=\ceu{\Nop}$
+    or~$p=\ceu{\Break}$'' as~``$p=\ceu{\Nop,\Break}$''.}
 \end{definition}
@@
 The use of qualifier ``irreducible'' in Definition~\ref{def.x.Hnst} is
-justified by Proposition~\ref{prop.x.irr-nst-i}.  The proposition
-establishes that if a finite number of applications of~$\nst$ results in an
-irreducible description, then that occurs exactly once, at some specific
-number~$i$.  The proof of the lemma follows directly from the definition
-of~$\nst$ by contradiction on the hypothesis that there is such~$k\ne{i}$.
-
-\begin{proposition}
+justified by Proposition~\ref{prop.x.irr-nst-i}, which states that if a
+finite number of applications of~$\nst$ results in an irreducible
+description, then that occurs exactly once, at some specific number~$i$.
+The proof of Proposition~\ref{prop.x.irr-nst-i} follows directly from the
+definition of~$\nst$ by contradiction on the hypothesis that there is
+such~$k\ne{i}$.
+
+\begin{restatable}{proposition}{propxirrnsti}
   \label{prop.x.irr-nst-i}
@@
   no~$\delta_\Hnst''$ such that~$\delta\nst[k]\delta''_\Hnst$.
-\end{proposition}
-
-The next lemma establishes that sequences of~$\nst$ transitions behave
-as expected regarding the order of evaluation of composition members.  Its
+\end{restatable}
+
+The next lemma establishes that sequences of~$\nst$ transitions behave as
+expected regarding the order of evaluation of composition branches.  Its
 proof follows by induction on~$i$.
 
-\begin{lemma}
+\begin{restatable}{lemma}{lemxpropsnsti}
   \label{lem.x.props-nst-i}\strut\\
@@
   \end{enumerate:a}
-\end{lemma}
+\end{restatable}
 %%
@@
 
-The syntactic restrictions discussed in Section~\ref{sec.sem}, regarding the
-body of~$\ceu{\Fin}$ and~$\ceu{\Loop}$ expressions, and their consequences,
-are formalized in the next assumption.
+The syntactic restriction discussed in Section~\ref{sec.ceu.evts}, regarding
+the body of loops, and the restriction mentioned earlier in this section,
+regarding the body of~$\ceu{\Fin}$ statements, are formalized in
+Assumption~\ref{ass.x.syn-rest} below.  These restrictions are essential to
+prove the next theorem, Theorem~\ref{thm.x.term-nst-*}.
 
@@
 number of~$\nst$ transitions eventually leads to a nested-irreducible
-description.  The proof is by induction on the structure of programs
-(members of set~$P$) and uses Lemma~\ref{lem.x.props-nst-i} and
-Assumption~\ref{ass.x.syn-rest}.
-
-\begin{theorem}
+description.  Put another way, it establishes that, for any input
+description~$\delta$, it is always possible to transform~$\delta$ in a
+nested-irreducible description~$\delta'$ by applying to it a sufficiently
+long sequence of~$\nst$ transitions.  The proof follows by induction on the
+structure of programs (members of set~$P$) and depends on
+Lemma~\ref{lem.x.props-nst-i} and Assumption~\ref{ass.x.syn-rest}.
+
+\begin{restatable}{theorem}{thmxtermnstx}
   \label{thm.x.term-nst-*}
@@
   that~$\delta\nst[*]\delta'_\Hnst$.
-\end{theorem}
+\end{restatable}
 

### 2018-04-15 06-appendix.tex
Cleanup Section 3 and Appendix A up to Theorem 3.8

@@
 
-\begin{definition}
-  \label{def.Hnst}
-  %%
-  A description~$\delta=\<p,n,e>$ is \emph{nested-irre\-ducible}
-  iff~$e\ne\nil$ or~$p=\ceu{\Nop},\ceu{\Break}$ or~$\isblocked(p,n)$~is
-  true.
-\end{definition}
-
-\begin{proposition}
-  \label{prop.irr-nst-i}
-  %%
-  If~$\delta\nst[i]\delta_\Hnst'$ then, for all~$k\ne{i}$, there is
-  no~$\delta_\Hnst''$ such that~$\delta\nst[k]\delta''_\Hnst$.
-\end{proposition}
+\propxirrnsti*
 \begin{proof}
@@
   There are two cases.
-  \begin{case}
-    \label{prop.irr-nst-i-case1}
+  \begin{casex}
+  \item\label{prop.x.irr-nst-i-case1}
     %%
@@
     \begin{equation}
-      \label{prop.irr-nst-i-eq1}
+      \label{prop.x.irr-nst-i-eq1}
       %%
@@
     that~$\delta'\nst[1]\delta_1'$, which
-    contradicts~\eqref{prop.irr-nst-i-eq1}.  Therefore, no such~$k$ can
+    contradicts~\eqref{prop.x.irr-nst-i-eq1}.  Therefore, no such~$k$ can
     exist.
-  \end{case}
-  \begin{case}
-    Suppose there are~$k<i$ and~$\delta''_\Hnst$ such
+    %%
+  \item Suppose there are~$k<i$ and~$\delta''_\Hnst$ such
     that~$\delta\nst[k]\delta''$.  Then, since~$i>k$, by
-    Case~\ref{prop.irr-nst-i-case1}, $\delta'$~cannot exist, which is
+    Case~\ref{prop.x.irr-nst-i-case1}, $\delta'$~cannot exist, which is
     absurd.  Therefore, the assumption that there is such~$k$ is
     false.\qedhere
-  \end{case}
+  \end{casex}
 \end{proof}
 
-\begin{lemma}
-  \label{lem.props-nst-i}\strut\\
-  %%
-  If~$\<p_1,n,e>\nst[i]\<p_1',n,e'>$, for any~$p_2$:
-  \begin{enumerate:a}
-  \item\label{lem.props-nst-i.a}
-    $\<\ceu{p_1;\,p_2},n,e>\nst[i]\<p_1';p_2,n,e'>$.
-    %%
-  \item\label{lem.props-nst-i.b}
-    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[i]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$.
-    %%
-  \item\label{lem.props-nst-i.c}
-    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$.
-    %%
-  \item\label{lem.props-nst-i.d}
-    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}'\AtOr{p_2}},n,e'>$.
-  \end{enumerate:a}
-  \smallskip
-  If~$\<p_2,n,e>\nst[i]\<p_2',n,e'>$, for any~$p_1$ such
-  that~$\isblocked(p_1,n)$:
-  \begin{enumerate:a}
-    \setcounter{enumi}{4}
-  \item\label{lem.props-nst-i.e}
-    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>$.
-    %%
-  \item\label{lem.props-nst-i.f}
-    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtOr{p_2'}},n,e'>$.
-  \end{enumerate:a}
-\end{lemma}
+\lemxpropsnsti*
 \begin{proof}
@@
     \begin{equation}
-      \label{lem.props-nst-i.a.eq1}
+      \label{lem.x.props-nst-i.a.eq1}
       \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[i-1]\<p_1',n,e'>\,,
@@
     for some~$i>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible, i.e.,
-    $e=\nil$ and~$p\ne{\ceu{\Nop},\ceu{\Break}}$ and~$\isblocked(p_1'',n)$
-    is false.  By~\eqref{lem.props-nst-i.a.eq1} and by~\R{seq-adv},
+    $e=\nil$ and~$p\ne{\ceu{\Nop},\ceu{\Break}}$
+    and~$\lnot\isblocked(p_1'',n)$.  By~\eqref{lem.x.props-nst-i.a.eq1} and
+    by~\R{seq-adv},
     \begin{equation}
-      \label{lem.props-nst-i.a.eq2}
+      \label{lem.x.props-nst-i.a.eq2}
       \<\ceu{p_1;\,p_2},n,e>\nst[1]\<\ceu{p_1'';\,p_2},n,e''>\,.
     \end{equation}
-    From~\eqref{lem.props-nst-i.a.eq1}, by the induction hypothesis,
+    From~\eqref{lem.x.props-nst-i.a.eq1}, by the induction hypothesis,
     \begin{equation}
-      \label{lem.props-nst-i.a.eq3}
+      \label{lem.x.props-nst-i.a.eq3}
       \<\ceu{p_1'';\,p_2},n,e''>\nst[i-1]\<\ceu{p_1';\,p_2},n,e'>\,.
     \end{equation}
-    From~\eqref{lem.props-nst-i.a.eq2} and~\eqref{lem.props-nst-i.a.eq3},
+    From~\eqref{lem.x.props-nst-i.a.eq2} and~\eqref{lem.x.props-nst-i.a.eq3},
     \[
@@
 
-  \item Similar to Case~\eqref{lem.props-nst-i.a}.
+  \item Similar to item~\ref{lem.x.props-nst-i.a}.
     %%
@@
     % \begin{equation}
-    %   \label{lem.props-nst-i.b.eq1}
+    %   \label{lem.x.props-nst-i.b.eq1}
     %   \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>\,,
@@
     % for some~$n>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible.
-    % By~\eqref{lem.props-nst-i.b.eq1} and by~\R{loop-adv},
+    % By~\eqref{lem.x.props-nst-i.b.eq1} and by~\R{loop-adv},
     % \begin{equation}
-    %   \label{lem.props-nst-i.b.eq2}
+    %   \label{lem.x.props-nst-i.b.eq2}
     %   \<\ceu{p_1\AtLoop{p_2}},n,e>\nst[1]\<\ceu{p_1''\Loop{p_2}},n,e''>\,.
     % \end{equation}
-    % From~\eqref{lem.props-nst-i.b.eq1}, by the induction hypothesis,
+    % From~\eqref{lem.x.props-nst-i.b.eq1}, by the induction hypothesis,
     % \begin{equation}
-    %   \label{lem.props-nst-i.b.eq3}
+    %   \label{lem.x.props-nst-i.b.eq3}
     %   \<\ceu{p_1''\AtLoop{p_2}},n,e''>
@@
     % \end{equation}
-    % From~\eqref{lem.props-nst-i.b.eq2} and~\eqref{lem.props-nst-i.b.eq3},
+    % From~\eqref{lem.x.props-nst-i.b.eq2} and~\eqref{lem.x.props-nst-i.b.eq3},
     % \[
@@
     % \begin{equation}
-    %   \label{lem.props-nst-i.c.eq1}
+    %   \label{lem.x.props-nst-i.c.eq1}
     %   \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>\,,
@@
     % for some~$n>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible.
-    % By~\eqref{lem.props-nst-i.c.eq1} and by~\R{and-adv1},
+    % By~\eqref{lem.x.props-nst-i.c.eq1} and by~\R{and-adv1},
     % \begin{equation}
-    %   \label{lem.props-nst-i.c.eq2}
+    %   \label{lem.x.props-nst-i.c.eq2}
     %   \<\ceu{{p_1}\AtAnd{p_2}},n,e>
@@
     % \end{equation}
-    % From~\eqref{lem.props-nst-i.c.eq1}, by the induction hypothesis,
+    % From~\eqref{lem.x.props-nst-i.c.eq1}, by the induction hypothesis,
     % \begin{equation}
-    %   \label{lem.props-nst-i.a.eq3}
+    %   \label{lem.x.props-nst-i.a.eq3}
     %   \<\ceu{{p_1''}\AtAnd{p_2}},n,e''>
@@
     % \end{equation}
-    % From~\eqref{lem.props-nst-i.c.eq2} and~\eqref{lem.props-nst-i.c.eq3},
+    % From~\eqref{lem.x.props-nst-i.c.eq2} and~\eqref{lem.x.props-nst-i.c.eq3},
     % \[
@@
     % \begin{equation}
-    %   \label{lem.props-nst-i.d.eq1}
+    %   \label{lem.x.props-nst-i.d.eq1}
     %   \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[n-1]\<p_1',n,e'>\,,
@@
     % for some~$n>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible.
-    % By~\eqref{lem.props-nst-i.d.eq1} and by~\R{or-adv1},
+    % By~\eqref{lem.x.props-nst-i.d.eq1} and by~\R{or-adv1},
     % \begin{equation}
-    %   \label{lem.props-nst-i.d.eq2}
+    %   \label{lem.x.props-nst-i.d.eq2}
     %   \<\ceu{{p_1}\AtOr{p_2}},n,e>
@@
     % \end{equation}
-    % From~\eqref{lem.props-nst-i.d.eq1}, by the induction hypothesis,
+    % From~\eqref{lem.x.props-nst-i.d.eq1}, by the induction hypothesis,
     % \begin{equation}
-    %   \label{lem.props-nst-i.a.eq3}
+    %   \label{lem.x.props-nst-i.a.eq3}
     %   \<\ceu{{p_1''}\AtOr{p_2}},n,e''>
@@
     % \end{equation}
-    % From~\eqref{lem.props-nst-i.d.eq2} and~\eqref{lem.props-nst-i.d.eq3},
+    % From~\eqref{lem.x.props-nst-i.d.eq2} and~\eqref{lem.x.props-nst-i.d.eq3},
     % \[
@@
     \begin{equation}
-      \label{lem.props-nst-i.e.eq1}
+      \label{lem.x.props-nst-i.e.eq1}
       \<p_2,n,e>\nst[1]\<p_2'',n,e''>\nst[i-1]\<p_2',n,e'>\,,
@@
     for some~$i>1$.  Then~$\<p_2'',n,e''>$ is not nested-irreducible.
-    By~\eqref{lem.props-nst-i.e.eq1} and by~\R{and-adv2},
+    By~\eqref{lem.x.props-nst-i.e.eq1} and by~\R{and-adv2},
     \begin{equation}
-      \label{lem.props-nst-i.e.eq2}
+      \label{lem.x.props-nst-i.e.eq2}
       \<\ceu{{p_1}\AtAnd{p_2}},n,e>
@@
     \end{equation}
-    From~\eqref{lem.props-nst-i.e.eq1}, by the induction hypothesis,
+    From~\eqref{lem.x.props-nst-i.e.eq1}, by the induction hypothesis,
     \begin{equation}
-      \label{lem.props-nst-i.e.eq3}
+      \label{lem.x.props-nst-i.e.eq3}
       \<\ceu{{p_1}\AtAnd{p_2''}},n,e''>
@@
     \end{equation}
-    From~\eqref{lem.props-nst-i.e.eq2} and~\eqref{lem.props-nst-i.e.eq3},
+    From~\eqref{lem.x.props-nst-i.e.eq2} and~\eqref{lem.x.props-nst-i.e.eq3},
     \[
@@
 
-  \item Similar to Case~\eqref{lem.props-nst-i.e}.\qedhere
+  \item Similar to item~\ref{lem.x.props-nst-i.e}.\qedhere
   \end{enumerate:a}
@@
 
-\begin{assumption}[Syntactic restrictions]\strut
-  \label{ass.syn-rest}
-  %%
-  \begin{enumerate:a}
-  \item\label{ass.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contain no
-    occurrences of~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$,
-    or~$\ceu{\Fin}$ expressions.  Consequently, for any~$n$,
-    \[
-      \<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>\,.
-    \]
-    %%
-  \item\label{ass.syn-rest.loop} If~$p=\ceu{\Loop{p_1}}$ then all execution
-    paths of~$p_1$ contain at least one occurrence of a
-    matching~$\ceu{\Break}$ or an~$\ceu{\AwaitExt}$ expression.
-    Consequently, for all~$n$, there are~$p_1'$ and~$e$ such that
-    \[
-      \<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>\,,
-    \]
-    where~$p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
-  \end{enumerate:a}
-\end{assumption}
-
-\begin{theorem}
-  \label{thm.term-nst-*}
-  %%
-  For any~$\delta$ there is a~$\delta'_\Hnst$ such
-  that~$\delta\nst[*]\delta'_\Hnst$.
-\end{theorem}
+\thmxtermnstx*
 \begin{proof}
@@
   indeed exists.
-  \begin{case}
-    $p=\ceu{\Mem(\Id)}$.
+  \begin{casex}
+  \item $p=\ceu{\Mem(\Id)}$.
     %%
@@
     \]
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{\EmitInt(e)}$.
+
+  \item $p=\ceu{\EmitInt(e)}$.
     %%
@@
     \]
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{\CanRun(n)}$.
+
+  \item $p=\ceu{\CanRun(n)}$.
     %%
@@
     \]
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{\IfElse{\Mem(\Id)}{p'}{p''}}$.
+
+  \item $p=\ceu{\IfElse{\Mem(\Id)}{p'}{p''}}$.
     %%
     There are two subcases.
-    \begin{subcase}
-      \label{thm.term-nst-*.if-true}
-      $\eval(\ceu{\Mem(\Id)})$~is true.
+    \begin{casex}
+    \item\label{thm.x.term-nst-*.if-true} $\eval(\ceu{\Mem(\Id)})$.
       %%
@@
       \end{align*}
-    \end{subcase}
-    \begin{subcase}
-      $\eval(\ceu{\Mem(\Id)})$~is false.
-      %%
-      Similar to Case~\ref{thm.term-nst-*.if-true}.
+      %%
+    \item$\lnot\eval(\ceu{\Mem(\Id)})$.
+      %%
+      Similar to Case~\ref{thm.x.term-nst-*.if-true}.
       %%
@@
       %%
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{p';\,p''}$.
-    %%
-    There are three subcases.
-    \begin{subcase}
-      \label{thm.term-nst-*.seq-nop}
-      $p'=\ceu{\Nop}$.
+    \end{casex}
+
+  \item $p=\ceu{p';\,p''}$.
+    %%
+    There are three subcases.
+    \begin{casex}
+    \item\label{thm.x.term-nst-*.seq-nop} $p'=\ceu{\Nop}$.
       %%
@@
       \]
-    \end{subcase}
-    \begin{subcase}
-      \label{thm.term-nst-*.seq-brk}
-      $p'=\ceu{\Break}$.
+      %%
+    \item\label{thm.x.term-nst-*.seq-brk} $p'=\ceu{\Break}$.
       %%
@@
       \]
-    \end{subcase}
-    \begin{subcase}
-      \label{thm.term-nst-*.seq-adv}
+      %%
+    \item\label{thm.x.term-nst-*.seq-adv}
       $p'\ne\ceu{\Nop},\ceu{\Break}$.
@@
       \]
-      By item~\eqref{lem.props-nst-i.a} of Lemma~\ref{lem.props-nst-i},
+      By item~\ref{lem.x.props-nst-i.a} of Lemma~\ref{lem.x.props-nst-i},
       \begin{equation}
-        \label{thm.term-nst-*.seq-adv.eq1}
+        \label{thm.x.term-nst-*.seq-adv.eq1}
         \<\ceu{p';\,p''},n,\nil>\nst[*]\<\ceu{p_1';\,p''},n,e>\,.
@@
       %%
-      \begin{subsubcase}
-        $e\ne\nil$.  Then, by the definition of~$\Hnst$,
+      \begin{casex}
+      \item $e\ne\nil$.  Then, by the definition of~$\Hnst$,
         description~$\<\ceu{p_1';\,p''},n,e>$ is nested-irreducible.
-      \end{subsubcase}
-      \begin{subsubcase}
-        $p_1'=\ceu{\Nop}$.
-        %%
-        From~\eqref{thm.term-nst-*.seq-adv.eq1},
+        %%
+      \item $p_1'=\ceu{\Nop}$.
+        %%
+        From~\eqref{thm.x.term-nst-*.seq-adv.eq1},
         \[
@@
         From this point on, this case is similar to
-        Case~\ref{thm.term-nst-*.seq-nop}.
-        %%
-      \end{subsubcase}
-      \begin{subsubcase}
-        $p_1'=\ceu{\Break}$.
-        %%
-        From~\eqref{thm.term-nst-*.seq-adv.eq1},
+        Case~\ref{thm.x.term-nst-*.seq-nop}.
+        %%
+      \item $p_1'=\ceu{\Break}$.
+        %%
+        From~\eqref{thm.x.term-nst-*.seq-adv.eq1},
         \[
@@
         From this point on, this case is similar to
-        Case~\ref{thm.term-nst-*.seq-brk}.
-      \end{subsubcase}
-      \begin{subsubcase}
-        $\isblocked(p_1',n)$~is true.
+        Case~\ref{thm.x.term-nst-*.seq-brk}.
+        %%
+      \item$\isblocked(p_1',n)$.
         %%
@@
         \]
-        Hence, from~\eqref{thm.term-nst-*.seq-adv.eq1} and by the
+        Hence, from~\eqref{thm.x.term-nst-*.seq-adv.eq1} and by the
         definition~$\Hnst$, description~$\<\ceu{p_1';\,p''},n,e>$ is
         nested-irreducible.
-      \end{subsubcase}
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    \label{thm.term-nst-*.loop}
+      \end{casex}
+    \end{casex}
+
+  \item\label{thm.x.term-nst-*.loop}
     $p=\ceu{\Loop{p'}}$.
     %%
-    Then, by item~\eqref{ass.syn-rest.loop} of
-    Assumption~\ref{ass.syn-rest},
-    \begin{equation}\label{thm.term-nst-*.loop-expd.eq1}
+    Then, by item~\ref{ass.x.syn-rest.loop} of
+    Assumption~\ref{ass.x.syn-rest},
+    \begin{equation}\label{thm.x.term-nst-*.loop-expd.eq1}
       \<\ceu{\Loop{p'}},n,\nil>\nst[*]\<p_1',n,e>\,,
@@
     or~$\isblocked(p_1',n)$.
-    \begin{subcase}
-      $p_1'=\ceu{\Break\AtLoop{p'}}$.
-      %%
-      From~\eqref{thm.term-nst-*.loop-expd.eq1}, by~\R{loop-brk},
+    \begin{casex}
+    \item$p_1'=\ceu{\Break\AtLoop{p'}}$.
+      %%
+      From~\eqref{thm.x.term-nst-*.loop-expd.eq1}, by~\R{loop-brk},
       \begin{align*}
@@
       \end{align*}
-    \end{subcase}
-    \begin{subcase}
-      $\isblocked(p_1',n)$ is true.  Hence,
-      from~\eqref{thm.term-nst-*.loop-expd.eq1} and by the definition
+      %%
+    \item$\isblocked(p_1',n)$.  Hence,
+      from~\eqref{thm.x.term-nst-*.loop-expd.eq1} and by the definition
       of~$\Hnst$, $\<p_1',n,e>_\Hnst$.
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{p'\AtLoop{p''}}$.
-    %%
-    There are three subcases.
-    \begin{subcase}
-      $p'=\ceu{\Nop}$.
+    \end{casex}
+
+  \item$p=\ceu{p'\AtLoop{p''}}$.
+    %%
+    There are three subcases.
+    \begin{casex}
+    \item$p'=\ceu{\Nop}$.
       %%
@@
       From this point on, this case is similar to
-      Case~\ref{thm.term-nst-*.loop}.
-    \end{subcase}
-    \begin{subcase}
-      $p'=\ceu{\Break}$.  Then, by~\R{loop-brk},
+      Case~\ref{thm.x.term-nst-*.loop}.
+      %%
+    \item$p'=\ceu{\Break}$.  Then, by~\R{loop-brk},
       \[
@@
       \]
-    \end{subcase}
-    \begin{subcase}
-      $p'\ne\ceu{\Nop},\ceu{\Break}$.  Then, by the induction hypothesis,
+      %%
+    \item$p'\ne\ceu{\Nop},\ceu{\Break}$.  Then, by the induction hypothesis,
       there are~$p_1'$ and~$e$ such that
@@
       \]
-      By item~\eqref{lem.props-nst-i.b} of Lemma~\ref{lem.props-nst-i},
+      By item~\ref{lem.x.props-nst-i.b} of Lemma~\ref{lem.x.props-nst-i},
       \[
@@
       nested-irreducible.  The rest of this proof is similar to that of
-      Case~\ref{thm.term-nst-*.seq-adv}.
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{{p'}\And{p''}}$.
+      Case~\ref{thm.x.term-nst-*.seq-adv}.
+    \end{casex}
+
+  \item$p=\ceu{{p'}\And{p''}}$.
     %%
@@
     From this point on, this case is similar to
-    Case~\ref{thm.term-nst-*.and}.
-  \end{case}
-
-  \begin{case}
-    \label{thm.term-nst-*.and}
+    Case~\ref{thm.x.term-nst-*.and}.
+    %%
+  \item\label{thm.x.term-nst-*.and}
     %%
@@
     There are two subcases.
-    \begin{subcase}
-      $\isblocked(p',n)$ is false.
-      %%
-      There are three subcases.
-      \begin{subsubcase}
-        \label{thm.term-nst-*.and-nop1}
+    \begin{casex}
+    \item$\lnot\isblocked(p',n)$.
+      %%
+      There are three subcases.
+      \begin{casex}
+      \item\label{thm.x.term-nst-*.and-nop1}
         $p'=\ceu{\Nop}$.
@@
         \]
-      \end{subsubcase}
-      \begin{subsubcase}
-        \label{thm.term-nst-*.and-brk1}
+        %%
+      \item\label{thm.x.term-nst-*.and-brk1}
         $p'=\ceu{\Break}$.
@@
         \begin{align}
-          \label{thm.term-nst-*.and-brk1.eq1}
+          \label{thm.x.term-nst-*.and-brk1.eq1}
           &\<\ceu{{\Break}\AtAnd{p''}},n,\nil>\\
@@
         \end{align}
-        From~\eqref{thm.term-nst-*.and-brk1.eq1}, by
-        item~\eqref{ass.syn-rest.fin} of Assumption~\ref{ass.syn-rest} and
-        by~\R{seq-nop},
+        From~\eqref{thm.x.term-nst-*.and-brk1.eq1}, by
+        item~\ref{ass.x.syn-rest.fin} of Assumption~\ref{ass.x.syn-rest}
+        and by~\R{seq-nop},
         \begin{align*}
@@
         \end{align*}
-      \end{subsubcase}
-      \begin{subsubcase}
-        \label{thm.term-nst-*.and-adv1}
+        %%
+      \item\label{thm.x.term-nst-*.and-adv1}
         $p'\ne\ceu{\Nop},\ceu{\Break}$.
@@
         \]
-        By item~\eqref{lem.props-nst-i.c} of Lemma~\ref{lem.props-nst-i},
+        By item~\ref{lem.x.props-nst-i.c} of Lemma~\ref{lem.x.props-nst-i},
         \[
@@
         \item If~$p_1'=\ceu{\Nop}$, this case is similar to
-          Case~\ref{thm.term-nst-*.and-nop1}.
+          Case~\ref{thm.x.term-nst-*.and-nop1}.
         \item If~$p_1'=\ceu{\Break}$, this case is similar to
-          Case~\ref{thm.term-nst-*.and-brk1}.
+          Case~\ref{thm.x.term-nst-*.and-brk1}.
         \item If~$\isblocked(p_1',n)$, this case is similar to
-          Case~\ref{thm.term-nst-*.and2}.
+          Case~\ref{thm.x.term-nst-*.and2}.
         \end{enumerate}
-      \end{subsubcase}
-    \end{subcase}
-    \begin{subcase}
-      \label{thm.term-nst-*.and2}
-      $\isblocked(p',n)$ is true.
-      %%
-      There are three subcases.
-      \begin{subsubcase}
-        \label{thm.term-nst-*.and-nop2}
+      \end{casex}
+      %%
+    \item\label{thm.x.term-nst-*.and2}
+      $\isblocked(p',n)$.
+      %%
+      There are three subcases.
+      \begin{casex}
+      \item\label{thm.x.term-nst-*.and-nop2}
         $p''=\ceu{\Nop}$.
@@
         \]
-      \end{subsubcase}
-      \begin{subsubcase}
-        \label{thm.term-nst-*.and-brk2}
+        %%
+      \item\label{thm.x.term-nst-*.and-brk2}
         $p''=\ceu{\Break}$.
@@
         From this point on, this case is similar to
-        Case~\ref{thm.term-nst-*.and-brk1}.
-      \end{subsubcase}
-      \begin{subsubcase}
-        \label{thm.term-nst-*.and-adv2}
+        Case~\ref{thm.x.term-nst-*.and-brk1}.
+        %%
+      \item\label{thm.x.term-nst-*.and-adv2}
         $p''\ne\ceu{\Nop},\ceu{\Break}$.
@@
         \]
-        By item~\eqref{lem.props-nst-i.e} of Lemma~\ref{lem.props-nst-i},
+        By item~\ref{lem.x.props-nst-i.e} of Lemma~\ref{lem.x.props-nst-i},
         \[
@@
         \item If~$p_1''=\ceu{\Nop}$, this case is similar to
-          Case~\ref{thm.term-nst-*.and-nop2}.
+          Case~\ref{thm.x.term-nst-*.and-nop2}.
         \item If~$p_1''=\ceu{\Break}$, this case is similar to
-          Case~\ref{thm.term-nst-*.and-brk2}.
+          Case~\ref{thm.x.term-nst-*.and-brk2}.
         \item If~$\isblocked(p_1'',n)$ then, as both sides are blocked, by
@@
         \end{enumerate}
-      \end{subsubcase}
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{{p'}\Or{p''}}$.
+      \end{casex}
+    \end{casex}
+
+  \item$p=\ceu{{p'}\Or{p''}}$.
     %%
@@
     From this point on, this case is similar to
-    Case~\ref{thm.term-nst-*.or}.
-  \end{case}
-
-  \begin{case}
-    \label{thm.term-nst-*.or}
+    Case~\ref{thm.x.term-nst-*.or}.
+
+  \item\label{thm.x.term-nst-*.or}
     $p=\ceu{{p'}\AtOr{p''}}$.
@@
     There are two subcases.
-    \begin{subcase}
-      $\isblocked(p',n)$~is false.
-      %%
-      There are three subcases.
-      \begin{subsubcase}
-        \label{thm.term-nst-*.or-nop1}
-        $p'=\ceu{\Nop}$.  Then, by~\R{or-nop1},
+    \begin{casex}
+    \item$\lnot\isblocked(p',n)$.
+      %%
+      There are three subcases.
+      \begin{casex}
+      \item\label{thm.x.term-nst-*.or-nop1} $p'=\ceu{\Nop}$.  Then,
+        by~\R{or-nop1},
         \begin{equation}
-          \label{thm.term-nst-*.or-nop1.eq1}
+          \label{thm.x.term-nst-*.or-nop1.eq1}
           \<\ceu{{\Nop}\AtOr{p''}},n,\nil>
@@
         \end{equation}
-        From~\eqref{thm.term-nst-*.or-nop1.eq1}, by
-        item~\eqref{ass.syn-rest.fin} Assumption~\ref{ass.syn-rest},
+        From~\eqref{thm.x.term-nst-*.or-nop1.eq1}, by
+        item~\ref{ass.x.syn-rest.fin} Assumption~\ref{ass.x.syn-rest},
         \[
@@
         \]
-      \end{subsubcase}
-      \begin{subsubcase}
-        \label{thm.term-nst-*.or-brk1}
+        %%
+      \item\label{thm.x.term-nst-*.or-brk1}
         $p'=\ceu{\Break}$.
         %%
-        Similar to Case~\ref{thm.term-nst-*.and-brk1}.
-      \end{subsubcase}
-      \begin{subsubcase}
-        $p'\ne\ceu{\Nop},\ceu{\Break}$.
-        %%
-        Similar to Case~\ref{thm.term-nst-*.and-adv1}.
-      \end{subsubcase}
-    \end{subcase}
-    \begin{subcase}
-      \label{thm.term-nst-*.or-adv1}
-      $\isblocked(p',n)$~is true.
-      %%
-      There are three subcases.
-      \begin{subsubcase}
-        $p''=\ceu{\Nop}$.
+        Similar to Case~\ref{thm.x.term-nst-*.and-brk1}.
+        %%
+      \item$p'\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{thm.x.term-nst-*.and-adv1}.
+      \end{casex}
+      %%
+    \item\label{thm.x.term-nst-*.or-adv1}$\isblocked(p',n)$.
+      %%
+      There are three subcases.
+      \begin{casex}
+      \item$p''=\ceu{\Nop}$.
         %%
@@
         \begin{equation}
-          \label{thm.term-nst-*.or-nop2.eq1}
+          \label{thm.x.term-nst-*.or-nop2.eq1}
           \<\ceu{p'\AtOr{\Nop}},n,\nil>
@@
         \end{equation}
-        From~\eqref{thm.term-nst-*.or-nop2.eq1}, by
-        item~\eqref{ass.syn-rest.fin} of Assumption~\ref{ass.syn-rest} and
+        From~\eqref{thm.x.term-nst-*.or-nop2.eq1}, by
+        item~\ref{ass.x.syn-rest.fin} of Assumption~\ref{ass.x.syn-rest} and
         by definition of~$\clear$,
@@
         \]
-      \end{subsubcase}
-      \begin{subsubcase}
-        $p''=\ceu{\Break}$.
-        %%
-        Similar to Case~\ref{thm.term-nst-*.and-brk2}.
-      \end{subsubcase}
-      \begin{subsubcase}
-        $p''\ne\ceu{\Nop},\ceu{\Break}$.
-        %%
-        Similar to Case~\ref{thm.term-nst-*.and-adv2}.\qedhere
-      \end{subsubcase}
-    \end{subcase}
-  \end{case}
+        %%
+      \item$p''=\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{thm.x.term-nst-*.and-brk2}.
+        %%
+      \item$p''\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{thm.x.term-nst-*.and-adv2}.\qedhere
+      \end{casex}
+    \end{casex}
+  \end{casex}
 \end{proof}
-
-\begin{definition}
-  \label{def.pot}
-  %%
-  The potency of a program~$p$ in reaction to event~$e$,
-  denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ expressions
-  that can be executed during a reaction of~$p$ to~$e$.
-
-  More formally,
-  \[
-    \pot(p,e)=\pot'(\bcast(p,e))\,,
-  \]
-  where~$\pot'$ is an auxiliary function that counts the number of
-  reachable~$\ceu{\EmitInt}$ expressions in the program resulting from the
-  broadcast of event~$e$ to~$p$.
-
-  The auxiliary function~$\pot'$ is defined by
-  clauses~\label{def.pot.first}--\label{def.pot.last}:
-  \begin{enumerate:a}
-  \item\label{def.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$;
-    %%
-  \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})
-    =\max\{\pot'(p_1),\pot'(p_2)\}$;
-    %%
-  \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$;
-    %%
-  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(p_1)+\pot'(p_2)$;
-    %%
-  \item$\pot'(\ceu{{p_1}\Or{p_2}})=\pot'(p_1)+\pot'(p_2)$;
-    %%
-  \item If~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
-    \begin{align*}
-      \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\
-      \pot'(\ceu{p_1\AtLoop{p_2}})&=
-      \begin{cases}
-        \pot'(p_1)              &\text{if\enspace(\dag)}\\
-        \pot'(p_1)+\pot'(p_2)   &\text{otherwise,}\\
-      \end{cases}
-    \end{align*}
-    where~(\dag) stands for: ``a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ occurs
-    in all execution paths of~$p_1$'';
-    %%
-  \item\label{def.pot.before-last} If~$p_1,p_2\ne\ceu{\Break}$,
-    \[
-      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,;
-    \]
-    %%
-  \item\label{def.pot.last} If~$p_1,p_2\ne\ceu{\Break}$
-    and~$p_1,p_2\ne\ceu{\Nop}$,
-    \[
-      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,;
-    \]
-    %%
-  \item Otherwise, if none
-    of~\eqref{def.pot.first}--\eqref{def.pot.before-last} applies,
-    $\pot(\_)=0$.
-  \end{enumerate:a}
-  %%
-  %%\gl{Should~$\pot'(\ceu{\protect\Every{e}\,{p_1}})\ne0$ above?}
-  %%
-\end{definition}
-
-\begin{definition}
-  \label{def.rank}
-  %%
-  The \emph{rank} of a description~$\delta=\<p,n,e>$,
-  denoted~$\rank(\delta)$, is a pair of nonnegative integers~$\<i,j>$ such
-  that
-  \begin{alignat*}{2}
-    i&=\pot(p,e) &\quad\text{and}\quad
-    j&=
-       \begin{cases}
-         n  &\text{if~$e=\nil$}\\
-         n+1&\text{otherwise\,.}
-       \end{cases}
-  \end{alignat*}
-\end{definition}
-
-\begin{definition}
-  \label{def.H}
-  %%
-  A description~$\delta$ is \emph{irreducible} (in symbols, $\delta_\#$) iff
-  it is nested-irreducible and its~$\rank(\delta)$ is~$\<i,0>$, for
-  some~$i\ge0$.
-\end{definition}
 
@@
     trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
-    nested-irreducible.  Then, by Theorem~\ref{thm.term-nst-*},
+    nested-irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
     $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
@@
           %%
-          Then, by~\R{out} and by Theorem~\ref{thm.term-nst-*}, there
+          Then, by~\R{out} and by Theorem~\ref{thm.x.term-nst-*}, there
           are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
@@
       %%
-      Then~$e=\nil$ and, by Theorems~\ref{thm.term-nst-*}
+      Then~$e=\nil$ and, by Theorems~\ref{thm.x.term-nst-*}
       and~\ref{thm.rank-nst-*}, there is a~$\delta'_\Hnst$ such

### 2018-04-16 03-semantics.tex
Continue refactoring of Section 3

@@
 
-The~$\ceu{\Mem(id)}$ primitive represents an assignment or system
-call that affects the memory location identified by~$id$.\footnotemark\
+The~$\ceu{\Mem(id)}$ primitive represents a read or write to the memory
+location identified by~$id$.\footnotemark\
 %
@@
 block~$\ceu{\Fin{p}}$ and the \texttt{@}-statements which do not exist in
-the concrete language.  (These result from the expansion of the transition
-rules to be presented.)
+the concrete language.  These result from the expansion of the transition
+rules to be presented.
 
@@
 emit-await pair.  In the concrete language, the \code{await} can be used as
-an expression which evaluates to the value stored in the emitted
+an expression which evaluates to the value stored in the corresponding emitted
 event, e.g., an
@@
 \subsection{Operational Semantics}
+\label{sec.sem.opsem}
 
@@
 transforms a triple consisting of a program~$p$, a stack level~$n$, and an
-emitted event~$e$ into a modified triple as follows:
+emitted event~$e$ into a possibly different triple, i.e.,
 \[
@@
 the last output description will block with a possibly modified program~$p'$ at stack
-level~0 and with no event emitted:
+level~0 and no event emitted:
 \[
@@
 and instantly broadcasts the event to the program, which means
-    (a)~awaking any active $\ceu{\AwaitExt}$ or $\ceu{\AwaitInt}$ statements
+    (i)~awaking any active $\ceu{\AwaitExt}$ or $\ceu{\AwaitInt}$ statements
     (see $\bcast$ in
         Figure~\ref{fig.bcast}),
-    (b)~creating a nested reaction by increasing the stack level, and, at the same time,
-    (c)~consuming the event ($e$ becomes~$\nil$).
+    (ii)~creating a nested reaction by increasing the stack level, and, at the same time,
+    (iii)~consuming the event ($e$ becomes~$\nil$).
 %
@@
 
-Given that our semantics focuses on control, rules \R{if-true} and
-\R{if-false} are the only to query~$\ceu{\Mem}$ as expressions.
+Rules \R{if-true} and \R{if-false} are the only rules that use~$\ceu{\Mem}$
+in a way that affects the control flow.
 %
@@
 %
-Then, if none of the trails is blocked in~$\ceu{\CanRun}$ statements, it means
-that the program cannot advance in the current reaction.
+Then, if none of the trails is blocked in~$\ceu{\CanRun}$, it means that the
+program cannot advance in the current reaction.
 %
@@
 \subsection{Properties}
+\label{sec.sem.props}
 
@@
 description either no rule is applicable or exactly one of them can be
-applied.  This means that the resulting transition relation~$\trans$ is in
-fact a partial function.
+applied (no choice involved).  This coupled with the fact that the output of
+every rule is a function of its input implies that transitions are
+deterministic: the same input description, if it can transition, will always
+result in the same output description.  In other words, the transition
+relation~$\trans$ is in fact a partial function.
 
@@
 inspection of rules~\R{push} and~\R{pop}.  The proof of
-Lemma~\ref{lem.x.det-nst} follows by induction on the structure
-of the derivation trees produced by the rules for~$\nst$.  Both lemmas are
-used in the proof of the next theorem.
+Lemma~\ref{lem.x.det-nst} follows by induction on the structure of the
+derivation trees produced by the rules for~$\nst$.  Both lemmas are used in
+the proof of the Theorem~\ref{thm.x.det}.
 
@@
 any given number~$i\ge0$ of applications of arbitrary transition rules,
-starting from the same input description, will always lead to the same output
-description.  In other words, any finite sequence of transitions behave
+starting from the same input description, will always lead to the same
+output description, i.e., any finite sequence of transitions behave
 deterministically.
@@
   iff~$e\ne\nil$ or~$p=\ceu{\Nop}$ or~$p=\ceu{\Break}$
-  or~$\isblocked(p,n)$.\footnote{We often abbreviate ``$p=\ceu{\Nop}$
+  or~$\isblocked(p,n)$.\footnote{We sometimes abbreviate ``$p=\ceu{\Nop}$
     or~$p=\ceu{\Break}$'' as~``$p=\ceu{\Nop,\Break}$''.}
@@
 
-The syntactic restriction discussed in Section~\ref{sec.ceu.evts}, regarding
-the body of loops, and the restriction mentioned earlier in this section,
-regarding the body of~$\ceu{\Fin}$ statements, are formalized in
-Assumption~\ref{ass.x.syn-rest} below.  These restrictions are essential to
-prove the next theorem, Theorem~\ref{thm.x.term-nst-*}.
+The syntactic restriction discussed in Section~\ref{sec.ceu.evts} regarding
+the body of loops and the restriction mentioned in
+Section~\ref{sec.sem.opsem} about the body of~$\ceu{\Fin}$ statements are
+formalized in Assumption~\ref{ass.x.syn-rest} below.  These restrictions are
+essential to prove the next theorem.
 
@@
   \item\label{ass.x.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contains
-    no occurrences of~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$,
-    or~$\ceu{\Fin}$ expressions.  Consequently, for any~$n$,
+    no occurrences of statements~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$,
+    $\ceu{\Every}$, or~$\ceu{\Fin}$.  And so, for any~$n$,
     $\<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>$.
@@
 general.  Before stating and proving it, we need to characterize irreducible
-descriptions in general.  This characterization depends on the notions of
-potency and rank.
+descriptions in general.  This characterization, Definition~\ref{def.x.H},
+and depends on the notions of potency and rank.
 
@@
   The \emph{potency} of a program~$p$ in reaction to event~$e$,
-  denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ expressions
-  that can be executed during a reaction of~$p$ to~$e$.
-
-  More formally, $\pot(p,e)=\pot'(\bcast(p,e))$ where~$\pot'$ is an
-  auxiliary function that counts the maximum number of
+  denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ statements
+  that can be executed in a reaction of~$p$ to~$e$, i.e.,
+  \[
+    \pot(p,e)=\pot'(\bcast(p,e))\,,
+  \]
+  where~$\pot'$ is an auxiliary function that counts the maximum number of
   reachable~$\ceu{\EmitInt}$ expressions in the program resulting from the
@@
 
-  The auxiliary function~$\pot'$ is defined by
-  clauses~\eqref{def.x.pot.first}--\eqref{def.x.pot.last}:
+  The function~$\pot'$ is defined by the following clauses:
   \begin{enumerate:a}
@@
     %%
-  \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})
-    =\max\{\pot'(p_1),\pot'(p_2)\}$.
+  \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})=$\\
+    \strut\quad$\max\{\pot'(p_1),\pot'(p_2)\}$.
     %%
@@
     %%
-  \item If~$p_1,p_2\ne\ceu{\Break}$.
+  \item If~$p_1,p_2\ne\ceu{\Break}$,
     \[
@@
   \item\label{def.x.pot.last} Otherwise, if none
-    of~\eqref{def.x.pot.first}--\eqref{def.x.pot.before-last} applies,
+    of~\ref{def.x.pot.first}--\ref{def.x.pot.before-last} applies,
     $\pot(\_)=0$.
@@
 
+Intuitively, the rank of a description~$\delta$ is a measure of the maximum
+amount of ``work'' (transitions) required to transform~$\delta$ into an
+irreducible description, as defined bellow.
+
 \begin{definition}
@@
 structure of~$\nst$ derivations.  Since Lemma~\ref{lem.x.rank-nst} is of
-fundamental importance for the termination result, we detail the proof of
-one of its cases.
+fundamental importance for the termination result, we present a sketch of its
+proof.
 

### 2018-04-16 03-semantics.tex
Finishes refactoring Sections 3 and 6

@@
 
-We will refer to the triples on the left-hand and right-hand sides of
+We refer to the triples on the left-hand and right-hand sides of
 symbol~$\trans$ as \emph{descriptions} (denoted~$\delta$).  The description
-on the left of~$\trans$ is called the \emph{input description},
-the one on its right is called the \emph{output description}.
+on the left of~$\trans$ is called the \emph{input description}, the one on
+its right is called the \emph{output description}.
 
@@
 
-At the beginning of a reaction to an input event~$id$, the input description is
-initialized with stack level~0 ($n=0$) and with the externally emitted event~$e=id$.
-%, but \code{emitInt} expressions can increase the stack level.
-At the end of a reaction, after an arbitrary but finite number of transitions,
-the last output description will block with a possibly modified program~$p'$ at stack
-level~0 and no event emitted:
+At the beginning of a reaction to an input event~$id$, the input description
+is initialized with stack level~0 ($n=0$) and with the externally emitted
+event~$e=\ceu{\Id}$.  At the end of a reaction, after an arbitrary but
+finite number of transitions, the last output description will block with a
+possibly modified program~$p'$ at stack level~0 and no event
+emitted\footnote{We write~$\trans[i]$ to mean~$i$ transitions in sequence,
+  and we write~$\trans[*]$ to mean a finite (possibly zero) number of
+  transitions in sequence.}:
 \[
@@
 that manipulate the stack level.
-\begin{align*}
-  &\AxiomC{$e\ne\nil$}
+\begin{gather*}
+  \AxiomC{$e\ne\nil$}
   \UnaryInfC{$\<p,n,e>\out\<\bcast(p,e),n+1,\nil>$}
@@
   %%
-  &\AxiomC{$n>0$}
+  \AxiomC{$n>0$}
   \AxiomC{$p=\ceu{\Nop} \vee p=\ceu{\Break} \vee \isblocked(p,n)$}
@@
   \Rtag{pop}
-\end{align*}
+\end{gather*}
 
@@
 
-A complete reaction consists of a series of transitions of the form
+A complete reaction consists of a series of transitions% of the form
 \begin{align*}
@@
 %
-where the~$\ast$'s stand for a finite (possibly zero) number of repetitions.
-First, a~$\outpush$ starts a nested reaction at level~1.
-Then, a series of alternations between zero or more~$\nst$ transitions (nested reactions) and a
-single~$\out$ transition (stack operation) takes place.
-Finally, a last~$\outpop$ transition decrements the
-stack level to~0 and terminates the reaction.
+First, a~$\outpush$ starts a nested reaction at level~1.  Then, a series of
+alternations between zero or more~$\nst$ transitions (nested reactions) and
+a single~$\out$ transition (stack operation) takes place.  Finally, a
+last~$\outpop$ transition decrements the stack level to~0 and terminates the
+reaction.
 
 The~$\nst$ rules for atoms are defined as follows:
+\begingroup
+\def\JOT{-.1\jot}
 \begin{align*}
@@
 \end{align*}
+\endgroup
 
@@
 \begingroup
-\def\JOT{.5\jot}
+\def\JOT{-.1\jot}
 \begin{gather*}
@@
 deterministic: the same input description, if it can transition, will always
-result in the same output description.  In other words, the transition
+result in the same output description.  Thus the transition
 relation~$\trans$ is in fact a partial function.
@@
 starting from the same input description, will always lead to the same
-output description, i.e., any finite sequence of transitions behave
-deterministically.
+output description.  In other words, any finite sequence of transitions
+behave deterministically.
 
@@
 \end{restatable}
-
 \begin{proof}
@@
 introduce some notation and establish some basic properties of
-transitions~$\nst$ and~$\out$.
+the transition relations~$\nst$ and~$\out$.
 
@@
 number of~$\nst$ transitions eventually leads to a nested-irreducible
-description.  Put another way, it establishes that, for any input
-description~$\delta$, it is always possible to transform~$\delta$ in a
-nested-irreducible description~$\delta'$ by applying to it a sufficiently
-long sequence of~$\nst$ transitions.  The proof follows by induction on the
-structure of programs (members of set~$P$) and depends on
-Lemma~\ref{lem.x.props-nst-i} and Assumption~\ref{ass.x.syn-rest}.
+description.  Hence, for any input description~$\delta$, it is always
+possible to transform~$\delta$ in a nested-irreducible description~$\delta'$
+by applying to it a sufficiently long sequence of~$\nst$ transitions.  The
+proof of the theorem follows by induction on the structure of programs
+(members of set~$P$) and depends on Lemma~\ref{lem.x.props-nst-i} and
+Assumption~\ref{ass.x.syn-rest}.
 
@@
 general.  Before stating and proving it, we need to characterize irreducible
-descriptions in general.  This characterization, Definition~\ref{def.x.H},
-and depends on the notions of potency and rank.
+descriptions in general.  This characterization, given in
+Definition~\ref{def.x.H}, depends on the notions of potency and rank.
 
@@
 
-  The function~$\pot'$ is defined by the following clauses:
+  Function~$\pot'$ is defined by the following clauses:
   \begin{enumerate:a}
@@
     %%
-  \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})=$\\
-    \strut\quad$\max\{\pot'(p_1),\pot'(p_2)\}$.
+  \item$\pot'(\ceu{\IfElse{\Mem(\Id)\!}{p_1\!}{p_2\!}})\!=\!
+    \max\{\pot'(p_1),\pot'(p_2)\}$.
     %%
@@
     %%
-  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(p_1)+\pot'(p_2)$.
-    %%
-  \item$\pot'(\ceu{{p_1}\Or{p_2}})=\pot'(p_1)+\pot'(p_2)$.
+  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(\ceu{{p_1}\Or{p_2}})=
+    \pot'(p_1)+\pot'(p_2)$.
     %%
@@
     \begin{align*}
-      \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\
+      \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\[-.5\jot]
       \pot'(\ceu{p_1\AtLoop{p_2}})&=
       \begin{cases}
-        \pot'(p_1)              &\text{if\enspace(\dag)}\\
-        \pot'(p_1)+\pot'(p_2)   &\text{otherwise,}\\
+        \pot'(p_1)              &\text{if\enspace(\dag)}\\[-.5\jot]
+        \pot'(p_1)+\pot'(p_2)   &\text{otherwise,}\\[-.5\jot]
       \end{cases}
@@
   \item If~$p_1,p_2\ne\ceu{\Break}$,
-    \[
-      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
-    \]
+    $\pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)$.
     %%
@@
     and~$p_1,p_2\ne\ceu{\Nop}$,
+    \vskip-.95em
     \[
@@
     of~\ref{def.x.pot.first}--\ref{def.x.pot.before-last} applies,
-    $\pot(\_)=0$.
+    $\pot'(\_)=0$.
   \end{enumerate:a}
@@
   \end{alignat*}
-\end{definition}
-
 Intuitively, the rank of a description~$\delta$ is a measure of the maximum
 amount of ``work'' (transitions) required to transform~$\delta$ into an
-irreducible description, as defined bellow.
+irreducible description, in the following sense.
+\end{definition}
 
@@
 proof of Lemma~\ref{lem.x.rank-nst}, however, is by induction on the
-structure of~$\nst$ derivations.  Since Lemma~\ref{lem.x.rank-nst} is of
-fundamental importance for the termination result, we present a sketch of its
-proof.
-
-\begin{lemma}\strut
+structure of~$\nst$ derivations.
+
+\begin{restatable}{lemma}{lemxrankout}\strut
   \label{lem.x.rank-out}
@@
   \begin{enumerate:a}
-  \item\label{lem.rank-out-push} If~$\delta\outpush\delta'$
+  \item\label{lem.x.rank-out-push} If~$\delta\outpush\delta'$
     then~$\rank(\delta)=\rank(\delta')$.
-  \item\label{lem.rank-out-pop} If~$\delta\outpop\delta'$
+  \item\label{lem.x.rank-out-pop} If~$\delta\outpop\delta'$
     then~$\rank(\delta)>\rank(\delta')$.
   \end{enumerate:a}
-\end{lemma}
-
-\begin{lemma}
+\end{restatable}
+
+%\vskip-.25em
+
+\begin{restatable}{lemma}{lemxranknst}
   \label{lem.x.rank-nst}
@@
   If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
-\end{lemma}
-\begin{proof}
-  By induction on the structure of~$\nst$ derivations.  Let
-  \[
-    \delta=\<p,n,e>
-    \quad\text{and}\quad
-    \delta'=\<p',n',e'>
-  \]
-  be descriptions with~$\rank(\delta)=\<i,j>$ and~$\rank(\delta')=\<i',j'>$
-  such that~$\delta\nst\delta'$.  Then there is a derivation~$\pi$ whose
-  conclusion is~$\delta\nst\delta'$, i.e.,
-  $\pi\Vdash\<p,n,e>\nst\<p',n',e'>$.
-  %%
-  % \[
-  %   \pi\Vdash\<p,n,e>\nst\<p',n',e'>\,.
-  % \]
-  %%
-  By definition of~$\nst$, $e=\nil$ and~$n=n'$.  Depending on the structure
-  of program~$p$, there are~11 distinct cases to be considered.  In each one
-  of them, we need to show that~$\rank(\delta)\ge\rank(\delta')$.
-
-  We proceed to prove the case where~$p=\ceu{{p_1}\AtLoop{p_2}}$, which
-  itself consists of three subcases.
-  \begin{case}
-    $p_1=\ceu{\Nop}$.
-    %%
-    Then~$\pi$ is an instance of~\R{loop-nop}.  i.e., its conclusion is
-    obtained by an application of this rule.  Hence~$p'=p_2$ and~$e'=\nil$.
-    Thus
-    \begin{align*}
-        \rank(\delta)=\<\pot'(p_1)+\pot'(p_2),n>
-                     \ge\<pot'(p_2),n>=\rank(\delta')\,.
-    \end{align*}
-  \end{case}
-  \begin{case}
-    $p_1=\ceu{\Break}$.
-    %%
-    Then~$\pi$ is an instance of~\R{loop-brk}.
-    Hence~$p'=p_1$ and~$e'=\nil$.  Thus
-    $\rank(\delta)=\rank(\delta')=\<0,n>$.
-    %%
-    % \[
-    %   \rank(\delta)=\rank(\delta')=\<0,n>\,.
-    % \]
-    %%
-  \end{case}
-  \begin{case}
-    $p_1\ne\ceu{\Nop},\ceu{\Break}$.
-    %%
-    Then~$\pi$ is an instance of~\R{loop-adv}.  Hence there is a
-    derivation~$\pi'$ such that
-    \begin{equation}
-      \label{lem.x.rank-nst-eq1}
-      \pi'\Vdash\<p_1,n,\nil>\nst\<p_1',n,e_1'>\,,
-    \end{equation}
-    for some~$p_1'$ and~$e_1'$.  Thus~$p'=\ceu{p_1'\AtLoop{\,p_2}}$
-    and~$e'=e_1'$.
-
-    There are two subcases.
-    \begin{subcase}
-      \label{lem.x.rank-nst-case1}
-      $\pot'(p)=\pot'(p_1)$.
-      %%
-      Then, by Definition~\ref{def.pot}, every execution path of~$p_1$
-      contains a (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
-      A single~$\nst$ cannot terminate the loop, since~$p_1\ne\ceu{\Break}$,
-      nor it can consume an~$\ceu{\AwaitExt}$, which means that all
-      execution paths in~$p_1'$ still contain a~$\ceu{\Break}$
-      or~$\ceu{\AwaitExt}$.  Hence~$\pot'(p')=\pot'(p_1')$.
-
-      There are two subcases.
-      \begin{subsubcase}
-        $e'=\nil$.
-        %%
-        Then $\rank(\delta)=\<pot'(p_1),n>$ and
-        $\rank(\delta')=\<pot'(p_1'),n>$.  From~\eqref{lem.x.rank-nst-eq1},
-        by the induction hypothesis, $\pot'(p_1)\ge\pot(p_1')$.  Thus
-        $\rank(\delta)\ge\rank(\delta')$.
-        %%
-        % \[
-        %   \rank(\delta)\ge\rank(\delta')\,.
-        % \]
-        %%
-      \end{subsubcase}
-      \begin{subsubcase}
-        $e'\ne\nil$.
-        %%
-        Then~$\pi'$ contains one application of~\R{emit-int} which consumes
-        one~$\ceu{\EmitInt(e')}$ expression from~$p_1$ and which implies
-        $\pot'(p_1)>\pot'(p_1')$.  Thus
-        \[
-          \hskip1em\rank(\delta)\!=\!\<\pot'(p_1),n>
-          >\<\pot'(p_1'),n+1>\!=\!\rank(\delta')\,.
-        \]
-        %%
-        % \begin{align*}
-        %   \rank(\delta)&=\<\pot'(p_1),n>\\
-        %                &>\<\pot'(p_1'),n+1>=\rank(\delta')\,.
-        % \end{align*}
-        %%
-      \end{subsubcase}
-    \end{subcase}
-    \begin{subsubcase}
-      $\pot'(p)=\pot'(p_1)+\pot'(p_2)$.
-      %%
-      Then some execution path in~$p_1$ does not contain a
-      (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
-      Since~$p_1\ne\ceu{\Nop}$, a single~$\nst$ cannot restart the loop,
-      which means that~$p_1'$ still contain some execution path in which
-      a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ does not occur.
-      Hence~$\pot'(p')=\pot'(p_1')+\pot'(p_2)$.  The rest of this proof is
-      similar to that of Case~\ref{lem.x.rank-nst-case1}.\qedhere
-    \end{subsubcase}
-  \end{case}
-\end{proof}
+\end{restatable}
 
 The next theorem is a generalization of Lemma~\ref{lem.x.rank-nst}
-for~$\nst[*]$ transitions.  Its proof is by induction on~$i$.
-
-\begin{theorem}
+for~$\nst[*]$.  Its proof follows from the lemma by induction on~$i$.
+
+\begin{restatable}{theorem}{thmxranknstx}
   \label{thm.x.rank-nst-*}
@@
   If~$\delta\nst[*]\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
-\end{theorem}
+\end{restatable}
 
 We now state and prove the main result of this section,
-Theorem~\ref{thm.x.term}, viz., the termination theorem for~$\trans[*]$.
-The idea of the proof is that a sufficiently large sequence of~$\nst$
-and~$\out$ transitions eventually decrease the rank of the current
-description until an irreducible description is reached.  This irreducible
-description is the final result of the reaction.
-%%
-\begin{theorem}[Termination]\label{thm.x.term}\strut\\
+Theorem~\ref{thm.x.term}, the termination theorem for~$\trans[*]$.  The idea
+of the proof is that a sufficiently large sequence of~$\nst$ and~$\out$
+transitions eventually decreases the rank of the current description until
+an irreducible description is reached.  This irreducible description is the
+final result of the reaction.
+%%
+\begin{restatable}[Termination]{theorem}{thmxterm}
+  \label{thm.x.term}
   %%
@@
   that~$\delta\trans[*]\delta'_\#$.
-\end{theorem}
+\end{restatable}
 \begin{proof}
@@
   $\rank(\delta)=\<i,j>$.
-  \begin{basis}
-    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$, as~$j=0$
-    implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be
-    applied).  There are two possibilities:  either~$\delta$ is
-    nested-irreducible or it is not.  In the first case, the theorem is
-    trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
-    nested-irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
-    $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
-    Theorem~\ref{thm.x.rank-nst-*},
-    $\<i,j>=\<0,0>\ge\rank(\delta')$,
-    %%
-    % \[
-    %   \<i,j>=\<0,0>\ge\rank(\delta')\,,
-    % \]
-    %%
-    which implies~$\rank(\delta')=\<0,0>$.
-  \end{basis}
-  \begin{induction}
-    Let~$\<i,j>\ne\<0,0>$.
-    %%
-    There are two subcases depending on whether or not $\delta$~is
-    nested-irreducible.
-    \begin{case}
-      \label{thm.x.term.Hnst}
-      $\delta$~is nested-irreducible.
-      %%
-      If~$j=0$ then, by Definition~\ref{def.x.H}, $\delta_\#$, and
-      thus~$\delta\trans[0]\delta_\#$.  Suppose that is not the case, i.e.,
-      that $j>0$.  Then there are two subcases.
-      \begin{subcase}
-        \label{thm.x.term.Hnst-j>0-nonnil}
-        $e\ne\nil$.
-        %%
-        Then, by~\R{out} and by Theorem~\ref{thm.x.term-nst-*}, there
+
+  For the basis, suppose $\<i,j>=\<0,0>$.  Then~$\delta$ cannot be advanced
+  by~$\out$, as~$j=0$ implies~$e=\nil$ and~$n=0$.  If~$\delta$ is
+  nested-irreducible, the theorem is trivially true,
+  as~$\delta\nst[0]\delta_\Hnst$ and~$\delta_\#$.  If~$\delta$ is not
+  nested-irreducible then, by Theorem~\ref{thm.x.term-nst-*},
+  $\delta\nst[*]\delta'_\Hnst$, for some~$\delta'_\Hnst$.  By
+  Theorem~\ref{thm.x.rank-nst-*}, $\rank(\delta)\ge\rank(\delta')$, which
+  implies~$\rank(\delta')=\<0,0>$, and so~$\delta'_\#$.
+
+  For the inductive step, suppose~$\<i,j>>\<0,0>$.
+  %%
+  Then, depending on whether or not $\delta$~is nested-irreducible, there
+  are two cases.
+  \begin{casex}
+  \item\label{thm.x.term.Hnst}$\delta$~is nested-irreducible.
+    %%
+    If~$j=0$, by Definition~\ref{def.x.H}, $\delta_\#$, and
+    so~$\delta\trans[0]\delta_\#$.  If~$j>0$, there are two subcases:
+    \begin{casex}
+    \item\label{thm.x.term.Hnst-j>0-nonnil}$e\ne\nil$.
+      %%
+      Then, by~\R{push} and by Theorem~\ref{thm.x.term-nst-*}, there
       are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
@@
       similar to that of Case~\ref{thm.x.term.Hnst-j>0-nil} below.
-        Otherwise, if~$e'\ne\nil$ then $i>i'$, since an~$\ceu{\EmitInt(e')}$
-        was consumed by the nested transitions.  Thus,
+      Otherwise, if~$e'\ne\nil$, then $i>i'$, as an~$\ceu{\EmitInt(e')}$ was
+      consumed by the nested transitions.  Thus,
       $\rank(\delta)>\rank(\delta')$.  By the induction hypothesis,
@@
       $\delta\trans[*]\delta''_\#$.
-      \end{subcase}
-      \begin{subcase}
-        \label{thm.x.term.Hnst-j>0-nil}
-        $e=\nil$.
-        %%
-        Then, since~$j>0$, $\delta\outpop\delta'$, for some~$\delta'$.  By
-        item~\eqref{lem.rank-out-pop} of Lemma~\ref{lem.rank-out},
-        $\rank(\delta)>\rank(\delta')$.  Hence, by the induction hypothesis,
-        there is a~$\delta''_\#$ such that~$\delta'\trans[*]\delta''_\#$.
-        Therefore, $\delta\trans[*]\delta''_\#$.
-      \end{subcase}
-    \end{case}
-    \begin{case}
-      $\delta$~is not nested-irreducible.
-      %%
-      Then~$e=\nil$ and, by Theorem~\ref{thm.x.term-nst-*}
+      %%
+    \item\label{thm.x.term.Hnst-j>0-nil}$e=\nil$.
+      %%
+      Then, as~$j>0$, $\delta\outpop\delta'$, for some~$\delta'$.  By
+      Lemma~\ref{lem.x.rank-out}, $\rank(\delta)>\rank(\delta')$.  Hence, by
+      the induction hypothesis, $\delta'\trans[*]\delta''_\#$, for
+      some~$\delta''_\#$.  And so, $\delta\trans[*]\delta''_\#$.
+    \end{casex}
+
+  \item$\delta$~is not nested-irreducible.
+    %%
+    Then~$e=\nil$ and, by Theorems~\ref{thm.x.term-nst-*}
     and~\ref{thm.x.rank-nst-*}, there is a~$\delta'_\Hnst$ such
     that~$\delta\nst[*]\delta'_\Hnst$
-      with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
-      is similar to that of Case~\ref{thm.x.term.Hnst} above.\qedhere
-    \end{case}
-  \end{induction}
+    with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof is
+    similar to that of Case~\ref{thm.x.term.Hnst} above.\qedhere
+  \end{casex}
 \end{proof}

### 2018-04-16 06-appendix.tex
Continue refactoring of Section 3
Finishes refactoring Sections 3 and 6

@@
     \end{equation}
-    From~\eqref{lem.x.props-nst-i.a.eq2} and~\eqref{lem.x.props-nst-i.a.eq3},
+    From~\eqref{lem.x.props-nst-i.a.eq2}
+    and~\eqref{lem.x.props-nst-i.a.eq3},
     \[
@@
     \end{equation}
-    From~\eqref{lem.x.props-nst-i.e.eq2} and~\eqref{lem.x.props-nst-i.e.eq3},
+    From~\eqref{lem.x.props-nst-i.e.eq2}
+    and~\eqref{lem.x.props-nst-i.e.eq3},
     \[
@@
 
-\begin{lemma}\strut
-  \label{lem.rank-out}
-  %%
-  \begin{enumerate:a}
-  \item\label{lem.rank-out-push} If~$\delta\outpush\delta'$
-    then~$\rank(\delta)=\rank(\delta')$.
-  \item\label{lem.rank-out-pop} If~$\delta\outpop\delta'$
-    then~$\rank(\delta)>\rank(\delta')$.
-  \end{enumerate:a}
-\end{lemma}
+\lemxrankout*
 \begin{proof}\strut
@@
     $e\ne\nil$, $p'=\bcast(p,e)$, $n'=n+1$, and~$e'=\nil$.  By
-    Definition~\ref{def.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
+    Definition~\ref{def.x.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
     as~$e'=\nil$ and~$n'=n+1$; hence~$j=j'$.
@@
       i&=\pot(p,e)
-         \tag*{by Definition~\ref{def.rank}}\\
+         \tag*{by Definition~\ref{def.x.rank}}\\
        &=\pot'(\bcast(p,e))
-         \tag*{by Definition~\ref{def.pot}}\\
+         \tag*{by Definition~\ref{def.x.pot}}\\
        &=\pot'(p')
@@
        &=\pot(p',e')
-         \tag*{by Definition~\ref{def.pot}}\\
+         \tag*{by Definition~\ref{def.x.pot}}\\
        &=i'
-         \tag*{by Definition~\ref{def.rank}}
+         \tag*{by Definition~\ref{def.x.rank}}
     \end{align*}
@@
     %%
-    By Definition~\ref{def.pot}, $\pot(\bcast(p,e))=\pot(\bcast(p',e'))$;
-    hence~$i=i'$.  And by Definition~\ref{def.rank}, $j=n$, as~$e=\nil$,
+    By Definition~\ref{def.x.pot}, $\pot(\bcast(p,e))=\pot(\bcast(p',e'))$;
+    hence~$i=i'$.  And by Definition~\ref{def.x.rank}, $j=n$, as~$e=\nil$,
     and~$j'=n-1$, as~$e'=\nil$ and~$n'=n-1$; hence~$j>j'$.
-    Therefore,~${\<i,j>}>{\<i',j'>}$.\qedhere
+    Therefore, ${\<i,j>}>{\<i',j'>}$.\qedhere
   \end{enumerate:a}
@@
 
-\begin{lemma}
-  \label{lem.rank-nst}
-  %%
-  If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
-\end{lemma}
+\lemxranknst*
 \begin{proof}
@@
   By definition of~$\nst$, $e=\nil$ and $n=n'$.  Depending on the structure
-  of program~$p$, there are~11 possibilities.  In each one of them, we show
+  of program~$p$, there are~11 possibilities.  In each one of them we show
   that~$\rank(\delta)\ge\rank(\delta')$.
-
-  \begin{case}
-    $p=\ceu{\Mem(id)}$.
+  \begin{casex}
+  \item$p=\ceu{\Mem(id)}$.
     %%
@@
     and~$e'=\nil$.  Thus $\rank(\delta)=\rank(\delta')=\<0,n>$.
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{\EmitInt(e_1)}$.
+
+  \item$p=\ceu{\EmitInt(e_1)}$.
     %%
@@
     \]
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{\CanRun(n)}$.
+
+  \item$p=\ceu{\CanRun(n)}$.
     %%
@@
     \]
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{\IfElse{p}{p_1}{p_2}}$.
-    %%
-    There are two subcases.
-    \begin{subcase}
-      \label{lem.rank-nst.if-true}
-      $\eval(\ceu{\Mem(\Id)})$~is true.
+
+  \item$p=\ceu{\IfElse{p}{p_1}{p_2}}$.
+    %%
+    There are two subcases, depending on the value of~$\ceu{\Mem(\Id)}$.
+    \begin{casex}
+    \item\label{lem.x.rank-nst.if-true}$\eval(\ceu{\Mem(\Id)})$.
       %%
@@
       \end{align*}
-    \end{subcase}
-    \begin{subcase}
-      $\eval(\ceu{\Mem(\Id)})$~is false.
-      %%
-      Similar to Case~\ref{lem.rank-nst.if-true}.
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{p_1;\,p_2}$.
-    %%
-    There are three subcases.
-    \begin{subcase}
-      \label{lem.rank-nst.seq-nop}
-      $p_1=\ceu{\Nop}$.
+      %%
+    \item$\lnot\eval(\ceu{\Mem(\Id)})$.
+      %%
+      Similar to Case~\ref{lem.x.rank-nst.if-true}.
+    \end{casex}
+
+  \item$p=\ceu{p_1;\,p_2}$.
+    %%
+    There are three subcases, depending on the structure of~$p_1$
+    \begin{casex}
+    \item\label{lem.x.rank-nst.seq-nop}$p_1=\ceu{\Nop}$.
       %%
@@
       \end{align*}
-    \end{subcase}
-    \begin{subcase}
-      \label{lem.rank-nst.seq-brk}
-      $p_1=\ceu{\Break}$.
+      %%
+    \item\label{lem.x.rank-nst.seq-brk}$p_1=\ceu{\Break}$.
       %%
@@
       \]
-    \end{subcase}
-    \begin{subcase}
-      \label{lem.rank-nst.seq-adv}
-      $p_1\ne\ceu{\Nop},\ceu{\Break}$.
+      %%
+    \item\label{lem.x.rank-nst.seq-adv}$p_1\ne\ceu{\Nop},\ceu{\Break}$.
       %%
@@
       \begin{equation}
-        \label{lem.rank-nst.seq-adv.eq1}
+        \label{lem.x.rank-nst.seq-adv.eq1}
         \rank(\<p_1,n,\nil>)\ge\rank(\<p_1',n,e_1'>)\,.
       \end{equation}
-      There are two subcases.
-      \begin{subsubcase}
-        $e'=\nil$.
+      There are two subcases, depending on whether or not~$e'$ is the empty
+      event.
+      \begin{casex}
+      \item$e'=\nil$
         %%
@@
         \end{align*}
-        By~\eqref{lem.rank-nst.seq-adv.eq1}, $\pot'(p_1)\ge\pot'(p_1')$.
+        By~\eqref{lem.x.rank-nst.seq-adv.eq1}, $\pot'(p_1)\ge\pot'(p_1')$.
         Thus
@@
         \]
-      \end{subsubcase}
-      \begin{subsubcase}
-        $e'\ne\nil$.
+        %%
+      \item$e'\ne\nil$.
         %%
         Then~$\pi'$ contains one application of~\R{emit-int}, which consumes
-        one~$\ceu{\EmitInt(e')}$ expression from~$p_1$ and implies
+        one~$\ceu{\EmitInt(e')}$ statement from~$p_1$ and implies
         $\pot'(p_1)>\pot'(p_1')$.  Thus
@@
         \end{align*}
-      \end{subsubcase}
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    \label{lem.rank-nst.loop-expd}
-    $p=\ceu{\Loop{p_1}}$.
+      \end{casex}
+    \end{casex}
+
+  \item\label{lem.x.rank-nst.loop-expd}$p=\ceu{\Loop{p_1}}$.
     %%
@@
     %%
-    By item~\eqref{ass.syn-rest.loop} of Assumption~\ref{ass.syn-rest}, all
-    execution paths of~$p_1$ contain at least one occurrence
+    By item~\ref{ass.x.syn-rest.loop} of Assumption~\ref{ass.x.syn-rest},
+    all execution paths of~$p_1$ contain at least one occurrence
     of~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$.  Thus, by condition~(\dag) in
-    Definition~\ref{def.pot},
+    Definition~\ref{def.x.pot},
     \[
@@
     \]
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{{p_1}\AtLoop{p_2}}$.
-    %%
-    There are three cases.
-    \begin{subcase}
-      $p_1=\ceu{\Nop}$.
-      %%
-      Similar to Case~\ref{lem.rank-nst.seq-nop}.
-    \end{subcase}
-    \begin{subcase}
-      $p_1=\ceu{\Break}$.
-      %%
-      Similar to Case~\ref{lem.rank-nst.seq-brk}.
-    \end{subcase}
-    \begin{subcase}
-      \label{lem.rank-nst.loop-adv}
-      $p_1\ne\ceu{\Nop},\ceu{\Break}$.
+
+  \item$p=\ceu{{p_1}\AtLoop{p_2}}$.
+    %%
+    There are three subcases, depending on the structure of~$p_1$.
+    \begin{casex}
+    \item$p_1=\ceu{\Nop}$.
+      %%
+      Similar to Case~\ref{lem.x.rank-nst.seq-nop}.
+      %%
+    \item$p_1=\ceu{\Break}$.
+      %%
+      Similar to Case~\ref{lem.x.rank-nst.seq-brk}.
+      %%
+    \item\label{lem.x.rank-nst.loop-adv}$p_1\ne\ceu{\Nop},\ceu{\Break}$.
       %%
@@
       % \begin{equation}
-      %   \label{lem.rank-nst.loop-adv-eq1}
+      %   \label{lem.x.rank-nst.loop-adv-eq1}
       %   \rank(\<p_1,n,\nil>)\ge\rank(\<p_1',n,e_1'>)\,.
@@
       %%
-      There are two subcases.
-      \begin{subsubcase}
-        $\pot'(p)=\pot'(p_1)$.
+      There are two subcases depending on the value of~$\pot'(p)$.
+      \begin{casex}
+      \item$\pot'(p)=\pot'(p_1)$.
         %%
@@
         Hence~$\pot'(p')=\pot'(p_1')$.  The rest of this proof is similar to
-        that of Case~\ref{lem.rank-nst.seq-adv}.
-      \end{subsubcase}
-      \begin{subsubcase}
-        $\pot'(p)=\pot'(p_1)+\pot'(p_2)$.
+        that of Case~\ref{lem.x.rank-nst.seq-adv}.
+        %%
+      \item$\pot'(p)=\pot'(p_1)+\pot'(p_2)$.
         %%
@@
         Hence~$\pot'(p')=\pot'(p_1')+\pot'(p_2)$.  The rest of this proof is
-        similar to that of Case~\ref{lem.rank-nst.seq-adv}.
-      \end{subsubcase}
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    \label{lem.rank-nst.and-expd}
-    $p=\ceu{{p_1}\And{p_2}}$.
+        similar to that of Case~\ref{lem.x.rank-nst.seq-adv}.
+      \end{casex}
+    \end{casex}
+
+  \item\label{lem.x.rank-nst.and-expd}$p=\ceu{{p_1}\And{p_2}}$.
     %%
@@
     \]
-  \end{case}
-
-  \begin{case}
-    \label{lem.rank-nst.and}
-    $p=\ceu{{p_1}\AtAnd{p_2}}$.
-    %%
-    There are two subcases.
-    \begin{subcase}
-      \label{lem.rank-nst.and1}
-      $\isblocked(p_1,n)$~is false.
-      %%
-      There are three subcases.
-      \begin{subsubcase}
-        \label{lem.rank-nst.and-nop1}
-        $p_1=\ceu{\Nop}$.
+
+  \item\label{lem.x.rank-nst.and}$p=\ceu{{p_1}\AtAnd{p_2}}$.
+    %%
+    There are two subcases, depending whether or not~$p_1$ is blocked at
+    stack level~$n$.
+    \begin{casex}
+    \item\label{lem.x.rank-nst.and1}$\lnot\isblocked(p_1,n)$.
+      %%
+      There are three subcases, depending on the structure of~$p_1$.
+      \begin{casex}
+      \item\label{lem.x.rank-nst.and-nop1}$p_1=\ceu{\Nop}$.
         %%
@@
         \]
-      \end{subsubcase}
-      \begin{subsubcase}
-        \label{lem.rank-nst.and-brk1}
+        %%
+      \item\label{lem.x.rank-nst.and-brk1}
         $p_1=\ceu{\Break}$.
@@
         Then~$\pi$ is an instance of~\R{and-brk1}.
-        Hence~$p'=\ceu{\clear(p_2);\,\Break}$ and~$e'=\nil$.  By item
-        \eqref{ass.syn-rest.fin} of Assumption~\ref{ass.syn-rest} and by the
-        definition of~$\clear$, $\clear(p_2)$ does not
+        Hence~$p'=\ceu{\clear(p_2);\,\Break}$ and~$e'=\nil$.  By
+        item~\ref{ass.x.syn-rest.fin} of Assumption~\ref{ass.x.syn-rest} and
+        by the definition of~$\clear$, $\clear(p_2)$ does not
         contain~$\ceu{\EmitInt}$ expressions.  Thus
@@
         \]
-      \end{subsubcase}
-      \begin{subsubcase}
-        \label{lem.rank-nst.and-adv1}
-        $p_1\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+      \item\label{lem.x.rank-nst.and-adv1}$p_1\ne\ceu{\Nop},\ceu{\Break}$.
         %%
@@
         precedence), the rest of this proof is similar to that of
-        Case~\ref{lem.rank-nst.seq-adv}.
-      \end{subsubcase}
-    \end{subcase}
-    \begin{subcase}
-      $\isblocked(p_1,n)$~is true.
-      %%
-      Similar to Case~\ref{lem.rank-nst.and1}
-    \end{subcase}
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{{p_1}\Or{p_2}}$.
+        Case~\ref{lem.x.rank-nst.seq-adv}.
+      \end{casex}
+      %%
+    \item$\isblocked(p_1,n)$.
+      %%
+      Similar to Case~\ref{lem.x.rank-nst.and1}
+    \end{casex}
+
+  \item$p=\ceu{{p_1}\Or{p_2}}$.
     %%
     Then~$\pi$ is an instance of~\R{or-expd}.
-    Hence~$p'=\ceu{{p_1}\AtOr(\CanRun(n);\,p_2)}$ and~$e'=\nil$.
-    Thus
+    Hence~$p'=\ceu{{p_1}\AtOr(\CanRun(n);\,p_2)}$ and~$e'=\nil$.  Thus
     \[
@@
     \]
-  \end{case}
-
-  \begin{case}
-    $p=\ceu{{p_1}\AtOr{p_2}}$.
-    %%
-    There are two subcases.
-    \begin{subcase}
-      \label{lem.rank-nst.or1}
-      $\isblocked(p_1,n)$ is false.
-      %%
-      There are three subcases.
-      \begin{subsubcase}
-        $p_1=\ceu{\Nop}$.
+
+  \item$p=\ceu{{p_1}\AtOr{p_2}}$.
+    %%
+    There are two subcases, depending whether or not~$p_1$ is block at stack
+    level~$n$.
+    \begin{casex}
+    \item\label{lem.x.rank-nst.or1}$\lnot\isblocked(p_1,n)$.
+      %%
+      There are three subcases, depending on the structure of~$p_1$.
+      \begin{casex}
+      \item$p_1=\ceu{\Nop}$.
         %%
         Then~$\pi$ is an instance of~\R{or-nop1}.  Hence~$p'=\clear(p_2)$
-        and~$e'=\nil$.  By item \eqref{ass.syn-rest.fin} of
-        Assumption~\ref{ass.syn-rest} and by the definition of~$\clear$,
+        and~$e'=\nil$.  By item~\ref{ass.x.syn-rest.fin} of
+        Assumption~\ref{ass.x.syn-rest} and by the definition of~$\clear$,
         $p'$ does not contain~$\ceu{\EmitInt}$ expressions.  Thus
@@
         \]
-      \end{subsubcase}
-      \begin{subsubcase}
-        $p_1=\ceu{\Break}$.
-        %%
-        Similar to Case~\ref{lem.rank-nst.and-brk1}.
-      \end{subsubcase}
-      \begin{subsubcase}
-        $p_1\ne\ceu{\Nop},\ceu{\Break}$.
-        %%
-        Similar to Case~\ref{lem.rank-nst.and-adv1}.
-      \end{subsubcase}
-    \end{subcase}
-    \begin{subcase}
-      $\isblocked(p_1,n)$ is true.
-      %%
-      Similar to Case~\ref{lem.rank-nst.or1}.\qedhere
-    \end{subcase}
-  \end{case}
+        %%
+      \item$p_1=\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{lem.x.rank-nst.and-brk1}.
+        %%
+      \item$p_1\ne\ceu{\Nop},\ceu{\Break}$.
+        %%
+        Similar to Case~\ref{lem.x.rank-nst.and-adv1}.
+      \end{casex}
+    \item$\isblocked(p_1,n)$.
+      %%
+      Similar to Case~\ref{lem.x.rank-nst.or1}.\qedhere
+    \end{casex}
+  \end{casex}
 \end{proof}
 
-\begin{theorem}
-  \label{thm.rank-nst-*}
-  %%
-  If~$\delta\nst[*]\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
-\end{theorem}
+\thmxranknstx*
 \begin{proof}
@@
   The theorem is trivially true for~$i=0$ and follows directly from
-  Lemma~\ref{lem.rank-nst} for~$i=1$.  Suppose
+  Lemma~\ref{lem.x.rank-nst} for~$i=1$.  Suppose
   $\delta\nst[1]\delta_1'\nst[i-1]\delta'$, for some~$i>1$ and~$\delta_1'$.
-  Thus, by Lemma~\ref{lem.rank-nst} and by the induction hypothesis,
+  Thus, by Lemma~\ref{lem.x.rank-nst} and by the induction hypothesis,
   \[
@@
 
-\begin{theorem}[Termination]\label{thm.term}\strut\\
-  %%
-  For any~$\delta$, there is a~$\delta'_\#$ such
-  that~$\delta\trans[*]\delta'_\#$.
-\end{theorem}
+\thmxterm*
 \begin{proof}
@@
     nested-irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
-    $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
-    Theorem~\ref{thm.rank-nst-*},
+    $\delta\nst[*]\delta'_\Hnst$, for some~$\delta'_\Hnst$.  By
+    Theorem~\ref{thm.x.rank-nst-*},
     \[
@@
     %%
-    There are two subcases.
-    \begin{case}
-      \label{thm.term.Hnst}
-      $\delta$~is nested-irreducible.
-      %%
-      There are two cases.
-      \begin{subcase}
-        \label{thm.term.Hnst-j0}
-        $j=0$.  By Definition~\ref{def.H}, $\delta_\#$.
+    There are two subcases, depending whether or not~$\delta$ is
+    nested-irreducible.
+    \begin{casex}
+    \item\label{thm.x.term.Hnst}$\delta$~is nested-irreducible.
+      %%
+      There are two cases, depending on the value of~$j$.
+      \begin{casex}
+      \item\label{thm.x.term.Hnst-j0}$j=0$.
+        %%
+        By Definition~\ref{def.x.H}, $\delta_\#$.
         Thus~$\delta\trans[0]\delta_\#$.
-      \end{subcase}
-      \begin{subcase}
-        \label{thm.term.Hnst-j>0}
-        $j>0$.
-        %%
-        There are two subcases.
-        \begin{subsubcase}
-          \label{thm.term.Hnst-j>0-nonnil}
-          $e\ne\nil$.
-          %%
-          Then, by~\R{out} and by Theorem~\ref{thm.x.term-nst-*}, there
+        %%
+      \item\label{thm.x.term.Hnst-j>0}$j>0$.
+        %%
+        There are two subcases, depending whether or not~$e$ is the empty
+        event.
+        \begin{casex}
+        \item\label{thm.x.term.Hnst-j>0-nonnil}$e\ne\nil$.
+          %%
+          Then, by~\R{push} and by Theorem~\ref{thm.x.term-nst-*}, there
           are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
@@
           \]
-          Thus, by item~\eqref{lem.rank-out-push} of
-          Lemma~\ref{lem.rank-out} and by Theorem~\ref{thm.rank-nst-*},
+          Thus, by item~\ref{lem.x.rank-out-push} of
+          Lemma~\ref{lem.x.rank-out} and by Theorem~\ref{thm.x.rank-nst-*},
           \begin{align*}
@@
           If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof
-          is similar to that of Case~\ref{thm.term.Hnst-j>0-nil}.
+          is similar to that of Case~\ref{thm.x.term.Hnst-j>0-nil}.
           Otherwise, if~$e'\ne\nil$ then $i>i'$, since
@@
           some~$\delta''_\#$.  Therefore, $\delta\trans[*]\delta''_\#$.
-        \end{subsubcase}
-        \begin{subsubcase}
-          \label{thm.term.Hnst-j>0-nil}
-          $e=\nil$.
-          %%
-          Then, since~$j>0$, $\delta\outpop\delta'$, for
-          some~$\delta''$.  By item~\eqref{lem.rank-out-pop} of
-          Lemma~\ref{lem.rank-out},
+          %%
+        \item\label{thm.x.term.Hnst-j>0-nil}$e=\nil$.
+          %%
+          Then, since~$j>0$, $\delta\outpop\delta'$, for some~$\delta''$.
+          By item~\ref{lem.x.rank-out-pop} of Lemma~\ref{lem.x.rank-out},
           \[
@@
           Therefore, $\delta\trans[*]\delta''_\#$.
-        \end{subsubcase}
-      \end{subcase}
-    \end{case}
-    \begin{case}
-      $\delta$~is not nested-irreducible.
+        \end{casex}
+      \end{casex}
+      %%
+    \item$\delta$~is not nested-irreducible.
       %%
       Then~$e=\nil$ and, by Theorems~\ref{thm.x.term-nst-*}
-      and~\ref{thm.rank-nst-*}, there is a~$\delta'_\Hnst$ such
+      and~\ref{thm.x.rank-nst-*}, there is a~$\delta'_\Hnst$ such
       that~$\delta\nst[*]\delta'_\Hnst$
       with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
-      is similar to that of Case~\ref{thm.term.Hnst}.\qedhere
-    \end{case}
+      is similar to that of Case~\ref{thm.x.term.Hnst}.\qedhere
+    \end{casex}
   \end{induction}

### 2018-04-16 06-appendix.tex
Fix inconsistencies in Section 6

@@
   two possibilities.
-  \begin{casex}
+  \begin{case}
   \item $e\ne\nil$.  Both transitions are applications of~\R{push}.
@@
     Hence $p_1=p_2=p$, $n_1=n_2=n-1$, and~$e_1=e_2=\nil$.\qedhere
-  \end{casex}
+  \end{case}
 \end{proof}
@@
   $\ceu{\AwaitInt}$, $\ceu{\Break}$, $\ceu{\Every}$, $\ceu{\Fin}$,
-  or~$\ceu{\Nop}$ statement as there are no~$\nst$ rules to transition such
+  or~$\ceu{\Nop}$ statement as there is no~$\nst$ rule to transition such
   programs.)
 
-  \begin{casex}
+  \begin{case}
   \item $p=\ceu{\Mem(\Id)}$.
@@
     There are two subcases.
-    \begin{casex}
+    \begin{case}
     \item$\eval(\ceu{\Mem(\Id)})$.
@@
       Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
-    \end{casex}
+    \end{case}
     %%
@@
     There are three subcases.
-    \begin{casex}
+    \begin{case}
     \item $p'=\ceu{\Nop}$.
@@
       and~$e_1=e_1'=e_2'=e_2$.
-    \end{casex}
+    \end{case}
     %%
@@
       and~$e_1=e_1'=e_2'=e_2$.
-    \end{casex}
+    \end{case}
   \item $p=\ceu{p'\And{p''}}$.
@@
     There are two subcases.
-    \begin{casex}
+    \begin{case}
     \item$\lnot\isblocked(p',n)$.
@@
       There are three subcases.
-      \begin{casex}
+      \begin{case}
       \item $p'=\ceu{\Nop}$.
@@
         and~$e_1=e_1'=e_2'=e_2$.
-      \end{casex}
+      \end{case}
       %%
@@
       There are three subcases.
-      \begin{casex}
+      \begin{case}
       \item $p''=\ceu{\Nop}$.
@@
         and~$e_1=e_1''=e_2''=e_2$.
-      \end{casex}
-    \end{casex}
+      \end{case}
+    \end{case}
     %%
@@
         Similar to Case~\ref{lem.x.det-nst.and-adv1}.
-      \end{casex}
+      \end{case}
       %%
@@
         %%
-      \end{casex}
-    \end{casex}
-  \end{casex}
+      \end{case}
+    \end{case}
+  \end{case}
 \end{proof}
@@
   There are two possibilities.
-  \begin{casex}
+  \begin{case}
   \item $\delta\out[1]\delta_1'$ and~$\delta\out[1]\delta_2'$.  Then, by
@@
     hypothesis, $\delta_1=\delta_2$.\qedhere
-  \end{casex}
+  \end{case}
 \end{proof}
@@
   There are two cases.
-  \begin{casex}
+  \begin{case}
   \item\label{prop.x.irr-nst-i-case1}
@@
     false.\qedhere
-  \end{casex}
+  \end{case}
 \end{proof}
@@
   indeed exists.
-  \begin{casex}
+  \begin{case}
   \item $p=\ceu{\Mem(\Id)}$.
@@
     There are two subcases.
-    \begin{casex}
+    \begin{case}
     \item\label{thm.x.term-nst-*.if-true} $\eval(\ceu{\Mem(\Id)})$.
@@
       %%
-    \end{casex}
+    \end{case}
 
@@
     There are three subcases.
-    \begin{casex}
+    \begin{case}
     \item\label{thm.x.term-nst-*.seq-nop} $p'=\ceu{\Nop}$.
@@
       %%
-      \begin{casex}
+      \begin{case}
       \item $e\ne\nil$.  Then, by the definition of~$\Hnst$,
@@
         nested-irreducible.
-      \end{casex}
-    \end{casex}
+      \end{case}
+    \end{case}
 
@@
     or~$\isblocked(p_1',n)$.
-    \begin{casex}
+    \begin{case}
     \item$p_1'=\ceu{\Break\AtLoop{p'}}$.
@@
       of~$\Hnst$, $\<p_1',n,e>_\Hnst$.
-    \end{casex}
+    \end{case}
 
@@
     There are three subcases.
-    \begin{casex}
+    \begin{case}
     \item$p'=\ceu{\Nop}$.
@@
       Case~\ref{thm.x.term-nst-*.seq-adv}.
-    \end{casex}
+    \end{case}
 
@@
       There are three subcases.
-      \begin{casex}
+      \begin{case}
       \item\label{thm.x.term-nst-*.and-nop1}
@@
         \end{enumerate}
-      \end{casex}
+      \end{case}
       %%
@@
       There are three subcases.
-      \begin{casex}
+      \begin{case}
       \item\label{thm.x.term-nst-*.and-nop2}
@@
         \end{enumerate}
-      \end{casex}
-    \end{casex}
+      \end{case}
+    \end{case}
 
@@
       There are three subcases.
-      \begin{casex}
+      \begin{case}
       \item\label{thm.x.term-nst-*.or-nop1} $p'=\ceu{\Nop}$.  Then,
@@
         Similar to Case~\ref{thm.x.term-nst-*.and-adv1}.
-      \end{casex}
+      \end{case}
       %%
@@
       There are three subcases.
-      \begin{casex}
+      \begin{case}
       \item$p''=\ceu{\Nop}$.
@@
         Similar to Case~\ref{thm.x.term-nst-*.and-adv2}.\qedhere
-      \end{casex}
-    \end{casex}
-  \end{casex}
+      \end{case}
+    \end{case}
+  \end{case}
 \end{proof}
@@
   that~$\rank(\delta)\ge\rank(\delta')$.
-  \begin{casex}
+  \begin{case}
   \item$p=\ceu{\Mem(id)}$.
@@
     %%
-    There are two subcases, depending on the value of~$\ceu{\Mem(\Id)}$.
-    \begin{casex}
+    There are two subcases.
+    \begin{case}
     \item\label{lem.x.rank-nst.if-true}$\eval(\ceu{\Mem(\Id)})$.
@@
       Similar to Case~\ref{lem.x.rank-nst.if-true}.
-    \end{casex}
+    \end{case}
 
@@
     %%
-    There are three subcases, depending on the structure of~$p_1$
-    \begin{casex}
+    There are three subcases.
+    \begin{case}
     \item\label{lem.x.rank-nst.seq-nop}$p_1=\ceu{\Nop}$.
@@
       \end{equation}
-      There are two subcases, depending on whether or not~$e'$ is the empty
-      event.
-      \begin{casex}
+      There are two subcases.
+      \begin{case}
       \item$e'=\nil$
@@
         \end{align*}
-      \end{casex}
-    \end{casex}
+      \end{case}
+    \end{case}
 
@@
     %%
-    There are three subcases, depending on the structure of~$p_1$.
-    \begin{casex}
+    There are three subcases.
+    \begin{case}
     \item$p_1=\ceu{\Nop}$.
@@
       %%
-      There are two subcases depending on the value of~$\pot'(p)$.
-      \begin{casex}
+      There are two subcases.
+      \begin{case}
       \item$\pot'(p)=\pot'(p_1)$.
@@
         similar to that of Case~\ref{lem.x.rank-nst.seq-adv}.
-      \end{casex}
-    \end{casex}
+      \end{case}
+    \end{case}
 
@@
     %%
-    There are two subcases, depending whether or not~$p_1$ is blocked at
+    There are two subcases.
     stack level~$n$.
-    \begin{casex}
+    \begin{case}
     \item\label{lem.x.rank-nst.and1}$\lnot\isblocked(p_1,n)$.
       %%
-      There are three subcases, depending on the structure of~$p_1$.
-      \begin{casex}
+      There are three subcases.
+      \begin{case}
       \item\label{lem.x.rank-nst.and-nop1}$p_1=\ceu{\Nop}$.
@@
         Case~\ref{lem.x.rank-nst.seq-adv}.
-      \end{casex}
+      \end{case}
       %%
@@
       Similar to Case~\ref{lem.x.rank-nst.and1}
-    \end{casex}
+    \end{case}
 
@@
     %%
-    There are two subcases, depending whether or not~$p_1$ is block at stack
-    level~$n$.
-    \begin{casex}
+    There are two subcases.
+    \begin{case}
     \item\label{lem.x.rank-nst.or1}$\lnot\isblocked(p_1,n)$.
       %%
-      There are three subcases, depending on the structure of~$p_1$.
-      \begin{casex}
+      There are three subcases.
+      \begin{case}
       \item$p_1=\ceu{\Nop}$.
@@
         Similar to Case~\ref{lem.x.rank-nst.and-adv1}.
-      \end{casex}
+      \end{case}
     \item$\isblocked(p_1,n)$.
@@
       Similar to Case~\ref{lem.x.rank-nst.or1}.\qedhere
-    \end{casex}
-  \end{casex}
+    \end{case}
+  \end{case}
 \end{proof}
@@
   $\rank(\delta)=\<i,j>$.
-  %%
-  \begin{basis}
+
+  \vskip\baselineskip
+  \noindent\emph{Basis}.
   If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$, as~$j=0$
-    implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be
-    applied).  There are two possibilities:  either~$\delta$ is
-    nested-irreducible or it is not.  In the first case, the theorem is
-    trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
+  implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be applied).
+  There are two possibilities:  either~$\delta$ is nested-irreducible or it
+  is not.  In the first case, the theorem is trivially true,
+  as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
   nested-irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
@@
   which implies~$\rank(\delta')=\<0,0>$.
-  \end{basis}
-  \begin{induction}
+
+  \vskip\baselineskip
+  \noindent\emph{Induction}.
   Let~$\<i,j>\ne\<0,0>$.
   %%
-    There are two subcases, depending whether or not~$\delta$ is
-    nested-irreducible.
-    \begin{casex}
+  There are two subcases.
+  \begin{case}
   \item\label{thm.x.term.Hnst}$\delta$~is nested-irreducible.
     %%
-      There are two cases, depending on the value of~$j$.
-      \begin{casex}
+    There are two subcases.
+    \begin{case}
     \item\label{thm.x.term.Hnst-j0}$j=0$.
@@
       %%
-        There are two subcases, depending whether or not~$e$ is the empty
+      There are two subcases.
       event.
-        \begin{casex}
+      \begin{case}
       \item\label{thm.x.term.Hnst-j>0-nonnil}$e\ne\nil$.
@@
         Therefore, $\delta\trans[*]\delta''_\#$.
-        \end{casex}
-      \end{casex}
+      \end{case}
+    \end{case}
     %%
@@
     is similar to that of Case~\ref{thm.x.term.Hnst}.\qedhere
-    \end{casex}
-  \end{induction}
+  \end{case}
 \end{proof}

### 2018-04-17 03-semantics.tex
Fix inconsistencies in Section 6
Make terminology uniform

@@
   to~$\ceu{\protect\Nop}$, and by unwinding $\ceu{\protect\Every}$
-  expressions.
+  statements.
   %%
@@
   %%
-  \space(iii)~Function~$\clear$ extracts~$\ceu{\protect\Fin}$ expressions in
+  \space(iii)~Function~$\clear$ extracts~$\ceu{\protect\Fin}$ statements in
   parallel and put their bodies in sequence.
@@
   where~$\pot'$ is an auxiliary function that counts the maximum number of
-  reachable~$\ceu{\EmitInt}$ expressions in the program resulting from the
+  reachable~$\ceu{\EmitInt}$ statements in the program resulting from the
   broadcast of event~$e$ to~$p$.
@@
   are two cases.
-  \begin{casex}
+  \begin{case}
   \item\label{thm.x.term.Hnst}$\delta$~is nested-irreducible.
@@
     so~$\delta\trans[0]\delta_\#$.  If~$j>0$, there are two subcases:
-    \begin{casex}
+    \begin{case}
     \item\label{thm.x.term.Hnst-j>0-nonnil}$e\ne\nil$.
@@
       some~$\delta''_\#$.  And so, $\delta\trans[*]\delta''_\#$.
-    \end{casex}
+    \end{case}
 
@@
     similar to that of Case~\ref{thm.x.term.Hnst} above.\qedhere
-  \end{casex}
+  \end{case}
 \end{proof}

### 2018-04-17 06-appendix.tex
Make terminology uniform

@@
         Then every execution path of~$p_1$ contains a~$\ceu{\Break}$
-        or~$\ceu{\AwaitExt}$ expression.  A single~$\nst$ cannot terminate
+        or~$\ceu{\AwaitExt}$ statement.  A single~$\nst$ cannot terminate
         the loop, since~$p_1\ne\ceu{\Break}$, nor can it consume
@@
         Then some execution path in~$p_1$ does not contain a~$\ceu{\Break}$
-        or~$\ceu{\AwaitExt}$ expression.  Since~$p_1\ne\ceu{\Nop}$, a
+        or~$\ceu{\AwaitExt}$ statement.  Since~$p_1\ne\ceu{\Nop}$, a
         single~$\nst$ cannot restart the loop, which means that~$p_1'$ still
@@
         by the definition of~$\clear$, $\clear(p_2)$ does not
-        contain~$\ceu{\EmitInt}$ expressions.  Thus
+        contain~$\ceu{\EmitInt}$ statements.  Thus
         \[
@@
         Assumption~\ref{ass.x.syn-rest} and by the definition of~$\clear$,
-        $p'$ does not contain~$\ceu{\EmitInt}$ expressions.  Thus
+        $p'$ does not contain~$\ceu{\EmitInt}$ statements.  Thus
         \[

### 2018-05-01 02-ceu.tex
Cleanup and compress stuff in Section 3 up to Determinism
Finishes refactoring Sections 3 and 6
Versão final. Apertando texto pra caber em 10 páginas...

@@
 %
-The finalization mechanism extends orthogonal abortion to also work with
+The finalization mechanism extends orthogonal abortion to %also work with
 activities that use stateful resources from the environment (such as files and
@@
 \footnote {
-The actual implementation of Céu supports a command-line option that accepts a
+The actual implementation of \CEU supports a command-line option that accepts a
 white list of library calls.
@@
 %
-\begin{enumerate}[i.]
+\begin{enumerate:i}
 \item The program initiates a ``boot reaction'' in a single trail (parallel
@@
       step~(ii).
-\end{enumerate}
+\end{enumerate:i}
 
@@
 occur simultaneously by definition.
-Hence, for the sequence of events \code{A->B}, \code{x} becomes \code{4}
-(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes \code{3}
-(\code{(1*2)+1}).
+Hence, for the sequence of events \code{A->B}, \code{x} becomes \code{4}% TODO-FULL
+%(\code{(1+1)*2})
+, while for \code{B->A}, \code{x} becomes \code{3}% TODO-FULL
+%(\code{(1*2)+1})
+.
 
@@
 
-However, aborting lines of execution that deal with external resources may lead
-to inconsistencies.
+However, aborting lines of execution that deal with resources may lead to
+inconsistencies.
 %
-For this reason, \CEU provides a \code{finalize} construct to unconditionally
+Therefore, \CEU provides a \code{finalize} construct to unconditionally
 execute a series of statements even if the enclosing block is externally
@@
 %
-In both cases, the code does not compile without the \code{finalize}
-construct.%
+In both cases, the code does not compile without the \code{finalize}.%
 \footnote{%

### 2018-05-01 04-related.tex
Versão final. Apertando texto pra caber em 10 páginas...

@@
 ~\cite{rp.rc,wsn.protothreads,wsn.sol,rp.synchc,rp.pretc}
-also rely on lexical scheduling to preserve determinism.
+rely on lexical scheduling to preserve determinism.
 %
@@
 is non-deterministic: \emph{``if there is no control dependency, as in
-\code{(call~f1()~||~call~f2())},
+{\scriptsize\code{(call~f1()||call~f2())}},
 the order is unspecified and it would be an error to rely on
@@
 %
-As far as we know, this extra safety level is unique to \CEU.
+As far as we know, this extra safety level is unique to \CEU among synchronous
+languages.

### 2018-05-01 index.tex
Add artifact evaluation
Revisões do artigo
Ajustes nos autores.
Finishes refactoring Sections 3 and 6
Versão final. Apertando texto pra caber em 10 páginas...

@@
-\documentclass[sigplan,anonymous,review,protrusion=true,expansion]{acmart}
+\documentclass[sigplan,protrusion=true,expansion,screen]{acmart}
+
+%\setlength{\belowcaptionskip}{-10pt}
 
@@
 %\setcopyright{acmlicensed}
-\setcopyright{rightsretained}
+%\setcopyright{rightsretained}
 %\setcopyright{usgov}
@@
 % DOI
-\acmDOI{10.475/123_4}
+%\acmDOI{10.475/123_4}
 
 % ISBN
-\acmISBN{123-4567-24-567/08/06}
+%\acmISBN{123-4567-24-567/08/06}
 
 %Conference
-\acmConference[LCTES'18]{ACM SIGPLAN/SIGBED}{June 2018}{Philadelphia, USA}
+%\acmConference[LCTES'18]{ACM SIGPLAN/SIGBED}{June 2018}{Philadelphia, USA}
 %\acmYear{1997}
@@
 %\acmBadgeR[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
+
+%%% If you see 'ACMUNKNOWN' in the 'setcopyright' statement below,
+%%% please first submit your publishing-rights agreement with ACM (follow link on submission page).
+%%% Then please update our instructions page and copy-and-paste the NEW commands into your article.
+%%% Please contact us in case of questions; allow up to 10 min for the system to propagate the information.
+%%%
+%%% The following is specific to LCTES'18 and the paper
+%%% 'A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language Céu'
+%%% by Guilherme F. Lima, Rodrigo C. M. Santos, Edward Hermann Haeusler, Roberto Ierusalimschy, and Francisco Sant'Anna.
+%%%
+%\setcopyright{ACMUNKNOWN}
+\setcopyright{acmcopyright}
+\acmPrice{15.00}
+\acmDOI{10.1145/3211332.3211334}
+\acmYear{2018}
+\copyrightyear{2018}
+\acmISBN{978-1-4503-5803-3/18/06}
+\acmConference[LCTES'18]{19th ACM SIGPLAN/SIGBED Conference on Languages, Compilers, and Tools for Embedded Systems}{June 19--20, 2018}{Philadelphia, PA, USA}
 
@@
 \email{rsantos@inf.puc-rio.br}
-\affiliation{PUC-Rio, Rio de Janeiro, Brazil}
+\affiliation{PUC-Rio, Brazil}
 %
@@
 \email{glima@inf.puc-rio.br}
-\affiliation{PUC-Rio, Rio de Janeiro, Brazil}
+\affiliation{PUC-Rio, Brazil}
 %
@@
 \email{francisco@ime.uerj.br}
-\affiliation{UERJ, Rio de Janeiro, Brazil}
+\affiliation{UERJ, Brazil}
 %
@@
 \email{roberto@inf.puc-rio.br}
-\affiliation{PUC-Rio, Rio de Janeiro, Brazil}
+\affiliation{PUC-Rio, Brazil}
 %
-\author[E.\,H.~Haeusler]{Edward Hermann Haeusler}
+\author[E.\,H.~Haeusler]{Edward H. Haeusler}
 \email{hermann@inf.puc-rio.br}
-\affiliation{PUC-Rio, Rio de Janeiro, Brazil}
+\affiliation{PUC-Rio, Brazil}
 
@@
 
+\balance
 \bibliographystyle{ACM-Reference-Format}
@@
 
+\appendix
+\clearpage
+\nobalance
 \input{06-appendix}
+\nobalance
+\clearpage
+\nobalance
+\input{07-artifact}
+\nobalance
 

### 2018-05-03 07-artifact.tex
Finishes refactoring Sections 3 and 6
Improve Artifact Evaluation based on the reviews.

@@
-\section{Artifact appendix}
+\section{Artifact Appendix}
 
@@
 The compiler/distribution of \CEU is available on GitHub:
+
 \url{https://github.com/fsantanna/ceu}
+
+The website of \CEU includes an introductory video and an online tutorial:
+
+\url{http://www.ceu-lang.org/}
+
+The language also provides extensive documentation online:
+
+\url{http://fsantanna.github.io/ceu/out/manual/v0.30/}
 
@@
 stats = {
-    count  = 3483,
-    trails = 9094,
-    bytes  = 52555128,
+    count = 3478,
+    trails = 9082,
+    bytes = 50803896,
     bcasts = 0,
-    visits = 4065937,
+    visits = 5189770,
 }
@@
 \subsection{Experiment customization}
+
+\subsubsection{Test Cases}
 
@@
 
+\subsubsection{Sample Programs}
+
+The distribution of \CEU comes with sample programs that can be executed as
+follows:
+
+\begin{verbatim}
+$ cd /<...>/ceu/               # change to the Céu repository
+$ make samples                  # execute all samples one by one
+\end{verbatim}
+
+\subsubsection{Single File}
+
+It is also possible to compile a single file with a program in \CEU:
+
+\begin{verbatim}
+$ cd /<...>/ceu/                         # change to the Céu repository
+$ make one CEU_SRC=/tmp/tst.ceu # compile a program
+$ /tmp/tst                      # execute it
+\end{verbatim}
+
+As an example, the file \code{/tmp/tst.ceu} may contain a program such as
+follows:
+
+\begin{verbatim}
+input int A; // two input events
+input int B;
+
+spawn do     // generates test inputs asynchronously
+    await async do
+        emit A(10);
+        emit B(100);
+    end
+end
+
+var int a = await A; // main program
+var int b = await B;
+
+escape a+b;  // 110 is the result of the program back to the OS
+\end{verbatim}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-05-03 04-related.tex
Adicionado Acknowledgments (grant da viagem).

@@
 events.
-In the former, \emph{``reactions to external and internal events (...) can be
-sensed only after completion of the step''}.
-In the latter, \emph{``the receiving state (of the internal event) acts here as
-a function''}.
+% TODO-FULL
+%In the former, \emph{``reactions to external and internal events (...) can be
+%sensed only after completion of the step''}.
+%In the latter, \emph{``the receiving state (of the internal event) acts here as
+%a function''}.
 Although the descriptions suggest a stack-based semantics, we are not aware of
-formalizations for these ideas for a deeper comparison with \CEU.
+formalizations or more precision for a deeper comparison with \CEU.
 
@@
 
+% TODO-FULL
 Regarding the integration with $C$ language-based environments, \CEU

### 2018-05-28 02-ceu.tex
(*) final version

@@
 
+In \CEU, any identifier prefixed with an underscore (e.g., \code{\_led}) is
+passed unchanged to the underlying~C compiler.
+%
+Therefore, access to~C is straightforward and syntactically traceable.
+%
+To ensure that programs operate under the synchronous hypothesis, the compiler
+environment should only provide access to~C operations that can be assumed to
+be instantaneous, such as non-blocking~I/O and simple data structure accessors.%
+\footnote {
+The actual implementation of \CEU supports a command-line option that accepts a
+white list of library calls.
+If a program tries to use a function not in the list, the compiler raises an
+error.
+}
+
 \begin{lstlisting}[
+  xleftmargin=1em,
   numbers=left,
@@
   float=h,
-  caption={A \CEU program that toggles a LED on radio receive,
-           terminating on a button press always with the LED~off.},
+  caption={A \CEU program that toggles a LED on receive,
+           terminating on a button always with the LED~off.},
   label={lst.ceu},
@@
 
-In \CEU, any identifier prefixed with an underscore (e.g., \code{\_led}) is
-passed unchanged to the underlying~C compiler.
-%
-Therefore, access to~C is straightforward and syntactically traceable.
-%
-To ensure that programs operate under the synchronous hypothesis, the compiler
-environment should only provide access to~C operations that can be assumed to
-be instantaneous, such as non-blocking~I/O and simple data structure accessors.%
-\footnote {
-The actual implementation of \CEU supports a command-line option that accepts a
-white list of library calls.
-If a program tries to use a function not in the list, the compiler raises an
-error.
-}
-
+\vspace{-0.5cm}
 \subsection{External and Internal Events}
@@
 
-\subsubsection*{External events and discrete time}
+\subsubsection{External Events and Discrete Time}
 
@@
 \begingroup
-\begin{figure}[h]
+\begin{figure}[b!]
 \centering
@@
 
-\subsubsection*{Internal events as subroutines}
+\subsubsection{Internal Events as Subroutines}
 
@@
 as illustrated in Listing~\ref{lst.sub}.
-
-\begin{lstlisting}[
-  numbers=left,
-  basicstyle=\ttfamily\small,
-  float=h,
-  caption={A ``subroutine'' that increments its argument.},
-  label={lst.sub},
-]
-event int* inc;       |\C{declares subroutine "inc"}|
-par/or do
-    var int* p;
-    every p in inc do |\C{implements "inc" through event iterator}|
-        *p = *p + 1;
-    end
-with
-    var int v = 1;
-    emit inc(&v);     |\C{calls "inc"}|
-    emit inc(&v);     |\C{calls "inc"}|
-    _assert(v==3);    |\C{asserts result after the two returns}|
-end
-\end{lstlisting}
-%\vskip-\baselineskip
 
@@
 invocation (ln.~10), and passes the assertion test (ln.~11).
+
+\begin{lstlisting}[
+  xleftmargin=1em,
+  numbers=left,
+  basicstyle=\ttfamily\small,
+  float=h,
+  caption={A ``subroutine'' that increments its argument.},
+  label={lst.sub},
+]
+event int* inc;       |\C{declares subroutine "inc"}|
+par/or do
+    var int* p;
+    every p in inc do |\C{implements "inc" as event iterator}|
+        *p = *p + 1;
+    end
+with
+    var int v = 1;
+    emit inc(&v);     |\C{calls "inc"}|
+    emit inc(&v);     |\C{calls "inc"}|
+    _assert(v==3);    |\C{asserts result after the two returns}|
+end
+\end{lstlisting}
+%\vskip-\baselineskip
 
@@
 \begingroup
-\begin{figure}[t]
+\begin{figure}[b!]
 \begin{minipage}[t]{0.45\linewidth}
@@
 
-\begin{figure}[t]
+%\vspace{2cm}
+\begin{figure}[b!]
 \begin{minipage}[t]{0.45\linewidth}

### 2018-05-28 06-appendix.tex
(*) final version

@@
   \item $e\ne\nil$.  Both transitions are applications of~\R{push}.
-    Hence~$p_1=p_2=\bcast(p,e)$, $n_1=n_2=n+1$, and~$e_1=e_2=\nil$.
+    Hence $p_1=p_2=\bcast(p,e)$, $n_1=n_2=n+1$, and~$e_1=e_2=\nil$.
     %%
@@
     i.e., their conclusions are obtained by an application of this rule.
-    Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
+    Hence $p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
     %%
@@
     Then~$\pi_1$ and~$\pi_2$ are instances of~\R{emit-int}.
-    Hence~$p_1=p_2=\ceu{\CanRun(n)}$ and~$e_1=e_2=e'$.
+    Hence $p_1=p_2=\ceu{\CanRun(n)}$ and~$e_1=e_2=e'$.
     %%
@@
     Then~$\pi_1$ and~$\pi_2$ are instances of~\R{can-run}.
-    Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
+    Hence $p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
     %%
@@
       Then~$\pi_1$ and~$\pi_2$ are instances of~\R{if-true}.
-      Hence~$p_1=p_2=p'$ and~$e_1=e_2=\nil$.
+      Hence $p_1=p_2=p'$ and~$e_1=e_2=\nil$.
       %%
@@
       Then~$\pi_1$ and~$\pi_2$ are instances of~\R{if-false}.
-      Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
+      Hence $p_1=p_2=p''$ and~$e_1=e_2=\nil$.
     \end{case}
@@
       Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-nop}.
-      Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
+      Hence $p_1=p_2=p''$ and~$e_1=e_2=\nil$.
       %%
@@
       Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-brk}.
-      Hence~$p_1=p_2=\ceu{\Break}$ and~$e_1=e_2=\nil$.
+      Hence $p_1=p_2=\ceu{\Break}$ and~$e_1=e_2=\nil$.
       %%
@@
       hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
-      Hence~$p_1=\ceu{p_1';p''}=\ceu{p_2';p''}=p_2$
+      Hence $p_1=\ceu{p_1';p''}=\ceu{p_2';p''}=p_2$
       and~$e_1=e_1'=e_2'=e_2$.
@@
     Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-expd}.
-    Hence~$p_1=p_2=\ceu{p'\AtLoop{p'}}$ and~$e_1=e_2=\nil$.
+    Hence $p_1=p_2=\ceu{p'\AtLoop{p'}}$ and~$e_1=e_2=\nil$.
     %%
@@
       Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-nop}.
-      Hence~$p_1=p_2=\ceu{\Loop{p''}}$ and~$e_1=e_2=\nil$.
+      Hence $p_1=p_2=\ceu{\Loop{p''}}$ and~$e_1=e_2=\nil$.
       %%
@@
       Then~$\pi_1$ and~$\pi_2$ are instances of~\R{loop-brk}.
-      Hence~$p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
+      Hence $p_1=p_2=\ceu{\Nop}$ and~$e_1=e_2=\nil$.
       %%
@@
       hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
-      Hence~$p_1=\ceu{p_1'\AtLoop{p''}}=\ceu{p_2'\AtLoop{p''}}=p_2$
+      Hence $p_1=\ceu{p_1'\AtLoop{p''}}=\ceu{p_2'\AtLoop{p''}}=p_2$
       and~$e_1=e_1'=e_2'=e_2$.
@@
     Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-expd}.
-    Hence~$p_1=p_2=\ceu{{p'}\AtAnd{(\CanRun(n);\,p'')}}$
+    Hence $p_1=p_2=\ceu{{p'}\AtAnd{(\CanRun(n);\,p'')}}$
     and~$e_1=e_2=\nil$.
@@
         Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop1}.
-        Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
+        Hence $p_1=p_2=p''$ and~$e_1=e_2=\nil$.
         %%
@@
         Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-brk1}.
-        Hence~$p_1=p_2=\ceu{\clear(p'');\Break}$ and~$e_1=e_2=\nil$.
+        Hence $p_1=p_2=\ceu{\clear(p'');\Break}$ and~$e_1=e_2=\nil$.
         %%
@@
         hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
-        Hence~$p_1=\ceu{{p_1'}\And{p''}}=\ceu{{p_2'}\And{p''}}=p_2$
+        Hence $p_1=\ceu{{p_1'}\And{p''}}=\ceu{{p_2'}\And{p''}}=p_2$
         and~$e_1=e_1'=e_2'=e_2$.
@@
         Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-nop2}.
-        Hence~$p_1=p_2=p'$ and~$e_1=e_2=\nil$.
+        Hence $p_1=p_2=p'$ and~$e_1=e_2=\nil$.
         %%
@@
         Then~$\pi_1$ and~$\pi_2$ are instances of~\R{and-brk2}.
-        Hence~$p_1=p_2=\ceu{\clear(p');\Break}$ and~$e_1=e_2=\nil$.
+        Hence $p_1=p_2=\ceu{\clear(p');\Break}$ and~$e_1=e_2=\nil$.
         %%
@@
         hypothesis, $p_1''=p_2''$ and~$e_1''=e_2''$.
-        Hence~$p_1=\ceu{{p'}\And{p_1''}}=\ceu{{p'}\And{p_2''}}=p_2$
+        Hence $p_1=\ceu{{p'}\And{p_1''}}=\ceu{{p'}\And{p_2''}}=p_2$
         and~$e_1=e_1''=e_2''=e_2$.
@@
     Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-expd}.
-    Hence~$p_1=p_2=\ceu{{p'}\AtOr{(\CanRun(n);\,p'')}}$
+    Hence $p_1=p_2=\ceu{{p'}\AtOr{(\CanRun(n);\,p'')}}$
     and~$e_1=e_2=\nil$.
@@
         Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-nop1}.
-        Hence~$p_1=p_2=\clear(p'')$ and~$e_1=e_2=\nil$.
+        Hence $p_1=p_2=\clear(p'')$ and~$e_1=e_2=\nil$.
         %%
@@
         Then~$\pi_1$ and~$\pi_2$ are instances of~\R{or-nop1}.
-        Hence~$p_1=p_2=\clear(p')$ and~$e_1=e_2=\nil$.
+        Hence $p_1=p_2=\clear(p')$ and~$e_1=e_2=\nil$.
         %%
@@
         \]
-        Hence, from~\eqref{thm.x.term-nst-*.seq-adv.eq1} and by the
+        Hence from~\eqref{thm.x.term-nst-*.seq-adv.eq1} and by the
         definition~$\Hnst$, description~$\<\ceu{p_1';\,p''},n,e>$ is
@@
       %%
-    \item$\isblocked(p_1',n)$.  Hence,
+    \item$\isblocked(p_1',n)$.  Hence
       from~\eqref{thm.x.term-nst-*.loop-expd.eq1} and by the definition
@@
     Definition~\ref{def.x.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
-    as~$e'=\nil$ and~$n'=n+1$; hence~$j=j'$.
+    as~$e'=\nil$ and~$n'=n+1$; hence $j=j'$.
     %%
@@
     By Definition~\ref{def.x.pot}, $\pot(\bcast(p,e))=\pot(\bcast(p',e'))$;
-    hence~$i=i'$.  And by Definition~\ref{def.x.rank}, $j=n$, as~$e=\nil$,
-    and~$j'=n-1$, as~$e'=\nil$ and~$n'=n-1$; hence~$j>j'$.
+    hence $i=i'$.  And by Definition~\ref{def.x.rank}, $j=n$, as~$e=\nil$,
+    and~$j'=n-1$, as~$e'=\nil$ and~$n'=n-1$; hence $j>j'$.
     Therefore, ${\<i,j>}>{\<i',j'>}$.\qedhere
@@
     %%
-    Then~$\pi$ is an instance of~\R{mem}.  Hence~$p'=\ceu{\Nop}$
+    Then~$\pi$ is an instance of~\R{mem}.  Hence $p'=\ceu{\Nop}$
     and~$e'=\nil$.  Thus $\rank(\delta)=\rank(\delta')=\<0,n>$.
@@
     %%
-    Then~$\pi$ is an instance of~\R{emit-int}.  Hence~$p'=\ceu{\CanRun}$
+    Then~$\pi$ is an instance of~\R{emit-int}.  Hence $p'=\ceu{\CanRun}$
     and~$e'=e_1\ne\nil$.
@@
     %%
-    Then~$\pi$ is an instance of~\R{can-run}.  Hence~$p'=\ceu{\Nop}$
+    Then~$\pi$ is an instance of~\R{can-run}.  Hence $p'=\ceu{\Nop}$
     and~$e'=\nil$.  Thus
@@
       %%
-      Then~$\pi$ is an instance of~\R{if-true}.  Hence~$p'=\ceu{p_1}$
+      Then~$\pi$ is an instance of~\R{if-true}.  Hence $p'=\ceu{p_1}$
       and~$e'=\nil$.  Thus
@@
       Then~$\pi$ is an instance of~\R{seq-nop}.
-      Hence~$p'=p_2$ and~$e'=\nil$.  Thus
+      Hence $p'=p_2$ and~$e'=\nil$.  Thus
       \begin{align*}
@@
       Then~$\pi$ is an instance of~\R{seq-brk}.
-      Hence~$p'=p_1$ and~$e'=\nil$.  Thus
+      Hence $p'=p_1$ and~$e'=\nil$.  Thus
       \[
@@
     Then~$\pi$ is an instance of~\R{loop-expd}.
-    Hence~$p'=\ceu{p_1\AtLoop{p_1}}$ and~$e'=\nil$.
+    Hence $p'=\ceu{p_1\AtLoop{p_1}}$ and~$e'=\nil$.
     %%
@@
         still contain a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$.
-        Hence~$\pot'(p')=\pot'(p_1')$.  The rest of this proof is similar to
+        Hence $\pot'(p')=\pot'(p_1')$.  The rest of this proof is similar to
         that of Case~\ref{lem.x.rank-nst.seq-adv}.
@@
         or~$\ceu{\AwaitExt}$ does not occur.
-        Hence~$\pot'(p')=\pot'(p_1')+\pot'(p_2)$.  The rest of this proof is
+        Hence $\pot'(p')=\pot'(p_1')+\pot'(p_2)$.  The rest of this proof is
         similar to that of Case~\ref{lem.x.rank-nst.seq-adv}.
@@
     Then~$\pi$ is an instance of~\R{and-expd}.
-    Hence~$p'=\ceu{{p_1}\AtAnd(\CanRun(n);\,p_2)}$ and~$e'=\nil$.
+    Hence $p'=\ceu{{p_1}\AtAnd(\CanRun(n);\,p_2)}$ and~$e'=\nil$.
     Thus
@@
         %%
-        Then~$\pi$ is an instance of~\R{and-nop1}.  Hence~$p'=p_2$
+        Then~$\pi$ is an instance of~\R{and-nop1}.  Hence $p'=p_2$
         and~$e'=\nil$.  Thus
@@
         Then~$\pi$ is an instance of~\R{and-brk1}.
-        Hence~$p'=\ceu{\clear(p_2);\,\Break}$ and~$e'=\nil$.  By
+        Hence $p'=\ceu{\clear(p_2);\,\Break}$ and~$e'=\nil$.  By
         item~\ref{ass.x.syn-rest.fin} of Assumption~\ref{ass.x.syn-rest} and
@@
     Then~$\pi$ is an instance of~\R{or-expd}.
-    Hence~$p'=\ceu{{p_1}\AtOr(\CanRun(n);\,p_2)}$ and~$e'=\nil$.  Thus
+    Hence $p'=\ceu{{p_1}\AtOr(\CanRun(n);\,p_2)}$ and~$e'=\nil$.  Thus
     \[
@@
         %%
-        Then~$\pi$ is an instance of~\R{or-nop1}.  Hence~$p'=\clear(p_2)$
+        Then~$\pi$ is an instance of~\R{or-nop1}.  Hence $p'=\clear(p_2)$
         and~$e'=\nil$.  By item~\ref{ass.x.syn-rest.fin} of
@@
 
+%\balance
+
 \thmxterm*
@@
   implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be applied).
-  There are two possibilities:  either~$\delta$ is nested-irreducible or it
+  There are two possibilities:  either~$\delta$ is nested irreducible or it
   is not.  In the first case, the theorem is trivially true,
   as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
-  nested-irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
+  nested irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
   $\delta\nst[*]\delta'_\Hnst$, for some~$\delta'_\Hnst$.  By
@@
         \]
-        Hence, by the induction hypothesis, there is a~$\delta''_\#$ such
+        Hence by the induction hypothesis, there is a~$\delta''_\#$ such
         that~$\delta'\trans[*]\delta''_\#$.
